### 2023-11-28 paper/paper.qmd
Start rewriting introduction

---
title: "A density ratio framework for evaluating the utility of synthetic data"
format:
  arxiv-pdf:
    keep-tex: true  
  arxiv-html: default
linenumbers: false
doublespacing: true
runninghead: "Density ratios for utility"
authorcols: true
link-citations: true
author:
  - name: Thom Benjamin Volker
    affiliations:
      - name: Utrecht University | Statistics Netherlands
        department: Methodology and Statistics | Methodology
        address: Padualaan 14
        city: Utrecht
        country: The Netherlands
        postal-code: "3584CH"
    orcid: 0000-0002-2408-7820
    email: t.b.volker@uu.nl
  - name: Peter-Paul de Wolf
    affiliations:
      - name: Statistics Netherlands
        department: Methodology
        address: PO Box 24500
        city: The Hague
        country: The Netherlands
        postal-code: "2490HA"
    email: pp.dewolf@cbs.nl
  - name: Erik-Jan van Kesteren
    affiliations:
      - name: Utrecht University
        department: Methodology and Statistics
        address: Padualaan 14
        city: Utrecht
        country: The Netherlands
        postal-code: "3584CH"
    email: e.vankesteren1@uu.nl
abstract: |
  TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO 
keywords: 
  - TODO
  - TODO
bibliography: bibliography.bib  
---

# Introduction

Openly accessible research data accelerates scientific progress tremendously [ADD REFERENCES].
Open data allows third-party researchers to answer research questions with already collected data, substantially lowering the costs of research. 
Sharing data in combination with code allows others to validate research findings and build upon the work.
Students can benefit from open data, as it fosters education with realistic data, as well as the general public, through stimulating citizen science projects. 
However, openly available research data is often at odds with privacy constraints on the dissemination of the data.
The information may cause harm to individuals or organizations if made publicly available. 
Additionally, sharing their information openly may withhold them from participating in future research.
These constraints have been named among the biggest hurdles in the advancement of computational social science [@lazer_css_2009], and among top reasons for companies to not share their data with researchers [@fpf_2017]. 
To overcome these obstacles, data providers can employ a suite of different disclosure limitation techniques before sharing data, for example top-coding, record-swapping or adding noise [e.g., @hundepool_disclosure_2012; @willenborg_elements_2012].

Recently, synthetic data has gained substantial traction as a means to disclosure limitation.
National statistical institutes and other government agencies have adopted the synthetic data framework to facilitate the use of their data [e.g., @SIPP_Beta_2006; @hawala_synthetic_2008; @drechsler2012]. 
Researchers started using synthetic data to train machine learning models [@nikolenko2021], and began to share synthetic versions of their research data to comply with open science standards [e.g., @vandewiel2023; @obermeyer2019; @zettler2021].
The idea of synthetic data is to replace some, or all, of the observed values in a data set by synthetic records that are generated from some model [e.g., @little_statistical_1993; @rubin_statistical_1993; @drechsler2011synthetic].
If only some values are replaced, disclosure risks can be reduced because the sensitive or identifying values do not correspond to their true values anymore. 
If all values are replaced, there is also no one-to-one mapping between the original and the synthetic data, further reducing the disclosure risk.

Many different methods have been proposed to generate synthetic data.
Traditionally, these were closely connected to methods used for multiple imputation of missing data, such as joint modelling [misdc2003], sequential regressions [@nowok2016] or fully conditional specification [@drechsler2011empirical; @volker2021].
The flexibility of sequential regressions and fully conditional specification is commonly combined with non-parametric imputation models, such as classification and regression trees [@reiter2005], random forests [@caiola2010] or support vector machines [@drechsler2010], to avoid distributional assumptions and easily model non-linear relationships. 
Recently, significant improvements in generative modelling sparked the scientific interest in synthetic data in the computer science community, leading to novel synthesis methods [e.g., @patki2016; @xu_ctgan_2019].
Combined with work on formal privacy guarantees such as differential privacy, this resulted in new models that explicitly control the level of privacy risk in synthesis methods [@jordon2018pategan; @Torkzadehmahani2019].

___TODO___

All methods for statistical disclosure limitation alter the data before these are provided to the public.
By doing so, the utility of the provided data is always lower than the utility of the original data, because some of the information in the data is sacrificed to protect the privacy of the respondents. 
The questions that naturally arise are how much information in the original data is actually sacrificed, and how useful the provided data are?
Answering this question allows researchers to decide what the altered data can and cannot be used for, and to evaluate the worth of conclusions drawn on the basis of these data.
After all, inferences from the altered data are valid only up to the extent that the perturbation methods approximate the true data-generating mechanism.
For data providers, a detailed assessment of the quality of the altered data can guide the procedure of altering the data. 
Statistical disclosure limitation is often an iterative process: some disclosure limitation technique is applied on the data, after which the result is investigated and modifications are made to applied process. 
Good measures of data quality are essential to determine the appropriate mechanisms used to protect the data, and can help to improve the utility of the data that will be disseminated.

In the statistical disclosure control literature, two different branches of utility measures have been distinguished: specific utility measures and general utility measures. 
_Add one/two sentences on the merit of visualization when assessing utility of altered data._
Specific utility measures focus on similarity of results obtained from analyses performed on the altered data and the original data. 
For example, after fitting the same analysis model on both data sets, one can calculate the confidence interval overlap of the estimated parameters [@karr_utility_2006]. 
Alternative measures are ellipsoidal overlap [@karr_utility_2006], which extends to confidence interval overlap to a measure that addresses the joint distribution of all model parameters simultaneous, the standardized absolute difference between estimates [@snoke_utility_2018], and the ratio of estimates for tabular count data [@taub2020impact].
As these measures quantify similarity between estimates from analyses performed on the observed and altered data, they are informed only to the extent that data users will recreate those analyses. 
This can be highly useful if the data is provided for reproducibility purposes (e.g., for third parties to evaluate analysis scripts).
However, the goal of distributing the protected data is often to allow researchers to do novel research with the data. 
In many practical situations, data providers thus have have only limited knowledge on the analyses that will be performed with the altered data. 
Covering the entire set of potentially relevant analyses is therefore not feasible.
If it was, the data providers could simply report the (potentially privacy-protected) results of those analyses performed on the real data, so that access to the (perturbed) data no longer yields additional benefits [for a similar argument, see @drechsler_utility_2022].
Additionally, similarity between results on the analyses that have been performed gives no guarantee that the results will also be similar for other analyses.
Hence, when determining how useful the altered data is for novel research, specific utility measures are only of limited use.

General utility measures attempt to capture how similar the multivariate distributions of the observed and altered data are.
This can be done by, for instance, estimating the Kullback-Leibler divergence between the distributions of the observed and altered data [@karr_utility_2006].
An alternative strategy is to try to discriminate between the observed and altered data, as is done with the $pMSE$ [@snoke_utility_2018; @woo_utility_2009].
In essence, the $pMSE$ quantifies how well one can predict whether observations are from the observed or the altered data. 
If better one can do this, the more pronounced the differences between the observed and altered data ought to be. 
However, various authors have criticized general utility measures for being too broad. 
That is, important discrepancies between the real and altered data might be missed, and an altered data set that is good in general (i.e., has high global utility) might still provide results that are far from the truth for some analyses [see, e.g., @drechsler_utility_2022].
Additionally, it is not straightforward to determine which prediction model to use for calculating the $pMSE$. 
Specifying a good prediction model in itself may be a challenging task, especially when the number of variables is large. 
When good models are available, different models, or even different choices of hyperparameters, may yield different results, potentially rendering ambiguity with respect to which altered data set is best.
Lastly, the output of global utility measures can be hard to interpret, and say little about the regions in which the synthetic data do not resemble the true data accurately enough.
That is, they give little guidance on how the quality of the altered data can be improved.

---

__Section 6: our contribution__

_Moet nog verder uitgewerkt worden_

1. We introduce density ratio estimation to the field of statistical disclosure control. Short remark on the idea that density ratio estimation is a complicated endeavor, especially if the goal is to compare distinct densities. Having to estimate just a single density (ratio) is generally much easier.

2. Note that density ratio estimation can capture specific and general utility measures into a common framework by being applicable on the level of the entire data, but also on the subset of variables that is relevant in an analysis. Additionally, note that confidence interval overlap, ellipsoidal overlap, but also $pMSE$ and Kullback-Leibler divergence, are closely related to density ratio estimation, and can be considered from this perspective.

3. Create a new utility metric based on density ratio estimation (probability with respect to some reference distribution as in permutation testing).

4. Because density ratio estimation can be difficult when there are many variables, we use dimension reduction techniques to capture most of the variability in the data in fewer dimensions on which density ratio estimation can be applied. A by-product of this is that the lower-dimensional subspace allows to create visualizations of deviations from the observed data.

5. Perform a simulation study to give indications about which methods to use (think about how to do this).

6. Implement all this in an R-package

__Section 6: outline article__

In the next section, we describe density ratio estimation and discuss how this method can be used as to measure utility.
Subsequently, we provide multiple examples that show how density ratio estimation works in the context of evaluating the quality of synthetic data. 
Hereafter, we show in multiple simulations that the method is superior (HOPEFULLY) to current global utility measures as the $pMSE$. 
Lastly, we discuss the advantages and disadvantages of density ratio estimation as a utility measure.

# Methodology

\@Peter-Paul: Eventueel een korte beschrijving van data perturbation techniques/synthetic data generation hier. Denk je dat dit wat toevoegt hier?

__Section 1: density ratio estimation__

In essence, the goal of utility measures is to quantify the similarity between the multivariate distribution of the observed data with the distribution of the altered data. 
If the used data perturbation techniques, or synthetic data generation models, approximate the distribution of the real data sufficiently, these distributions should be highly similar, and analyses on the two data sets should give similar results. 
However, estimating the probability distribution of a data set is known to be one of the most complicated challenges in statistics [E.G. Vapnik 1998].
Estimating the probability distribution for both observed and altered data can lead to errors in both, artificially magnifying discrepancies between the two. 
Hence, subsequent comparisons will be affected by these errors. 
The procedure can be simplified by using density ratio estimation, because this only requires to estimate a single density.

<!-- Check section 10.2 in density ratio estimation in machine learning.  -->
<!-- Two-sample test/homogeneity test (Kullback, 1959): test whether probability distributions be behind two sets of samples are equivalent. -->

<!-- "A standard approach to the two-sample test is to estimate a divergence between two probability distributions (Keziou & Leoni-Aubin, 2005; Kanamori et al., 2011a). A key observation is that a general class of divergences (Ali & Silvey, 1966; Csiszár, 1967), including the Kullback-Leibler divergene (Kullback & Leibler, 1951) and the Pearson divergence (Pearson 1900) can be approximated accurately via density-ratio estimation (Nguyen et al., 2010; Sugiyama et al., 2011c), resulting in better test accuracy than estimating the distributions separately." -->

Introduce density ratio estimation as a utility measure. 
What does this measure mean/how to interpret it. 
How to make decisions based on this measure.

Say something on whether (and if so, how) categorical variables can be incorporated as well.

__Section 3: theoretical comparison with conventional approaches for general utility assessment__

Relate density ratio estimation to specific and general utility measures.
Pick one/two specific utility measures and relate these to density ratio estimation (ratio of estimates seems straightforward, as well as confidence interval overlap and ellipsoidal overlap).

Relate density ratio estimation to $pMSE$ and KL divergence (to some extent, both are generalizations of density ratio estimation, or at least are conceptually similar).
Give some more information on the $pMSE$, describe what it shortcomings are. 
The quality of the $pMSE$ highly depends on the model used to calculate the propensity scores. 
Perhaps give an example of logistic regression, which basically estimates whether the conditional mean of the observed and altered data is the same. 
Explain how density ratio estimation can overcome the shortcomings of the previously mentioned methods.

__Section 4: Dimension reduction for utility__

The difficulty of density ratio estimation increases with the dimensionality of the data. 
Therefore, we follow previous recommendations to incorporate dimensionality reduction techniques in density ratio estimation.

Shortly name examples of dimensionality reduction techniques (i.e., PCA; LFDA or UMAP).

A useful by-product of dimension reduction is that it allows to create visualizations, and these visualizations can be used to get more insight in discrepancies between observed and altered data. 
Show what such visualizations can look like, and how they can help.

# Simulations

## Small illustration / example  with multivariate Gaussian distributions.

1. Simple, multivariate normal simulation (e.g., two correlation structures, two sample sizes, so $2 \times 2$ full factorial design); basically similar to what we did already.

## More complex simulation, more variables, non-linearities, perhaps using real data.

2. More advanced simulation (e.g., some non-linearities, different sample sizes)

Have to think about this in more detail still.

# Real data example

Clinical records heart-failure data? Misschien ook niet, nog over nadenken.

Exemplify how utility measures could (should!) be used to improve the quality of the altered data (e.g., illustrate how models can be adjusted iteratively based on utility assessment).

_Some notes to self_

Current ways to assess the utility?

- pMSE - logistic, regression, CART models (Snoke, Raab, Nowok, Dibben & Slavkovic, 2018; General and specific utility measures for synthetic data AND Woo, Reiter, Oganian & Karr, 2009; Global measures of data utlity for microdata masked for disclosure limitation)

- Kullback-Leiber divergence (Karr, Kohnen, Oganian, Reiter & Sanil, 2006; A framework for evaluating the utility of data altered to protect confidentiality).

- According to multiple authors, both specific and general utility measures have important drawbacks (see Drechsler Utility PSD; cites others). Narrow measures potentially focus on analyses that are not relevant for the end user, and do not generalize to the analyses that are relevant. Global utility measures are generally too broad, and important deviations in the synthetic data might be missed. Moreover, the measures are typically hard to interpret.

- See Drechsler for a paragraph on fit for purpose measures, that lie between general and specific utility measures (i.e., plausibility checks such as non-negativity; goodness of fit measures as $\chi^2$ for cross-tabulations; Kolmogorov-Smirnov).

- Drechsler also illustrates that the standardized $pMSE$ has substantial flaws, as the results are highly dependent on the model used to estimate the propensity scores, and unable to detect important differences in the utility for most of the model specifications. Hence, it is claimed that a thorough assessment of utility is required.

Things to add in new version:
- High dimensional example
- Expansion of discussion points:
	empirical example with weighted analyses
	individual data utility
- Privacy remark?
- When does it not work?

# Methodology

TO DO

# Simulations

TO DO

# Real data example

TO DO

# Results

TO DO

# Discussion and conclusion

TO DO

# References {.unnumbered}

::: {#refs}
:::

### 2023-11-28 paper/paper.tex
Start rewriting introduction

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Latin Modern Roman}
  \setmathfont[]{Latin Modern Math}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A density ratio framework for evaluating the utility of synthetic data},
  pdfauthor={Thom Benjamin Volker; Peter-Paul de Wolf; Erik-Jan van Kesteren},
  pdfkeywords={TODO, TODO},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\usepackage{setspace}
\doublespacing
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Density ratios for utility }
\title{A density ratio framework for evaluating the utility of synthetic
data}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Thom Benjamin
Volker}~\orcidlink{0000-0002-2408-7820}\\Methodology and Statistics
\textbar{} Methodology\\Utrecht University \textbar{} Statistics
Netherlands\\Utrecht,\ 3584CH\\\href{mailto:t.b.volker@uu.nl}{t.b.volker@uu.nl}\asep\textbf{Peter-Paul
de Wolf}\\Methodology\\Statistics Netherlands\\The
Hague,\ 2490HA\\\href{mailto:pp.dewolf@cbs.nl}{pp.dewolf@cbs.nl}\asep\textbf{Erik-Jan
van Kesteren}\\Methodology and Statistics\\Utrecht
University\\Utrecht,\ 3584CH\\\href{mailto:e.vankesteren1@uu.nl}{e.vankesteren1@uu.nl}}
\date{}
\begin{document}
\maketitle
\begin{abstract}
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
TODO TODO TODO
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
TODO \sep 
TODO

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable, frame hidden]}{\end{tcolorbox}}\fi

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Openly accessible research data accelerates scientific progress
tremendously {[}ADD REFERENCES{]}. Open data allows third-party
researchers to answer research questions with already collected data,
substantially lowering the costs of research. Sharing data in
combination with code allows others to validate research findings and
build upon the work. Students can benefit from open data, as it fosters
education with realistic data, as well as the general public, through
stimulating citizen science projects. However, openly available research
data is often at odds with privacy constraints on the dissemination of
the data. The information may cause harm to individuals or organizations
if made publicly available. Additionally, sharing their information
openly may withhold them from participating in future research. These
constraints have been named among the biggest hurdles in the advancement
of computational social science
(\protect\hyperlink{ref-lazer_css_2009}{Lazer et al. 2009}), and among
top reasons for companies to not share their data with researchers
(\protect\hyperlink{ref-fpf_2017}{Future of Privacy Forum 2017}). To
overcome these obstacles, data providers can employ a suite of different
disclosure limitation techniques before sharing data, for example
top-coding, record-swapping or adding noise (e.g.,
\protect\hyperlink{ref-hundepool_disclosure_2012}{Hundepool et al.
2012}; \protect\hyperlink{ref-willenborg_elements_2012}{Willenborg and
De Waal 2012}).

Recently, synthetic data has gained substantial traction as a means to
disclosure limitation. National statistical institutes and other
government agencies have adopted the synthetic data framework to
facilitate the use of their data (e.g.,
\protect\hyperlink{ref-SIPP_Beta_2006}{Abowd, Stinson, and Benedetto
2006}; \protect\hyperlink{ref-hawala_synthetic_2008}{Hawala 2008};
\protect\hyperlink{ref-drechsler2012}{Drechsler 2012}). Researchers
started using synthetic data to train machine learning models
(\protect\hyperlink{ref-nikolenko2021}{Nikolenko 2021}), and began to
share synthetic versions of their research data to comply with open
science standards (e.g., \protect\hyperlink{ref-vandewiel2023}{van de
Wiel et al. 2023}; \protect\hyperlink{ref-obermeyer2019}{Obermeyer et
al. 2019}; \protect\hyperlink{ref-zettler2021}{Zettler et al. 2021}).
The idea of synthetic data is to replace some, or all, of the observed
values in a data set by synthetic records that are generated from some
model (e.g., \protect\hyperlink{ref-little_statistical_1993}{Little
1993}; \protect\hyperlink{ref-rubin_statistical_1993}{Rubin 1993};
\protect\hyperlink{ref-drechsler2011synthetic}{Drechsler 2011}). If only
some values are replaced, disclosure risks can be reduced because the
sensitive or identifying values do not correspond to their true values
anymore. If all values are replaced, there is also no one-to-one mapping
between the original and the synthetic data, further reducing the
disclosure risk.

Many different methods have been proposed to generate synthetic data.
Traditionally, these were closely connected to methods used for multiple
imputation of missing data, such as joint modelling {[}misdc2003{]},
sequential regressions (\protect\hyperlink{ref-nowok2016}{Nowok, Raab,
and Dibben 2016}) or fully conditional specification
(\protect\hyperlink{ref-drechsler2011empirical}{Drechsler and Reiter
2011}; \protect\hyperlink{ref-volker2021}{Volker and Vink 2021}). The
flexibility of sequential regressions and fully conditional
specification is commonly combined with non-parametric imputation
models, such as classification and regression trees
(\protect\hyperlink{ref-reiter2005}{Reiter 2005}), random forests
(\protect\hyperlink{ref-caiola2010}{Caiola and Reiter 2010}) or support
vector machines (\protect\hyperlink{ref-drechsler2010}{Drechsler 2010}),
to avoid distributional assumptions and easily model non-linear
relationships. Recently, significant improvements in generative
modelling sparked the scientific interest in synthetic data in the
computer science community, leading to novel synthesis methods (e.g.,
\protect\hyperlink{ref-patki2016}{Patki, Wedge, and Veeramachaneni
2016}; \protect\hyperlink{ref-xu_ctgan_2019}{Xu et al. 2019}). Combined
with work on formal privacy guarantees such as differential privacy,
this resulted in new models that explicitly control the level of privacy
risk in synthesis methods
(\protect\hyperlink{ref-jordon2018pategan}{Jordon, Yoon, and Schaar
2019}; \protect\hyperlink{ref-Torkzadehmahani2019}{Torkzadehmahani,
Kairouz, and Paten 2019}).

\textbf{\emph{TODO}}

All methods for statistical disclosure limitation alter the data before
these are provided to the public. By doing so, the utility of the
provided data is always lower than the utility of the original data,
because some of the information in the data is sacrificed to protect the
privacy of the respondents. The questions that naturally arise are how
much information in the original data is actually sacrificed, and how
useful the provided data are? Answering this question allows researchers
to decide what the altered data can and cannot be used for, and to
evaluate the worth of conclusions drawn on the basis of these data.
After all, inferences from the altered data are valid only up to the
extent that the perturbation methods approximate the true
data-generating mechanism. For data providers, a detailed assessment of
the quality of the altered data can guide the procedure of altering the
data. Statistical disclosure limitation is often an iterative process:
some disclosure limitation technique is applied on the data, after which
the result is investigated and modifications are made to applied
process. Good measures of data quality are essential to determine the
appropriate mechanisms used to protect the data, and can help to improve
the utility of the data that will be disseminated.

In the statistical disclosure control literature, two different branches
of utility measures have been distinguished: specific utility measures
and general utility measures. \emph{Add one/two sentences on the merit
of visualization when assessing utility of altered data.} Specific
utility measures focus on similarity of results obtained from analyses
performed on the altered data and the original data. For example, after
fitting the same analysis model on both data sets, one can calculate the
confidence interval overlap of the estimated parameters
(\protect\hyperlink{ref-karr_utility_2006}{Karr et al. 2006}).
Alternative measures are ellipsoidal overlap
(\protect\hyperlink{ref-karr_utility_2006}{Karr et al. 2006}), which
extends to confidence interval overlap to a measure that addresses the
joint distribution of all model parameters simultaneous, the
standardized absolute difference between estimates
(\protect\hyperlink{ref-snoke_utility_2018}{Snoke et al. 2018}), and the
ratio of estimates for tabular count data
(\protect\hyperlink{ref-taub2020impact}{\textbf{taub2020impact?}}). As
these measures quantify similarity between estimates from analyses
performed on the observed and altered data, they are informed only to
the extent that data users will recreate those analyses. This can be
highly useful if the data is provided for reproducibility purposes
(e.g., for third parties to evaluate analysis scripts). However, the
goal of distributing the protected data is often to allow researchers to
do novel research with the data. In many practical situations, data
providers thus have have only limited knowledge on the analyses that
will be performed with the altered data. Covering the entire set of
potentially relevant analyses is therefore not feasible. If it was, the
data providers could simply report the (potentially privacy-protected)
results of those analyses performed on the real data, so that access to
the (perturbed) data no longer yields additional benefits (for a similar
argument, see \protect\hyperlink{ref-drechsler_utility_2022}{Drechsler
2022}). Additionally, similarity between results on the analyses that
have been performed gives no guarantee that the results will also be
similar for other analyses. Hence, when determining how useful the
altered data is for novel research, specific utility measures are only
of limited use.

General utility measures attempt to capture how similar the multivariate
distributions of the observed and altered data are. This can be done by,
for instance, estimating the Kullback-Leibler divergence between the
distributions of the observed and altered data
(\protect\hyperlink{ref-karr_utility_2006}{Karr et al. 2006}). An
alternative strategy is to try to discriminate between the observed and
altered data, as is done with the \(pMSE\)
(\protect\hyperlink{ref-snoke_utility_2018}{Snoke et al. 2018};
\protect\hyperlink{ref-woo_utility_2009}{\textbf{woo\_utility\_2009?}}).
In essence, the \(pMSE\) quantifies how well one can predict whether
observations are from the observed or the altered data. If better one
can do this, the more pronounced the differences between the observed
and altered data ought to be. However, various authors have criticized
general utility measures for being too broad. That is, important
discrepancies between the real and altered data might be missed, and an
altered data set that is good in general (i.e., has high global utility)
might still provide results that are far from the truth for some
analyses (see, e.g.,
\protect\hyperlink{ref-drechsler_utility_2022}{Drechsler 2022}).
Additionally, it is not straightforward to determine which prediction
model to use for calculating the \(pMSE\). Specifying a good prediction
model in itself may be a challenging task, especially when the number of
variables is large. When good models are available, different models, or
even different choices of hyperparameters, may yield different results,
potentially rendering ambiguity with respect to which altered data set
is best. Lastly, the output of global utility measures can be hard to
interpret, and say little about the regions in which the synthetic data
do not resemble the true data accurately enough. That is, they give
little guidance on how the quality of the altered data can be improved.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\textbf{Section 6: our contribution}

\emph{Moet nog verder uitgewerkt worden}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We introduce density ratio estimation to the field of statistical
  disclosure control. Short remark on the idea that density ratio
  estimation is a complicated endeavor, especially if the goal is to
  compare distinct densities. Having to estimate just a single density
  (ratio) is generally much easier.
\item
  Note that density ratio estimation can capture specific and general
  utility measures into a common framework by being applicable on the
  level of the entire data, but also on the subset of variables that is
  relevant in an analysis. Additionally, note that confidence interval
  overlap, ellipsoidal overlap, but also \(pMSE\) and Kullback-Leibler
  divergence, are closely related to density ratio estimation, and can
  be considered from this perspective.
\item
  Create a new utility metric based on density ratio estimation
  (probability with respect to some reference distribution as in
  permutation testing).
\item
  Because density ratio estimation can be difficult when there are many
  variables, we use dimension reduction techniques to capture most of
  the variability in the data in fewer dimensions on which density ratio
  estimation can be applied. A by-product of this is that the
  lower-dimensional subspace allows to create visualizations of
  deviations from the observed data.
\item
  Perform a simulation study to give indications about which methods to
  use (think about how to do this).
\item
  Implement all this in an R-package
\end{enumerate}

\textbf{Section 6: outline article}

In the next section, we describe density ratio estimation and discuss
how this method can be used as to measure utility. Subsequently, we
provide multiple examples that show how density ratio estimation works
in the context of evaluating the quality of synthetic data. Hereafter,
we show in multiple simulations that the method is superior (HOPEFULLY)
to current global utility measures as the \(pMSE\). Lastly, we discuss
the advantages and disadvantages of density ratio estimation as a
utility measure.

\hypertarget{methodology}{%
\section{Methodology}\label{methodology}}

@Peter-Paul: Eventueel een korte beschrijving van data perturbation
techniques/synthetic data generation hier. Denk je dat dit wat toevoegt
hier?

\textbf{Section 1: density ratio estimation}

In essence, the goal of utility measures is to quantify the similarity
between the multivariate distribution of the observed data with the
distribution of the altered data. If the used data perturbation
techniques, or synthetic data generation models, approximate the
distribution of the real data sufficiently, these distributions should
be highly similar, and analyses on the two data sets should give similar
results. However, estimating the probability distribution of a data set
is known to be one of the most complicated challenges in statistics
{[}E.G. Vapnik 1998{]}. Estimating the probability distribution for both
observed and altered data can lead to errors in both, artificially
magnifying discrepancies between the two. Hence, subsequent comparisons
will be affected by these errors. The procedure can be simplified by
using density ratio estimation, because this only requires to estimate a
single density.

Introduce density ratio estimation as a utility measure. What does this
measure mean/how to interpret it. How to make decisions based on this
measure.

Say something on whether (and if so, how) categorical variables can be
incorporated as well.

\textbf{Section 3: theoretical comparison with conventional approaches
for general utility assessment}

Relate density ratio estimation to specific and general utility
measures. Pick one/two specific utility measures and relate these to
density ratio estimation (ratio of estimates seems straightforward, as
well as confidence interval overlap and ellipsoidal overlap).

Relate density ratio estimation to \(pMSE\) and KL divergence (to some
extent, both are generalizations of density ratio estimation, or at
least are conceptually similar). Give some more information on the
\(pMSE\), describe what it shortcomings are. The quality of the \(pMSE\)
highly depends on the model used to calculate the propensity scores.
Perhaps give an example of logistic regression, which basically
estimates whether the conditional mean of the observed and altered data
is the same. Explain how density ratio estimation can overcome the
shortcomings of the previously mentioned methods.

\textbf{Section 4: Dimension reduction for utility}

The difficulty of density ratio estimation increases with the
dimensionality of the data. Therefore, we follow previous
recommendations to incorporate dimensionality reduction techniques in
density ratio estimation.

Shortly name examples of dimensionality reduction techniques (i.e., PCA;
LFDA or UMAP).

A useful by-product of dimension reduction is that it allows to create
visualizations, and these visualizations can be used to get more insight
in discrepancies between observed and altered data. Show what such
visualizations can look like, and how they can help.

\hypertarget{simulations}{%
\section{Simulations}\label{simulations}}

\hypertarget{small-illustration-example-with-multivariate-gaussian-distributions.}{%
\subsection{Small illustration / example with multivariate Gaussian
distributions.}\label{small-illustration-example-with-multivariate-gaussian-distributions.}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simple, multivariate normal simulation (e.g., two correlation
  structures, two sample sizes, so \(2 \times 2\) full factorial
  design); basically similar to what we did already.
\end{enumerate}

\hypertarget{more-complex-simulation-more-variables-non-linearities-perhaps-using-real-data.}{%
\subsection{More complex simulation, more variables, non-linearities,
perhaps using real
data.}\label{more-complex-simulation-more-variables-non-linearities-perhaps-using-real-data.}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  More advanced simulation (e.g., some non-linearities, different sample
  sizes)
\end{enumerate}

Have to think about this in more detail still.

\hypertarget{real-data-example}{%
\section{Real data example}\label{real-data-example}}

Clinical records heart-failure data? Misschien ook niet, nog over
nadenken.

Exemplify how utility measures could (should!) be used to improve the
quality of the altered data (e.g., illustrate how models can be adjusted
iteratively based on utility assessment).

\emph{Some notes to self}

Current ways to assess the utility?

\begin{itemize}
\item
  pMSE - logistic, regression, CART models (Snoke, Raab, Nowok, Dibben
  \& Slavkovic, 2018; General and specific utility measures for
  synthetic data AND Woo, Reiter, Oganian \& Karr, 2009; Global measures
  of data utlity for microdata masked for disclosure limitation)
\item
  Kullback-Leiber divergence (Karr, Kohnen, Oganian, Reiter \& Sanil,
  2006; A framework for evaluating the utility of data altered to
  protect confidentiality).
\item
  According to multiple authors, both specific and general utility
  measures have important drawbacks (see Drechsler Utility PSD; cites
  others). Narrow measures potentially focus on analyses that are not
  relevant for the end user, and do not generalize to the analyses that
  are relevant. Global utility measures are generally too broad, and
  important deviations in the synthetic data might be missed. Moreover,
  the measures are typically hard to interpret.
\item
  See Drechsler for a paragraph on fit for purpose measures, that lie
  between general and specific utility measures (i.e., plausibility
  checks such as non-negativity; goodness of fit measures as \(\chi^2\)
  for cross-tabulations; Kolmogorov-Smirnov).
\item
  Drechsler also illustrates that the standardized \(pMSE\) has
  substantial flaws, as the results are highly dependent on the model
  used to estimate the propensity scores, and unable to detect important
  differences in the utility for most of the model specifications.
  Hence, it is claimed that a thorough assessment of utility is
  required.
\end{itemize}

Things to add in new version: - High dimensional example - Expansion of
discussion points: empirical example with weighted analyses individual
data utility - Privacy remark? - When does it not work?

\hypertarget{methodology-1}{%
\section{Methodology}\label{methodology-1}}

TO DO

\hypertarget{simulations-1}{%
\section{Simulations}\label{simulations-1}}

TO DO

\hypertarget{real-data-example-1}{%
\section{Real data example}\label{real-data-example-1}}

TO DO

\hypertarget{results}{%
\section{Results}\label{results}}

TO DO

\hypertarget{discussion-and-conclusion}{%
\section{Discussion and conclusion}\label{discussion-and-conclusion}}

TO DO

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-SIPP_Beta_2006}{}}%
Abowd, John M., Martha Stinson, and Gary Benedetto. 2006. {``Final
Report to the Social Security Administration on the {SIPP/SSA/IRS}
Public Use File Project.''} Longitudinal Employer-Household Dynamics
Program, U.S. Bureau of the Census, Washington, DC.
\url{https://ecommons.cornell.edu/bitstream/handle/1813/43929/SSAfinal.pdf?sequence=3\&isAllowed=y}.

\leavevmode\vadjust pre{\hypertarget{ref-caiola2010}{}}%
Caiola, Gregory, and Jerome P. Reiter. 2010. {``Random Forests for
Generating Partially Synthetic, Categorical Data.''} \emph{Transactions
on Data Privacy} 3: 27--42.
\url{https://doi.org/10.5555/1747335.1747337}.

\leavevmode\vadjust pre{\hypertarget{ref-drechsler2010}{}}%
Drechsler, Jörg. 2010. {``Using Support Vector Machines for Generating
Synthetic Datasets.''} In \emph{Privacy in Statistical Databases},
edited by Josep Domingo-Ferrer and Emmanouil Magkos, 148--61. Berlin,
Heidelberg: Springer Berlin Heidelberg.
\url{https://doi.org/10.1007/978-3-642-15838-4_14}.

\leavevmode\vadjust pre{\hypertarget{ref-drechsler2011synthetic}{}}%
---------. 2011. \emph{Synthetic Datasets for Statistical Disclosure
Control: Theory and Implementation}. New York: Springer Science \&
Business Media. \url{https://doi.org/10.1007/978-1-4614-0326-5}.

\leavevmode\vadjust pre{\hypertarget{ref-drechsler2012}{}}%
---------. 2012. {``New Data Dissemination Approaches in Old Europe
{\textendash} Synthetic Datasets for a German Establishment Survey.''}
\emph{Journal of Applied Statistics} 39 (2): 243--65.
\url{https://doi.org/10.1080/02664763.2011.584523}.

\leavevmode\vadjust pre{\hypertarget{ref-drechsler_utility_2022}{}}%
---------. 2022. {``Challenges in Measuring Utility for Fully Synthetic
Data.''} In \emph{Privacy in Statistical Databases}, edited by Josep
Domingo-Ferrer and Maryline Laurent, 220--33. Cham: Springer
International Publishing.
\url{https://doi.org/10.1007/978-3-031-13945-1_16}.

\leavevmode\vadjust pre{\hypertarget{ref-drechsler2011empirical}{}}%
Drechsler, Jörg, and Jerome P Reiter. 2011. {``An Empirical Evaluation
of Easily Implemented, Nonparametric Methods for Generating Synthetic
Datasets.''} \emph{Computational Statistics \& Data Analysis} 55 (12):
3232--43.

\leavevmode\vadjust pre{\hypertarget{ref-fpf_2017}{}}%
Future of Privacy Forum. 2017. {``Understanding Corporate Data Sharing
Decisions: Practices, Challenges, and Opportunities for Sharing
Corporate Data with Researchers.''}

\leavevmode\vadjust pre{\hypertarget{ref-hawala_synthetic_2008}{}}%
Hawala, Sam. 2008. \emph{Producing Partially Synthetic Data to Avoid
Disclosure}.
\url{http://www.asasrms.org/Proceedings/y2008/Files/301018.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-hundepool_disclosure_2012}{}}%
Hundepool, Anco, Josep Domingo-Ferrer, Luisa Franconi, Sarah Giessing,
Eric Schulte Nordholt, Keith Spicer, and Peter-Paul De Wolf. 2012.
\emph{Statistical Disclosure Control}. John Wiley \& Sons.
\url{https://doi.org/10.1002/9781118348239}.

\leavevmode\vadjust pre{\hypertarget{ref-jordon2018pategan}{}}%
Jordon, James, Jinsung Yoon, and Mihaela van der Schaar. 2019.
{``{PATE}-{GAN}: Generating Synthetic Data with Differential Privacy
Guarantees.''} In \emph{International Conference on Learning
Representations}. \url{https://openreview.net/forum?id=S1zk9iRqF7}.

\leavevmode\vadjust pre{\hypertarget{ref-karr_utility_2006}{}}%
Karr, Alan F., Christine N. Kohnen, Anna Oganian, Jerome P. Reiter, and
Ashish P. Sanil. 2006. {``A Framework for Evaluating the Utility of Data
Altered to Protect Confidentiality.''} \emph{The American Statistician}
60 (3): 224--32. \url{https://doi.org/10.1198/000313006X124640}.

\leavevmode\vadjust pre{\hypertarget{ref-lazer_css_2009}{}}%
Lazer, David, Alex Pentland, Lada Adamic, Sinan Aral, Albert-László
Barabási, Devon Brewer, Nicholas Christakis, et al. 2009.
{``Computational Social Science.''} \emph{Science} 323 (5915): 721--23.
\url{https://doi.org/10.1126/science.1167742}.

\leavevmode\vadjust pre{\hypertarget{ref-little_statistical_1993}{}}%
Little, Roderick J. A. 1993. {``Statistical Analysis of Masked Data.''}
\emph{Journal of Official Statistics} 9 (2): 407--7.
\url{https://www.scb.se/contentassets/ca21efb41fee47d293bbee5bf7be7fb3/statistical-analysis-of-masked-data.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-nikolenko2021}{}}%
Nikolenko, Sergey I. 2021. \emph{Synthetic Data for Deep Learning}.
Springer International Publishing.
\url{https://doi.org/10.1007/978-3-030-75178-4}.

\leavevmode\vadjust pre{\hypertarget{ref-nowok2016}{}}%
Nowok, Beata, Gillian M. Raab, and Chris Dibben. 2016.
{``{\emph{Synthpop}:} Bespoke Creation of Synthetic Data in
{\emph{R}}.''} \emph{Journal of Statistical Software} 74 (11).
\url{https://doi.org/10.18637/jss.v074.i11}.

\leavevmode\vadjust pre{\hypertarget{ref-obermeyer2019}{}}%
Obermeyer, Ziad, Brian Powers, Christine Vogeli, and Sendhil
Mullainathan. 2019. {``Dissecting Racial Bias in an Algorithm Used to
Manage the Health of Populations.''} \emph{Science} 366 (6464): 447--53.
\url{https://doi.org/10.1126/science.aax2342}.

\leavevmode\vadjust pre{\hypertarget{ref-patki2016}{}}%
Patki, Neha, Roy Wedge, and Kalyan Veeramachaneni. 2016. {``The
Synthetic Data Vault.''} \emph{2016 IEEE International Conference on
Data Science and Advanced Analytics (DSAA)}, October.
\url{https://doi.org/10.1109/dsaa.2016.49}.

\leavevmode\vadjust pre{\hypertarget{ref-reiter2005}{}}%
Reiter, Jerome P. 2005. {``Using CART to Generate Partially Synthetic
Public Use Microdata.''} \emph{Journal of Official Statistics} 21 (3):
441--62.
\url{https://www.scb.se/contentassets/ca21efb41fee47d293bbee5bf7be7fb3/using-cart-to-generate-partially-synthetic-public-use-microdata.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-rubin_statistical_1993}{}}%
Rubin, Donald B. 1993. {``Statistical Disclosure Limitation.''}
\emph{Journal of Official Statistics} 9 (2): 461--68.
\url{https://www.scb.se/contentassets/ca21efb41fee47d293bbee5bf7be7fb3/discussion-statistical-disclosure-limitation2.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-snoke_utility_2018}{}}%
Snoke, Joshua, Gillian M. Raab, Beata Nowok, Chris Dibben, and
Aleksandra Slavkovic. 2018. {``General and Specific Utility Measures for
Synthetic Data.''} \emph{Journal of the Royal Statistical Society.
Series A (Statistics in Society)} 181 (3): pp. 663--688.
\url{https://doi.org/10.1111/rssa.12358}.

\leavevmode\vadjust pre{\hypertarget{ref-Torkzadehmahani2019}{}}%
Torkzadehmahani, Reihaneh, Peter Kairouz, and Benedict Paten. 2019.
{``{DP-CGAN}: Differentially Private Synthetic Data and Label
Generation.''} In \emph{Proceedings of the IEEE/CVF Conference on
Computer Vision and Pattern Recognition (CVPR) Workshops}.
\url{https://doi.org/10.1109/cvprw.2019.00018}.

\leavevmode\vadjust pre{\hypertarget{ref-vandewiel2023}{}}%
van de Wiel, Mark A., Gwenaël G. R. Leday, Jeroen Hoogland, Martijn W.
Heymans, Erik W. van Zwet, and Ailko H. Zwinderman. 2023. {``Think
Before You Shrink: Alternatives to Default Shrinkage Methods Can Improve
Prediction Accuracy, Calibration and Coverage.''}
\url{https://doi.org/10.48550/ARXIV.2301.09890}.

\leavevmode\vadjust pre{\hypertarget{ref-volker2021}{}}%
Volker, Thom Benjamin, and Gerko Vink. 2021. {``Anonymiced Shareable
Data: Using Mice to Create and Analyze Multiply Imputed Synthetic
Datasets.''} \emph{Psych} 3 (4): 703--16.
\url{https://doi.org/10.3390/psych3040045}.

\leavevmode\vadjust pre{\hypertarget{ref-willenborg_elements_2012}{}}%
Willenborg, Leon, and Ton De Waal. 2012. \emph{Elements of Statistical
Disclosure Control}. Springer Science \& Business Media.
\url{https://doi.org/10.1007/978-1-4613-0121-9}.

\leavevmode\vadjust pre{\hypertarget{ref-xu_ctgan_2019}{}}%
Xu, Lei, Maria Skoularidou, Alfredo Cuesta-Infante, and Kalyan
Veeramachaneni. 2019. {``Modeling Tabular Data Using Conditional GAN.''}
In \emph{Advances in Neural Information Processing Systems}, edited by
H. Wallach, H. Larochelle, A. Beygelzimer, F. dAlché-Buc, E. Fox, and R.
Garnett. Vol. 32. Curran Associates, Inc.
\url{https://proceedings.neurips.cc/paper_files/paper/2019/file/254ed7d2de3b23ab10936522dd547b78-Paper.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-zettler2021}{}}%
Zettler, Ingo, Christoph Schild, Lau Lilleholt, Lara Kroencke, Till
Utesch, Morten Moshagen, Robert Böhm, Mitja D. Back, and Katharina
Geukes. 2021. {``The Role of Personality in COVID-19-Related
Perceptions, Evaluations, and Behaviors: Findings Across Five Samples,
Nine Traits, and 17 Criteria.''} \emph{Social Psychological and
Personality Science} 13 (1): 299--310.
\url{https://doi.org/10.1177/19485506211001680}.

\end{CSLReferences}

\end{document}

### 2024-04-26 paper/paper.qmd
Update paper

@@
     keep-tex: true
+    include-in-header:
+      - file: preamble.tex
   arxiv-html: default
+number-sections: true
 linenumbers: false
@@
     email: e.vankesteren1@uu.nl
-abstract: |
-  TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
-keywords: 
-  - TODO
-  - TODO
+abstract: TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
+keywords: |
+  synthetic data, utility, density ratio, privacy, disclosure limitation
 bibliography: bibliography.bib  
@@
 
-Openly accessible research data accelerates scientific progress tremendously [ADD REFERENCES].
-Open data allows third-party researchers to answer research questions with already collected data, substantially lowering the costs of research. 
-Sharing data in combination with code allows others to validate research findings and build upon the work.
-Students can benefit from open data, as it fosters education with realistic data, as well as the general public, through stimulating citizen science projects. 
-However, openly available research data is often at odds with privacy constraints on the dissemination of the data.
-The information may cause harm to individuals or organizations if made publicly available. 
-Additionally, sharing their information openly may withhold them from participating in future research.
-These constraints have been named among the biggest hurdles in the advancement of computational social science [@lazer_css_2009], and among top reasons for companies to not share their data with researchers [@fpf_2017].
-To overcome these obstacles, data providers can employ a suite of different disclosure limitation techniques before sharing data, for example top-coding, record-swapping or adding noise [e.g., @hundepool_disclosure_2012; @willenborg_elements_2012].
+Openly accessible research data accelerates scientific progress tremendously.
+Open data allows third-party researchers to answer research questions with already collected data, freeing up resources that would otherwise be devoted to data collection [@ramachandran_open_2021]. 
+Sharing data in combination with code allows others to validate research findings and build upon the work [@obels_analysis_2020; @crosas_automating_2015].
+Students can benefit from open data, as it fosters education with realistic data [@atenas_open_2015], as well as the general public, through stimulating citizen science projects [@newman_future_2012].
+However, making data openly available is often (rightfully) hampered by official legislation, like the General Data Protection Regulation [GDPR\; @gdpr], and general privacy concerns. 
+In the worst case, sharing data may cause harm to individuals or organizations, which may withhold these entities from participating in future research.
+These privacy constraints have been named among the biggest hurdles in the advancement of computational social science [@lazer_css_2009], and among top reasons for companies to not share their data with researchers [@fpf_2017].
 
-Recently, synthetic data has gained substantial traction as a means to disclosure limitation.
-National statistical institutes and other government agencies have adopted the synthetic data framework to facilitate the use of their data [e.g., @SIPP_Beta_2006; @hawala_synthetic_2008; @drechsler2012].
-Researchers started using synthetic data to train machine learning models [@nikolenko2021], and began to share synthetic versions of their research data to comply with open science standards [e.g., @vandewiel2023; @obermeyer2019; @zettler2021].
-The idea of synthetic data is to replace some, or all, of the observed values in a data set by synthetic records that are generated from some model [e.g., @little_statistical_1993; @rubin_statistical_1993; @drechsler2011synthetic].
+Multiple approaches exist to balance the benefits of open data with potential privacy risks. 
+Traditionally, data providers employed a suite of different disclosure limitation techniques before sharing the data, such as top-coding, record-swapping or adding noise [e.g., @hundepool_disclosure_2012; @willenborg_elements_2001].
+More recently, synthetic data has gained traction as a means to disseminate private data [@SIPP_Beta_2006; @hawala_synthetic_2008; @drechsler2012; @vandewiel2023; @obermeyer2019; @zettler2021], although the conceptual framework traces back to the previous century [@little_statistical_1993; @rubin_statistical_1993].
+Simply put, the idea of synthetic data is to replace some, or all, of the observed values in a data set by synthetic values that are generated from some model [e.g., @drechsler2011synthetic].
 If only some values are replaced, disclosure risks can be reduced because the sensitive or identifying values do not correspond to their true values anymore. 
 If all values are replaced, there is also no one-to-one mapping between the original and the synthetic data, further reducing the disclosure risk.
+However, an increase in privacy typically comes at the cost of a decrease in utility. 
+As more of the data is altered, the quality of the released data becomes more sensitive to the suitability of the generative model. 
+Regardless of the approach to disclosure limitation, if the technique used to generate or alter the data does not align with the intricacies of the problem at hand, the utility of the released data will be further reduced than necessary.
 
-Many different methods have been proposed to generate synthetic data.
-Traditionally, these were closely connected to methods used for multiple imputation of missing data, such as joint modelling [misdc2003], sequential regressions [@nowok2016] or fully conditional specification [@drechsler2011empirical; @volker2021].
-The flexibility of sequential regressions and fully conditional specification is commonly combined with non-parametric imputation models, such as classification and regression trees [@reiter2005], random forests [@caiola2010] or support vector machines [@drechsler2010], to avoid distributional assumptions and easily model non-linear relationships.
-Recently, significant improvements in generative modelling sparked the scientific interest in synthetic data in the computer science community, leading to novel synthesis methods [e.g., @patki2016; @xu_ctgan_2019].
-Combined with work on formal privacy guarantees such as differential privacy, this resulted in new models that explicitly control the level of privacy risk in synthesis methods [@jordon2018pategan; @Torkzadehmahani2019].
+Given that all disclosure limitation techniques reduce the utility of the data, the challenge that arises is how to determine whether the released data still has acceptable utility.
+Alternatively, one might consider different disclosure limitation techniques that all satisfy the defined privacy restrictions, and employ the one which yields data with the highest utility.
+That is, given a set of methods that all meet the privacy restrictions, one may aim to maximize the utility of the released data. 
+Both strategies require a reliable and encompassing measure of data utility that allows to evaluate the quality of the released data, and that allows to compare different disclosure limitation techniques and/or synthesis models in terms of utility.
+Moreover, adequate utility measures often guide the synthesis process, by providing detailed feedback on important discrepancies between the original and synthetic data.
+Lastly, good utility measures help the data user in determining what the synthetic data can and cannot be used for.
 
-___TODO___
+In the synthetic data field, three classes of utility measures have been distinguished [see @drechsler2023 for a thorough review]: fit-for-purpose measures, analysis-specific utility measures and global utility measures.
+Fit-for-purpose measures are often the first step in assessing the quality of the synthetic data. 
+These typically involve comparing the univariate distributions of the observed and synthetic data (for example using visualization techniques or goodness-of-fit measures). 
+Although these measures provide an initial impression of the quality of the synthesis models used, this picture is by definition limited, because only one or two variables are assessed at the same time.
+Hence, complex relationships between variables will always be out of scope.
+Such relationships may be captured by analysis-specific utility measures, which quantify whether analyses on synthetic data provide results that are comparable to results from the same analysis performed on the observed data.
+These measures can, for example, evaluate how similar the coefficients of a regression model are [e.g., using the confidence interval overlap\; @karr_utility_2006], or whether prediction models trained on the synthetic and observed data perform comparably in terms of evaluation metrics.
+However, analysis-specific utility generally does not carry over: high specific utility for one analysis does not at all imply high utility for another analysis. 
+Since data providers typically do not know which analyses will be performed with the synthetic data, it is impossible to provide analysis-specific utility measures for all potentially relevant analyses [see also @drechsler_utility_2022].
 
-All methods for statistical disclosure limitation alter the data before these are provided to the public.
-By doing so, the utility of the provided data is always lower than the utility of the original data, because some of the information in the data is sacrificed to protect the privacy of the respondents.
-The questions that naturally arise are how much information in the original data is actually sacrificed, and how useful the provided data are?
-Answering this question allows researchers to decide what the altered data can and cannot be used for, and to evaluate the worth of conclusions drawn on the basis of these data.
-After all, inferences from the altered data are valid only up to the extent that the perturbation methods approximate the true data-generating mechanism.
-For data providers, a detailed assessment of the quality of the altered data can guide the procedure of altering the data. 
-Statistical disclosure limitation is often an iterative process: some disclosure limitation technique is applied on the data, after which the result is investigated and modifications are made to applied process.
-Good measures of data quality are essential to determine the appropriate mechanisms used to protect the data, and can help to improve the utility of the data that will be disseminated.
+Global utility measures may overcome the shortcomings of the previous approaches, as they evaluate the discrepancy between the entire multivariate distribution of the observed and synthetic data.
+As such, global utility measures yields the most promising class of utility measures, because if the observed and synthetic data have similar (multivariate) distributions, all potential analyses should yield similar results.
+Global utility can be evaluated using some divergence measure, such as the Kullback-Leibler divergence [@karr_utility_2006], or by evaluating whether the observed and synthetic data are distinguishable using a classification model [a technique called $pMSE$\; @Woo_global_2009; @snoke_utility_2018].
+However, a common critique of global utility measures is that they tend to be too general [@drechsler_utility_2022].
+That is, analyses on a synthetic data set that is overall quite similar to the observed data (i.e., has high global utility), may still yield results that are far from the results obtained from the real data.
+Also, commonly used methods for estimating the $pMSE$, as logistic regression and classification and regression trees, tend to become less reliable as the dimensionality of the data increases, and are vulnerable to model misspecification [@drechsler_utility_2022].
+Lastly, the output of global utility measures can be hard to interpret, and say little about the regions in which the synthetic data do not resemble the true data accurately enough.
 
-In the statistical disclosure control literature, two different branches of utility measures have been distinguished: specific utility measures and general utility measures. 
-_Add one/two sentences on the merit of visualization when assessing utility of altered data._
-Specific utility measures focus on similarity of results obtained from analyses performed on the altered data and the original data. 
-For example, after fitting the same analysis model on both data sets, one can calculate the confidence interval overlap of the estimated parameters [@karr_utility_2006]. 
-Alternative measures are ellipsoidal overlap [@karr_utility_2006], which extends to confidence interval overlap to a measure that addresses the joint distribution of all model parameters simultaneous, the standardized absolute difference between estimates [@snoke_utility_2018], and the ratio of estimates for tabular count data [@taub2020impact].
-As these measures quantify similarity between estimates from analyses performed on the observed and altered data, they are informed only to the extent that data users will recreate those analyses. 
-This can be highly useful if the data is provided for reproducibility purposes (e.g., for third parties to evaluate analysis scripts).
-However, the goal of distributing the protected data is often to allow researchers to do novel research with the data. 
-In many practical situations, data providers thus have have only limited knowledge on the analyses that will be performed with the altered data. 
-Covering the entire set of potentially relevant analyses is therefore not feasible.
-If it was, the data providers could simply report the (potentially privacy-protected) results of those analyses performed on the real data, so that access to the (perturbed) data no longer yields additional benefits [for a similar argument, see @drechsler_utility_2022].
-Additionally, similarity between results on the analyses that have been performed gives no guarantee that the results will also be similar for other analyses.
-Hence, when determining how useful the altered data is for novel research, specific utility measures are only of limited use.
+To overcome the issues related to traditional global utility measures, we propose to use the density ratio estimation framework [@sugiyama_suzuki_kanamori_2012] as a way of evaluating utility. 
+Intuitively, if two data sets have similar multivariate distributions, the density ratio should be close to one over the range of the data. 
+If the distributions of the observed and synthetic data are very different, the density ratio should be far from one at those regions where the distributions differ.
+As density estimation is known to be a difficult problem, the density ratio estimation framework provides techniques to directly estimate the density ratio, rather than the two separate densities, in a non-parametric way [@sugiyama_suzuki_kanamori_2012].
+These non-parametric estimation techniques come with automatic model specification, which mitigates the issue of model specification. 
+This functionality is implemented in the `R`-package `densityratio` [@densityratio].
+Importantly, the density ratio is estimated over the entire range of the data, which provides measures of utility at every (possible) point in the data space.
+This point-specific quantification of utility turns out to be a useful side-product, as it allows to reweigh analyses on synthetic data when further improving the utility directly is not possible.
 
-General utility measures attempt to capture how similar the multivariate distributions of the observed and altered data are.
-This can be done by, for instance, estimating the Kullback-Leibler divergence between the distributions of the observed and altered data [@karr_utility_2006].
-An alternative strategy is to try to discriminate between the observed and altered data, as is done with the $pMSE$ [@snoke_utility_2018; @woo_utility_2009].
-In essence, the $pMSE$ quantifies how well one can predict whether observations are from the observed or the altered data. 
-If better one can do this, the more pronounced the differences between the observed and altered data ought to be. 
-However, various authors have criticized general utility measures for being too broad. 
-That is, important discrepancies between the real and altered data might be missed, and an altered data set that is good in general (i.e., has high global utility) might still provide results that are far from the truth for some analyses [see, e.g., @drechsler_utility_2022].
-Additionally, it is not straightforward to determine which prediction model to use for calculating the $pMSE$. 
-Specifying a good prediction model in itself may be a challenging task, especially when the number of variables is large. 
-When good models are available, different models, or even different choices of hyperparameters, may yield different results, potentially rendering ambiguity with respect to which altered data set is best.
-Lastly, the output of global utility measures can be hard to interpret, and say little about the regions in which the synthetic data do not resemble the true data accurately enough.
-That is, they give little guidance on how the quality of the altered data can be improved.
+In the remainder of the article, we introduce the density ratio framework and the associated estimation techniques, and connect the framework to traditional utility measures as the $pMSE$ and the Kullback-Leibler divergence.
+We then present simulations to demonstrate the performance of density ratio estimation in stylized settings and compare it to traditional utility measures.
+Subsequently, we apply density ratio estimation in a case study where we evaluate the utility of multiple synthetic versions of the U.S. Current Population Survey.
+We conclude with a discussion of the results, highlight the strengths and weaknesses of the density ratio framework, and provide recommendations for future research.
 
----
+# Background
 
-__Section 6: our contribution__
+Over the years, many methods have been introduced to generate synthetic data, all with the aim of providing a suitable balance between privacy and utility.
+These methods can be relatively simple, such as a sequence of generalized linear models [e.g., @reiter_releasing_2004], or as complex as deep learning models with thousands of parameters [e.g., @xu_ctgan_2019], with many options in between.
+However, the complexity of the generation process is not necessarily a good indicator of the quality of the synthetic data.
+That is, relatively simple methods could still capture the most important aspects of the data that complex methods fail to capture (and vice versa).
+Hence, data providers typically do not know which synthesis method will provide the highest utility a priori, and might compare multiple synthesis strategies to determine which data set will be released.
+Good global utility measures can help in this process, by allowing to quantify the quality of the candidate synthetic data sets.^[We focus on global utility measures, because in many situations the data provider does not know which analysis will be performed with the synthetic data.
+If the data provider knows for which purposes the data will be used, analysis-specific utility measures may be more informative.]
+Moreover, such global utility measures may guide the synthesis process itself, if they provide sufficiently specific information about the degree of misfit of the synthetic data.
+In the upcoming section, we provide an overview of existing global utility measures, and introduce the density ratio framework as an encompassing approach to evaluating global utility.
 
-_Moet nog verder uitgewerkt worden_
+<!-- Many different methods have been proposed to generate synthetic data. -->
+<!-- Traditionally, these were closely connected to methods used for multiple imputation of missing data, such as joint modelling [misdc2003], sequential regressions [@nowok2016] or fully conditional specification [@drechsler2011empirical; @volker2021].
+-->
+<!-- The flexibility of sequential regressions and fully conditional specification is commonly combined with non-parametric imputation models, such as classification and regression trees [@reiter2005], random forests [@caiola2010] or support vector machines [@drechsler2010], to avoid distributional assumptions and easily model non-linear relationships.
+-->
+<!-- Recently, significant improvements in generative modelling sparked the scientific interest in synthetic data in the computer science community, leading to novel synthesis methods [e.g., @patki2016; @xu_ctgan_2019].
+-->
+<!-- Combined with work on formal privacy guarantees such as differential privacy, this resulted in new models that explicitly control the level of privacy risk in synthesis methods [@jordon2018pategan; @Torkzadehmahani2019].
+-->
+
+## Existing global utility measures
+
+Global utility measures typically attempt to quantify the distributional similarity between the observed and synthetic data samples.
+The intuition is that if two data sets have similar distributions, the data sets can be used for the same purposes, and analyses on the two data sets should yield similar results.
+A common way to formalize distributional similarity is through the Kullback-Leibler (KL) divergence, as proposed in @karr_utility_2006.
+The KL-divergence measures the relative entropy from the probability distribution of the observed data $\pobs(\bx)$ to the probability distribution of the synthetic data $\psyn(\bx)$ (with $\bx \in \mathbb{R}^{n \times d}$), and is defined as
+$$
+D_{\text{KL}}(\psyn || \pobs) = 
+\int \psyn(\bx) \log \frac{\psyn(\bx)}{\pobs(\bx)} \text{d} \bx.
+$${#eq-kl-div}
+@karr_utility_2006 argue that the KL-divergence can be constructed from density estimators $\hat{p}_\text{obs}$ and $\hat{p}_\text{syn}$, after which the integral can be approximated using numerical quadrature.
+Alternatively, if the observed and synthetic data are both (multivariate) normally distributed, the KL-divergence can be computed analytically.
+Both implementations might be unfeasible: the combination of density estimation with numerical quadrature is cumbersome in high dimensional settings, and assuming multivariate normality might be too restrictive.
+Yet, we show in later sections that the density ratio estimation framework gives rives to an alternative way of estimating the KL-divergence that overcomes these challenges.
+
+An alternative way to evaluate whether two distributions are statistically indistinguishable is by evaluating whether a classification model can tell samples from the two distributions apart [see @kim_classification_2021, who formalize the connection between classification accuracy and two-sample testing].
+In the context of synthetic data, this implies that if a classification model can distinguish observed from synthetic samples, the distributional similarity is low, and so is the global utility. 
+If a classifier cannot distinguish between the observed and synthetic data, one would conclude that the global utility is high.
+The propensity score mean-squared error ($pMSE$), introduced by @Woo_global_2009 and further developed in @snoke_utility_2018, formalizes this intuition. 
+Let $I_i$ denote an indicator variable that equals $1$ if observation $i$ ($i \in 1, \dots, N$, $N = \nobs + \nsyn$) belongs to the synthetic data, and $0$ otherwise.
+We then train a classifier that outputs the predicted probability of observation $i$ being a synthetic record $\hat{s}_i$ based on the observation's scores on the variables (this can be the set of all variables, but also a subset).
+From these, we can calculate the utility statistic 
+$$
+pMSE = \frac{1}{N} \sum_{i=1}^N \Big(\hat{s}_i - \frac{\nsyn}{N}\Big)^2,
+$$ {#eq-pMSE}
+which ought to be smaller when the synthetic data is more like the observed data.
+Crucially, the $pMSE$ depends on the classification model used and increases in the flexibility of the classification model, making it prone to overfitting and hard to interpret.
+To combat these issues, @snoke_utility_2018 suggests to compare the $pMSE$-value with its expectation under the null distribution.
+Provided that the classification model is a logistic regression model with $k$ parameters (including the intercept), @snoke_utility_2018 show that the expected $pMSE$ is given by
+$$
+\mathbb{E}[pMSE] = \Big(\frac{k-1}{N}\Big) \Big(\frac{\nobs}{N}\Big)^2
+\Big(\frac{\nsyn}{N}\Big).
+$$
+For other classification models, the expectation can be approximated through a resampling procedure. 
+Accordingly, the $pMSE$-ratio is given by
+$$
+pMSE\text{-ratio} = \frac{pMSE}{\mathbb{E}[pMSE]},
+$$
+with values smaller than $10$ deemed acceptable [@raab2021assessing\; although the authors remark that the smaller the better].
+Apart from the $pMSE$, several other measures exist that can be constructed from the estimated propensity scores, such as the percentage of records correctly predicted [@raab2021assessing] or the Kolmogorov-Smirnov statistic [@Bowen_differentially_2021], both which are strongly correlated with the $pMSE$ [@raab2021assessing].
+
+Due to its intuitive nature, multiple studies advice the use of the $pMSE$ as a promising technique to evaluate the quality of synthetic data [e.g., @raab2017guidelines; @raab2021assessing; @hu_advancing_2024].
+Yet, it is not free of criticism. 
+The usefulness of the $pMSE$ hinges on choosing a model that can capture the important intricacies of the observed data. 
+@drechsler_utility_2022 illustrated that the utility score is highly dependent on the model used to estimate the propensity scores, and that clear improvements in synthesis models are not necessarily picked up in the $pMSE$.
+Moreover, the $pMSE$ is prone to overfitting and users may find it difficult to select an appropriate model for estimating the propensity scores.
+Hence, current methods to evaluate global utility are either difficult to estimate, especially in high-dimensional situations, or depend strongly on model specification.
+Density ratio estimation might alleviate these issues.
+In what follows, we explain how the density ratio can be estimated, and how it can be used to evaluate the utility of synthetic data.
+
+## Density ratio estimation
+
+The density ratio estimation framework was originally developed in the machine learning community for the comparison of two probability distributions [for an overview, see @sugiyama_suzuki_kanamori_2012].
+The framework has been shown to be applicable to prediction [@sugiyama_conditional_2010; @sugiyama_classification_2010], outlier detection [@shohei_dre_outlier_2008], change-point detection in time-series [@liu_change_2013], importance weighting under domain adaptation [i.e., sample selection bias\; @kanamori_ulsif_2009], and two-sample homogeneity tests [@sugiyama_lstst_2011].
+The general idea of density ratio estimation is depicted in @fig-dr-plot, and boils down to comparing two distributions by modelling the density ratio $r(\bx)$ between the probability distributions of the numerator samples, taken from the synthetic data distribution, $\psyn(\bx)$, and the denominator samples, taken from the observed data distribution, $\pobs(\bx)$, such that
+$$
+r(\bx) = \frac{\psyn(\bx)}{\pobs(\bx)}.
+$$ {#eq-dr}
+
+This specification has the intuitive interpretation that in locations where the density ratio takes large values, too many synthetic observations will be generated in that region and at the locations where the density ratio is small, there will be too few synthetic observations, both relative to the observed data.
+An intuitive approach to estimating $r(\bx)$ from samples of $\pobs(\bx)$ and $\psyn(\bx)$ would be to estimate the observed and synthetic data density separately, for example using kernel density estimation [e.g., see @Scott1992 for an overview], and subsequently compute the ratio of these estimated densities.
+However, density estimation is one of the hardest tasks in statistical learning, unavoidably leading to estimation errors for both densities, especially in high dimensions.
+When subsequently taking the ratio of the estimated densities, estimation errors tend to be magnified. 
+Direct density ratio estimation avoids this issue by specifying and estimating a model directly for the ratio without first estimating the separate densities.
+Extensive simulations on a wide variety of tasks showed that this approach typically outperforms density ratio estimation through naive kernel density estimation, especially when the dimensionality of the data increases [e.g., @Kanamori2012; @shohei_dre_outlier_2008; @kanamori_ulsif_2009].
+
+```{r}
+#| label: fig-dr-plot
+#| echo: false
+#| warning: false
+#| message: false
+#| fig-dpi: 500
+#| fig-cap: "Example of the true and estimated density ratio of two normal distributions with different means and variances (i.e., $\\psyn(\\bx) = N(0,1)$ and $\\pobs(\\bx) = N(1,2)$).
+The function $r(\\bx) = \\psyn(\\bx)/\\pobs(\\bx)$ denotes the true density ratio, the function $\\hat{r}(\\bx)$ denotes an estimate of the density ratio based on $\\nsyn = \\nobs = 200$ samples from each distribution obtained with unconstrained Least-Squares Importance Fitting (uLSIF).
+Note that the density ratio is itself not a proper density."
+#| out-width: 100%
+#| fig-height: 1.8
+#| fig-pos: t
+#| dev: cairo_pdf
+
+library(ggplot2)
+library(patchwork)
+extrafont::loadfonts(quiet = TRUE)
+
+set.seed(12)
+nu <- rnorm(200, 0, 1)
+de <- rnorm(200, 1, sqrt(2))
+rhat <- densityratio::ulsif(nu, de)
+
+ggplot() +
+  stat_function(aes(col = "A"), 
+                fun = dnorm, args = list(mean = 0, sd = 1),
+                size = 0.5) +
+  stat_function(aes(col = "B"),
+                fun = dnorm, args = list(mean = 1, sd = sqrt(2)),
+                size = 0.5) +
+  scale_color_brewer(labels = c("A" = expression(italic(p)[syn](x)),
+                                "B" = expression(italic(p)[obs](x))),
+                     palette = "Set2",
+                     name = "") +
+  xlim(-4, 6) +
+  ylim(0, 0.5) +
+  ylab("Density") +
+  theme_minimal() +
+  theme(legend.position = "bottom",
+        text = element_text(family = "LM Roman 10"),
+        legend.background = element_blank()) +
+ggplot() +
+  stat_function(aes(col = "C"),
+                fun = ~ dnorm(.x, 0, 1) / dnorm(.x, 1, sqrt(2)),
+                size = 0.5) +
+  stat_function(aes(col = "D"),
+                fun = ~ predict(rhat, newdata = .x),
+                size = 0.5) +
+  scale_color_manual(labels = c("C" = expression(italic(r(x))),
+                                "D" = expression(italic(hat(r)(x)))),
+                     values = c("black", "steelblue3"),
+                     name = "") +
+  xlim(-4, 6) +
+  ylim(0, 2.5) +
+  theme_minimal() +
+  ylab("Density ratio") +
+  theme(legend.position = "bottom",
+        text = element_text(family = "LM Roman 10"))
+```
+
+### Estimating the density ratio
+
+Over the past years, several methods for direct density ratio estimation have been developed.
+A large class of these methods attempt to directly minimize the error between the true density ratio $r(\bx)$ and its estimate $\hat{r}(\bx)$.
+Following this approach, we define a loss function $\mathcal{L}(r(\bx), \hat{r}(\bx))$ that measures the discrepancy between the true and estimated density ratio. 
+To give an example, consider the following loss based on the squared error
+$$
+\begin{aligned}
+\mathcal{L}_S(r(\bx), \hat{r}(\bx)) &= 
+\frac{1}{2} \int (\hat{r}(\bx) - r(\bx))^2 \pobs(\bx) \text{d}\bx \\
+&= \frac{1}{2} \int \hat{r}(\bx)^2 \pobs(\bx) \text{d}\bx - 
+\int \hat{r}(\bx) r(\bx) \pobs(\bx) \text{d}\bx + 
+\frac{1}{2} \int r(\bx)^2 \pobs(\bx) \text{d}\bx.
+\end{aligned}
+$$ {#eq-squared-error}
+The second term can be rewritten, because the denominator in $r(\bx)$ cancels with $\pobs(\bx)$, while the third term is a constant with respect to the parameters in the density ratio model and can thus be ignored.
+Hence, we are left with the following loss function to minimize
+$$
+\mathcal{L}'_S (r(\bx), \hat{r}(\bx)) = \frac{1}{2} \int \hat{r}(\bx)^2 \pobs(\bx) \text{d}\bx - \int \hat{r}(\bx) \psyn(\bx) \text{d}\bx,
+$$ {#eq-squared-error-loss}
+which is termed least squares importance fitting (LSIF) by @kanamori_ulsif_2009.
+
+This approach can be straightforwardly generalized to a wide class of loss functions that fall under the family of Bregman divergences [see @sugiyama_bregman_2012; @mohamed2017learning]. 
+Then, a general class of losses is encompassed by the expression
+$$
+\mathcal{L}_f (r(\bx), \hat{r}(\bx)) =
+\int \Big(
+f(r(\bx)) - f(\hat{r}(\bx)) - f'(\hat{r}(\bx))(r(\bx) - \hat{r}(\bx))
+\Big) \pobs(\bx) \text{d}\bx,
+$$ {#eq-bregman-loss}
+where $f$ is a differentiable and strictly convex function with derivative $f'$.
+Then, ignoring the terms independent of $\hat{r}(\bx)$ and noting that $r(\bx)\pobs(\bx) = \psyn(\bx)$, we obtain the following objective
+$$
+\mathcal{L}'_f (r(\bx), \hat{r}(\bx)) = \int 
+\Big( 
+f'(\hat{r}(\bx)) \hat{r}(\bx) - f(\hat{r}(\bx))
+\Big)  \pobs(\bx) \text{d}\bx
+- \int f'(\hat{r}(\bx)) \psyn(\bx)
+\text{d}\bx.
+$$ {#eq-bregman-objective}
+Minimizing this loss for different functions $f$ yields different estimators for the density ratio, that focus on different regions of the density ratio.^[It is easy to see that using $f(x) = \frac{1}{2}(x-1)^2$ turns the Bregman divergence (@eq-bregman-objective) into the squared error (@eq-squared-error-loss).]
+Specifically, some estimators place more emphasis on accurately modelling the regions in which the true density ratio is large, but less emphasis on accurately estimating the smaller density ratios, and vice versa [see @sugiyama_bregman_2012; @menon2016dreloss].
+Importantly, @sugiyama_bregman_2012 show that the Bregman divergence minimization approach to density ratio estimation is equivalent to estimating the $f$-divergence [@ali_silvey_divergence_1966] between the synthetic and observed data distributions.
+Hence, when estimating the density ratio, one implicitly estimates the divergence between the synthetic and observed data distributions.
+
+After defining a loss function, we need a model for the density ratio function $\hat{r}(\bx)$. 
+There are many possibilities to specify this model, but a common choice is to use a linear model for the density ratio [e.g., @huang_kmm_2006; @kanamori_ulsif_2009; @izbicki_dre_2014; @gruber2024overcoming].
+That is, we define the density ratio model as
+$$
+\hat{r}(\bx) = \boldsymbol{\varphi}(\bx) \boldsymbol{\theta},
+$$ {#eq-dr-model}
+where $\boldsymbol{\varphi}(\bx)$ is a basis function vector, that transforms the data from a $p$-dimensional to a $b$-dimensional space, and $\boldsymbol{\theta}$ is a $b$-dimensional parameter vector.
+Although the model is linear in its parameters, the density ratio itself is typically a non-linear function of the data due to the basis functions.
+The parameter vector $\boldsymbol{\theta}$ is estimated to minimize the discrepancy with the true density ratio using the loss function $\mathcal{L}(r(\bx), \hat{r}(\bx))$ in @eq-bregman-objective.
+
+Also for the basis functions, several specifications are possible, ranging from an identity function [@qin_inferences_1998] to normalizing flows [@choi_featurized_2021] or neural networks [@tiao2018dre; @uehara2016generative].
+However, a commonly used basis function in the density ratio literature is the Gaussian kernel function [e.g., @huang_kmm_2006; @sugiyama_kliep_2007; @kanamori_ulsif_2009; @liu_change_2013; @gruber2024overcoming], which conveniently allows to model the non-linear density ratio using models that are linear in their parameters.
+The Gaussian kernel function is defined as
+$$
+\boldsymbol{\varphi}(\bx) = \mathcal{K}(\bx_i, \boldsymbol{c}_j) = 
+\exp\left(
+- \frac{\lVert\bx - \boldsymbol{c}_j \rVert^2}{2 \sigma^2}
+\right),
+$$
+where $\boldsymbol{c}_j$ ($j \in 1, \dots, J$) denotes the centers of the Gaussian kernel functions, and $\sigma$ controls the kernel width [i.e., it defines over which distance differences between the observations and the centers are considered relevant\; for more information on kernel functions, see, e.g., @murphy_pmlintro_2022].
+The appropriate width of the kernel can be determined using cross-validation. 
+The centers are typically sampled from the data (in our case, the observed and synthetic data can both be used, but it is also possible to solely use the synthetic samples).
+
+After defining the model and the loss function, the density ratio can be easily estimated. 
+The `densityratio` package in R [@densityratio] provides an implementation of commonly used loss functions with a Gaussian kernel basis function. 
+The package comes with an easy-to-use interface, automatic cross-validation of hyperparameters and builds on `C++` for fast computation. 
+However, for increased flexibility with respect to loss functions and basis functions, one could use classic optimization techniques as gradient descent or (quasi-)Newton methods, for example as implemented in the `optim` function in the statistical software `R` [@R].
+Both approaches are briefly illustrated in Appendix [-@sec-app-A] (TODO).
+
+### Evaluating data utility with the density ratio
+
+After estimating the density ratio, the information provided by the density ratio can be summarized in a divergence measure, using the fact that estimating the density ratio is equivalent to estimating the $f$-divergence between the synthetic and observed data distributions [@sugiyama_bregman_2012].
+Accordingly, we can use the estimated density ratio function to estimate an $f$-divergence between the observed and synthetic data, and use this divergence measure as a utility measure for the synthetic data.
+Although this divergence statistic is difficult to interpret in an absolute sense, it can be used as a relative measure of quality of the synthetic data.
+That is, for different synthetic data sets, we can calculate the divergence to the observed data, and compare these values to determine which synthetic data set is most similar to the observed data.
+In a more formal way, the $f$-divergence gives rise to a test statistic that can be used to test the null hypothesis that the synthetic data is generated from the same distribution as the observed data.
+The corresponding $p$-value can be calculated by comparing the observed test statistic to the null distribution of test statistics obtained using random permutations of the observed and synthetic data [see, for example, @sugiyama_lstst_2011; @Wornowizki2016; @kanamori_divergence_2012].
+Lastly, the density ratio function itself can be plotted against individual variables or pairs of variables to discover whether particular variables are particularly poorly captured in the synthetic data.
+Each of these strategies can be readily executed using the `densityratio` package in `R` [@densityratio], and will be further illustrated in the upcoming sections.
+
+**Ik twijfel nog een beetje over de volgende zaken, en voordat ik alles ga zitten uitwerken hoor ik graag eerst jullie mening:**
 
 1.
-We introduce density ratio estimation to the field of statistical disclosure control.
-Short remark on the idea that density ratio estimation is a complicated endeavor, especially if the goal is to compare distinct densities.
-Having to estimate just a single density (ratio) is generally much easier.
+Ik heb nu het deel over regularisatie eruit gelaten, omdat ik vond dat het wat afleid van de main ideas, maar kan deze ook gewoon toevoegen aan @eq-bregman-objective.
+Het kan ook casual opgemerkt worden wanneer het daadwerkelijk gebruikt wordt.
+Wat vinden jullie?
 
 2.
-Note that density ratio estimation can capture specific and general utility measures into a common framework by being applicable on the level of the entire data, but also on the subset of variables that is relevant in an analysis.
-Additionally, note that confidence interval overlap, ellipsoidal overlap, but also $pMSE$ and Kullback-Leibler divergence, are closely related to density ratio estimation, and can be considered from this perspective.
+Dimension reduction: ik was niet zo zeker hoe ik dit er direct moest bouwen, want effectief is het niet veel anders dan de huidige aanpak, alleen met een andere basis function (namelijk $\phi(x) = \mathcal{K}(x_i^T U, c_j^T U)$), waarbij U ook parameters heeft die geschat worden.
+Anyway, ik kan aan het eind een sectie maken met dat DRE nog steeds lastig kan zijn in hoge dimensies, en dat daar oplossingen voor zijn (bijvoorbeeld een LFDA-stap voor ulsif, of LHSS, of the spectral density ratio estimation methode (laatste twee zitten allebei in het package)).
+Het punt is vooral dat ik niet wil dat het teveel afleidt van het main punt, namelijk density ratio estimation, maar qua content zou het misschien een eigen sectie verdienen, ook omdat het niet direct enorm aansluit bij de twee secties hierboven.
 
-3. Create a new utility metric based on density ratio estimation (probability with respect to some reference distribution as in permutation testing).
+<!-- The difficulty of density ratio estimation increases with the dimensionality of the data.  -->
+<!-- Therefore, we follow previous recommendations to incorporate dimensionality reduction techniques in density ratio estimation.  -->
+
+<!-- Shortly name examples of dimensionality reduction techniques (i.e., PCA; LFDA or UMAP). -->
+
+<!-- A useful by-product of dimension reduction is that it allows to create visualizations, and these visualizations can be used to get more insight in discrepancies between observed and altered data.
+-->
+<!-- Show what such visualizations can look like, and how they can help. -->
+
+3.
+Denken jullie dat dit een goede plek is om wat dieper in te gaan op mogelijke voordelen van density ratio estimation?
+Ik dacht dat na sectie 2.2.2 een voor de hand liggende plek zou zijn om hier wat dieper op in te gaan, en dan specifiek de volgende punten te benoemen: utility quantifications at every point in multivariate space, easy and flexible model specification through automatic model selection with cross-validation, potential for importance weighting, wellicht dan ook direct het punt van dimension reduction hierbij pakken.
 
 4.
-Because density ratio estimation can be difficult when there are many variables, we use dimension reduction techniques to capture most of the variability in the data in fewer dimensions on which density ratio estimation can be applied.
-A by-product of this is that the lower-dimensional subspace allows to create visualizations of deviations from the observed data.
+Ik wil graag nog ergens iets zeggen over de relaties tussen density ratio estimation en de andere methodes (KL-divergence, pMSE), maar ik vraag me een beetje af of dat niet beter in de discussie kan.
 
-5. Perform a simulation study to give indications about which methods to use (think about how to do this).
+<!-- Relate density ratio estimation to $pMSE$ and KL divergence (to some extent, both are generalizations of density ratio estimation, or at least are conceptually similar). -->
+<!-- Give some more information on the $pMSE$, describe what it shortcomings are.  -->
+<!-- The quality of the $pMSE$ highly depends on the model used to calculate the propensity scores.  -->
+<!-- Perhaps give an example of logistic regression, which basically estimates whether the conditional mean of the observed and altered data is the same.  -->
+<!-- Explain how density ratio estimation can overcome the shortcomings of the previously mentioned methods. -->
 
-6. Implement all this in an R-package
+5.
+Is dit echt de goede plaats om iets te zeggen over categorische variabelen?
+Deze kunnen namelijk gewoon geïncludeerd worden, maar met een Gaussian kernel wordt dat misschien een beetje awkward.
+Desalniettemin kan het prima, en kunnen we er ook voor kiezen om in de discussie te benoemen dat we een oplossing hebben gekozen, maar dat er ook andere opties zijn (bijv.
+andere kernels).
 
-__Section 6: outline article__
+# Numerical illustrations: Simulation studies
 
-In the next section, we describe density ratio estimation and discuss how this method can be used as to measure utility.
-Subsequently, we provide multiple examples that show how density ratio estimation works in the context of evaluating the quality of synthetic data. 
-Hereafter, we show in multiple simulations that the method is superior (HOPEFULLY) to current global utility measures as the $pMSE$. 
-Lastly, we discuss the advantages and disadvantages of density ratio estimation as a utility measure.
+To evaluate the performance of density ratio estimation and compare it to existing methods, we conduct a series of simulation studies. 
+We consider a variety of scenarios, including univariate and multivariate models, varying correlations, number of variables, and sample sizes.
+To perform density ratio estimation, we use unconstrained Least-Squares Importance Fitting [uLSIF\; @kanamori_ulsif_2009] as implemented in the `densityratio` package in `R` [@densityratio], because it fast and has been shown to perform well in a variety of settings [ADD CITATIONS <!-- TODO: ADD CITATIONS-->].
+We compare the performance of uLSIF to the performance of the $p$-MSE as implemented in the `R`-package `synthpop`.
 
-# Methodology
+## Univariate simulations
 
-\@Peter-Paul: Eventueel een korte beschrijving van data perturbation techniques/synthetic data generation hier. Denk je dat dit wat toevoegt hier?
+For the sake of illustrational clarity, we first evaluate the performance of density ratio estimation in a univariate setting where its performance can be evaluated using visualizations.
+In the simulations, we generate data according to four data-generating mechanisms, (1) a Laplace distribution, (2) a log-normal distribution, (3) a location-scale $t$-distribution and  (4) a normal distribution.
+Subsequently, we approximate the true data generating mechanism using a Gaussian model with the same mean and variance as the original data. 
+This setting is similar to situations that are commonly encountered in the synthetic data field, in the sense that the true data generating mechanism is unknown and needs to be approximated using a simpler approximating model.
+The exact specifications of the data generating mechanisms are as follows:
 
-__Section 1: density ratio estimation__
+1. Laplace distribution with location parameter $\mu = 1$ and scale parameter $b = 1$.
+2. Log-normal distribution with log-mean parameter $\mu_{\text{log}} = \log \{1/\sqrt{3} \}$ and log-standard deviation parameter $\sigma_\text{log} = \sqrt{\log 3}$.
+3. Location-scale $t$-distribution with location parameter $\mu = 1$, scale parameter $\tau^2 = 1$ and degrees of freedom $\nu = 4$. 
+4. Normal distribution with mean $\mu = 1$ and variance $\sigma^2 = 2$.
 
-In essence, the goal of utility measures is to quantify the similarity between the multivariate distribution of the observed data with the distribution of the altered data. 
-If the used data perturbation techniques, or synthetic data generation models, approximate the distribution of the real data sufficiently, these distributions should be highly similar, and analyses on the two data sets should give similar results.
-However, estimating the probability distribution of a data set is known to be one of the most complicated challenges in statistics [E.G. Vapnik 1998].
-Estimating the probability distribution for both observed and altered data can lead to errors in both, artificially magnifying discrepancies between the two. 
-Hence, subsequent comparisons will be affected by these errors. 
-The procedure can be simplified by using density ratio estimation, because this only requires to estimate a single density.
+These data generating mechanisms are chosen such that they all have the same population mean $\mu = 1$ and variance $\sigma^2 = 2$.
+The approximating model is a Gaussian distribution with mean $\mu = 1$ and variance $\sigma^2 = 2$.
+This distribution has the same mean and variance, but differs in higher-order moments, except in the fourth scenario. 
+In the last simulation, we model the data generating mechanism correctly to obtain some understanding into how density ratio estimation behaves under a well-specified synthesis model.
+In all scenarios, we generate $500$ data sets with $\nobs = 250$ observations from the true data generating mechanism, and $\nsyn = 250$ synthetic observations from the approximating Gaussian model. 
+The density ratio model is estimated using uLSIF with a Gaussian kernel and an $L_2$-regularization parameter.
+Both the kernel bandwidth and the regularization parameter are selected using cross-validation, using the default settings as implemented in the `densityratio` package.
+
+```{r}
+#| label: load-functions
+#| eval: true
+#| include: false
+
+source("../code/functions.R")
+```
+
+```{r}
+#| label: fig-densities-sim1
+#| eval: true
+#| echo: false
+#| warning: false
+#| message: false
+#| fig-dpi: 500
+#| fig-pos: t
+#| out-width: 100%
+#| fig-height: 4
+#| dev: cairo_pdf
+#| fig-cap: "True and synthetic data densities for the four simulations with Laplace, Log-normal, location-scale $t$- and Normal densities.
+All data-generating mechanisms have the same mean $\\mu = 1$ and variance $\\sigma^2 = 2$.
+Note that the true and synthetic data density in the bottom right panel are completely overlapping."
+
+mu <- 1
+sigma <- sqrt(2)
+
+library(ggplot2)
+library(patchwork)
+library(tidyr)
+library(dplyr)
+library(purrr)
+
+ggplot() +
+  geom_function(aes(col = "A"), 
+                fun = dlaplace, args = list(mu = mu, sigma = sigma),
+                size = 0.5) +
+  geom_function(aes(col = "B"), 
+                fun = dnorm, args = list(mean = mu, sd = sigma),
+                size = 0.5) +
+  xlim(-3, 5) +
+  ylim(0, 1.2) +
+  scale_color_brewer(labels = c("A" = "Laplace", "B" = "Normal"),
+                     palette = "Set2",
+                     name = "") +
+  ylab("Density") +
+  theme_minimal() +
+  theme(legend.position = "bottom",
+        text = element_text(family = "LM Roman 10"),
+        legend.background = element_blank()) +
+  ggplot() +
+  geom_function(aes(col = "A"), 
+                fun = dlnorm, args = list(meanlog = log(mu^2 / sqrt(mu^2 + sigma^2)), 
+                                          sdlog = sqrt(log(1 + sigma^2/mu^2))),
+                size = 0.5) +
+  geom_function(aes(col = "B"), 
+                fun = dnorm, args = list(mean = mu, sd = sigma),
+                size = 0.5) +
+  xlim(-3, 5) +
+  ylim(0, 1.2) +
+  scale_color_brewer(labels = c("A" = "Log-normal", "B" = "Normal"),
+                     palette = "Set2",
+                     name = "") +
+  ylab("Density") +
+  theme_minimal() +
+  theme(legend.position = "bottom",
+        text = element_text(family = "LM Roman 10"),
+        legend.background = element_blank()) +
+  ggplot() +
+  geom_function(aes(col = "A"), 
+                fun = ~dt(.x - mu, df = 2 * sigma^2 / (sigma^2 - 1)),
+                size = 0.5) +
+  geom_function(aes(col = "B"), 
+                fun = dnorm, args = list(mean = mu, sd = sigma),
+                size = 0.5) +
+  xlim(-3, 5) +
+  ylim(0, 1.2) +
+  scale_color_brewer(labels = c("A" = expression(italic(lst)), "B" = "Normal"),
+                     palette = "Set2",
+                     name = "") +
+  ylab("Density") +
+  theme_minimal() +
+  theme(legend.position = "bottom",
+        text = element_text(family = "LM Roman 10"),
+        legend.background = element_blank()) +
+  ggplot() +
+  geom_function(aes(col = "A"), 
+                fun = dnorm, args = list(mean = mu, sd = sigma),
+                size = 0.5) +
+  geom_function(aes(col = "B"), 
+                fun = dnorm, args = list(mean = mu, sd = sigma),
+                size = 0.5) +
+  xlim(-3, 5) +
+  ylim(0, 1.2) +
+  scale_color_brewer(labels = c("A" = "Normal", "B" = "Normal"),
+                     palette = "Set2",
+                     name = "") +
+  ylab("Density") +
+  theme_minimal() +
+  theme(legend.position = "bottom",
+        legend.background = element_blank(),
+        text = element_text(family = "LM Roman 10"))
+```
+
+```{r}
+#| label: code-sim1
+#| eval: false
+#| include: false
+
+source("../code/sim1.R")
+```
+
+```{r}
+#| label: fig-sim1-results
+#| eval: true
+#| echo: false
+#| warning: false
+#| message: false
+#| fig-dpi: 500
+#| fig-pos: t
+#| out-width: 100%
+#| fig-height: 4
+#| dev: cairo_pdf
+#| fig-cap: "Estimated density ratios by unconstrained least-squares importance fitting in four univariate examples: A Laplace distribution, a log-normal distribution, a location-scale $t$-distribution and a normal distribution, all approximated by a normal distribution with the same mean and variance as the sample from the true distribution."
+
+load("../results/sim1.RData")
+
+# evaluation data for plotting
+x_eval <- seq(-3, 5, length.out = 1600)
+
+# Create figure
+sims |>
+  mutate(model = factor(model,
+                        levels = c("laplace", "lognormal", "t", "normal"),
+                        labels = c("Laplace", "Log-normal", "italic(t)", "Normal"),
+                        ordered = TRUE),
+         xpreds = list(x_eval),
+         ypreds = map2(r, xpreds, ~predict(.x, .y))) |>
+  unnest(cols = c(xpreds, ypreds)) |>
+  ggplot(aes(x = xpreds, y = ypreds, group = interaction(model, sim))) +
+  geom_line(alpha = 0.1, col = "steelblue3") +
+  geom_line(data = true_ratio, aes(x = xpreds, y = ypreds, group = NULL), col = "black") +
+  facet_wrap(~model, labeller = label_parsed) +
+  ylim(0, 6) +
+  theme_minimal() +
+  xlab(expression(italic(x))) +
+  ylab(expression(hat(italic(r))(x)))
+```
+
+TODO: Add interpretation of output; add significance tests and compare with $pMSE$ and kolmogorov-smirnov.
+
+## Multivariate simulations
+
+TODO: Adjust section title
+
+Currently, my plan is as follows, specify a (two/four) true data generating mechanism:
+A multivariate normal distribution with a given correlation structure, 7 and 17 variables each for two different sample size (say n = 500 and n = 2000). 
+Append the multivariate normal distribution with three variables with a different distribution that depend on the other variables through some non-linear function.
+Have three synthetic data generating mechanisms, a simple multivariate normal distribution with the same means and variances but zero covariances, a model that uses a multivariate normal distribution with correlation structure taken from the real data but misses the non-linear effects (and marginal distributions of these variables), and a model that is equivalent to the true model.
+Then, evaluate the quality of the synthetic data sets using density ratio estimation and the $pMSE$, and see which of the two ranks the models correctly most of the time.
+
+# Application: Synthetic data generation for the U.S. Current Population Survey
+
+ADD TEXT
+
+```{r}
+#| label: fig-application-distributions
+#| eval: true
+#| echo: false
+#| warning: false
+#| message: false
+#| fig-dpi: 500
+#| fig-pos: t
+#| out-width: 100%
+#| fig-height: 5
+#| dev: cairo_pdf
+#| fig-cap: "Real and synthetic data distributions for the variables age, household income (income), household property taxes (tax) and social security benefits (social security) on a cubic root scale."
+
+load("../results/application.RData")
+
+comb_df <- bind_rows(
+  Real = df,
+  Naive = bind_rows(synlist$unadj$syn), 
+  Transformed = bind_rows(synlist$trans$syn),
+  `Semi-continuous` = bind_rows(synlist$semi$syn),
+  `Smoothed CART` = bind_rows(synlist$smoothed$syn),
+  .id = "Data"
+) |>
+  select(Data,
+         Age = age, 
+         Income = income, 
+         Tax = tax, 
+         `Social security` = ss) |>
+  mutate(Data = factor(Data, levels = c("Real", "Naive", "Transformed", "Semi-continuous", "Smoothed CART")),
+         RealSyn = ifelse(Data == "Real", 1, 2) |> factor(labels = c("Real", "Synthetic")),
+         across(c(Age, Income, Tax, `Social security`), ~abs(.x)^{1/3} * sign(.x))) |>
+  tidyr::pivot_longer(cols = c(Age, Income, Tax, `Social security`), names_to = "Variable")
+
+purrr::map(c("Age", "Income", "Social security", "Tax"), ~
+      ggplot(comb_df |> filter(Variable == .x), aes(x = value, fill = RealSyn, after_stat(density))) +
+      geom_histogram(col = "black", bins = 20) +
+      scale_fill_brewer(palette = "Set2") + 
+      facet_wrap(~Data, ncol = 5) + 
+      theme_minimal() +
+      ylab(.x) +
+      theme(legend.position = "none", 
+            axis.title.x = element_blank(), 
+            strip.text.x = element_text(size = 8),
+            text = element_text(family = "LM Roman 10"))) |>
+  patchwork::wrap_plots(nrow = 5)
+```
+
+```{r}
+#| label: fig-application-pearson-div
+#| eval: true
+#| echo: false
+#| warning: false
+#| message: false
+#| fig-dpi: 500
+#| fig-pos: t
+#| out-width: 100%
+#| fig-height: 2.5
+#| dev: cairo_pdf
+#| fig-cap: "Pearson divergence estimates after different synthesis strategies for the separate variables and the synthetic data as a whole."
+
+bind_rows(PE, PE_allvars) |>
+  select(-c(cart, smoothed_semi)) |>
+  mutate(variable = factor(1:5, labels = c("Age", "Income", "Social security", "Tax", "All"))) |>
+  pivot_longer(cols = c(unadj, trans, semi, smoothed)) |>
+  mutate(`Synthesis method` = factor(name, levels = c("unadj", "trans", "semi", "smoothed"),
+                                     labels = c("Naïve", "Transformed", "Semi-continuous", "Smoothed CART"))) |>
+  ggplot(aes(x = variable, y = value, col = `Synthesis method`, shape = `Synthesis method`, 
+             group = `Synthesis method`)) +
+  geom_point(size = 3) +
+  # geom_line(alpha = 0.1, linetype = "dashed") +
+  scale_y_log10() +
+  theme_minimal() +
+  scale_color_brewer(palette = "Set1") +
+  ylab("Pearson divergence") +
+  xlab("Variable") +
+  theme(legend.position = "bottom",
+        text = element_text(family = "LM Roman 10", size = 9))
+```
+
+TODO: Expand example
+
+High-dimensional density ratio estimation
+
+Use density ratio values for weighted analysis
+
+Incorporate point on individual data utility
+
+- High dimensional example
+- Expansion of discussion points:
+	empirical example with weighted analyses
+	individual data utility
+
+<!-- __Section 1: density ratio estimation__ -->
+
+<!-- However, estimating the probability distribution of a data set is known to be one of the most complicated challenges in statistics [E.G. Vapnik 1998]. -->
+<!-- Estimating the probability distribution for both observed and altered data can lead to errors in both, artificially magnifying discrepancies between the two.  -->
+<!-- Hence, subsequent comparisons will be affected by these errors.  -->
+<!-- The procedure can be simplified by using density ratio estimation, because this only requires to estimate a single density. -->
 
@@
 
-Introduce density ratio estimation as a utility measure. 
-What does this measure mean/how to interpret it. 
-How to make decisions based on this measure.
+# Discussion and conclusion
 
-Say something on whether (and if so, how) categorical variables can be incorporated as well.
+Adapt from previous version
 
-__Section 3: theoretical comparison with conventional approaches for general utility assessment__
-
-Relate density ratio estimation to specific and general utility measures.
-Pick one/two specific utility measures and relate these to density ratio estimation (ratio of estimates seems straightforward, as well as confidence interval overlap and ellipsoidal overlap).
-
-Relate density ratio estimation to $pMSE$ and KL divergence (to some extent, both are generalizations of density ratio estimation, or at least are conceptually similar).
-Give some more information on the $pMSE$, describe what it shortcomings are. 
-The quality of the $pMSE$ highly depends on the model used to calculate the propensity scores. 
-Perhaps give an example of logistic regression, which basically estimates whether the conditional mean of the observed and altered data is the same. 
-Explain how density ratio estimation can overcome the shortcomings of the previously mentioned methods.
-
-__Section 4: Dimension reduction for utility__
-
-The difficulty of density ratio estimation increases with the dimensionality of the data. 
-Therefore, we follow previous recommendations to incorporate dimensionality reduction techniques in density ratio estimation.
-
-Shortly name examples of dimensionality reduction techniques (i.e., PCA; LFDA or UMAP).
-
-A useful by-product of dimension reduction is that it allows to create visualizations, and these visualizations can be used to get more insight in discrepancies between observed and altered data. 
-Show what such visualizations can look like, and how they can help.
-
-# Simulations
-
-## Small illustration / example  with multivariate Gaussian distributions.
-
-1. Simple, multivariate normal simulation (e.g., two correlation structures, two sample sizes, so $2 \times 2$ full factorial design); basically similar to what we did already.
-
-## More complex simulation, more variables, non-linearities, perhaps using real data.
-
-2. More advanced simulation (e.g., some non-linearities, different sample sizes)
-
-Have to think about this in more detail still.
-
-# Real data example
-
-Clinical records heart-failure data? Misschien ook niet, nog over nadenken.
-
-Exemplify how utility measures could (should!) be used to improve the quality of the altered data (e.g., illustrate how models can be adjusted iteratively based on utility assessment).
-
-_Some notes to self_
-
-Current ways to assess the utility?
-
-- pMSE - logistic, regression, CART models (Snoke, Raab, Nowok, Dibben & Slavkovic, 2018; General and specific utility measures for synthetic data AND Woo, Reiter, Oganian & Karr, 2009; Global measures of data utlity for microdata masked for disclosure limitation)
-
-- Kullback-Leiber divergence (Karr, Kohnen, Oganian, Reiter & Sanil, 2006; A framework for evaluating the utility of data altered to protect confidentiality).
-
-- According to multiple authors, both specific and general utility measures have important drawbacks (see Drechsler Utility PSD; cites others).
-Narrow measures potentially focus on analyses that are not relevant for the end user, and do not generalize to the analyses that are relevant.
-Global utility measures are generally too broad, and important deviations in the synthetic data might be missed.
-Moreover, the measures are typically hard to interpret.
-
-- See Drechsler for a paragraph on fit for purpose measures, that lie between general and specific utility measures (i.e., plausibility checks such as non-negativity; goodness of fit measures as $\chi^2$ for cross-tabulations; Kolmogorov-Smirnov).
-
-- Drechsler also illustrates that the standardized $pMSE$ has substantial flaws, as the results are highly dependent on the model used to estimate the propensity scores, and unable to detect important differences in the utility for most of the model specifications.
-Hence, it is claimed that a thorough assessment of utility is required.
-
-Things to add in new version:
-- High dimensional example
-- Expansion of discussion points:
-	empirical example with weighted analyses
-	individual data utility
 - Privacy remark?
@@
 
-# Methodology
+Density ratios can be used directly to train generative models, see:
+[Mohamed & Lakshminarayanan (2016)](https://arxiv.org/pdf/1610.03483) and [Uehara, Sata, Suzuki, Nakayama & Matsuo (2016)](https://arxiv.org/pdf/1610.02920.pdf)
 
-TO DO
+Density ratio estimation is equivalent to class probability estimation in classification problems, only differing in the loss function that is employed (assuming the same model class is used for density ratio estimation and classification).
+See [Menon & Ong, 2016](https://proceedings.mlr.press/v48/menon16.pdf)
 
-# Simulations
-
-TO DO
-
-# Real data example
-
-TO DO
-
-# Results
-
-TO DO
-
-# Discussion and conclusion
-
-TO DO
+Logistic regression achieves the minimum asymptotic variance for correctly specified models (Qin, Biometrika 1998), but is not reliable for misspecified models (Kanamori, Suzuki, Sugiyama, IECE, 2010)
 
@@
 :::
+
+\setcounter{section}{0}
+\renewcommand{\thesection}{\Alph{section}}
+
+\setcounter{table}{0}
+\renewcommand{\thetable}{A\arabic{table}}
+
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{A\arabic{figure}}
+
+# Density ratio estimation in `R` {#sec-app-A}
+
+TO DO
+
+Small example of ulsif in `densityratio`
+
+Do the same with `optim`
+
+Do the same for a different loss function

### 2024-04-26 paper/paper.tex
Update paper

@@
 
+\newcommand{\Xobs}{\mathbf{X}_{\text{obs}}}
+\newcommand{\Xsyn}{\mathbf{X}_{\text{syn}}}
+\newcommand{\dist}{\mathcal{P}}
+\newcommand{\pobs}{p_{\text{obs}}}
+\newcommand{\psyn}{p_{\text{syn}}}
+\newcommand{\nobs}{n_{\text{obs}}}
+\newcommand{\nsyn}{n_{\text{syn}}}
+\newcommand{\bx}{\mathbf{x}}
 \usepackage{arxiv}
@@
   pdfauthor={Thom Benjamin Volker; Peter-Paul de Wolf; Erik-Jan van Kesteren},
-  pdfkeywords={TODO, TODO},
+  pdfkeywords={synthetic data, utility, density ratio, privacy,
+disclosure limitation},
   colorlinks=true,
@@
 TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
-TODO TODO TODO
+TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
+TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
 \end{abstract}
@@
 \def\sep{\textbullet\ }
-TODO \sep 
-TODO
-
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable, frame hidden]}{\end{tcolorbox}}\fi
+synthetic data, utility, density ratio, privacy, disclosure
+limitation \sep 
+synthetic data, utility, density ratio, privacy, disclosure limitation
+
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, enhanced]}{\end{tcolorbox}}\fi
 
@@
 Openly accessible research data accelerates scientific progress
-tremendously {[}ADD REFERENCES{]}. Open data allows third-party
-researchers to answer research questions with already collected data,
-substantially lowering the costs of research. Sharing data in
-combination with code allows others to validate research findings and
-build upon the work. Students can benefit from open data, as it fosters
-education with realistic data, as well as the general public, through
-stimulating citizen science projects. However, openly available research
-data is often at odds with privacy constraints on the dissemination of
-the data. The information may cause harm to individuals or organizations
-if made publicly available. Additionally, sharing their information
-openly may withhold them from participating in future research. These
-constraints have been named among the biggest hurdles in the advancement
-of computational social science
+tremendously. Open data allows third-party researchers to answer
+research questions with already collected data, freeing up resources
+that would otherwise be devoted to data collection
+(\protect\hyperlink{ref-ramachandran_open_2021}{Ramachandran, Bugbee,
+and Murphy 2021}). Sharing data in combination with code allows others
+to validate research findings and build upon the work
+(\protect\hyperlink{ref-obels_analysis_2020}{Obels et al. 2020};
+\protect\hyperlink{ref-crosas_automating_2015}{Crosas et al. 2015}).
+Students can benefit from open data, as it fosters education with
+realistic data (\protect\hyperlink{ref-atenas_open_2015}{Atenas,
+Havemann, and Priego 2015}), as well as the general public, through
+stimulating citizen science projects
+(\protect\hyperlink{ref-newman_future_2012}{Newman et al. 2012}).
+However, making data openly available is often (rightfully) hampered by
+official legislation, like the General Data Protection Regulation (GDPR;
+\protect\hyperlink{ref-gdpr}{European Parliament and Council of the
+European Union 2016}), and general privacy concerns. In the worst case,
+sharing data may cause harm to individuals or organizations, which may
+withhold these entities from participating in future research. These
+privacy constraints have been named among the biggest hurdles in the
+advancement of computational social science
 (\protect\hyperlink{ref-lazer_css_2009}{Lazer et al. 2009}), and among
 top reasons for companies to not share their data with researchers
-(\protect\hyperlink{ref-fpf_2017}{Future of Privacy Forum 2017}). To
-overcome these obstacles, data providers can employ a suite of different
-disclosure limitation techniques before sharing data, for example
-top-coding, record-swapping or adding noise (e.g.,
+(\protect\hyperlink{ref-fpf_2017}{Future of Privacy Forum 2017}).
+
+Multiple approaches exist to balance the benefits of open data with
+potential privacy risks. Traditionally, data providers employed a suite
+of different disclosure limitation techniques before sharing the data,
+such as top-coding, record-swapping or adding noise (e.g.,
 \protect\hyperlink{ref-hundepool_disclosure_2012}{Hundepool et al.
-2012}; \protect\hyperlink{ref-willenborg_elements_2012}{Willenborg and
-De Waal 2012}).
-
-Recently, synthetic data has gained substantial traction as a means to
-disclosure limitation. National statistical institutes and other
-government agencies have adopted the synthetic data framework to
-facilitate the use of their data (e.g.,
-\protect\hyperlink{ref-SIPP_Beta_2006}{Abowd, Stinson, and Benedetto
+2012}; \protect\hyperlink{ref-willenborg_elements_2001}{Willenborg and
+De Waal 2001}). More recently, synthetic data has gained traction as a
+means to disseminate private data
+(\protect\hyperlink{ref-SIPP_Beta_2006}{Abowd, Stinson, and Benedetto
 2006}; \protect\hyperlink{ref-hawala_synthetic_2008}{Hawala 2008};
-\protect\hyperlink{ref-drechsler2012}{Drechsler 2012}). Researchers
-started using synthetic data to train machine learning models
-(\protect\hyperlink{ref-nikolenko2021}{Nikolenko 2021}), and began to
-share synthetic versions of their research data to comply with open
-science standards (e.g., \protect\hyperlink{ref-vandewiel2023}{van de
-Wiel et al. 2023}; \protect\hyperlink{ref-obermeyer2019}{Obermeyer et
-al. 2019}; \protect\hyperlink{ref-zettler2021}{Zettler et al. 2021}).
-The idea of synthetic data is to replace some, or all, of the observed
-values in a data set by synthetic records that are generated from some
-model (e.g., \protect\hyperlink{ref-little_statistical_1993}{Little
-1993}; \protect\hyperlink{ref-rubin_statistical_1993}{Rubin 1993};
-\protect\hyperlink{ref-drechsler2011synthetic}{Drechsler 2011}). If only
-some values are replaced, disclosure risks can be reduced because the
-sensitive or identifying values do not correspond to their true values
-anymore. If all values are replaced, there is also no one-to-one mapping
-between the original and the synthetic data, further reducing the
-disclosure risk.
-
-Many different methods have been proposed to generate synthetic data.
-Traditionally, these were closely connected to methods used for multiple
-imputation of missing data, such as joint modelling {[}misdc2003{]},
-sequential regressions (\protect\hyperlink{ref-nowok2016}{Nowok, Raab,
-and Dibben 2016}) or fully conditional specification
-(\protect\hyperlink{ref-drechsler2011empirical}{Drechsler and Reiter
-2011}; \protect\hyperlink{ref-volker2021}{Volker and Vink 2021}). The
-flexibility of sequential regressions and fully conditional
-specification is commonly combined with non-parametric imputation
-models, such as classification and regression trees
-(\protect\hyperlink{ref-reiter2005}{Reiter 2005}), random forests
-(\protect\hyperlink{ref-caiola2010}{Caiola and Reiter 2010}) or support
-vector machines (\protect\hyperlink{ref-drechsler2010}{Drechsler 2010}),
-to avoid distributional assumptions and easily model non-linear
-relationships. Recently, significant improvements in generative
-modelling sparked the scientific interest in synthetic data in the
-computer science community, leading to novel synthesis methods (e.g.,
-\protect\hyperlink{ref-patki2016}{Patki, Wedge, and Veeramachaneni
-2016}; \protect\hyperlink{ref-xu_ctgan_2019}{Xu et al. 2019}). Combined
-with work on formal privacy guarantees such as differential privacy,
-this resulted in new models that explicitly control the level of privacy
-risk in synthesis methods
-(\protect\hyperlink{ref-jordon2018pategan}{Jordon, Yoon, and Schaar
-2019}; \protect\hyperlink{ref-Torkzadehmahani2019}{Torkzadehmahani,
-Kairouz, and Paten 2019}).
-
-\textbf{\emph{TODO}}
-
-All methods for statistical disclosure limitation alter the data before
-these are provided to the public. By doing so, the utility of the
-provided data is always lower than the utility of the original data,
-because some of the information in the data is sacrificed to protect the
-privacy of the respondents. The questions that naturally arise are how
-much information in the original data is actually sacrificed, and how
-useful the provided data are? Answering this question allows researchers
-to decide what the altered data can and cannot be used for, and to
-evaluate the worth of conclusions drawn on the basis of these data.
-After all, inferences from the altered data are valid only up to the
-extent that the perturbation methods approximate the true
-data-generating mechanism. For data providers, a detailed assessment of
-the quality of the altered data can guide the procedure of altering the
-data. Statistical disclosure limitation is often an iterative process:
-some disclosure limitation technique is applied on the data, after which
-the result is investigated and modifications are made to applied
-process. Good measures of data quality are essential to determine the
-appropriate mechanisms used to protect the data, and can help to improve
-the utility of the data that will be disseminated.
-
-In the statistical disclosure control literature, two different branches
-of utility measures have been distinguished: specific utility measures
-and general utility measures. \emph{Add one/two sentences on the merit
-of visualization when assessing utility of altered data.} Specific
-utility measures focus on similarity of results obtained from analyses
-performed on the altered data and the original data. For example, after
-fitting the same analysis model on both data sets, one can calculate the
-confidence interval overlap of the estimated parameters
-(\protect\hyperlink{ref-karr_utility_2006}{Karr et al. 2006}).
-Alternative measures are ellipsoidal overlap
-(\protect\hyperlink{ref-karr_utility_2006}{Karr et al. 2006}), which
-extends to confidence interval overlap to a measure that addresses the
-joint distribution of all model parameters simultaneous, the
-standardized absolute difference between estimates
-(\protect\hyperlink{ref-snoke_utility_2018}{Snoke et al. 2018}), and the
-ratio of estimates for tabular count data
-(\protect\hyperlink{ref-taub2020impact}{\textbf{taub2020impact?}}). As
-these measures quantify similarity between estimates from analyses
-performed on the observed and altered data, they are informed only to
-the extent that data users will recreate those analyses. This can be
-highly useful if the data is provided for reproducibility purposes
-(e.g., for third parties to evaluate analysis scripts). However, the
-goal of distributing the protected data is often to allow researchers to
-do novel research with the data. In many practical situations, data
-providers thus have have only limited knowledge on the analyses that
-will be performed with the altered data. Covering the entire set of
-potentially relevant analyses is therefore not feasible. If it was, the
-data providers could simply report the (potentially privacy-protected)
-results of those analyses performed on the real data, so that access to
-the (perturbed) data no longer yields additional benefits (for a similar
-argument, see \protect\hyperlink{ref-drechsler_utility_2022}{Drechsler
-2022}). Additionally, similarity between results on the analyses that
-have been performed gives no guarantee that the results will also be
-similar for other analyses. Hence, when determining how useful the
-altered data is for novel research, specific utility measures are only
-of limited use.
-
-General utility measures attempt to capture how similar the multivariate
-distributions of the observed and altered data are. This can be done by,
-for instance, estimating the Kullback-Leibler divergence between the
-distributions of the observed and altered data
-(\protect\hyperlink{ref-karr_utility_2006}{Karr et al. 2006}). An
-alternative strategy is to try to discriminate between the observed and
-altered data, as is done with the \(pMSE\)
-(\protect\hyperlink{ref-snoke_utility_2018}{Snoke et al. 2018};
-\protect\hyperlink{ref-woo_utility_2009}{\textbf{woo\_utility\_2009?}}).
-In essence, the \(pMSE\) quantifies how well one can predict whether
-observations are from the observed or the altered data. If better one
-can do this, the more pronounced the differences between the observed
-and altered data ought to be. However, various authors have criticized
-general utility measures for being too broad. That is, important
-discrepancies between the real and altered data might be missed, and an
-altered data set that is good in general (i.e., has high global utility)
-might still provide results that are far from the truth for some
-analyses (see, e.g.,
+\protect\hyperlink{ref-drechsler2012}{Drechsler 2012};
+\protect\hyperlink{ref-vandewiel2023}{van de Wiel et al. 2023};
+\protect\hyperlink{ref-obermeyer2019}{Obermeyer et al. 2019};
+\protect\hyperlink{ref-zettler2021}{Zettler et al. 2021}), although the
+conceptual framework traces back to the previous century
+(\protect\hyperlink{ref-little_statistical_1993}{Little 1993};
+\protect\hyperlink{ref-rubin_statistical_1993}{Rubin 1993}). Simply put,
+the idea of synthetic data is to replace some, or all, of the observed
+values in a data set by synthetic values that are generated from some
+model (e.g., \protect\hyperlink{ref-drechsler2011synthetic}{Drechsler
+2011}). If only some values are replaced, disclosure risks can be
+reduced because the sensitive or identifying values do not correspond to
+their true values anymore. If all values are replaced, there is also no
+one-to-one mapping between the original and the synthetic data, further
+reducing the disclosure risk. However, an increase in privacy typically
+comes at the cost of a decrease in utility. As more of the data is
+altered, the quality of the released data becomes more sensitive to the
+suitability of the generative model. Regardless of the approach to
+disclosure limitation, if the technique used to generate or alter the
+data does not align with the intricacies of the problem at hand, the
+utility of the released data will be further reduced than necessary.
+
+Given that all disclosure limitation techniques reduce the utility of
+the data, the challenge that arises is how to determine whether the
+released data still has acceptable utility. Alternatively, one might
+consider different disclosure limitation techniques that all satisfy the
+defined privacy restrictions, and employ the one which yields data with
+the highest utility. That is, given a set of methods that all meet the
+privacy restrictions, one may aim to maximize the utility of the
+released data. Both strategies require a reliable and encompassing
+measure of data utility that allows to evaluate the quality of the
+released data, and that allows to compare different disclosure
+limitation techniques and/or synthesis models in terms of utility.
+Moreover, adequate utility measures often guide the synthesis process,
+by providing detailed feedback on important discrepancies between the
+original and synthetic data. Lastly, good utility measures help the data
+user in determining what the synthetic data can and cannot be used for.
+
+In the synthetic data field, three classes of utility measures have been
+distinguished (see \protect\hyperlink{ref-drechsler2023}{Drechsler and
+Haensch 2023} for a thorough review): fit-for-purpose measures,
+analysis-specific utility measures and global utility measures.
+Fit-for-purpose measures are often the first step in assessing the
+quality of the synthetic data. These typically involve comparing the
+univariate distributions of the observed and synthetic data (for example
+using visualization techniques or goodness-of-fit measures). Although
+these measures provide an initial impression of the quality of the
+synthesis models used, this picture is by definition limited, because
+only one or two variables are assessed at the same time. Hence, complex
+relationships between variables will always be out of scope. Such
+relationships may be captured by analysis-specific utility measures,
+which quantify whether analyses on synthetic data provide results that
+are comparable to results from the same analysis performed on the
+observed data. These measures can, for example, evaluate how similar the
+coefficients of a regression model are (e.g., using the confidence
+interval overlap; \protect\hyperlink{ref-karr_utility_2006}{Karr et al.
+2006}), or whether prediction models trained on the synthetic and
+observed data perform comparably in terms of evaluation metrics.
+However, analysis-specific utility generally does not carry over: high
+specific utility for one analysis does not at all imply high utility for
+another analysis. Since data providers typically do not know which
+analyses will be performed with the synthetic data, it is impossible to
+provide analysis-specific utility measures for all potentially relevant
+analyses (see also
 \protect\hyperlink{ref-drechsler_utility_2022}{Drechsler 2022}).
-Additionally, it is not straightforward to determine which prediction
-model to use for calculating the \(pMSE\). Specifying a good prediction
-model in itself may be a challenging task, especially when the number of
-variables is large. When good models are available, different models, or
-even different choices of hyperparameters, may yield different results,
-potentially rendering ambiguity with respect to which altered data set
-is best. Lastly, the output of global utility measures can be hard to
-interpret, and say little about the regions in which the synthetic data
-do not resemble the true data accurately enough. That is, they give
-little guidance on how the quality of the altered data can be improved.
-
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
-\textbf{Section 6: our contribution}
-
-\emph{Moet nog verder uitgewerkt worden}
+
+Global utility measures may overcome the shortcomings of the previous
+approaches, as they evaluate the discrepancy between the entire
+multivariate distribution of the observed and synthetic data. As such,
+global utility measures yields the most promising class of utility
+measures, because if the observed and synthetic data have similar
+(multivariate) distributions, all potential analyses should yield
+similar results. Global utility can be evaluated using some divergence
+measure, such as the Kullback-Leibler divergence
+(\protect\hyperlink{ref-karr_utility_2006}{Karr et al. 2006}), or by
+evaluating whether the observed and synthetic data are distinguishable
+using a classification model (a technique called \(pMSE\);
+\protect\hyperlink{ref-Woo_global_2009}{Woo et al. 2009};
+\protect\hyperlink{ref-snoke_utility_2018}{Snoke et al. 2018}). However,
+a common critique of global utility measures is that they tend to be too
+general (\protect\hyperlink{ref-drechsler_utility_2022}{Drechsler
+2022}). That is, analyses on a synthetic data set that is overall quite
+similar to the observed data (i.e., has high global utility), may still
+yield results that are far from the results obtained from the real data.
+Also, commonly used methods for estimating the \(pMSE\), as logistic
+regression and classification and regression trees, tend to become less
+reliable as the dimensionality of the data increases, and are vulnerable
+to model misspecification
+(\protect\hyperlink{ref-drechsler_utility_2022}{Drechsler 2022}).
+Lastly, the output of global utility measures can be hard to interpret,
+and say little about the regions in which the synthetic data do not
+resemble the true data accurately enough.
+
+To overcome the issues related to traditional global utility measures,
+we propose to use the density ratio estimation framework
+(\protect\hyperlink{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki,
+and Kanamori 2012a}) as a way of evaluating utility. Intuitively, if two
+data sets have similar multivariate distributions, the density ratio
+should be close to one over the range of the data. If the distributions
+of the observed and synthetic data are very different, the density ratio
+should be far from one at those regions where the distributions differ.
+As density estimation is known to be a difficult problem, the density
+ratio estimation framework provides techniques to directly estimate the
+density ratio, rather than the two separate densities, in a
+non-parametric way
+(\protect\hyperlink{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki,
+and Kanamori 2012a}). These non-parametric estimation techniques come
+with automatic model specification, which mitigates the issue of model
+specification. This functionality is implemented in the
+\texttt{R}-package \texttt{densityratio}
+(\protect\hyperlink{ref-densityratio}{Volker 2023}). Importantly, the
+density ratio is estimated over the entire range of the data, which
+provides measures of utility at every (possible) point in the data
+space. This point-specific quantification of utility turns out to be a
+useful side-product, as it allows to reweigh analyses on synthetic data
+when further improving the utility directly is not possible.
+
+In the remainder of the article, we introduce the density ratio
+framework and the associated estimation techniques, and connect the
+framework to traditional utility measures as the \(pMSE\) and the
+Kullback-Leibler divergence. We then present simulations to demonstrate
+the performance of density ratio estimation in stylized settings and
+compare it to traditional utility measures. Subsequently, we apply
+density ratio estimation in a case study where we evaluate the utility
+of multiple synthetic versions of the U.S. Current Population Survey. We
+conclude with a discussion of the results, highlight the strengths and
+weaknesses of the density ratio framework, and provide recommendations
+for future research.
+
+\hypertarget{background}{%
+\section{Background}\label{background}}
+
+Over the years, many methods have been introduced to generate synthetic
+data, all with the aim of providing a suitable balance between privacy
+and utility. These methods can be relatively simple, such as a sequence
+of generalized linear models (e.g.,
+\protect\hyperlink{ref-reiter_releasing_2004}{Reiter 2004}), or as
+complex as deep learning models with thousands of parameters (e.g.,
+\protect\hyperlink{ref-xu_ctgan_2019}{Xu et al. 2019}), with many
+options in between. However, the complexity of the generation process is
+not necessarily a good indicator of the quality of the synthetic data.
+That is, relatively simple methods could still capture the most
+important aspects of the data that complex methods fail to capture (and
+vice versa). Hence, data providers typically do not know which synthesis
+method will provide the highest utility a priori, and might compare
+multiple synthesis strategies to determine which data set will be
+released. Good global utility measures can help in this process, by
+allowing to quantify the quality of the candidate synthetic data
+sets.\footnote{We focus on global utility measures, because in many
+  situations the data provider does not know which analysis will be
+  performed with the synthetic data. If the data provider knows for
+  which purposes the data will be used, analysis-specific utility
+  measures may be more informative.} Moreover, such global utility
+measures may guide the synthesis process itself, if they provide
+sufficiently specific information about the degree of misfit of the
+synthetic data. In the upcoming section, we provide an overview of
+existing global utility measures, and introduce the density ratio
+framework as an encompassing approach to evaluating global utility.
+
+\hypertarget{existing-global-utility-measures}{%
+\subsection{Existing global utility
+measures}\label{existing-global-utility-measures}}
+
+Global utility measures typically attempt to quantify the distributional
+similarity between the observed and synthetic data samples. The
+intuition is that if two data sets have similar distributions, the data
+sets can be used for the same purposes, and analyses on the two data
+sets should yield similar results. A common way to formalize
+distributional similarity is through the Kullback-Leibler (KL)
+divergence, as proposed in Karr et al.
+(\protect\hyperlink{ref-karr_utility_2006}{2006}). The KL-divergence
+measures the relative entropy from the probability distribution of the
+observed data \(\pobs(\bx)\) to the probability distribution of the
+synthetic data \(\psyn(\bx)\) (with
+\(\bx \in \mathbb{R}^{n \times d}\)), and is defined as
+\begin{equation}\protect\hypertarget{eq-kl-div}{}{
+D_{\text{KL}}(\psyn || \pobs) = 
+\int \psyn(\bx) \log \frac{\psyn(\bx)}{\pobs(\bx)} \text{d} \bx.
+}\label{eq-kl-div}\end{equation} Karr et al.
+(\protect\hyperlink{ref-karr_utility_2006}{2006}) argue that the
+KL-divergence can be constructed from density estimators
+\(\hat{p}_\text{obs}\) and \(\hat{p}_\text{syn}\), after which the
+integral can be approximated using numerical quadrature. Alternatively,
+if the observed and synthetic data are both (multivariate) normally
+distributed, the KL-divergence can be computed analytically. Both
+implementations might be unfeasible: the combination of density
+estimation with numerical quadrature is cumbersome in high dimensional
+settings, and assuming multivariate normality might be too restrictive.
+Yet, we show in later sections that the density ratio estimation
+framework gives rives to an alternative way of estimating the
+KL-divergence that overcomes these challenges.
+
+An alternative way to evaluate whether two distributions are
+statistically indistinguishable is by evaluating whether a
+classification model can tell samples from the two distributions apart
+(see \protect\hyperlink{ref-kim_classification_2021}{Kim et al. 2021},
+who formalize the connection between classification accuracy and
+two-sample testing). In the context of synthetic data, this implies that
+if a classification model can distinguish observed from synthetic
+samples, the distributional similarity is low, and so is the global
+utility. If a classifier cannot distinguish between the observed and
+synthetic data, one would conclude that the global utility is high. The
+propensity score mean-squared error (\(pMSE\)), introduced by Woo et al.
+(\protect\hyperlink{ref-Woo_global_2009}{2009}) and further developed in
+Snoke et al. (\protect\hyperlink{ref-snoke_utility_2018}{2018}),
+formalizes this intuition. Let \(I_i\) denote an indicator variable that
+equals \(1\) if observation \(i\) (\(i \in 1, \dots, N\),
+\(N = \nobs + \nsyn\)) belongs to the synthetic data, and \(0\)
+otherwise. We then train a classifier that outputs the predicted
+probability of observation \(i\) being a synthetic record \(\hat{s}_i\)
+based on the observation's scores on the variables (this can be the set
+of all variables, but also a subset). From these, we can calculate the
+utility statistic \begin{equation}\protect\hypertarget{eq-pMSE}{}{
+pMSE = \frac{1}{N} \sum_{i=1}^N \Big(\hat{s}_i - \frac{\nsyn}{N}\Big)^2,
+}\label{eq-pMSE}\end{equation} which ought to be smaller when the
+synthetic data is more like the observed data. Crucially, the \(pMSE\)
+depends on the classification model used and increases in the
+flexibility of the classification model, making it prone to overfitting
+and hard to interpret. To combat these issues, Snoke et al.
+(\protect\hyperlink{ref-snoke_utility_2018}{2018}) suggests to compare
+the \(pMSE\)-value with its expectation under the null distribution.
+Provided that the classification model is a logistic regression model
+with \(k\) parameters (including the intercept), Snoke et al.
+(\protect\hyperlink{ref-snoke_utility_2018}{2018}) show that the
+expected \(pMSE\) is given by \[
+\mathbb{E}[pMSE] = \Big(\frac{k-1}{N}\Big) \Big(\frac{\nobs}{N}\Big)^2
+\Big(\frac{\nsyn}{N}\Big).
+\] For other classification models, the expectation can be approximated
+through a resampling procedure. Accordingly, the \(pMSE\)-ratio is given
+by \[
+pMSE\text{-ratio} = \frac{pMSE}{\mathbb{E}[pMSE]},
+\] with values smaller than \(10\) deemed acceptable
+(\protect\hyperlink{ref-raab2021assessing}{Gillian M. Raab, Nowok, and
+Dibben 2021}; although the authors remark that the smaller the better).
+Apart from the \(pMSE\), several other measures exist that can be
+constructed from the estimated propensity scores, such as the percentage
+of records correctly predicted
+(\protect\hyperlink{ref-raab2021assessing}{Gillian M. Raab, Nowok, and
+Dibben 2021}) or the Kolmogorov-Smirnov statistic
+(\protect\hyperlink{ref-Bowen_differentially_2021}{Bowen, Liu, and Su
+2021}), both which are strongly correlated with the \(pMSE\)
+(\protect\hyperlink{ref-raab2021assessing}{Gillian M. Raab, Nowok, and
+Dibben 2021}).
+
+Due to its intuitive nature, multiple studies advice the use of the
+\(pMSE\) as a promising technique to evaluate the quality of synthetic
+data (e.g., \protect\hyperlink{ref-raab2017guidelines}{Gillian M. Raab,
+Nowok, and Dibben 2017};
+\protect\hyperlink{ref-raab2021assessing}{Gillian M. Raab, Nowok, and
+Dibben 2021}; \protect\hyperlink{ref-hu_advancing_2024}{Hu and Bowen
+2024}). Yet, it is not free of criticism. The usefulness of the \(pMSE\)
+hinges on choosing a model that can capture the important intricacies of
+the observed data. Drechsler
+(\protect\hyperlink{ref-drechsler_utility_2022}{2022}) illustrated that
+the utility score is highly dependent on the model used to estimate the
+propensity scores, and that clear improvements in synthesis models are
+not necessarily picked up in the \(pMSE\). Moreover, the \(pMSE\) is
+prone to overfitting and users may find it difficult to select an
+appropriate model for estimating the propensity scores. Hence, current
+methods to evaluate global utility are either difficult to estimate,
+especially in high-dimensional situations, or depend strongly on model
+specification. Density ratio estimation might alleviate these issues. In
+what follows, we explain how the density ratio can be estimated, and how
+it can be used to evaluate the utility of synthetic data.
+
+\hypertarget{density-ratio-estimation}{%
+\subsection{Density ratio estimation}\label{density-ratio-estimation}}
+
+The density ratio estimation framework was originally developed in the
+machine learning community for the comparison of two probability
+distributions (for an overview, see
+\protect\hyperlink{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki,
+and Kanamori 2012a}). The framework has been shown to be applicable to
+prediction (\protect\hyperlink{ref-sugiyama_conditional_2010}{Sugiyama
+et al. 2010};
+\protect\hyperlink{ref-sugiyama_classification_2010}{Sugiyama 2010}),
+outlier detection (\protect\hyperlink{ref-shohei_dre_outlier_2008}{Hido
+et al. 2008}), change-point detection in time-series
+(\protect\hyperlink{ref-liu_change_2013}{Liu et al. 2013}), importance
+weighting under domain adaptation (i.e., sample selection bias;
+\protect\hyperlink{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama
+2009}), and two-sample homogeneity tests
+(\protect\hyperlink{ref-sugiyama_lstst_2011}{Sugiyama et al. 2011}). The
+general idea of density ratio estimation is depicted in
+Figure~\ref{fig-dr-plot}, and boils down to comparing two distributions
+by modelling the density ratio \(r(\bx)\) between the probability
+distributions of the numerator samples, taken from the synthetic data
+distribution, \(\psyn(\bx)\), and the denominator samples, taken from
+the observed data distribution, \(\pobs(\bx)\), such that
+\begin{equation}\protect\hypertarget{eq-dr}{}{
+r(\bx) = \frac{\psyn(\bx)}{\pobs(\bx)}.
+}\label{eq-dr}\end{equation}
+
+This specification has the intuitive interpretation that in locations
+where the density ratio takes large values, too many synthetic
+observations will be generated in that region and at the locations where
+the density ratio is small, there will be too few synthetic
+observations, both relative to the observed data. An intuitive approach
+to estimating \(r(\bx)\) from samples of \(\pobs(\bx)\) and
+\(\psyn(\bx)\) would be to estimate the observed and synthetic data
+density separately, for example using kernel density estimation (e.g.,
+see \protect\hyperlink{ref-Scott1992}{Scott 1992} for an overview), and
+subsequently compute the ratio of these estimated densities. However,
+density estimation is one of the hardest tasks in statistical learning,
+unavoidably leading to estimation errors for both densities, especially
+in high dimensions. When subsequently taking the ratio of the estimated
+densities, estimation errors tend to be magnified. Direct density ratio
+estimation avoids this issue by specifying and estimating a model
+directly for the ratio without first estimating the separate densities.
+Extensive simulations on a wide variety of tasks showed that this
+approach typically outperforms density ratio estimation through naive
+kernel density estimation, especially when the dimensionality of the
+data increases (e.g., \protect\hyperlink{ref-Kanamori2012}{Kanamori,
+Suzuki, and Sugiyama 2012b};
+\protect\hyperlink{ref-shohei_dre_outlier_2008}{Hido et al. 2008};
+\protect\hyperlink{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama
+2009}).
+
+\begin{figure}[t]
+
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-dr-plot-1.pdf}
+
+}
+
+\caption{\label{fig-dr-plot}Example of the true and estimated density
+ratio of two normal distributions with different means and variances
+(i.e., \(\psyn(\bx) = N(0,1)\) and \(\pobs(\bx) = N(1,2)\)). The
+function \(r(\bx) = \psyn(\bx)/\pobs(\bx)\) denotes the true density
+ratio, the function \(\hat{r}(\bx)\) denotes an estimate of the density
+ratio based on \(\nsyn = \nobs = 200\) samples from each distribution
+obtained with unconstrained Least-Squares Importance Fitting (uLSIF).
+Note that the density ratio is itself not a proper density.}
+
+\end{figure}
+
+\hypertarget{estimating-the-density-ratio}{%
+\subsubsection{Estimating the density
+ratio}\label{estimating-the-density-ratio}}
+
+Over the past years, several methods for direct density ratio estimation
+have been developed. A large class of these methods attempt to directly
+minimize the error between the true density ratio \(r(\bx)\) and its
+estimate \(\hat{r}(\bx)\). Following this approach, we define a loss
+function \(\mathcal{L}(r(\bx), \hat{r}(\bx))\) that measures the
+discrepancy between the true and estimated density ratio. To give an
+example, consider the following loss based on the squared error
+\begin{equation}\protect\hypertarget{eq-squared-error}{}{
+\begin{aligned}
+\mathcal{L}_S(r(\bx), \hat{r}(\bx)) &= 
+\frac{1}{2} \int (\hat{r}(\bx) - r(\bx))^2 \pobs(\bx) \text{d}\bx \\
+&= \frac{1}{2} \int \hat{r}(\bx)^2 \pobs(\bx) \text{d}\bx - 
+\int \hat{r}(\bx) r(\bx) \pobs(\bx) \text{d}\bx + 
+\frac{1}{2} \int r(\bx)^2 \pobs(\bx) \text{d}\bx.
+\end{aligned}
+}\label{eq-squared-error}\end{equation} The second term can be
+rewritten, because the denominator in \(r(\bx)\) cancels with
+\(\pobs(\bx)\), while the third term is a constant with respect to the
+parameters in the density ratio model and can thus be ignored. Hence, we
+are left with the following loss function to minimize
+\begin{equation}\protect\hypertarget{eq-squared-error-loss}{}{
+\mathcal{L}'_S (r(\bx), \hat{r}(\bx)) = \frac{1}{2} \int \hat{r}(\bx)^2 \pobs(\bx) \text{d}\bx - \int \hat{r}(\bx) \psyn(\bx) \text{d}\bx,
+}\label{eq-squared-error-loss}\end{equation} which is termed least
+squares importance fitting (LSIF) by Kanamori, Hido, and Sugiyama
+(\protect\hyperlink{ref-kanamori_ulsif_2009}{2009}).
+
+This approach can be straightforwardly generalized to a wide class of
+loss functions that fall under the family of Bregman divergences (see
+\protect\hyperlink{ref-sugiyama_bregman_2012}{Sugiyama, Suzuki, and
+Kanamori 2012b}; \protect\hyperlink{ref-mohamed2017learning}{Mohamed and
+Lakshminarayanan 2017}). Then, a general class of losses is encompassed
+by the expression
+\begin{equation}\protect\hypertarget{eq-bregman-loss}{}{
+\mathcal{L}_f (r(\bx), \hat{r}(\bx)) =
+\int \Big(
+f(r(\bx)) - f(\hat{r}(\bx)) - f'(\hat{r}(\bx))(r(\bx) - \hat{r}(\bx))
+\Big) \pobs(\bx) \text{d}\bx,
+}\label{eq-bregman-loss}\end{equation} where \(f\) is a differentiable
+and strictly convex function with derivative \(f'\). Then, ignoring the
+terms independent of \(\hat{r}(\bx)\) and noting that
+\(r(\bx)\pobs(\bx) = \psyn(\bx)\), we obtain the following objective
+\begin{equation}\protect\hypertarget{eq-bregman-objective}{}{
+\mathcal{L}'_f (r(\bx), \hat{r}(\bx)) = \int 
+\Big( 
+f'(\hat{r}(\bx)) \hat{r}(\bx) - f(\hat{r}(\bx))
+\Big)  \pobs(\bx) \text{d}\bx
+- \int f'(\hat{r}(\bx)) \psyn(\bx)
+\text{d}\bx.
+}\label{eq-bregman-objective}\end{equation} Minimizing this loss for
+different functions \(f\) yields different estimators for the density
+ratio, that focus on different regions of the density ratio.\footnote{It
+  is easy to see that using \(f(x) = \frac{1}{2}(x-1)^2\) turns the
+  Bregman divergence (Equation~\ref{eq-bregman-objective}) into the
+  squared error (Equation~\ref{eq-squared-error-loss}).} Specifically,
+some estimators place more emphasis on accurately modelling the regions
+in which the true density ratio is large, but less emphasis on
+accurately estimating the smaller density ratios, and vice versa (see
+\protect\hyperlink{ref-sugiyama_bregman_2012}{Sugiyama, Suzuki, and
+Kanamori 2012b}; \protect\hyperlink{ref-menon2016dreloss}{Menon and Ong
+2016}). Importantly, Sugiyama, Suzuki, and Kanamori
+(\protect\hyperlink{ref-sugiyama_bregman_2012}{2012b}) show that the
+Bregman divergence minimization approach to density ratio estimation is
+equivalent to estimating the \(f\)-divergence
+(\protect\hyperlink{ref-ali_silvey_divergence_1966}{Ali and Silvey
+1966}) between the synthetic and observed data distributions. Hence,
+when estimating the density ratio, one implicitly estimates the
+divergence between the synthetic and observed data distributions.
+
+After defining a loss function, we need a model for the density ratio
+function \(\hat{r}(\bx)\). There are many possibilities to specify this
+model, but a common choice is to use a linear model for the density
+ratio (e.g., \protect\hyperlink{ref-huang_kmm_2006}{Huang et al. 2006};
+\protect\hyperlink{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama
+2009}; \protect\hyperlink{ref-izbicki_dre_2014}{Izbicki, Lee, and
+Schafer 2014}; \protect\hyperlink{ref-gruber2024overcoming}{Gruber et
+al. 2024}). That is, we define the density ratio model as
+\begin{equation}\protect\hypertarget{eq-dr-model}{}{
+\hat{r}(\bx) = \boldsymbol{\varphi}(\bx) \boldsymbol{\theta},
+}\label{eq-dr-model}\end{equation} where \(\boldsymbol{\varphi}(\bx)\)
+is a basis function vector, that transforms the data from a
+\(p\)-dimensional to a \(b\)-dimensional space, and
+\(\boldsymbol{\theta}\) is a \(b\)-dimensional parameter vector.
+Although the model is linear in its parameters, the density ratio itself
+is typically a non-linear function of the data due to the basis
+functions. The parameter vector \(\boldsymbol{\theta}\) is estimated to
+minimize the discrepancy with the true density ratio using the loss
+function \(\mathcal{L}(r(\bx), \hat{r}(\bx))\) in
+Equation~\ref{eq-bregman-objective}.
+
+Also for the basis functions, several specifications are possible,
+ranging from an identity function
+(\protect\hyperlink{ref-qin_inferences_1998}{Qin 1998}) to normalizing
+flows (\protect\hyperlink{ref-choi_featurized_2021}{Choi, Liao, and
+Ermon 2021}) or neural networks
+(\protect\hyperlink{ref-tiao2018dre}{Tiao 2018};
+\protect\hyperlink{ref-uehara2016generative}{Uehara et al. 2016}).
+However, a commonly used basis function in the density ratio literature
+is the Gaussian kernel function (e.g.,
+\protect\hyperlink{ref-huang_kmm_2006}{Huang et al. 2006};
+\protect\hyperlink{ref-sugiyama_kliep_2007}{Sugiyama et al. 2007};
+\protect\hyperlink{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama
+2009}; \protect\hyperlink{ref-liu_change_2013}{Liu et al. 2013};
+\protect\hyperlink{ref-gruber2024overcoming}{Gruber et al. 2024}), which
+conveniently allows to model the non-linear density ratio using models
+that are linear in their parameters. The Gaussian kernel function is
+defined as \[
+\boldsymbol{\varphi}(\bx) = \mathcal{K}(\bx_i, \boldsymbol{c}_j) = 
+\exp\left(
+- \frac{\lVert\bx - \boldsymbol{c}_j \rVert^2}{2 \sigma^2}
+\right),
+\] where \(\boldsymbol{c}_j\) (\(j \in 1, \dots, J\)) denotes the
+centers of the Gaussian kernel functions, and \(\sigma\) controls the
+kernel width (i.e., it defines over which distance differences between
+the observations and the centers are considered relevant; for more
+information on kernel functions, see, e.g.,
+\protect\hyperlink{ref-murphy_pmlintro_2022}{Murphy 2022}). The
+appropriate width of the kernel can be determined using
+cross-validation. The centers are typically sampled from the data (in
+our case, the observed and synthetic data can both be used, but it is
+also possible to solely use the synthetic samples).
+
+After defining the model and the loss function, the density ratio can be
+easily estimated. The \texttt{densityratio} package in R
+(\protect\hyperlink{ref-densityratio}{Volker 2023}) provides an
+implementation of commonly used loss functions with a Gaussian kernel
+basis function. The package comes with an easy-to-use interface,
+automatic cross-validation of hyperparameters and builds on \texttt{C++}
+for fast computation. However, for increased flexibility with respect to
+loss functions and basis functions, one could use classic optimization
+techniques as gradient descent or (quasi-)Newton methods, for example as
+implemented in the \texttt{optim} function in the statistical software
+\texttt{R} (\protect\hyperlink{ref-R}{R Core Team 2023}). Both
+approaches are briefly illustrated in Appendix \ref{sec-app-A} (TODO).
+
+\hypertarget{evaluating-data-utility-with-the-density-ratio}{%
+\subsubsection{Evaluating data utility with the density
+ratio}\label{evaluating-data-utility-with-the-density-ratio}}
+
+After estimating the density ratio, the information provided by the
+density ratio can be summarized in a divergence measure, using the fact
+that estimating the density ratio is equivalent to estimating the
+\(f\)-divergence between the synthetic and observed data distributions
+(\protect\hyperlink{ref-sugiyama_bregman_2012}{Sugiyama, Suzuki, and
+Kanamori 2012b}). Accordingly, we can use the estimated density ratio
+function to estimate an \(f\)-divergence between the observed and
+synthetic data, and use this divergence measure as a utility measure for
+the synthetic data. Although this divergence statistic is difficult to
+interpret in an absolute sense, it can be used as a relative measure of
+quality of the synthetic data. That is, for different synthetic data
+sets, we can calculate the divergence to the observed data, and compare
+these values to determine which synthetic data set is most similar to
+the observed data. In a more formal way, the \(f\)-divergence gives rise
+to a test statistic that can be used to test the null hypothesis that
+the synthetic data is generated from the same distribution as the
+observed data. The corresponding \(p\)-value can be calculated by
+comparing the observed test statistic to the null distribution of test
+statistics obtained using random permutations of the observed and
+synthetic data (see, for example,
+\protect\hyperlink{ref-sugiyama_lstst_2011}{Sugiyama et al. 2011};
+\protect\hyperlink{ref-Wornowizki2016}{Wornowizki and Fried 2016};
+\protect\hyperlink{ref-kanamori_divergence_2012}{Kanamori, Suzuki, and
+Sugiyama 2012a}). Lastly, the density ratio function itself can be
+plotted against individual variables or pairs of variables to discover
+whether particular variables are particularly poorly captured in the
+synthetic data. Each of these strategies can be readily executed using
+the \texttt{densityratio} package in \texttt{R}
+(\protect\hyperlink{ref-densityratio}{Volker 2023}), and will be further
+illustrated in the upcoming sections.
+
+\textbf{Ik twijfel nog een beetje over de volgende zaken, en voordat ik
+alles ga zitten uitwerken hoor ik graag eerst jullie mening:}
 
@@
 \item
-  We introduce density ratio estimation to the field of statistical
-  disclosure control. Short remark on the idea that density ratio
-  estimation is a complicated endeavor, especially if the goal is to
-  compare distinct densities. Having to estimate just a single density
-  (ratio) is generally much easier.
+  Ik heb nu het deel over regularisatie eruit gelaten, omdat ik vond dat
+  het wat afleid van de main ideas, maar kan deze ook gewoon toevoegen
+  aan Equation~\ref{eq-bregman-objective}. Het kan ook casual opgemerkt
+  worden wanneer het daadwerkelijk gebruikt wordt. Wat vinden jullie?
 \item
-  Note that density ratio estimation can capture specific and general
-  utility measures into a common framework by being applicable on the
-  level of the entire data, but also on the subset of variables that is
-  relevant in an analysis. Additionally, note that confidence interval
-  overlap, ellipsoidal overlap, but also \(pMSE\) and Kullback-Leibler
-  divergence, are closely related to density ratio estimation, and can
-  be considered from this perspective.
+  Dimension reduction: ik was niet zo zeker hoe ik dit er direct moest
+  bouwen, want effectief is het niet veel anders dan de huidige aanpak,
+  alleen met een andere basis function (namelijk
+  \(\phi(x) = \mathcal{K}(x_i^T U, c_j^T U)\)), waarbij U ook parameters
+  heeft die geschat worden. Anyway, ik kan aan het eind een sectie maken
+  met dat DRE nog steeds lastig kan zijn in hoge dimensies, en dat daar
+  oplossingen voor zijn (bijvoorbeeld een LFDA-stap voor ulsif, of LHSS,
+  of the spectral density ratio estimation methode (laatste twee zitten
+  allebei in het package)). Het punt is vooral dat ik niet wil dat het
+  teveel afleidt van het main punt, namelijk density ratio estimation,
+  maar qua content zou het misschien een eigen sectie verdienen, ook
+  omdat het niet direct enorm aansluit bij de twee secties hierboven.
+\end{enumerate}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{2}
 \item
-  Create a new utility metric based on density ratio estimation
-  (probability with respect to some reference distribution as in
-  permutation testing).
+  Denken jullie dat dit een goede plek is om wat dieper in te gaan op
+  mogelijke voordelen van density ratio estimation? Ik dacht dat na
+  sectie 2.2.2 een voor de hand liggende plek zou zijn om hier wat
+  dieper op in te gaan, en dan specifiek de volgende punten te benoemen:
+  utility quantifications at every point in multivariate space, easy and
+  flexible model specification through automatic model selection with
+  cross-validation, potential for importance weighting, wellicht dan ook
+  direct het punt van dimension reduction hierbij pakken.
 \item
-  Because density ratio estimation can be difficult when there are many
-  variables, we use dimension reduction techniques to capture most of
-  the variability in the data in fewer dimensions on which density ratio
-  estimation can be applied. A by-product of this is that the
-  lower-dimensional subspace allows to create visualizations of
-  deviations from the observed data.
+  Ik wil graag nog ergens iets zeggen over de relaties tussen density
+  ratio estimation en de andere methodes (KL-divergence, pMSE), maar ik
+  vraag me een beetje af of dat niet beter in de discussie kan.
+\end{enumerate}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{4}
+\tightlist
 \item
-  Perform a simulation study to give indications about which methods to
-  use (think about how to do this).
-\item
-  Implement all this in an R-package
+  Is dit echt de goede plaats om iets te zeggen over categorische
+  variabelen? Deze kunnen namelijk gewoon geïncludeerd worden, maar met
+  een Gaussian kernel wordt dat misschien een beetje awkward.
+  Desalniettemin kan het prima, en kunnen we er ook voor kiezen om in de
+  discussie te benoemen dat we een oplossing hebben gekozen, maar dat er
+  ook andere opties zijn (bijv. andere kernels).
 \end{enumerate}
 
-\textbf{Section 6: outline article}
-
-In the next section, we describe density ratio estimation and discuss
-how this method can be used as to measure utility. Subsequently, we
-provide multiple examples that show how density ratio estimation works
-in the context of evaluating the quality of synthetic data. Hereafter,
-we show in multiple simulations that the method is superior (HOPEFULLY)
-to current global utility measures as the \(pMSE\). Lastly, we discuss
-the advantages and disadvantages of density ratio estimation as a
-utility measure.
-
-\hypertarget{methodology}{%
-\section{Methodology}\label{methodology}}
-
-@Peter-Paul: Eventueel een korte beschrijving van data perturbation
-techniques/synthetic data generation hier. Denk je dat dit wat toevoegt
-hier?
-
-\textbf{Section 1: density ratio estimation}
-
-In essence, the goal of utility measures is to quantify the similarity
-between the multivariate distribution of the observed data with the
-distribution of the altered data. If the used data perturbation
-techniques, or synthetic data generation models, approximate the
-distribution of the real data sufficiently, these distributions should
-be highly similar, and analyses on the two data sets should give similar
-results. However, estimating the probability distribution of a data set
-is known to be one of the most complicated challenges in statistics
-{[}E.G. Vapnik 1998{]}. Estimating the probability distribution for both
-observed and altered data can lead to errors in both, artificially
-magnifying discrepancies between the two. Hence, subsequent comparisons
-will be affected by these errors. The procedure can be simplified by
-using density ratio estimation, because this only requires to estimate a
-single density.
-
-Introduce density ratio estimation as a utility measure. What does this
-measure mean/how to interpret it. How to make decisions based on this
-measure.
-
-Say something on whether (and if so, how) categorical variables can be
-incorporated as well.
-
-\textbf{Section 3: theoretical comparison with conventional approaches
-for general utility assessment}
-
-Relate density ratio estimation to specific and general utility
-measures. Pick one/two specific utility measures and relate these to
-density ratio estimation (ratio of estimates seems straightforward, as
-well as confidence interval overlap and ellipsoidal overlap).
-
-Relate density ratio estimation to \(pMSE\) and KL divergence (to some
-extent, both are generalizations of density ratio estimation, or at
-least are conceptually similar). Give some more information on the
-\(pMSE\), describe what it shortcomings are. The quality of the \(pMSE\)
-highly depends on the model used to calculate the propensity scores.
-Perhaps give an example of logistic regression, which basically
-estimates whether the conditional mean of the observed and altered data
-is the same. Explain how density ratio estimation can overcome the
-shortcomings of the previously mentioned methods.
-
-\textbf{Section 4: Dimension reduction for utility}
-
-The difficulty of density ratio estimation increases with the
-dimensionality of the data. Therefore, we follow previous
-recommendations to incorporate dimensionality reduction techniques in
-density ratio estimation.
-
-Shortly name examples of dimensionality reduction techniques (i.e., PCA;
-LFDA or UMAP).
-
-A useful by-product of dimension reduction is that it allows to create
-visualizations, and these visualizations can be used to get more insight
-in discrepancies between observed and altered data. Show what such
-visualizations can look like, and how they can help.
-
-\hypertarget{simulations}{%
-\section{Simulations}\label{simulations}}
-
-\hypertarget{small-illustration-example-with-multivariate-gaussian-distributions.}{%
-\subsection{Small illustration / example with multivariate Gaussian
-distributions.}\label{small-illustration-example-with-multivariate-gaussian-distributions.}}
+\hypertarget{numerical-illustrations-simulation-studies}{%
+\section{Numerical illustrations: Simulation
+studies}\label{numerical-illustrations-simulation-studies}}
+
+To evaluate the performance of density ratio estimation and compare it
+to existing methods, we conduct a series of simulation studies. We
+consider a variety of scenarios, including univariate and multivariate
+models, varying correlations, number of variables, and sample sizes. To
+perform density ratio estimation, we use unconstrained Least-Squares
+Importance Fitting (uLSIF;
+\protect\hyperlink{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama
+2009}) as implemented in the \texttt{densityratio} package in \texttt{R}
+(\protect\hyperlink{ref-densityratio}{Volker 2023}), because it fast and
+has been shown to perform well in a variety of settings {[}ADD CITATIONS
+{]}. We compare the performance of uLSIF to the performance of the
+\(p\)-MSE as implemented in the \texttt{R}-package \texttt{synthpop}.
+
+\hypertarget{univariate-simulations}{%
+\subsection{Univariate simulations}\label{univariate-simulations}}
+
+For the sake of illustrational clarity, we first evaluate the
+performance of density ratio estimation in a univariate setting where
+its performance can be evaluated using visualizations. In the
+simulations, we generate data according to four data-generating
+mechanisms, (1) a Laplace distribution, (2) a log-normal distribution,
+(3) a location-scale \(t\)-distribution and (4) a normal distribution.
+Subsequently, we approximate the true data generating mechanism using a
+Gaussian model with the same mean and variance as the original data.
+This setting is similar to situations that are commonly encountered in
+the synthetic data field, in the sense that the true data generating
+mechanism is unknown and needs to be approximated using a simpler
+approximating model. The exact specifications of the data generating
+mechanisms are as follows:
 
@@
 \item
-  Simple, multivariate normal simulation (e.g., two correlation
-  structures, two sample sizes, so \(2 \times 2\) full factorial
-  design); basically similar to what we did already.
+  Laplace distribution with location parameter \(\mu = 1\) and scale
+  parameter \(b = 1\).
+\item
+  Log-normal distribution with log-mean parameter
+  \(\mu_{\text{log}} = \log \{1/\sqrt{3} \}\) and log-standard deviation
+  parameter \(\sigma_\text{log} = \sqrt{\log 3}\).
+\item
+  Location-scale \(t\)-distribution with location parameter \(\mu = 1\),
+  scale parameter \(\tau^2 = 1\) and degrees of freedom \(\nu = 4\).
+\item
+  Normal distribution with mean \(\mu = 1\) and variance
+  \(\sigma^2 = 2\).
 \end{enumerate}
 
-\hypertarget{more-complex-simulation-more-variables-non-linearities-perhaps-using-real-data.}{%
-\subsection{More complex simulation, more variables, non-linearities,
-perhaps using real
-data.}\label{more-complex-simulation-more-variables-non-linearities-perhaps-using-real-data.}}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{1}
+These data generating mechanisms are chosen such that they all have the
+same population mean \(\mu = 1\) and variance \(\sigma^2 = 2\). The
+approximating model is a Gaussian distribution with mean \(\mu = 1\) and
+variance \(\sigma^2 = 2\). This distribution has the same mean and
+variance, but differs in higher-order moments, except in the fourth
+scenario. In the last simulation, we model the data generating mechanism
+correctly to obtain some understanding into how density ratio estimation
+behaves under a well-specified synthesis model. In all scenarios, we
+generate \(500\) data sets with \(\nobs = 250\) observations from the
+true data generating mechanism, and \(\nsyn = 250\) synthetic
+observations from the approximating Gaussian model. The density ratio
+model is estimated using uLSIF with a Gaussian kernel and an
+\(L_2\)-regularization parameter. Both the kernel bandwidth and the
+regularization parameter are selected using cross-validation, using the
+default settings as implemented in the \texttt{densityratio} package.
+
+\begin{figure}[t]
+
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-densities-sim1-1.pdf}
+
+}
+
+\caption{\label{fig-densities-sim1}True and synthetic data densities for
+the four simulations with Laplace, Log-normal, location-scale \(t\)- and
+Normal densities. All data-generating mechanisms have the same mean
+\(\mu = 1\) and variance \(\sigma^2 = 2\). Note that the true and
+synthetic data density in the bottom right panel are completely
+overlapping.}
+
+\end{figure}
+
+\begin{figure}[t]
+
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-sim1-results-1.pdf}
+
+}
+
+\caption{\label{fig-sim1-results}Estimated density ratios by
+unconstrained least-squares importance fitting in four univariate
+examples: A Laplace distribution, a log-normal distribution, a
+location-scale \(t\)-distribution and a normal distribution, all
+approximated by a normal distribution with the same mean and variance as
+the sample from the true distribution.}
+
+\end{figure}
+
+TODO: Add interpretation of output; add significance tests and compare
+with \(pMSE\) and kolmogorov-smirnov.
+
+\hypertarget{multivariate-simulations}{%
+\subsection{Multivariate simulations}\label{multivariate-simulations}}
+
+TODO: Adjust section title
+
+Currently, my plan is as follows, specify a (two/four) true data
+generating mechanism: A multivariate normal distribution with a given
+correlation structure, 7 and 17 variables each for two different sample
+size (say n = 500 and n = 2000). Append the multivariate normal
+distribution with three variables with a different distribution that
+depend on the other variables through some non-linear function. Have
+three synthetic data generating mechanisms, a simple multivariate normal
+distribution with the same means and variances but zero covariances, a
+model that uses a multivariate normal distribution with correlation
+structure taken from the real data but misses the non-linear effects
+(and marginal distributions of these variables), and a model that is
+equivalent to the true model. Then, evaluate the quality of the
+synthetic data sets using density ratio estimation and the \(pMSE\), and
+see which of the two ranks the models correctly most of the time.
+
+\hypertarget{application-synthetic-data-generation-for-the-u.s.-current-population-survey}{%
+\section{Application: Synthetic data generation for the U.S. Current
+Population
+Survey}\label{application-synthetic-data-generation-for-the-u.s.-current-population-survey}}
+
+ADD TEXT
+
+\begin{figure}[t]
+
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-application-distributions-1.pdf}
+
+}
+
+\caption{\label{fig-application-distributions}Real and synthetic data
+distributions for the variables age, household income (income),
+household property taxes (tax) and social security benefits (social
+security) on a cubic root scale.}
+
+\end{figure}
+
+\begin{figure}[t]
+
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-application-pearson-div-1.pdf}
+
+}
+
+\caption{\label{fig-application-pearson-div}Pearson divergence estimates
+after different synthesis strategies for the separate variables and the
+synthetic data as a whole.}
+
+\end{figure}
+
+TODO: Expand example
+
+High-dimensional density ratio estimation
+
+Use density ratio values for weighted analysis
+
+Incorporate point on individual data utility
+
+\begin{itemize}
 \tightlist
 \item
-  More advanced simulation (e.g., some non-linearities, different sample
-  sizes)
-\end{enumerate}
-
-Have to think about this in more detail still.
-
-\hypertarget{real-data-example}{%
-\section{Real data example}\label{real-data-example}}
-
-Clinical records heart-failure data? Misschien ook niet, nog over
-nadenken.
-
-Exemplify how utility measures could (should!) be used to improve the
-quality of the altered data (e.g., illustrate how models can be adjusted
-iteratively based on utility assessment).
-
-\emph{Some notes to self}
-
-Current ways to assess the utility?
-
-\begin{itemize}
+  High dimensional example
 \item
-  pMSE - logistic, regression, CART models (Snoke, Raab, Nowok, Dibben
-  \& Slavkovic, 2018; General and specific utility measures for
-  synthetic data AND Woo, Reiter, Oganian \& Karr, 2009; Global measures
-  of data utlity for microdata masked for disclosure limitation)
-\item
-  Kullback-Leiber divergence (Karr, Kohnen, Oganian, Reiter \& Sanil,
-  2006; A framework for evaluating the utility of data altered to
-  protect confidentiality).
-\item
-  According to multiple authors, both specific and general utility
-  measures have important drawbacks (see Drechsler Utility PSD; cites
-  others). Narrow measures potentially focus on analyses that are not
-  relevant for the end user, and do not generalize to the analyses that
-  are relevant. Global utility measures are generally too broad, and
-  important deviations in the synthetic data might be missed. Moreover,
-  the measures are typically hard to interpret.
-\item
-  See Drechsler for a paragraph on fit for purpose measures, that lie
-  between general and specific utility measures (i.e., plausibility
-  checks such as non-negativity; goodness of fit measures as \(\chi^2\)
-  for cross-tabulations; Kolmogorov-Smirnov).
-\item
-  Drechsler also illustrates that the standardized \(pMSE\) has
-  substantial flaws, as the results are highly dependent on the model
-  used to estimate the propensity scores, and unable to detect important
-  differences in the utility for most of the model specifications.
-  Hence, it is claimed that a thorough assessment of utility is
-  required.
+  Expansion of discussion points: empirical example with weighted
+  analyses individual data utility
 \end{itemize}
-
-Things to add in new version: - High dimensional example - Expansion of
-discussion points: empirical example with weighted analyses individual
-data utility - Privacy remark? - When does it not work?
-
-\hypertarget{methodology-1}{%
-\section{Methodology}\label{methodology-1}}
-
-TO DO
-
-\hypertarget{simulations-1}{%
-\section{Simulations}\label{simulations-1}}
-
-TO DO
-
-\hypertarget{real-data-example-1}{%
-\section{Real data example}\label{real-data-example-1}}
-
-TO DO
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-TO DO
 
@@
 
-TO DO
+Adapt from previous version
+
+\begin{itemize}
+\tightlist
+\item
+  Privacy remark?
+\item
+  When does it not work?
+\end{itemize}
+
+Density ratios can be used directly to train generative models, see:
+\href{https://arxiv.org/pdf/1610.03483}{Mohamed \& Lakshminarayanan
+(2016)} and \href{https://arxiv.org/pdf/1610.02920.pdf}{Uehara, Sata,
+Suzuki, Nakayama \& Matsuo (2016)}
+
+Density ratio estimation is equivalent to class probability estimation
+in classification problems, only differing in the loss function that is
+employed (assuming the same model class is used for density ratio
+estimation and classification). See
+\href{https://proceedings.mlr.press/v48/menon16.pdf}{Menon \& Ong, 2016}
+
+Logistic regression achieves the minimum asymptotic variance for
+correctly specified models (Qin, Biometrika 1998), but is not reliable
+for misspecified models (Kanamori, Suzuki, Sugiyama, IECE, 2010)
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-caiola2010}{}}%
-Caiola, Gregory, and Jerome P. Reiter. 2010. {``Random Forests for
-Generating Partially Synthetic, Categorical Data.''} \emph{Transactions
-on Data Privacy} 3: 27--42.
-\url{https://doi.org/10.5555/1747335.1747337}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-drechsler2010}{}}%
-Drechsler, Jörg. 2010. {``Using Support Vector Machines for Generating
-Synthetic Datasets.''} In \emph{Privacy in Statistical Databases},
-edited by Josep Domingo-Ferrer and Emmanouil Magkos, 148--61. Berlin,
-Heidelberg: Springer Berlin Heidelberg.
-\url{https://doi.org/10.1007/978-3-642-15838-4_14}.
+\leavevmode\vadjust pre{\hypertarget{ref-ali_silvey_divergence_1966}{}}%
+Ali, S. M., and S. D. Silvey. 1966. {``A General Class of Coefficients
+of Divergence of One Distribution from Another.''} \emph{Journal of the
+Royal Statistical Society. Series B (Methodological)} 28 (1): 131--42.
+\url{https://doi.org/10.1111/j.2517-6161.1966.tb00626.x}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-atenas_open_2015}{}}%
+Atenas, Javiera, Leo Havemann, and Ernesto Priego. 2015. {``Open Data as
+Open Educational Resources: Towards Transversal Skills and Global
+Citizenship.''} \emph{Open Praxis} 7 (4): 377--89.
+\url{https://www.learntechlib.org/p/161986}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Bowen_differentially_2021}{}}%
+Bowen, Claire McKay, Fang Liu, and Bingyue Su. 2021. {``Differentially
+Private Data Release via Statistical Election to Partition
+Sequentially.''} \emph{METRON} 79 (1): 1--31.
+\url{https://doi.org/10.1007/s40300-021-00201-0}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-choi_featurized_2021}{}}%
+Choi, Kristy, Madeline Liao, and Stefano Ermon. 2021. {``Featurized
+Density Ratio Estimation.''} In \emph{Proceedings of the Thirty-Seventh
+Conference on Uncertainty in Artificial Intelligence}, edited by Cassio
+de Campos and Marloes H. Maathuis, 161:172--82. Proceedings of Machine
+Learning Research. PMLR.
+\url{https://proceedings.mlr.press/v161/choi21a.html}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-crosas_automating_2015}{}}%
+Crosas, Mercè, Gary King, James Honaker, and Latanya Sweeney. 2015.
+{``Automating Open Science for Big Data.''} \emph{The ANNALS of the
+American Academy of Political and Social Science} 659 (1): 260--73.
+\url{https://doi.org/10.1177/0002716215570847}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-drechsler2011synthetic}{}}%
----------. 2011. \emph{Synthetic Datasets for Statistical Disclosure
-Control: Theory and Implementation}. New York: Springer Science \&
-Business Media. \url{https://doi.org/10.1007/978-1-4614-0326-5}.
+Drechsler, Jörg. 2011. \emph{Synthetic Datasets for Statistical
+Disclosure Control: Theory and Implementation}. New York: Springer
+Science \& Business Media.
+\url{https://doi.org/10.1007/978-1-4614-0326-5}.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-drechsler2011empirical}{}}%
-Drechsler, Jörg, and Jerome P Reiter. 2011. {``An Empirical Evaluation
-of Easily Implemented, Nonparametric Methods for Generating Synthetic
-Datasets.''} \emph{Computational Statistics \& Data Analysis} 55 (12):
-3232--43.
+\leavevmode\vadjust pre{\hypertarget{ref-drechsler2023}{}}%
+Drechsler, Jörg, and Anna-Carolina Haensch. 2023. {``30 Years of
+Synthetic Data.''} \url{https://doi.org/10.48550/ARXIV.2304.02107}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gdpr}{}}%
+European Parliament, and Council of the European Union. 2016.
+{``Regulation ({EU}) 2016/679 of the {European} {Parliament} and of the
+{Council}. Of 27 {April} 2016 on the Protection of Natural Persons with
+Regard to the Processing of Personal Data and on the Free Movement of
+Such Data, and Repealing {Directive} 95/46/{EC} ({General} {Data}
+{Protection} {Regulation}).''} OJ L 119, 4.5.2016, p. 1--88. May 4,
+2016. \url{https://eur-lex.europa.eu/eli/reg/2016/679/oj}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-gruber2024overcoming}{}}%
+Gruber, Lukas, Markus Holzleitner, Johannes Lehner, Sepp Hochreiter, and
+Werner Zellinger. 2024. {``Overcoming Saturation in Density Ratio
+Estimation by Iterated Regularization.''}
+\url{https://doi.org/10.48550/arXiv.2402.13891}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-hawala_synthetic_2008}{}}%
@@
 \url{http://www.asasrms.org/Proceedings/y2008/Files/301018.pdf}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-shohei_dre_outlier_2008}{}}%
+Hido, Shohei, Yuta Tsuboi, Hisashi Kashima, Masashi Sugiyama, and
+Takafumi Kanamori. 2008. {``Inlier-Based Outlier Detection via Direct
+Density Ratio Estimation.''} In \emph{2008 Eighth IEEE International
+Conference on Data Mining}, edited by Fosca Giannotti, Dimitrios
+Gunopulos, Franco Turini, Carlo Zaniolo, Naren Ramakrishnan, and Xindong
+Wu, 223--32. \url{https://doi.org/10.1109/ICDM.2008.49}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hu_advancing_2024}{}}%
+Hu, Jingchen, and Claire McKay Bowen. 2024. {``Advancing Microdata
+Privacy Protection: A Review of Synthetic Data Methods.''} \emph{WIREs
+Computational Statistics} 16 (1): e1636.
+https://doi.org/\url{https://doi.org/10.1002/wics.1636}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-huang_kmm_2006}{}}%
+Huang, Jiayuan, Alexander J. Smola, Arthur Gretton, Karsten M.
+Borgwardt, and Bernhard Schölkopf. 2006. {``Correcting Sample Selection
+Bias by Unlabeled Data.''} In \emph{Advances in Neural Information
+Processing Systems}, edited by B. Schölkopf, J. Platt, and T. Hoffman.
+Vol. 19. MIT Press.
+\url{https://proceedings.neurips.cc/paper_files/paper/2006/file/a2186aa7c086b46ad4e8bf81e2a3a19b-Paper.pdf}.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-jordon2018pategan}{}}%
-Jordon, James, Jinsung Yoon, and Mihaela van der Schaar. 2019.
-{``{PATE}-{GAN}: Generating Synthetic Data with Differential Privacy
-Guarantees.''} In \emph{International Conference on Learning
-Representations}. \url{https://openreview.net/forum?id=S1zk9iRqF7}.
+\leavevmode\vadjust pre{\hypertarget{ref-izbicki_dre_2014}{}}%
+Izbicki, Rafael, Ann Lee, and Chad Schafer. 2014. {``{High-Dimensional
+Density Ratio Estimation with Extensions to Approximate Likelihood
+Computation}.''} In \emph{Proceedings of the Seventeenth International
+Conference on Artificial Intelligence and Statistics}, edited by Samuel
+Kaski and Jukka Corander, 33:420--29. Proceedings of Machine Learning
+Research. Reykjavik, Iceland: PMLR.
+\url{https://proceedings.mlr.press/v33/izbicki14.html}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kanamori_ulsif_2009}{}}%
+Kanamori, Takafumi, Shohei Hido, and Masashi Sugiyama. 2009. {``A
+Least-Squares Approach to Direct Importance Estimation.''} \emph{Journal
+of Machine Learning Research} 10 (48): 1391--1445.
+\url{http://jmlr.org/papers/v10/kanamori09a.html}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kanamori_divergence_2012}{}}%
+Kanamori, Takafumi, Taiji Suzuki, and Masashi Sugiyama. 2012a. {``\(f\)
+-Divergence Estimation and Two-Sample Homogeneity Test Under
+Semiparametric Density-Ratio Models.''} \emph{IEEE Transactions on
+Information Theory} 58 (2): 708--20.
+\url{https://doi.org/10.1109/TIT.2011.2163380}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Kanamori2012}{}}%
+---------. 2012b. {``Statistical Analysis of Kernel-Based Least-Squares
+Density-Ratio Estimation.''} \emph{Machine Learning} 86 (3): 335--67.
+\url{https://doi.org/10.1007/s10994-011-5266-3}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-kim_classification_2021}{}}%
+Kim, Ilmun, Aaditya Ramdas, Aarti Singh, and Larry Wasserman. 2021.
+{``{Classification accuracy as a proxy for two-sample testing}.''}
+\emph{The Annals of Statistics} 49 (1): 411--34.
+\url{https://doi.org/10.1214/20-AOS1962}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-lazer_css_2009}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-nikolenko2021}{}}%
-Nikolenko, Sergey I. 2021. \emph{Synthetic Data for Deep Learning}.
-Springer International Publishing.
-\url{https://doi.org/10.1007/978-3-030-75178-4}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nowok2016}{}}%
-Nowok, Beata, Gillian M. Raab, and Chris Dibben. 2016.
-{``{\emph{Synthpop}:} Bespoke Creation of Synthetic Data in
-{\emph{R}}.''} \emph{Journal of Statistical Software} 74 (11).
-\url{https://doi.org/10.18637/jss.v074.i11}.
+\leavevmode\vadjust pre{\hypertarget{ref-liu_change_2013}{}}%
+Liu, Song, Makoto Yamada, Nigel Collier, and Masashi Sugiyama. 2013.
+{``Change-Point Detection in Time-Series Data by Relative Density-Ratio
+Estimation.''} \emph{Neural Networks} 43: 72--83.
+\url{https://doi.org/10.1016/j.neunet.2013.01.012}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-menon2016dreloss}{}}%
+Menon, Aditya, and Cheng Soon Ong. 2016. {``Linking Losses for Density
+Ratio and Class-Probability Estimation.''} In \emph{Proceedings of the
+33rd International Conference on Machine Learning}, edited by Maria
+Florina Balcan and Kilian Q. Weinberger, 48:304--13. Proceedings of
+Machine Learning Research. New York, New York, USA: PMLR.
+\url{https://proceedings.mlr.press/v48/menon16.html}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mohamed2017learning}{}}%
+Mohamed, Shakir, and Balaji Lakshminarayanan. 2017. {``Learning in
+Implicit Generative Models.''}
+\url{https://doi.org/10.48550/arXiv.1610.03483}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-murphy_pmlintro_2022}{}}%
+Murphy, Kevin P. 2022. \emph{Probabilistic Machine Learning: An
+Introduction}. MIT Press. \href{https://probml.ai}{probml.ai}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-newman_future_2012}{}}%
+Newman, Greg, Andrea Wiggins, Alycia Crall, Eric Graham, Sarah Newman,
+and Kevin Crowston. 2012. {``The Future of Citizen Science: Emerging
+Technologies and Shifting Paradigms.''} \emph{Frontiers in Ecology and
+the Environment} 10 (6): 298--304. \url{https://doi.org/10.1890/110294}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-obels_analysis_2020}{}}%
+Obels, Pepijn, Daniël Lakens, Nicholas A. Coles, Jaroslav Gottfried, and
+Seth A. Green. 2020. {``Analysis of Open Data and Computational
+Reproducibility in Registered Reports in Psychology.''} \emph{Advances
+in Methods and Practices in Psychological Science} 3 (2): 229--37.
+\url{https://doi.org/10.1177/2515245920918872}.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-patki2016}{}}%
-Patki, Neha, Roy Wedge, and Kalyan Veeramachaneni. 2016. {``The
-Synthetic Data Vault.''} \emph{2016 IEEE International Conference on
-Data Science and Advanced Analytics (DSAA)}, October.
-\url{https://doi.org/10.1109/dsaa.2016.49}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-reiter2005}{}}%
-Reiter, Jerome P. 2005. {``Using CART to Generate Partially Synthetic
-Public Use Microdata.''} \emph{Journal of Official Statistics} 21 (3):
-441--62.
-\url{https://www.scb.se/contentassets/ca21efb41fee47d293bbee5bf7be7fb3/using-cart-to-generate-partially-synthetic-public-use-microdata.pdf}.
+\leavevmode\vadjust pre{\hypertarget{ref-qin_inferences_1998}{}}%
+Qin, Jing. 1998. {``{Inferences for case-control and semiparametric
+two-sample density ratio models}.''} \emph{Biometrika} 85 (3): 619--30.
+\url{https://doi.org/10.1093/biomet/85.3.619}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-R}{}}%
+R Core Team. 2023. \emph{R: A Language and Environment for Statistical
+Computing}. Vienna, Austria: R Foundation for Statistical Computing.
+\url{https://www.R-project.org/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-raab2017guidelines}{}}%
+Raab, Gillian M., Beata Nowok, and Chris Dibben. 2017. {``Guidelines for
+Producing Useful Synthetic Data.''}
+\url{https://arxiv.org/abs/1712.04078}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-raab2021assessing}{}}%
+Raab, Gillian M, Beata Nowok, and Chris Dibben. 2021. {``Assessing,
+Visualizing and Improving the Utility of Synthetic Data.''}
+\url{https://doi.org/10.48550/arXiv.2109.12717}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ramachandran_open_2021}{}}%
+Ramachandran, Rahul, Kaylin Bugbee, and Kevin Murphy. 2021. {``From Open
+Data to Open Science.''} \emph{Earth and Space Science} 8 (5):
+e2020EA001562. \url{https://doi.org/10.1029/2020EA001562}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-reiter_releasing_2004}{}}%
+Reiter, Jerome P. 2004. {``{Releasing Multiply Imputed, Synthetic Public
+use Microdata: An Illustration and Empirical Study}.''} \emph{Journal of
+the Royal Statistical Society Series A: Statistics in Society} 168 (1):
+185--205. \url{https://doi.org/10.1111/j.1467-985X.2004.00343.x}.
 
@@
 \url{https://www.scb.se/contentassets/ca21efb41fee47d293bbee5bf7be7fb3/discussion-statistical-disclosure-limitation2.pdf}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Scott1992}{}}%
+Scott, David W. 1992. \emph{Multivariate Density Estimation: Theory,
+Practice, and Visualization}. Wiley.
+\url{https://doi.org/10.1002/9780470316849}.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Torkzadehmahani2019}{}}%
-Torkzadehmahani, Reihaneh, Peter Kairouz, and Benedict Paten. 2019.
-{``{DP-CGAN}: Differentially Private Synthetic Data and Label
-Generation.''} In \emph{Proceedings of the IEEE/CVF Conference on
-Computer Vision and Pattern Recognition (CVPR) Workshops}.
-\url{https://doi.org/10.1109/cvprw.2019.00018}.
+\leavevmode\vadjust pre{\hypertarget{ref-sugiyama_classification_2010}{}}%
+Sugiyama, Masashi. 2010. {``Superfast-Trainable Multi-Class
+Probabilistic Classifier by Least-Squares Posterior Fitting.''}
+\emph{IEICE Transactions on Information and Systems} E93-D (10).
+\url{https://doi.org/10.1587/transinf.E93.D.2690}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sugiyama_kliep_2007}{}}%
+Sugiyama, Masashi, Shinichi Nakajima, Hisashi Kashima, Paul Buenau, and
+Motoaki Kawanabe. 2007. {``Direct Importance Estimation with Model
+Selection and Its Application to Covariate Shift Adaptation.''} In
+\emph{Advances in Neural Information Processing Systems}, edited by J.
+Platt, D. Koller, Y. Singer, and S. Roweis. Vol. 20. Curran Associates,
+Inc.
+\url{https://proceedings.neurips.cc/paper_files/paper/2007/file/be83ab3ecd0db773eb2dc1b0a17836a1-Paper.pdf}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sugiyama_lstst_2011}{}}%
+Sugiyama, Masashi, Taiji Suzuki, Yuta Itoh, Takafumi Kanamori, and
+Manabu Kimura. 2011. {``Least-Squares Two-Sample Test.''} \emph{Neural
+Networks} 24 (7): 735--51.
+\url{https://doi.org/10.1016/j.neunet.2011.04.003}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sugiyama_suzuki_kanamori_2012}{}}%
+Sugiyama, Masashi, Taiji Suzuki, and Takafumi Kanamori. 2012a.
+\emph{Density Ratio Estimation in Machine Learning}. Cambridge
+University Press. \url{https://doi.org/10.1017/CBO9781139035613}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sugiyama_bregman_2012}{}}%
+---------. 2012b. {``Density-Ratio Matching Under the Bregman
+Divergence: A Unified Framework of Density-Ratio Estimation.''}
+\emph{Annals of the Institute of Statistical Mathematics} 64 (5):
+1009--44. \url{https://doi.org/10.1007/s10463-011-0343-8}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sugiyama_conditional_2010}{}}%
+Sugiyama, Masashi, Ichiro Takeuchi, Taiji Suzuki, Takafumi Kanamori,
+Hirotaka Hachiya, and Daisuke Okanohara. 2010. {``Conditional Density
+Estimation via Least-Squares Density Ratio Estimation.''} In
+\emph{Proceedings of the Thirteenth International Conference on
+Artificial Intelligence and Statistics}, edited by Yee Whye Teh and Mike
+Titterington, 9:781--88. Proceedings of Machine Learning Research. Chia
+Laguna Resort, Sardinia, Italy: PMLR.
+\url{https://proceedings.mlr.press/v9/sugiyama10a.html}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tiao2018dre}{}}%
+Tiao, Louis C. 2018. {``{D}ensity {R}atio {E}stimation for {KL}
+{D}ivergence {M}inimization Between {I}mplicit {D}istributions.''}
+\emph{Tiao.io}.
+\url{https://tiao.io/post/density-ratio-estimation-for-kl-divergence-minimization-between-implicit-distributions/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-uehara2016generative}{}}%
+Uehara, Masatoshi, Issei Sato, Masahiro Suzuki, Kotaro Nakayama, and
+Yutaka Matsuo. 2016. {``Generative Adversarial Nets from a Density Ratio
+Estimation Perspective.''}
+\url{https://doi.org/10.48550/arXiv.1610.02920}.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-volker2021}{}}%
-Volker, Thom Benjamin, and Gerko Vink. 2021. {``Anonymiced Shareable
-Data: Using Mice to Create and Analyze Multiply Imputed Synthetic
-Datasets.''} \emph{Psych} 3 (4): 703--16.
-\url{https://doi.org/10.3390/psych3040045}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-willenborg_elements_2012}{}}%
-Willenborg, Leon, and Ton De Waal. 2012. \emph{Elements of Statistical
+\leavevmode\vadjust pre{\hypertarget{ref-densityratio}{}}%
+Volker, Thom Benjamin. 2023. {``Densityratio: Direct Estimation of the
+Ratio of Densities of Two Groups of Observations.''}
+\url{https://github.com/thomvolker/densityratio}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-willenborg_elements_2001}{}}%
+Willenborg, Leon, and Ton De Waal. 2001. \emph{Elements of Statistical
 Disclosure Control}. Springer Science \& Business Media.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Woo_global_2009}{}}%
+Woo, Mi-Ja, Jerome P. Reiter, Anna Oganian, and Alan F. Karr. 2009.
+{``Global Measures of Data Utility for Microdata Masked for Disclosure
+Limitation.''} \emph{Journal of Privacy and Confidentiality} 1 (1).
+\url{https://doi.org/10.29012/jpc.v1i1.568}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Wornowizki2016}{}}%
+Wornowizki, Max, and Roland Fried. 2016. {``Two-Sample Homogeneity Tests
+Based on Divergence Measures.''} \emph{Computational Statistics} 31 (1):
+291--313. \url{https://doi.org/10.1007/s00180-015-0633-3}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-xu_ctgan_2019}{}}%
 Xu, Lei, Maria Skoularidou, Alfredo Cuesta-Infante, and Kalyan
-Veeramachaneni. 2019. {``Modeling Tabular Data Using Conditional GAN.''}
-In \emph{Advances in Neural Information Processing Systems}, edited by
-H. Wallach, H. Larochelle, A. Beygelzimer, F. dAlché-Buc, E. Fox, and R.
-Garnett. Vol. 32. Curran Associates, Inc.
+Veeramachaneni. 2019. \emph{Modeling Tabular Data Using Conditional
+GAN}. Edited by H. Wallach, H. Larochelle, A. Beygelzimer, F.
+dAlché-Buc, E. Fox, and R. Garnett. Vol. 32. Curran Associates, Inc.
 \url{https://proceedings.neurips.cc/paper_files/paper/2019/file/254ed7d2de3b23ab10936522dd547b78-Paper.pdf}.
@@
 
+\setcounter{section}{0}
+\renewcommand{\thesection}{\Alph{section}}
+
+\setcounter{table}{0}
+\renewcommand{\thetable}{A\arabic{table}}
+
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{A\arabic{figure}}
+
+\hypertarget{sec-app-A}{%
+\section{\texorpdfstring{Density ratio estimation in
+\texttt{R}}{Density ratio estimation in R}}\label{sec-app-A}}
+
+TO DO
+
+Small example of ulsif in \texttt{densityratio}
+
+Do the same with \texttt{optim}
+
+Do the same for a different loss function
+
 \end{document}

### 2024-08-23 paper/paper.qmd
Update paper

@@
       - file: preamble.tex
+    header-includes:
+      - \usepackage{float}
+      - \floatplacement{table}{t}
   arxiv-html: default
@@
     email: e.vankesteren1@uu.nl
-abstract: TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
+abstract: | 
+  Synthetic data generation is a promising technique to facilitate the use of sensitive data while mitigating the risk of privacy breaches.
+However, for synthetic data to be useful in downstream analysis tasks, it needs to be of sufficient quality.
+Various methods have been proposed to measure the utility of synthetic data, but their results are often incomplete or even misleading.
+In this paper, we propose using density ratio estimation to improve quality evaluation for synthetic data, and thereby the quality of synthesized datasets.
+  We show how this framework relates to and builds on existing measures, yielding global and local utility measures that are informative and easy to interpret.
+We develop an estimator which requires little to no manual tuning due to automatic selection of a nonparametric density ratio model.
+Through simulations, we find that density ratio estimation yields more accurate estimates of global utility than established procedures.
+A real-world data application demonstrates how the density ratio can guide refinements of synthesis models and can be used to improve downstream analyses.
+We conclude that density ratio estimation is a valuable tool in synthetic data generation workflows and provide these methods in the accessible open source R-package densityratio.
 keywords: |
@@
 bibliography: bibliography.bib
+params:
+  runsims: false
 ---
@@
 # Introduction
+
+```{r}
+#| label: setup
+knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
+
+chunk_hook  <- knitr::knit_hooks$get("chunk")
+knitr::knit_hooks$set(chunk = function(x, options) {
+  x <- chunk_hook(x, options)
+  paste0("\\linespread{1}\n", x, "\n\n\\linespread{2}")
+})
+```
 
@@
 Global utility measures typically attempt to quantify the distributional similarity between the observed and synthetic data samples.
-The intuition is that if two data sets have similar distributions, the data sets can be used for the same purposes, and analyses on the two data sets should yield similar results.
-A common way to formalize distributional similarity is through the Kullback-Leibler (KL) divergence, as proposed in @karr_utility_2006.
-The KL-divergence measures the relative entropy from the probability distribution of the observed data $\pobs(\bx)$ to the probability distribution of the synthetic data $\psyn(\bx)$ (with $\bx \in \mathbb{R}^{n \times d}$), and is defined as
-$$
-D_{\text{KL}}(\psyn || \pobs) = 
-\int \psyn(\bx) \log \frac{\psyn(\bx)}{\pobs(\bx)} \text{d} \bx.
-$${#eq-kl-div}
-@karr_utility_2006 argue that the KL-divergence can be constructed from density estimators $\hat{p}_\text{obs}$ and $\hat{p}_\text{syn}$, after which the integral can be approximated using numerical quadrature.
-Alternatively, if the observed and synthetic data are both (multivariate) normally distributed, the KL-divergence can be computed analytically.
-Both implementations might be unfeasible: the combination of density estimation with numerical quadrature is cumbersome in high dimensional settings, and assuming multivariate normality might be too restrictive.
-Yet, we show in later sections that the density ratio estimation framework gives rives to an alternative way of estimating the KL-divergence that overcomes these challenges.
-
-An alternative way to evaluate whether two distributions are statistically indistinguishable is by evaluating whether a classification model can tell samples from the two distributions apart [see @kim_classification_2021, who formalize the connection between classification accuracy and two-sample testing].
-In the context of synthetic data, this implies that if a classification model can distinguish observed from synthetic samples, the distributional similarity is low, and so is the global utility. 
+The intuition is that two data sets are realizations from the same underlying distribution, the data sets can be used for the same purposes, and analyses on the two data sets should yield similar results.
+One way to evaluate distributional similarity is by assessing whether a classification model can tell samples from the two distributions apart [see @kim_classification_2021, who formalize the connection between classification accuracy and two-sample testing].
+Hence, if a classification model can distinguish between the observed and synthetic data with high accuracy, the distributional similarity is low, and so is the global utility. 
 If a classifier cannot distinguish between the observed and synthetic data, one would conclude that the global utility is high.
@@
 Let $I_i$ denote an indicator variable that equals $1$ if observation $i$ ($i \in 1, \dots, N$, $N = \nobs + \nsyn$) belongs to the synthetic data, and $0$ otherwise.
-We then train a classifier that outputs the predicted probability of observation $i$ being a synthetic record $\hat{s}_i$ based on the observation's scores on the variables (this can be the set of all variables, but also a subset).
+We then train a classifier that outputs the predicted probability of observation $i$ being a synthetic record $\hat{\pi}_i$ based on the observation's scores on the variables (this can be the set of all variables, but also a subset).
 From these, we can calculate the utility statistic 
 $$
-pMSE = \frac{1}{N} \sum_{i=1}^N \Big(\hat{s}_i - \frac{\nsyn}{N}\Big)^2,
+pMSE = \frac{1}{N} \sum_{i=1}^N \Big(\hat{\pi}_i - \frac{\nsyn}{N}\Big)^2,
 $$ {#eq-pMSE}
@@
 Crucially, the $pMSE$ depends on the classification model used and increases in the flexibility of the classification model, making it prone to overfitting and hard to interpret.
-To combat these issues, @snoke_utility_2018 suggests to compare the $pMSE$-value with its expectation under the null distribution.
+To combat these issues, @snoke_utility_2018 suggests to compare the $pMSE$-value with its expected value under the null hypothesis that the observed and synthetic data are indistinguishable.
 Provided that the classification model is a logistic regression model with $k$ parameters (including the intercept), @snoke_utility_2018 show that the expected $pMSE$ is given by
@@
 $$
-pMSE\text{-ratio} = \frac{pMSE}{\mathbb{E}[pMSE]},
+pMSE\text{-ratio} = \frac{pMSE}{\mathbb{E}[pMSE]}.
 $$
-with values smaller than $10$ deemed acceptable [@raab2021assessing\; although the authors remark that the smaller the better].
-Apart from the $pMSE$, several other measures exist that can be constructed from the estimated propensity scores, such as the percentage of records correctly predicted [@raab2021assessing] or the Kolmogorov-Smirnov statistic [@Bowen_differentially_2021], both which are strongly correlated with the $pMSE$ [@raab2021assessing].
+Apart from the $pMSE$, several other measures can be constructed from the estimated propensity scores, such as the percentage of records correctly predicted [@raab2021assessing] or the Kolmogorov-Smirnov statistic [@Bowen_differentially_2021], both which are strongly correlated with the $pMSE$ [@raab2021assessing].
 
@@
 @drechsler_utility_2022 illustrated that the utility score is highly dependent on the model used to estimate the propensity scores, and that clear improvements in synthesis models are not necessarily picked up in the $pMSE$.
-Moreover, the $pMSE$ is prone to overfitting and users may find it difficult to select an appropriate model for estimating the propensity scores.
-Hence, current methods to evaluate global utility are either difficult to estimate, especially in high-dimensional situations, or depend strongly on model specification.
-Density ratio estimation might alleviate these issues.
-In what follows, we explain how the density ratio can be estimated, and how it can be used to evaluate the utility of synthetic data.
+Moreover, selecting an appropriate propensity score model may be difficult, as the challenges associated with model selection, such as the bias-variance trade-off, typically apply.
+
+Another way to formalize distributional similarity is through the Kullback-Leibler (KL) divergence, as proposed in @karr_utility_2006.
+The KL-divergence measures the relative entropy from the probability distribution of the observed data $\pobs(\bx)$ to the probability distribution of the synthetic data $\psyn(\bx)$ (with $\bx \in \mathbb{R}^{d}$), and is defined as^[In contrast to @karr_utility_2006, we define the KL-divergence as the relative entropy of $\pobs$ with respect to $\psyn$, such that it is consistent with our later formulations.]
+$$
+D_{\text{KL}}(\pobs || \psyn) = 
+\int \pobs(\bx) \log \frac{\pobs(\bx)}{\psyn(\bx)} \text{d} \bx.
+$${#eq-kl-div}
+@karr_utility_2006 argue that the KL-divergence can be computed by approximating the integral using density estimators $\hat{p}_\text{obs}$ and $\hat{p}_\text{syn}$. 
+An implementation of this approach for divergence estimation has been described in @wang_divergence_2009, where the densities are estimated using nearest-neighbor density estimation. 
+Another approach posed by @karr_utility_2006 is to assume multivariate normality for both $\psyn$ and $\pobs$, and calculate the KL-divergence in closed form from the sample means and covariance matrices.
+Although all these approaches are easy to implement, they have their limitations. 
+Density estimation is challenging in high-dimensional settings, and assuming multivariate normality might be too restrictive. 
+We show in later sections that the KL-divergence naturally fits into the density ratio estimation framework.
 
@@
 The framework has been shown to be applicable to prediction [@sugiyama_conditional_2010; @sugiyama_classification_2010], outlier detection [@shohei_dre_outlier_2008], change-point detection in time-series [@liu_change_2013], importance weighting under domain adaptation [i.e., sample selection bias\; @kanamori_ulsif_2009], and two-sample homogeneity tests [@sugiyama_lstst_2011].
-The general idea of density ratio estimation is depicted in @fig-dr-plot, and boils down to comparing two distributions by modelling the density ratio $r(\bx)$ between the probability distributions of the numerator samples, taken from the synthetic data distribution, $\psyn(\bx)$, and the denominator samples, taken from the observed data distribution, $\pobs(\bx)$, such that
+The general idea of density ratio estimation is depicted in @fig-dr-plot, and boils down to comparing two distributions by modelling the density ratio $r(\bx)$.
+In our case, we define $r(\bx)$ as the ratio between the probability distributions of the numerator samples, taken from the observed data distribution, $\pobs(\bx)$, and the denominator samples, taken from the synthetic data distribution, $\psyn(\bx)$, such that
 $$
-r(\bx) = \frac{\psyn(\bx)}{\pobs(\bx)}.
+r(\bx) = \frac{\pobs(\bx)}{\psyn(\bx)}.
 $$ {#eq-dr}
 
-This specification has the intuitive interpretation that in locations where the density ratio takes large values, too many synthetic observations will be generated in that region and at the locations where the density ratio is small, there will be too few synthetic observations, both relative to the observed data.
-An intuitive approach to estimating $r(\bx)$ from samples of $\pobs(\bx)$ and $\psyn(\bx)$ would be to estimate the observed and synthetic data density separately, for example using kernel density estimation [e.g., see @Scott1992 for an overview], and subsequently compute the ratio of these estimated densities.
-However, density estimation is one of the hardest tasks in statistical learning, unavoidably leading to estimation errors for both densities, especially in high dimensions.
+This specification has the intuitive interpretation that in locations where the density ratio takes values larger than 1, too few synthetic observations are generated in that region.
+Conversely, in locations where the density ratio takes values smaller than one, too many synthetic observations are generated in that region (see @fig-dr-plot).
+We put $\pobs(\bx)$ in the numerator such that it plays the role of the actual distribution, while $\psyn(\bx)$ serves as the approximating distribution (we briefly come back to this in the next section).
+A straightforward approach to estimating $r(\bx)$ from samples of $\pobs(\bx)$ and $\psyn(\bx)$ would be to estimate the observed and synthetic data density separately, for example using kernel density estimation [e.g., see @Scott1992 for an overview], and subsequently compute the ratio of these estimated densities.
+However, density estimation is one of the hardest tasks in statistical learning, unavoidably leading to estimation errors for both densities, especially in high dimensions [@sugiyama_suzuki_kanamori_2012].
 When subsequently taking the ratio of the estimated densities, estimation errors tend to be magnified. 
 Direct density ratio estimation avoids this issue by specifying and estimating a model directly for the ratio without first estimating the separate densities.
-Extensive simulations on a wide variety of tasks showed that this approach typically outperforms density ratio estimation through naive kernel density estimation, especially when the dimensionality of the data increases [e.g., @Kanamori2012; @shohei_dre_outlier_2008; @kanamori_ulsif_2009].
+Extensive simulations on a wide variety of tasks showed that this approach typically outperforms density ratio estimation through kernel density estimation for each distribution separately, especially when the dimensionality of the data increases [e.g., @Kanamori2012; @shohei_dre_outlier_2008; @kanamori_ulsif_2009].
 
@@
 #| fig-cap: "Example of the true and estimated density ratio of two normal distributions with different means and variances (i.e., $\\psyn(\\bx) = N(0,1)$ and $\\pobs(\\bx) = N(1,2)$).
-The function $r(\\bx) = \\psyn(\\bx)/\\pobs(\\bx)$ denotes the true density ratio, the function $\\hat{r}(\\bx)$ denotes an estimate of the density ratio based on $\\nsyn = \\nobs = 200$ samples from each distribution obtained with unconstrained Least-Squares Importance Fitting (uLSIF).
+The function $r(\\bx) = \\pobs(\\bx)/\\psyn(\\bx)$ denotes the true density ratio, the function $\\hat{r}(\\bx)$ denotes an estimate of the density ratio based on $\\nsyn = \\nobs = 200$ samples from each distribution obtained with unconstrained Least-Squares Importance Fitting (uLSIF).
 Note that the density ratio is itself not a proper density."
@@
 ggplot() +
-  stat_function(aes(col = "A"), 
+  stat_function(aes(col = "A", linetype = "A"), 
                 fun = dnorm, args = list(mean = 0, sd = 1),
                 size = 0.5) +
-  stat_function(aes(col = "B"),
+  stat_function(aes(col = "B", linetype = "B"),
                 fun = dnorm, args = list(mean = 1, sd = sqrt(2)),
                 size = 0.5) +
-  scale_color_brewer(labels = c("A" = expression(italic(p)[syn](x)),
-                                "B" = expression(italic(p)[obs](x))),
+  scale_color_brewer(labels = c("A" = expression(italic(p)[obs](x)),
+                                "B" = expression(italic(p)[syn](x))),
                      palette = "Set2",
+                     name = "") +
+  scale_linetype_manual(labels = c("A" = expression(italic(p)[obs](x)),
+                                  "B" = expression(italic(p)[syn](x))),
+                        values = c("solid", "dashed"),
                         name = "") +
@@
   theme(legend.position = "bottom",
-        text = element_text(family = "LM Roman 10"),
+        text = element_text(family = "LM Roman 10", size = 9),
         legend.background = element_blank()) +
 ggplot() +
-  stat_function(aes(col = "C"),
+  geom_hline(mapping = aes(yintercept = 1), alpha = 0.3, size = 0.3) +
+  stat_function(aes(col = "C", linetype = "C"),
                 fun = ~ dnorm(.x, 0, 1) / dnorm(.x, 1, sqrt(2)),
                 size = 0.5) +
-  stat_function(aes(col = "D"),
+  stat_function(aes(col = "D", linetype = "D"),
                 fun = ~ predict(rhat, newdata = .x),
@@
                      name = "") +
+  scale_linetype_manual(labels = c("C" = expression(italic(r(x))),
+                                   "D" = expression(italic(hat(r)(x)))),
+                        values = c("solid", "dashed"),
+                        name = "") +  
   xlim(-4, 6) +
@@
   theme(legend.position = "bottom",
-        text = element_text(family = "LM Roman 10"))
+        text = element_text(family = "LM Roman 10", size = 9)) &
+  theme(plot.margin = unit(c(0.2, 0.2, 0, 0.2), "cm"),
+        legend.margin = unit(c(0.1, 0.1, 0.1, 0.1), "cm"))
 ```
@@
 Over the past years, several methods for direct density ratio estimation have been developed.
-A large class of these methods attempt to directly minimize the error between the true density ratio $r(\bx)$ and its estimate $\hat{r}(\bx)$.
+A large class of these methods attempt to directly minimize the error between the true density ratio $r(\bx)$ and a density ratio model $\hat{r}(\bx)$.
 Following this approach, we define a loss function $\mathcal{L}(r(\bx), \hat{r}(\bx))$ that measures the discrepancy between the true and estimated density ratio. 
@@
 \begin{aligned}
-\mathcal{L}_S(r(\bx), \hat{r}(\bx)) &= 
-\frac{1}{2} \int (\hat{r}(\bx) - r(\bx))^2 \pobs(\bx) \text{d}\bx \\
-&= \frac{1}{2} \int \hat{r}(\bx)^2 \pobs(\bx) \text{d}\bx - 
-\int \hat{r}(\bx) r(\bx) \pobs(\bx) \text{d}\bx + 
-\frac{1}{2} \int r(\bx)^2 \pobs(\bx) \text{d}\bx.
+\mathcal{L}^*_S(r(\bx), \hat{r}(\bx)) &= 
+\frac{1}{2} \int (\hat{r}(\bx) - r(\bx))^2 \psyn(\bx) \text{d}\bx \\
+&= \frac{1}{2} \int \hat{r}(\bx)^2 \psyn(\bx) \text{d}\bx - 
+\int \hat{r}(\bx) r(\bx) \psyn(\bx) \text{d}\bx + 
+\frac{1}{2} \int r(\bx)^2 \psyn(\bx) \text{d}\bx.
 \end{aligned}
 $$ {#eq-squared-error}
-The second term can be rewritten, because the denominator in $r(\bx)$ cancels with $\pobs(\bx)$, while the third term is a constant with respect to the parameters in the density ratio model and can thus be ignored.
+The second term can be rewritten, because the denominator in $r(\bx)$ cancels with $\psyn(\bx)$, while the third term is a constant with respect to the parameters in the density ratio model and can thus be ignored.
 Hence, we are left with the following loss function to minimize
 $$
-\mathcal{L}'_S (r(\bx), \hat{r}(\bx)) = \frac{1}{2} \int \hat{r}(\bx)^2 \pobs(\bx) \text{d}\bx - \int \hat{r}(\bx) \psyn(\bx) \text{d}\bx,
+\mathcal{L}_S (r(\bx), \hat{r}(\bx)) = \frac{1}{2} \int \hat{r}(\bx)^2 \psyn(\bx) \text{d}\bx - \int \hat{r}(\bx) \pobs(\bx) \text{d}\bx,
 $$ {#eq-squared-error-loss}
@@
 $$
-\mathcal{L}_f (r(\bx), \hat{r}(\bx)) =
+\mathcal{L}^*_f (r(\bx), \hat{r}(\bx)) =
 \int \Big(
 f(r(\bx)) - f(\hat{r}(\bx)) - f'(\hat{r}(\bx))(r(\bx) - \hat{r}(\bx))
-\Big) \pobs(\bx) \text{d}\bx,
+\Big) \psyn(\bx) \text{d}\bx,
 $$ {#eq-bregman-loss}
 where $f$ is a differentiable and strictly convex function with derivative $f'$.
-Then, ignoring the terms independent of $\hat{r}(\bx)$ and noting that $r(\bx)\pobs(\bx) = \psyn(\bx)$, we obtain the following objective
+Then, ignoring the terms independent of $\hat{r}(\bx)$ and noting that $r(\bx)\psyn(\bx) = \pobs(\bx)$, we obtain the following objective
 $$
-\mathcal{L}'_f (r(\bx), \hat{r}(\bx)) = \int 
+\mathcal{L}_f (r(\bx), \hat{r}(\bx)) = \int 
 \Big( 
 f'(\hat{r}(\bx)) \hat{r}(\bx) - f(\hat{r}(\bx))
-\Big)  \pobs(\bx) \text{d}\bx
-- \int f'(\hat{r}(\bx)) \psyn(\bx)
+\Big)  \psyn(\bx) \text{d}\bx
+- \int f'(\hat{r}(\bx)) \pobs(\bx)
 \text{d}\bx.
 $$ {#eq-bregman-objective}
-Minimizing this loss for different functions $f$ yields different estimators for the density ratio, that focus on different regions of the density ratio.^[It is easy to see that using $f(x) = \frac{1}{2}(x-1)^2$ turns the Bregman divergence (@eq-bregman-objective) into the squared error (@eq-squared-error-loss).]
-Specifically, some estimators place more emphasis on accurately modelling the regions in which the true density ratio is large, but less emphasis on accurately estimating the smaller density ratios, and vice versa [see @sugiyama_bregman_2012; @menon2016dreloss].
-Importantly, @sugiyama_bregman_2012 show that the Bregman divergence minimization approach to density ratio estimation is equivalent to estimating the $f$-divergence [@ali_silvey_divergence_1966] between the synthetic and observed data distributions.
-Hence, when estimating the density ratio, one implicitly estimates the divergence between the synthetic and observed data distributions.
+Minimizing this loss over $\hat{r}(\bx)$ for different functions $f$ yields different estimators for the density ratio, that focus on different regions of the density ratio.^[It is easy to see that using $f(x) = \frac{1}{2}(x-1)^2$ turns the Bregman divergence (@eq-bregman-loss) into the squared error (@eq-squared-error).]
+Specifically, some estimators place more emphasis on accurately modelling the regions in which the true density ratio is large, but less emphasis on accurately estimating the density ratio in regions where it takes small values, and vice versa [see @sugiyama_bregman_2012; @menon2016dreloss].
+Importantly, @sugiyama_bregman_2012 show that the Bregman divergence minimization approach to density ratio estimation is equivalent to estimating an $f$-divergence [@ali_silvey_divergence_1966] of the observed data distribution from the synthetic data distribution.
+Hence, when estimating the density ratio, one implicitly estimates the divergence between the observed and synthetic data distributions.
 
@@
 $$ {#eq-dr-model}
-where $\boldsymbol{\varphi}(\bx)$ is a basis function vector, that transforms the data from a $p$-dimensional to a $b$-dimensional space, and $\boldsymbol{\theta}$ is a $b$-dimensional parameter vector.
+where $\boldsymbol{\varphi}(\bx)$ is a basis function vector (optionally including an intercept term), that transforms the data from a $p$-dimensional to a $b$-dimensional space, and $\boldsymbol{\theta}$ is a $b$-dimensional parameter vector.
 Although the model is linear in its parameters, the density ratio itself is typically a non-linear function of the data due to the basis functions.
 The parameter vector $\boldsymbol{\theta}$ is estimated to minimize the discrepancy with the true density ratio using the loss function $\mathcal{L}(r(\bx), \hat{r}(\bx))$ in @eq-bregman-objective.
+Potentially, the loss function can be extended with a regularization term (e.g., a $L_1$ or $L_2$ norm on the parameters of the model), to decrease the flexibility of the density ratio.
+Parameter estimation can be carried out using standard optimization techniques (see Appendix [-@sec-app-A] for a brief demonstration in `R`).
 
 Also for the basis functions, several specifications are possible, ranging from an identity function [@qin_inferences_1998] to normalizing flows [@choi_featurized_2021] or neural networks [@tiao2018dre; @uehara2016generative].
-However, a commonly used basis function in the density ratio literature is the Gaussian kernel function [e.g., @huang_kmm_2006; @sugiyama_kliep_2007; @kanamori_ulsif_2009; @liu_change_2013; @gruber2024overcoming], which conveniently allows to model the non-linear density ratio using models that are linear in their parameters.
+A commonly used basis function in the density ratio literature is the Gaussian kernel function [e.g., @huang_kmm_2006; @sugiyama_kliep_2007; @kanamori_ulsif_2009; @liu_change_2013; @gruber2024overcoming].
 The Gaussian kernel function is defined as
 $$
-\boldsymbol{\varphi}(\bx) = \mathcal{K}(\bx_i, \boldsymbol{c}_j) = 
+\boldsymbol{\varphi}(\bx) = \mathcal{K}(\bx, \boldsymbol{c}_j) = 
 \exp\left(
@@
 $$
-where $\boldsymbol{c}_j$ ($j \in 1, \dots, J$) denotes the centers of the Gaussian kernel functions, and $\sigma$ controls the kernel width [i.e., it defines over which distance differences between the observations and the centers are considered relevant\; for more information on kernel functions, see, e.g., @murphy_pmlintro_2022].
-The appropriate width of the kernel can be determined using cross-validation. 
-The centers are typically sampled from the data (in our case, the observed and synthetic data can both be used, but it is also possible to solely use the synthetic samples).
+where $\boldsymbol{c}_j$ ($j \in 1, \dots, b$) denotes the centers of the Gaussian kernel functions, and $\sigma$ controls the kernel width [i.e., it defines over which distance differences between the observations and the centers are considered relevant\; for more information on kernel functions, see, e.g., @murphy_pmlintro_2022].
+The appropriate width of the kernel and the optimal value of the regularization parameter can be determined using cross-validation. 
+The centers are typically sampled from the data (in our case, both the observed and synthetic data).
 
-After defining the model and the loss function, the density ratio can be easily estimated. 
+After defining the model and the loss function, the density ratio can be estimated. 
 The `densityratio` package in R [@densityratio] provides an implementation of commonly used loss functions with a Gaussian kernel basis function.
-The package comes with an easy-to-use interface, automatic cross-validation of hyperparameters and builds on `C++` for fast computation. 
-However, for increased flexibility with respect to loss functions and basis functions, one could use classic optimization techniques as gradient descent or (quasi-)Newton methods, for example as implemented in the `optim` function in the statistical software `R` [@R].
-Both approaches are briefly illustrated in Appendix [-@sec-app-A] (TODO).
+The package comes with an easy-to-use interface, automatic cross-validation of hyperparameters and builds on `C++` for fast computation (see Appendix [-@sec-app-A] for a brief demonstration).
+The density ratio estimation framework thus equips researchers with a measure of utility which is accurate, fast to compute and does not require user specification of the model.
 
@@
 
-After estimating the density ratio, the information provided by the density ratio can be summarized in a divergence measure, using the fact that estimating the density ratio is equivalent to estimating the $f$-divergence between the synthetic and observed data distributions [@sugiyama_bregman_2012].
-Accordingly, we can use the estimated density ratio function to estimate an $f$-divergence between the observed and synthetic data, and use this divergence measure as a utility measure for the synthetic data.
-Although this divergence statistic is difficult to interpret in an absolute sense, it can be used as a relative measure of quality of the synthetic data.
+Having estimated the density ratio, we can evaluate both global and local utility of the synthetic data.
+Like other global utility measures, the density ratio can be used to construct a single discrepancy measure that quantifies the degree of misfit of the synthetic data. 
+The density ratio can be used to directly calculate some $f$-divergence between the observed and synthetic data distributions, which serves as a measure of utility for the synthetic data. 
+Although this divergence statistic is difficult to interpret in an absolute sense, it can be used as a relative measure of the quality of the synthetic data. 
 That is, for different synthetic data sets, we can calculate the divergence to the observed data, and compare these values to determine which synthetic data set is most similar to the observed data.
-In a more formal way, the $f$-divergence gives rise to a test statistic that can be used to test the null hypothesis that the synthetic data is generated from the same distribution as the observed data.
-The corresponding $p$-value can be calculated by comparing the observed test statistic to the null distribution of test statistics obtained using random permutations of the observed and synthetic data [see, for example, @sugiyama_lstst_2011; @Wornowizki2016; @kanamori_divergence_2012].
-Lastly, the density ratio function itself can be plotted against individual variables or pairs of variables to discover whether particular variables are particularly poorly captured in the synthetic data.
-Each of these strategies can be readily executed using the `densityratio` package in `R` [@densityratio], and will be further illustrated in the upcoming sections.
+More formally, the $f$-divergence can be used in a hypothesis test to determine whether the synthetic data is generated from the same distribution as the observed data.
+The corresponding $p$-value can be calculated by comparing the observed divergence measure to the null distribution of divergence measures obtained using random permutations of the observed and synthetic data [see, for example, @sugiyama_lstst_2011; @Wornowizki2016; @kanamori_divergence_2012].
 
-**Ik twijfel nog een beetje over de volgende zaken, en voordat ik alles ga zitten uitwerken hoor ik graag eerst jullie mening:**
+An appealing characteristic of the density ratio framework is that it also readily allows for evaluating local utility. 
+As said, the estimated density ratio shows in which regions the synthetic data diverges from the observed data. 
+By inspecting which observations have a high, or small, estimated density ratio, we can identify the regions in which the synthetic data is most different from the observed data.
+Visual inspection can aid this process, for example by plotting the estimated density ratio against variables or pairs of variables in the data.
+On an even finer scale, the density ratio can be used to identify individual observations that are far from what can be expected on the basis of the observed data. 
+This information could identify flaws in the synthetic data model, or could be used directly to remove such observations from the synthetic data. 
+An additional advantage of the density ratio lies in the fact that for every synthetic observation, the predicted density ratio value can be regarded as an importance weight. 
+Releasing the predicted density ratio values for each synthetic observation allows data users to reweigh analyses on the synthetic data.
+We illustrate these features in @sec-app.
 
-1.
-Ik heb nu het deel over regularisatie eruit gelaten, omdat ik vond dat het wat afleid van de main ideas, maar kan deze ook gewoon toevoegen aan @eq-bregman-objective.
-Het kan ook casual opgemerkt worden wanneer het daadwerkelijk gebruikt wordt.
-Wat vinden jullie?
-
-2.
-Dimension reduction: ik was niet zo zeker hoe ik dit er direct moest bouwen, want effectief is het niet veel anders dan de huidige aanpak, alleen met een andere basis function (namelijk $\phi(x) = \mathcal{K}(x_i^T U, c_j^T U)$), waarbij U ook parameters heeft die geschat worden.
-Anyway, ik kan aan het eind een sectie maken met dat DRE nog steeds lastig kan zijn in hoge dimensies, en dat daar oplossingen voor zijn (bijvoorbeeld een LFDA-stap voor ulsif, of LHSS, of the spectral density ratio estimation methode (laatste twee zitten allebei in het package)).
-Het punt is vooral dat ik niet wil dat het teveel afleidt van het main punt, namelijk density ratio estimation, maar qua content zou het misschien een eigen sectie verdienen, ook omdat het niet direct enorm aansluit bij de twee secties hierboven.
-
-<!-- The difficulty of density ratio estimation increases with the dimensionality of the data.  -->
-<!-- Therefore, we follow previous recommendations to incorporate dimensionality reduction techniques in density ratio estimation.  -->
-
-<!-- Shortly name examples of dimensionality reduction techniques (i.e., PCA; LFDA or UMAP). -->
-
-<!-- A useful by-product of dimension reduction is that it allows to create visualizations, and these visualizations can be used to get more insight in discrepancies between observed and altered data.
--->
-<!-- Show what such visualizations can look like, and how they can help. -->
-
-3.
-Denken jullie dat dit een goede plek is om wat dieper in te gaan op mogelijke voordelen van density ratio estimation?
-Ik dacht dat na sectie 2.2.2 een voor de hand liggende plek zou zijn om hier wat dieper op in te gaan, en dan specifiek de volgende punten te benoemen: utility quantifications at every point in multivariate space, easy and flexible model specification through automatic model selection with cross-validation, potential for importance weighting, wellicht dan ook direct het punt van dimension reduction hierbij pakken.
-
-4.
-Ik wil graag nog ergens iets zeggen over de relaties tussen density ratio estimation en de andere methodes (KL-divergence, pMSE), maar ik vraag me een beetje af of dat niet beter in de discussie kan.
-
-<!-- Relate density ratio estimation to $pMSE$ and KL divergence (to some extent, both are generalizations of density ratio estimation, or at least are conceptually similar). -->
-<!-- Give some more information on the $pMSE$, describe what it shortcomings are.  -->
-<!-- The quality of the $pMSE$ highly depends on the model used to calculate the propensity scores.  -->
-<!-- Perhaps give an example of logistic regression, which basically estimates whether the conditional mean of the observed and altered data is the same.  -->
-<!-- Explain how density ratio estimation can overcome the shortcomings of the previously mentioned methods. -->
-
-5.
-Is dit echt de goede plaats om iets te zeggen over categorische variabelen?
-Deze kunnen namelijk gewoon geïncludeerd worden, maar met een Gaussian kernel wordt dat misschien een beetje awkward.
-Desalniettemin kan het prima, en kunnen we er ook voor kiezen om in de discussie te benoemen dat we een oplossing hebben gekozen, maar dat er ook andere opties zijn (bijv.
-andere kernels).
+A final advantage of the density ratio estimation framework lies in its flexibility.
+The combination of a non-parametric kernel-based model with automatic cross-validation for hyperparameter selection (e.g., kernel and regularization parameters) allows for flexible estimation of the density ratio.
+This flexibility reduces the burden on the user in terms of model specification when evaluating the utility of the synthetic data. 
+All functionality described above, including visualization, divergence estimation (and testing), extracting importance weights and automatic cross-validation for hyperparameter selection, is readily available in the `densityratio` `R`-package [@densityratio].
+The package provides fast and efficient algorithms for density ratio estimation, and uses default implementations that work across a wide range of data distributions, which will be shown in the upcoming simulations.
 
@@
 
-To evaluate the performance of density ratio estimation and compare it to existing methods, we conduct a series of simulation studies. 
-We consider a variety of scenarios, including univariate and multivariate models, varying correlations, number of variables, and sample sizes.
-To perform density ratio estimation, we use unconstrained Least-Squares Importance Fitting [uLSIF\; @kanamori_ulsif_2009] as implemented in the `densityratio` package in `R` [@densityratio], because it fast and has been shown to perform well in a variety of settings [ADD CITATIONS <!-- TODO: ADD CITATIONS-->].
-We compare the performance of uLSIF to the performance of the $p$-MSE as implemented in the `R`-package `synthpop`.
+To evaluate the performance of density ratio estimation and compare it to existing methods, we conduct a series of simulation studies (all code is available on GitHub^[[https://github.com/thomvolker/dr-utility](https://github.com/thomvolker/dr-utility)]).
+First, we illustrate the approach in a univariate simulation, after which we evaluate the performance of the approach in a multivariate setting, varying the distributions, number of variables and sample sizes.
+To perform density ratio estimation, we use unconstrained Least-Squares Importance Fitting [uLSIF\; @kanamori_ulsif_2009] as implemented in the `densityratio` package in `R` [@densityratio], because it fast and has been shown to perform well in a variety of settings [e.g., @kanamori_ulsif_2009; @li_application_2010; @Kanamori2012].
+We compare the performance of uLSIF to the performance of state-of-the-art software implementations of the $pMSE$ [@synthpop] and Kullback-Leibler divergence estimation through $k$-nearest neighbor density estimation [@kldest].
+Note that Kullback-Leibler divergence estimation through $k$-nearest neighbor density estimation implicitly performs direct density ratio estimation, and the procedure can be regarded as a special case of density ratio estimation (see Appendix [-@sec-app-B]).
 
-## Univariate simulations
+## Univariate illustrations {#sec-univ-sims}
 
-For the sake of illustrational clarity, we first evaluate the performance of density ratio estimation in a univariate setting where its performance can be evaluated using visualizations.
-In the simulations, we generate data according to four data-generating mechanisms, (1) a Laplace distribution, (2) a log-normal distribution, (3) a location-scale $t$-distribution and  (4) a normal distribution.
+We first illustrate the behavior of density ratio estimation in a univariate setting, which simplifies visualization of the estimated density ratios. 
+Specifically, we consider four settings in which we generate the real data from a distribution, and generate synthetic data from an approximating Gaussian distribution with the same mean and variance as the true data generating mechanism.
+The real data is generated from (1) a Laplace distribution, (2) a log-normal distribution, (3) a location-scale $t$-distribution and  (4) a normal distribution (see @fig-densities-sim1).
 Subsequently, we approximate the true data generating mechanism using a Gaussian model with the same mean and variance as the original data. 
-This setting is similar to situations that are commonly encountered in the synthetic data field, in the sense that the true data generating mechanism is unknown and needs to be approximated using a simpler approximating model.
+This setting is similar to situations commonly encountered in the synthetic data field, in the sense that the true data generating mechanism is unknown and needs to be approximated using a simpler approximating model.
 The exact specifications of the data generating mechanisms are as follows:
@@
 1. Laplace distribution with location parameter $\mu = 1$ and scale parameter $b = 1$.
-2. Log-normal distribution with log-mean parameter $\mu_{\text{log}} = \log \{1/\sqrt{3} \}$ and log-standard deviation parameter $\sigma_\text{log} = \sqrt{\log 3}$.
+2. Log-normal distribution with log-mean parameter $\mu_{\text{log}} = \log \{1/\sqrt{3} \}$ and log-variance parameter $\sigma^2_\text{log} = \log 3$.
 3. Location-scale $t$-distribution with location parameter $\mu = 1$, scale parameter $\tau^2 = 1$ and degrees of freedom $\nu = 4$. 
@@
 These data generating mechanisms are chosen such that they all have the same population mean $\mu = 1$ and variance $\sigma^2 = 2$.
-The approximating model is a Gaussian distribution with mean $\mu = 1$ and variance $\sigma^2 = 2$.
-This distribution has the same mean and variance, but differs in higher-order moments, except in the fourth scenario. 
-In the last simulation, we model the data generating mechanism correctly to obtain some understanding into how density ratio estimation behaves under a well-specified synthesis model.
-In all scenarios, we generate $500$ data sets with $\nobs = 250$ observations from the true data generating mechanism, and $\nsyn = 250$ synthetic observations from the approximating Gaussian model. 
-The density ratio model is estimated using uLSIF with a Gaussian kernel and an $L_2$-regularization parameter.
-Both the kernel bandwidth and the regularization parameter are selected using cross-validation, using the default settings as implemented in the `densityratio` package.
+The approximating Gaussian distribution has mean $\mu = 1$ and variance $\sigma^2 = 2$.
+Hence, the synthetic data has the same population mean and variance, but differs in higher-order moments, except in the fourth scenario, where it is equal to true data generating mechanism.
+In all scenarios, we generate $1000$ data sets with $\nobs = 250$ observations from the true data generating mechanism, and $\nsyn = 250$ synthetic observations from the approximating Gaussian model. 
+The density ratio model is estimated using uLSIF with the default Gaussian kernel and $L_2$-penalty on the parameter vector $\boldsymbol{\theta}$.
+Both the kernel bandwidth and the regularization parameter are selected using cross-validation, using the default settings in the `densityratio` package.
 
@@
 
-source("../code/functions.R")
+source("code/functions.R")
 ```
@@
 #| out-width: 100%
-#| fig-height: 4
+#| fig-height: 3
 #| dev: cairo_pdf
@@
         text = element_text(family = "LM Roman 10"),
-        legend.background = element_blank()) +
+        legend.background = element_blank(),
+        plot.margin = unit(c(0.1, 0.2, 0.1, 0.2), "cm"),
+        legend.margin = unit(c(0.1, 0.1, 0.1, 0.1), "cm")) +
   ggplot() +
@@
         legend.background = element_blank(),
-        text = element_text(family = "LM Roman 10"))
+        text = element_text(family = "LM Roman 10"),
+        plot.margin = unit(c(0.1, 0.2, 0.1, 0.2), "cm"),
+        legend.margin = unit(c(0.1, 0.1, 0.1, 0.1), "cm")) &
+  theme(plot.margin = unit(c(0.1, 0.2, 0.1, 0.2), "cm"),
+        legend.margin = unit(c(0.1, 0.1, 0.1, 0.1), "cm"))
 ```
@@
 #| label: code-sim1
-#| eval: false
+#| eval: !expr params$runsims
 #| include: false
 
-source("../code/sim1.R")
+source("code/sim1.R")
 ```
@@
 #| out-width: 100%
-#| fig-height: 4
-#| dev: cairo_pdf
-#| fig-cap: "Estimated density ratios by unconstrained least-squares importance fitting in four univariate examples: A Laplace distribution, a log-normal distribution, a location-scale $t$-distribution and a normal distribution, all approximated by a normal distribution with the same mean and variance as the sample from the true distribution."
+#| fig-height: 3
+#| dev: ragg_png
+#| fig-cap: "Estimated density ratios by unconstrained least-squares importance fitting in four univariate examples: A Laplace distribution, a log-normal distribution, a location-scale $t$-distribution and a normal distribution, all approximated by a normal distribution with the same mean and variance as the sample from the true distribution.
+Note that the mass of the synthetic data distribution in the tails (smaller than $-2$ or greater than $4$) is smaller than $0.034$."
 
-load("../results/sim1.RData")
-
-# evaluation data for plotting
-x_eval <- seq(-3, 5, length.out = 1600)
+load("results/sim1.RData")
 
@@
 sims |>
-  mutate(model = factor(model,
-                        levels = c("laplace", "lognormal", "t", "normal"),
-                        labels = c("Laplace", "Log-normal", "italic(t)", "Normal"),
-                        ordered = TRUE),
-         xpreds = list(x_eval),
-         ypreds = map2(r, xpreds, ~predict(.x, .y))) |>
-  unnest(cols = c(xpreds, ypreds)) |>
+  unnest(c(xpreds, ypreds)) |>
+  group_by(model, xpreds) |>
+  mutate(ymean = mean(ypreds)) |>
   ggplot(aes(x = xpreds, y = ypreds, group = interaction(model, sim))) +
-  geom_line(alpha = 0.1, col = "steelblue3") +
-  geom_line(data = true_ratio, aes(x = xpreds, y = ypreds, group = NULL), col = "black") +
+  geom_line(aes(col = "A", linetype = "A"), alpha = 0.05) +
+  geom_line(aes(y = ymean, group = model, linetype = "B", col = "B"),
+            size = 0.3) +
+  geom_line(data = true_ratio |> unnest(c(xpreds, ypreds)), 
+            aes(x = xpreds, y = ypreds, group = NULL, linetype = "C", col = "C")) +
   facet_wrap(~model, labeller = label_parsed) +
-  ylim(0, 6) +
   theme_minimal() +
   xlab(expression(italic(x))) +
-  ylab(expression(hat(italic(r))(x)))
+  ylab(expression(hat(italic(r))(x))) +
+  scale_linetype_manual(
+    values = c("A" = "solid", "B" = "dashed", "C" = "solid"),
+    labels = c("A" = "Estimated density ratio",
+               "B" = "Average estimated density ratio",
+               "C" = "True density ratio"),
+    name = "") +
+  scale_color_manual(
+    name = "",
+    values = c("A" = "steelblue3", "B" = "black", "C" = "black"),
+    labels = c("A" = "Estimated density ratio",
+               "B" = "Average estimated density ratio",
+               "C" = "True density ratio")
+  ) +
+  scale_y_continuous(breaks = 0:6, limits = c(0,6)) +
+  theme(legend.position = "bottom",
+        text = element_text(family = "LM Roman 10", size = 8),
+        legend.margin = margin(0,0,0,0),
+        legend.box.margin = margin(-10, -10, 0, -10)) +
+  guides(colour = guide_legend(override.aes = list(alpha = 1)))
 ```
 
-TODO: Add interpretation of output; add significance tests and compare with $pMSE$ and kolmogorov-smirnov.
+@fig-sim1-results shows that the estimated density ratios (the blue lines) follow the general trend of the true density ratio (the solid black line) for each data-generating mechanism.
+Especially in the center of each panel, approximately between $-2$ and $4$, the average density ratio (the dashed black line) closely matches the true density ratio.
+Moreover, all estimated density ratios model the peak in the center of each panel, where the true data generating distribution has more mass than the synthetic data distribution  (except the bottom right panel, where the true and synthetic data distributions are the same).
+When moving from the middle to the sides, where the synthetic data distribution has more mass than the true data generating distributions, the estimated density ratios still follow the true density ratios, in the sense that they are typically smaller than $1$.
+In the tails of the distribution, where the synthetic data distribution has almost no mass ($P(x \leq -2) = P(x \geq 4) \approx 0.017$), estimating the density ratio becomes difficult, leading to estimated density ratios that diverge from the true density ratios
+In these regions, the kernel function yields low values for most observations, which, combined with the $L_2$-penalty on the density ratio parameters, yields a downward bias for the estimated density ratio.
+
+When assessing the panels individually, the following observations can be made. 
+In the top-left panel, showing the estimated density ratio of the Laplace-distributed samples over the synthetic Gaussian-distributed samples, the peak in the center is well approximated. 
+Yet, the discontinuity in the peak of the Laplace distribution cannot be captured by a Gaussian kernel, which by definition models the density ratio as a smooth function.
+The same can be said about the log-normal distribution, where the cut-off at $0$ is not captured exactly. 
+Moreover, this panel clearly shows the shrinkage due to the regularization parameter. 
+When there are no samples in one of the two distributions, the regularization has relatively more influence and shrinks the density ratio parameters towards zero.
+Accordingly, the estimated density ratio is shrunk towards the estimated intercept parameter.
+The same behavior can be seen in the bottom-left panel, showing the estimated density ratios of location-scale $t$-distribution over the Gaussian samples. 
+In the centre of the distributions, the density ratio is well approximated, but the density ratio is consistently underestimated in the tails of the distribution. 
+Lastly, it can be seen that, although the estimated density ratios follow the general trends of the true density ratios, there is also some variability around the true values, yielding rather wiggly lines in some simulations.
+
+Finally, we briefly compare the performance of uLSIF to $k$-nearest neighbor density ratio estimation, which is implicitly used in Kullback-Leibler divergence estimation through $k$-nearest neighbor density estimation (see Appendix [-@sec-app-B] for details).
+We set the number of neighbors to $k = 15 \approx \sqrt n$ [as suggested by @loftsgaarden_nonparametric_1965].
+@fig-AppB shows that the estimated density ratios by uLSIF are less variable than the estimated density ratios by $k$-nearest neighbor density ratio estimation.
+Moreover, the estimated density ratios by $k$-nearest neighbor density ratio estimation have similar bias in the tails of the distribution, where data is scarce. 
+We conclude that uLSIF is able to provide more accurate estimates of the density ratio if there are enough samples in the regions of interest.
 
@@
 
-TODO: Adjust section title
+To further investigate the performance of density ratio estimation, we conducted a simulation study in a multivariate setting to compare against existing global utility measures.
+The goal of these simulations is to evaluate whether density ratio estimation is able to capture improvements in the synthetic data model, and is able to reflect these improvements in the utility statistics.
+In what follows, we describe the data-generating mechanism, the synthetic data models, the utility measures used to evaluate the quality of the synthetic data models and the results of the simulations.
 
-Currently, my plan is as follows, specify a (two/four) true data generating mechanism:
-A multivariate normal distribution with a given correlation structure, 7 and 17 variables each for two different sample size (say n = 500 and n = 2000). 
-Append the multivariate normal distribution with three variables with a different distribution that depend on the other variables through some non-linear function.
-Have three synthetic data generating mechanisms, a simple multivariate normal distribution with the same means and variances but zero covariances, a model that uses a multivariate normal distribution with correlation structure taken from the real data but misses the non-linear effects (and marginal distributions of these variables), and a model that is equivalent to the true model.
-Then, evaluate the quality of the synthetic data sets using density ratio estimation and the $pMSE$, and see which of the two ranks the models correctly most of the time.
+### True data-generating mechanism
 
-# Application: Synthetic data generation for the U.S. Current Population Survey
+We generate the real data according to three data-generating mechanisms, consisting of $D \in \{5, 25, 50\}$ variables. 
+When $D = 5$, the first four variables are multivariate normally distributed. 
+These normally distributed variables have mean $\mu_d = 0$, variance $\sigma^2_{d,d} = 1$ and covariance $\sigma^2_{d,d'} = 0.5$, for $d = 1, \dots, D-1$. 
+The fifth variable is non-linearly related to the first variable, such that $X_5 \sim \mathcal{N}(X_1^2, \hat{\text{Var}}(X_1^2))$.
+When $D = 25$ or $D = 50$, the first $20$ and $45$ variables are distributed normally, again with zero mean, unit variance and covariance $\sigma^2_{d,d'} = 0.5$ for $d = 1, \dots, D-5$. 
+The last five variables have a non-linear relationship with the first five variables, defined as $X_{D-5+j} \sim \mathcal{N}(X_j^{j+1}, \hat{\text{Var}}(X_j^j))$, for $j = 1, \dots, 5$. 
+Hence, the first variable with a non-linear relationship is obtained by raising the first variable to the power $2$, the second is obtained by raising the second variable to the power $3$, and so on, with an additional variance term that scales with the variability due to raising terms to a higher power.
+For each number of variables, we generate $1000$ data sets, with three different numbers of observations ($n \in \{100, 1000, 2000\}$).
 
-ADD TEXT
+### Synthetic data generation
+
+We generate synthetic data via three mechanisms, each incrementally closer to the true data generating mechanism. 
+The first synthetic data model is a multivariate normal distribution with mean and variance estimated from the real data, but neither covariances nor non-linear relationships are taken into account. 
+The second synthetic data model is a multivariate normal distribution with means, variances and covariances estimated from the real data, but without the non-linear relationships. 
+The final, and correctly specified, synthetic data model estimates the means, variances, covariances and non-linear parameters from the data.
+Each synthetic data set has the same number of observations as the observed data.
+
+### Utility measures
+
+The goal of the simulation study is to evaluate whether the global utility measures are able to detect the improvements in the synthesis models.
+To this end, we use three different utility measures, and two different specifications of each measure. 
+First, we use uLSIF with automatic hyperparameter selection through cross-validation (the default in the `densityratio`-package). 
+We summarize the estimated utility of the synthetic data into a single number using the Pearson divergence [$PE$\; @sugiyama_lstst_2011]. 
+Additionally, we use the Kullback-Leibler divergence based on $k$-nearest neighbor density ratio estimation as implemented in the `R`-package `kldest` [@kldest], varying the number of neighbors between $k = 1$ (the default in the software) and $k = \sqrt{n}$ [as advocated by @loftsgaarden_nonparametric_1965].
+Finally, we use the $pMSE$-ratio, based on a CART model using the default settings in `synthpop` [@synthpop] and a logistic regression model without interactions, as not all combinations of number of variables and samples allow to include interactions in the model.
+
+### Results
+
+```{r}
+#| label: code-sim2
+#| eval: !expr params$runsims
+#| include: false
+
+source("code/sim2.R")
+```
+
+```{r}
+#| label: tbl-sim2-results
+#| eval: true
+#| echo: false
+#| warning: false
+#| message: false
+#| tbl-cap: "Proportion of simulations in which the true synthetic data model order is captured by the utility measure.
+The column $PE$ refers to the Pearson divergence estimated using uLSIF, $KL_1$ and $KL_{\\sqrt{n}}$ refer to the Kullback-Leibler divergence estimated using $k$-nearest neighbor density ratio estimation with respectively $k = 1$ and $k = \\sqrt{n}$, and $pMSE_{\\text{CART}}$ and $pMSE_{\\text{logit}}$ refer to the $pMSE$-ratio estimated using a CART model and a logistic regression model, respectively.
+\\vspace{0.4cm}"
+#| tbl-colwidths: [0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2]
+
+dat <- readRDS("results/sim2.rds")
+
+dat |>
+  unnest_wider(out, simplify = FALSE) |>
+  group_by(N, P) |>
+  summarize(
+    across(c(pe, spmse_cart, spmse_logit, kl1, kln), 
+           .fns = ~map_lgl(.x, \(x) x[1] > x[2] & x[2] > x[3]) |> mean())
+  ) |>
+  select(`$n$` = N, `$D$` = P,
+         `$PE$` = pe, 
+         `$KL_1$` = kl1,
+         `$KL_{\\sqrt{n}}$` = kln, 
+         `$pMSE_{\\text{CART}}$` = spmse_cart,
+         `$pMSE_{\\text{logit}}$` = spmse_logit) |>
+  knitr::kable(
+    format = "latex", 
+    align = c("l", "l", "c", "c", "c", "c", "c", "c"),
+    escape = FALSE,
+    booktabs = TRUE,
+    linesep = c("", "", "\\addlinespace"),
+    digits = 3
+  ) |>
+  kableExtra::kable_styling(full_width = FALSE) |>
+  kableExtra::column_spec(3:8, width = "15mm")
+```
+
+@tbl-sim2-results shows the proportion of simulations in which the utility measures correctly order the synthetic data sets in terms of quality of the synthesis model. 
+That is, the first synthesis model, using only the means and variances of each variable, should have the lowest utility (highest divergence or $pMSE$-ratio), followed by the second synthesis model, using the correct means, variances and covariances but not the non-linear relationships, and the third, correctly specified synthesis model should obtain the highest utility score.
+@tbl-sim2-results shows that the $KL$-divergence based on $k = \sqrt{n}$ neighbors correctly ranks the synthetic data sets most often, followed by the $KL$-divergence based on $k = 1$ neighbors and the uLSIF model.
+Unless the sample size is small, these approaches correctly rank the synthetic data sets in almost all instances.
+Both the $KL$-divergence and the density ratio based Pearson divergence clearly outperform the common $pMSE$.
+Additionally, @tbl-sim2-results shows that the performance of all utility measures improves with the sample size, and decreases when the dimensionality of the data increases.
+The only exception to the latter trend is the logistic regression based $pMSE$-ratio model, which consistently improves with the dimensionality of the data).
+
+To zoom in on the behavior of the three methods, we visualize the results of uLSIF, the $KL$-divergence based on $k = \sqrt{n}$ neighbors, and the $pMSE$-ratio based on a CART model for $n = 1000$ and $D = 50$ in @fig-sim2-results.
+For $PE$ and $KL_{\sqrt{n}}$, the improvements in the synthesis model are consistently picked up by the utility measures, whereas for $pMSE_\text{cart}$, the estimated changes in utility are much more variable across the simulations.
+Additionally, the figures show that the improvement in utility due to modelling the covariances rather than only the means and variances is much larger than the improvement due to also modelling the non-linear relationships.
+That is, the increase in utility is much more pronounced when moving from the first to the second synthetic data set than moving from the second to the third synthetic data set.
+
+```{r}
+#| label: fig-sim2-results
+#| eval: true
+#| echo: false
+#| warning: false
+#| message: false
+#| fig-dpi: 500
+#| fig-pos: t
+#| out-width: 100%
+#| fig-height: 2.5
+#| dev: ragg_png
+#| fig-cap: Utility measures for $n = 1000$ and $D = 25$ for $PE$, $KL_{\sqrt{n}}$ and $pMSE_{\text{cart}}$ for $1000$ simulations of the three synthetic data models.
+
+p1 <- dat |>
+  filter(N == 1000, P == 50) |>
+  unnest_wider(out, simplify = FALSE) |>
+  select(sim, N, P, pe) |>
+  unnest(pe) |>
+  pivot_longer(cols = starts_with("V"), names_to = "Synthetic data model", values_to = "PE") |>
+  mutate(`Synthetic data model` = factor(`Synthetic data model`, 
+                                         levels = c("V1", "V2", "V3"),
+                                         labels = c("M1", "M2", "M3"))) |>
+  ggplot(aes(x = `Synthetic data model`, y = PE, group = sim)) +
+  geom_point(alpha = 0.1) +
+  geom_line(alpha = 0.01) +
+  theme_minimal()
+
+p2 <- dat |>
+  filter(N == 1000, P == 50) |>
+  unnest_wider(out, simplify = FALSE) |>
+  select(sim, N, P, kln) |>
+  unnest(kln) |>
+  pivot_longer(cols = starts_with("V"), 
+               names_to = "Synthetic data model", 
+               values_to = "KL") |>
+  mutate(`Synthetic data model` = factor(`Synthetic data model`, 
+                                         levels = c("V1", "V2", "V3"),
+                                         labels = c("M1", "M2", "M3"))) |>
+  ggplot(aes(x = `Synthetic data model`, y = KL, group = sim)) +
+  geom_point(alpha = 0.1) +
+  geom_line(alpha = 0.01) +
+  theme_minimal()
+
+p3 <- dat |>
+  filter(N == 1000, P == 50) |>
+  unnest_wider(out, simplify = FALSE) |>
+  select(sim, N, P, spmse_cart) |>
+  unnest(spmse_cart) |>
+  pivot_longer(cols = starts_with("V"), 
+               names_to = "Synthetic data model", 
+               values_to = "pMSE") |>
+  mutate(`Synthetic data model` = factor(`Synthetic data model`, 
+                                         levels = c("V1", "V2", "V3"),
+                                         labels = c("M1", "M2", "M3"))) |>
+  ggplot(aes(x = `Synthetic data model`, y = pMSE, group = sim)) +
+  geom_point(alpha = 0.1) +
+  geom_line(alpha = 0.01) +
+  theme_minimal()
+
+p1 + p2 + p3 +
+  plot_layout(axis_titles = "collect")
+```
+
+These results have a remarkable implication for the evaluation of synthetic data generation methods. 
+When the aim is to rank candidate synthetic data sets using a global utility statistic, using a rather flexible model seems advisable.
+In @sec-univ-sims and Appendix [-@sec-app-B], we showed that the $k$-nearest neighbor density ratio estimate is substantially more variable than the uLSIF-based estimate, but when the aim is to quantify the utility of the synthetic data, this variability does not seem to be problematic.
+Rather, these simulations seem to suggest that a more flexible model is better able to detect discrepancies between the observed and synthetic data.
+Obtaining a stable and accurate estimate of the density ratio (or, by analogy, of the propensity scores in the $pMSE$) does not seem to be required when one is solely interested in globally ranking different synthetic data sets.
+However, local information on the misfit of the synthetic data is valuable in itself, which we illustrate in the next section.
+
+# Application: Synthetic data generation for the U.S. Current Population Survey {#sec-app}
+
+```{r}
+#| label: code-application
+#| eval: !expr params$runsims
+#| include: false
+
+source("code/application.R")
+```
+
+```{r}
+#| label: load-app-data
+#| eval: true
+#| echo: false
+#| warning: false
+#| message: false
+
+load("results/application.RData")
+```
+
+To further illustrate the use of density ratio estimation for synthetic data utility using the March 2000 U.S. Current Population Survey (CPS).
+The data contains the continuous variables age, household income, household property taxes and social security payments, and the categorical variables sex, race, marital status and education level measured on $n_\text{obs} = 5000$ individuals.
+We compare two different synthesis strategies that increasingly tailor toward the data at hand [using the `R`-package `synthpop`\; @synthpop]. 
+Note that all continuous variables except age are substantially skewed. 
+Moreover, the variables household property taxes and social security payments have a point-mass at zero (see the first column in @fig-application-distributions).
+To account for the skewness, we transform the continuous variables to cubic root scale ($f(x) = |x|^{1/3}\cdot\text{sign}(x)$).
+Accordingly, the first modelling strategies applies a linear regression model on the transformed continuous variables. 
+The categorical variables are modelled using (multinomial) logistic regression models. 
+The second modelling strategy extends the first strategy by applying a semi-continuous model to the transformed variables household property taxes and social security payments. 
+Here, the point mass is modelled first using a logistic regression model, after which the non-zero values are synthesized using a linear regression model. 
+All other variables are modelled according to the first strategy.
+We generate $m = 5$ synthetic data sets for both strategies, each with $n_\text{syn} = 5000$ observations, and evaluate their utility using unconstrained least-squares importance fitting.
+We emphasize that throughout the illustration, we use the same default settings as in the previous sections.
+
+## Evaluating global utility using density ratio estimation
 
@@
 #| out-width: 100%
-#| fig-height: 5
+#| fig-height: 3
 #| dev: cairo_pdf
-#| fig-cap: "Real and synthetic data distributions for the variables age, household income (income), household property taxes (tax) and social security benefits (social security) on a cubic root scale."
-
-load("../results/application.RData")
+#| fig-cap: "Real and synthetic data distributions for the variables household property taxes and social security benefits (social security) for the transformed and semi-continuous synthesis strategy.
+The panel titles display the estimated Pearson divergence for each variable between the observed and synthetic data, estimated using a density ratio model for each variable separately.
+Note that the y-axis is displayed on a square-root scale to enhance visibility."
 
@@
   Real = df,
-  Naive = bind_rows(synlist$unadj$syn), 
   Transformed = bind_rows(synlist$trans$syn),
   `Semi-continuous` = bind_rows(synlist$semi$syn),
-  `Smoothed CART` = bind_rows(synlist$smoothed$syn),
   .id = "Data"
@@
   select(Data,
-         Age = age, 
-         Income = income, 
          Tax = tax, 
          `Social security` = ss) |>
-  mutate(Data = factor(Data, levels = c("Real", "Naive", "Transformed", "Semi-continuous", "Smoothed CART")),
-         RealSyn = ifelse(Data == "Real", 1, 2) |> factor(labels = c("Real", "Synthetic")),
-         across(c(Age, Income, Tax, `Social security`), ~abs(.x)^{1/3} * sign(.x))) |>
-  tidyr::pivot_longer(cols = c(Age, Income, Tax, `Social security`), names_to = "Variable")
+  mutate(Data = factor(Data, levels = c("Real", "Transformed", "Semi-continuous")),
+         RealSyn = ifelse(Data == "Real", 1, 2) |> factor(labels = c("Real", "Synthetic"))) |> #,
+#         across(c(Tax, `Social security`), ~abs(.x)^{1/3}*sign(.x))) |>
+  tidyr::pivot_longer(cols = c(Tax, `Social security`), names_to = "Variable")
 
-purrr::map(c("Age", "Income", "Social security", "Tax"), ~
+varPEs <- map(PE, ~map_dbl(.x, mean) |> round(2) |> format(nsmall = 2))
+varPEs <- map(varPEs, ~setNames(.x, c("Age", "Household income", "Social security", "Tax")))
+
+PEs <- map(PE_allvars, \(x) map_dbl(x, ~summary(.x)$PE))
+
+labs <- list(
+  "Social security" = labeller(Data = c(
+    Real = "Real",
+    Transformed = paste0("Transformed (PE = ", varPEs$trans[["Social security"]], ")"),
+    `Semi-continuous` = paste0("Semi-continuous (PE = ", varPEs$semi[["Social security"]], ")")
+  )),
+  "Tax" = labeller(Data = c(
+    Real = "Real",
+    Transformed = paste0("Transformed (PE = ", varPEs$trans[["Tax"]], ")"),
+    `Semi-continuous` = paste0("Semi-continuous (PE = ", varPEs$semi[["Tax"]], ")")
+  ))
+)
+
+purrr::map(c("Social security", "Tax"), ~
       ggplot(comb_df |> filter(Variable == .x), aes(x = value, fill = RealSyn, after_stat(density))) +
-      geom_histogram(col = "black", bins = 20) +
+      geom_histogram(col = "black", binwidth = 2000, boundary = 0, closed = "left") +
       scale_fill_brewer(palette = "Set2") + 
-      facet_wrap(~Data, ncol = 5) + 
+      facet_wrap(~Data, ncol = 3, labeller = labs[[.x]]) + 
       theme_minimal() +
+      scale_y_continuous(transform = "sqrt") +
       ylab(.x) +
@@
             axis.title.x = element_blank(), 
-            strip.text.x = element_text(size = 8),
-            text = element_text(family = "LM Roman 10"))) |>
-  patchwork::wrap_plots(nrow = 5)
+            axis.text.x = element_text(size = 6),
+            text = element_text(family = "LM Roman 10", size = 8))) |>
+  patchwork::wrap_plots(nrow = 2)
 ```
 
+@fig-application-distributions displays the distributions of the variables social security payments (Social security) and household property taxes (Tax) for the real data and the synthetic data according to the transformed and semi-continuous modelling strategies.
+The figure shows that the semi-continuous modelling strategy fits the observed data better than the transformed modelling strategy, as it better captures the point mass and does not yield an excessive number of negative values.
+Note that there are only minor differences for the other variables, as there are no differences in synthesis methods. 
+When evaluating the global utility of the two synthetic data sets, the improvement in synthesis models is reflected in the Pearson divergence. 
+That is, the transformed modelling strategy has a higher average estimated Pearson divergence ($\bar{PE}_\text{transformed} = `r round(mean(PEs$trans), 3) |> format(nsmall = 3)`$, $\text{range} = [`r round(min(PEs$trans), 3) |> format(nsmall = 3)`, `r round(max(PEs$trans), 3) |> format(nsmall = 3)`]$) than the semi-continuous modelling strategy ($\bar{PE}_\text{semi-continuous} = `r round(mean(PEs$semi), 3) |> format(nsmall = 3)`$, $\text{range} = [`r round(min(PEs$semi), 3) |> format(nsmall = 3)`, `r round(max(PEs$semi), 3) |> format(nsmall = 3)`]$).
+Although the difference may seem small in an absolute sense, it is substantial in a relative sense.
+Additionally, the variability in estimated Pearson divergences for each synthesis strategy is much smaller than the difference between the two strategies. 
+Finally, we remark that when estimating the density ratio on the level of the individual variables, the semi-continuous modelling strategy obtains substantially lower Pearson divergences than the transformed modelling strategy [see the panel titles in Figure @fig-application-distributions].
+Hence, we conclude that the Pearson divergence serves as an adequate measure of synthetic data utility.
+
+## Density ratios as local utility estimates
+
 ```{r}
-#| label: fig-application-pearson-div
+#| label: fig-app-utility-results
 #| eval: true
@@
 #| out-width: 100%
-#| fig-height: 2.5
-#| dev: cairo_pdf
-#| fig-cap: "Pearson divergence estimates after different synthesis strategies for the separate variables and the synthetic data as a whole."
+#| fig-height: 4
+#| dev: ragg_png
+#| fig-cap: "Estimated density ratios (on log-scale) for the first synthetic data set under the two synthesis strategies plotted against the variables social security payments (Social security) and household property taxes (Tax) for the observed and synthetic data.
+Note that the density ratios are estimated using a single model for the entire data set."
 
-bind_rows(PE, PE_allvars) |>
-  select(-c(cart, smoothed_semi)) |>
-  mutate(variable = factor(1:5, labels = c("Age", "Income", "Social security", "Tax", "All"))) |>
-  pivot_longer(cols = c(unadj, trans, semi, smoothed)) |>
-  mutate(`Synthesis method` = factor(name, levels = c("unadj", "trans", "semi", "smoothed"),
-                                     labels = c("Naïve", "Transformed", "Semi-continuous", "Smoothed CART"))) |>
-  ggplot(aes(x = variable, y = value, col = `Synthesis method`, shape = `Synthesis method`, 
-             group = `Synthesis method`)) +
-  geom_point(size = 3) +
-  # geom_line(alpha = 0.1, linetype = "dashed") +
-  scale_y_log10() +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  ylab("Pearson divergence") +
-  xlab("Variable") +
-  theme(legend.position = "bottom",
-        text = element_text(family = "LM Roman 10", size = 9))
+p1 <- densityratio::plot_univariate(PE_allvars$trans[[1]], vars = c("ss", "tax"))
+p2 <- densityratio::plot_univariate(PE_allvars$semi[[1]], vars = c("ss", "tax"))
+
+clean_plots <- function(p) {
+  p$layers[[1]]$aes_params$size <- 0.3
+  p$layers[[1]]$aes_params$alpha <- 0.2
+  p + 
+    guides(
+      colour = guide_legend(override.aes = list(size = 0.5, alpha = 1)),
+    ) +
+    theme(legend.position = "bottom") +
+    scale_colour_viridis_d("Data", option = "cividis", 
+                           labels = c(numerator = "Observed", denominator = "Synthetic"))
+}
+
+clean_plots(p1[[1]]) + 
+  lims(x = c(-24000, 51000),
+       y = c(-.8, .8)) +
+  xlab("Social security") +
+  ggtitle("Transformed") +
+  clean_plots(p1[[2]]) + 
+  lims(x = c(-1000, 100000),
+       y = c(-.8, .8)) +
+  xlab("Tax") +
+  ylab("") +
+  ggtitle("") +
+  clean_plots(p2[[1]]) + 
+  lims(x = c(-24000, 51000),
+       y = c(-.8, .8)) +
+  xlab("Social security") +
+  ggtitle("Semi-continuous") +
+  clean_plots(p2[[2]]) + 
+  lims(x = c(-1000, 100000),
+       y = c(-.8, .8)) +
+  xlab("Tax") +
+  ylab("") +
+  ggtitle("") +
+  plot_layout(nrow = 2, guides = "collect") &
+  theme(title = element_text(size = 8),
+        legend.position = "bottom",
+        legend.margin = margin(0,0,0,0),
+        legend.box.margin = margin(-10, -10, 0, -10),
+        axis.text.x = element_text(size = 6),
+        text = element_text(family = "LM Roman 10", size = 8))
 ```
 
-TODO: Expand example
+Besides quantifying the global utility of synthetic data, the density ratio also provides local information about the utility of synthetic data. 
+First, the estimated density ratio can provide an intuition about which variables are not modelled adequately.
+That is, by plotting the estimated density ratio against each variable or pairs of variables, one can identify variables for which there is a clear pattern in the density ratio values.
+@fig-app-utility-results shows the estimated density ratio values (on a log-scale) for the variables social security payments and household property taxes for the first synthetic data set under both synthesis strategies.
+First, the figure shows that the spread in density ratio values is substantially larger for the transformed modelling strategy than for the semi-continuous modelling strategy. 
+Second, the estimated density ratios indicate that the transformed modelling strategy misses the point mass at zero for social security, but generated too many values slightly larger than zero. 
+Moreover, figure shows that the transformed modelling strategy yielded too little mass at the values around $10000$, as indicated by the spike in density ratio values around $10000$.
+Plotting the density ratio against household property taxes is less insightful, as the difference in marginal distributions is relatively small.
+For the semi-continuous modelling strategy, there is some spread in density ratio values, but there is no clear pattern, and the synthetic and observed data points are mostly overlapping. 
+Hence, marginally, the density ratio values do not reveal a substantial misfit of the semi-continuous strategy.
 
-High-dimensional density ratio estimation
+Additionally, we note that in the tails of the distribution, the estimated density ratios are regularized, such that the high values for tax in the observed data do not yield extremely high density ratio values.
+From a utility perspective, this might be undesirable, as misfit in the tails of the distribution might go by unnoticed. 
+However, from a privacy perspective, this regularization is beneficial, as it prevents the density ratio from becoming very large around the location of outlying observed data points, potentially revealing the location of these points.
+The regularization thus ensures that the density ratio is not overly sensitive to individual observations.
 
-Use density ratio values for weighted analysis
+Finally, we illustrate another use of the density ratio as a local utility measure, namely, using the density ratio as importance weights in downstream analysis tasks.
+In this case, we consider the synthetic data as a sample that suffers from selection bias, and we use the estimated density ratio as importance weights to correct for this bias when making inferences.
+Using importance weighting after density ratio estimation, it is possible to obtain consistent estimates of the population parameters [@kanamori_ulsif_2009; @Kanamori2012], even when the synthesis method is not consistent.
+Consider the case that we deem our semi-continuous synthetic data model sufficient from both a privacy and utility perspective.
+Perhaps using more complex synthesis models would yield higher utility, but at the cost of a higher disclosure risk.
+To further improve the utility of the semi-continuous synthetic data, we can release the estimated density ratio values, such that a third-party can use these values as importance weights in their analysis.
+Releasing these weights obviously adds to the disclosure risk as more information about the observed data is provided. 
+However, the density ratio values might be less disclosive then synthetic data generated with a more complex synthesis model.
+Although this is no formal guarantee, the regularization in the density ratio model limits the influence each observation has on the density ratio values. 
+We return to this issue in the discussion.
 
-Incorporate point on individual data utility
+```{r}
+#| label: tbl-application-regression
+#| eval: true
+#| echo: false
+#| warning: false
+#| message: false
+#| tbl-cap: "Linear regression coefficients for the observed data, synthetic data and reweighted synthetic data using the density ratio as importance weights.
+Boldface text indicates which coefficient has smaller absolute bias relative to the observed data."
 
-- High dimensional example
-- Expansion of discussion points:
-	empirical example with weighted analyses
-	individual data utility
+rowlabels <- c(
+  "Intercept",
+  "Age",
+  "Education: High school",
+  "Education: Associate or Bachelor's degree",
+  "Education: Master's degree or higher",
+  "Marital: Separated",
+  "Marital: Widowed",
+  "Marital: Single",
+  "Marital: Widowed or divorced",
+  "Race: Non-white",
+  "Sex: Female",
+  "log(Tax + 1)",
+  "log(Social security + 1)"
+)
+
+coefs |>
+  mutate(bias_syn = bias$syn_semi,
+         bias_adj = bias$syn_adj,
+         syn_semi = kableExtra::cell_spec(round(syn_semi, 3) |> format(nsmall = 3), 
+                                          "latex", 
+                                          bold = abs(bias_syn) < abs(bias_adj)),
+         syn_adj = kableExtra::cell_spec(round(syn_adj, 3) |> format(nsmall = 3),
+                                         "latex", 
+                                         bold = abs(bias_adj) < abs(bias_syn))) |>
+  select("Observed" = true,
+         "Synthetic" = syn_semi,
+         "Reweighted" = syn_adj) |>
+  tibble::rownames_to_column("Variable") |>
+  mutate(Variable = factor(Variable, levels = unique(Variable), 
+                           labels = rowlabels)) |>
+  knitr::kable(digits = 3, booktabs = TRUE, linesep="", format = "latex", escape = FALSE)
+```
+
+To demonstrate the idea, we fit a linear regression model predicting the logarithm of household income using all predictors in the data set (i.e., age, household property taxes, social security payments, sex, race, marital status and education level) on the observed data and the semi-continuous synthetic data.
+Subsequently, we estimate the model on the semi-continuous synthetic data using the estimated density ratio values as importance weights. 
+@tbl-application-regression shows the coefficients of these regression models averaged over the $m=5$ synthetic datasets, where boldface text denotes the coefficient with smallest absolute bias compared to the estimate obtained from the observed data.
+The table shows that although the synthetic regression coefficients are quite close to the observed data coefficients, the reweighed coefficients further improve the accuracy. 
+Especially for the continuous predictors, age, household property taxes and social security payments, using the density ratio weights improves the estimates.
+Averaging the absolute normalized bias over all coefficients and all synthetic data sets yields an improvement from approximately $\text{avg}(|\beta_\text{obs}-\beta_\text{syn}|/SE_\text{obs}) = `r sprintf(mean(abs(bias$syn_semi)), fmt = '%.2f')`$ to $\text{avg}(|\beta_\text{obs}-\beta_\text{syn-adj}|/SE_\text{obs}) = `r sprintf(mean(abs(bias$syn_adj)), fmt = '%.2f')`$ after reweighing.
+Hence, besides providing a measure of global utility, the estimated density ratio can be used to further improve the utility of the synthetic data.
 
@@
 
-Adapt from previous version
+High quality utility measures are essential when creating synthetic data for use in downstream tasks. 
+These measures provide an insight into the quality of the synthetic data, and can be used to guide the synthesis process, fostering higher quality synthetic data.
+In this paper, we provided a framework for evaluating synthetic data utility through density ratio estimation.
+Our simulations and illustration show that the density estimation framework can be used in a wide variety of settings, and yields more accurate results than established procedures for evaluating synthetic data utility under a variety of synthetic data models and sample sizes, all using the same default hyperparameter specifications.
+Moreover, we showed that the estimated density ratio, besides providing a global utility measure, serves as a measure of local utility, and can even be used directly to further improve specific utility of the synthetic data.
+We have made these methods openly available through the `densityratio` package in `R` [@densityratio].
 
-- Privacy remark?
-- When does it not work?
+Throughout the paper, we have contrasted the density ratio framework with existing utility measures as the $pMSE$ and the Kullback-Leibler divergence.
+Both measures can be considered special cases of the density ratio framework.
+The Kullback-Leibler divergence explicitly contains the density ratio in its definition, and Appendix [-@sec-app-B] describes how the $k$-nearest neighbor approach to $KL$-divergence estimation gives rise to a direct estimate of the density ratio.
+The $pMSE$ can also be seen as a special case of the density ratio framework, by noting that the predicted probabilities are easily transformed into density ratios [see @sugiyama_suzuki_kanamori_2012].
+In fact, @menon2016dreloss show that density ratio estimation is equivalent to class probability estimation (as performed in the $pMSE$), in the sense that minimizing a class probability estimation loss is equivalent to minimizing a Bregman divergence to the true density ratio function.
+These connections raise the question which model class, and which loss function, is most appropriate when the goal is to evaluate the utility of synthetic data.
 
-Density ratios can be used directly to train generative models, see:
-[Mohamed & Lakshminarayanan (2016)](https://arxiv.org/pdf/1610.03483) and [Uehara, Sata, Suzuki, Nakayama & Matsuo (2016)](https://arxiv.org/pdf/1610.02920.pdf)
+While we leave a formal evaluation of this question to future work, we can provide some intuitions. 
+First, utility measures ought to be robust, in the sense that they should not be too sensitive to extreme values in the observed data, as this may pose a privacy risk.
+That is, extreme cases typically bear higher disclosure risk as they do not blend into the crowd [@gehrke_crowd_2012].
+If the utility measure is too sensitive to these extreme cases, small discrepancies that serve to protect privacy may seem to reflect low quality synthetic data.
+@sugiyama_bregman_2012 showed that uLSIF is relatively robust to outliers compared to density ratio estimation using the $KL$-loss, but the robustness can further be improved by using different loss functions [such as Basu's power divergence\; @basu_power_1998].
+Second, utility measures should work well in a wide-variety of settings, so users do not need to fine-tune utility measures and focus on the problem at hand.
+To this end, non-parametric methods may be preferred over parametric methods, as they do not hinge upon an appropriate model specification.
+Although correctly specified parametric models are more efficient in a statistical sense, they are less reliable when the model is misspecified [@sugiyama_suzuki_kanamori_2012].
+Throughout the paper, we have shown that uLSIF serves as a suitable compromise between flexibility and robustness, although the KL-divergence using $k$-nearest neighbor density ratio estimation may be more sensitive to discrepancies between the observed and synthetic data.
 
-Density ratio estimation is equivalent to class probability estimation in classification problems, only differing in the loss function that is employed (assuming the same model class is used for density ratio estimation and classification).
-See [Menon & Ong, 2016](https://proceedings.mlr.press/v48/menon16.pdf)
+By viewing synthetic data utility through the lens of density ratio estimation, several appealing properties are obtained.
+First, the density ratio perspective encompasses global and local utility in a single framework. 
+The estimated density ratio gives rise to global utility measures, but also allows for the identification of variables that have not been modelled adequately. 
+Additionally, the estimated density ratio can be used directly to improve the utility of the synthetic data, by reweighing the synthetic data to better match the observed data.
+These findings are supported by other research showing that synthetic data utility can be improved by importance weighting [@ghalebikesabi_dpiw_2022].
+The proposed framework provide a flexible class of models that are easily implemented through the `densityratio` package [@densityratio], that tailor to a wide variety of settings. 
+Moreover, several extensions are available to deal with high-dimensional data, enabling the evaluation of the quality of synthetic data with many variables, such as spectral density ratio estimation [@izbicki_dre_2014] and least-squares heterodistributional subspace search [@sugiyama_lhss_2011], both implemented in the `densityratio` package.
+Finally, the framework can be used directly to create synthetic data, as the density ratio can serve the role of discriminator in generative models [@uehara2016generative].
 
-Logistic regression achieves the minimum asymptotic variance for correctly specified models (Qin, Biometrika 1998), but is not reliable for misspecified models (Kanamori, Suzuki, Sugiyama, IECE, 2010)
+Despite the appealing characteristics of density ratio estimation, several open questions remain. 
+Kernel methods, as typically employed in density ratio estimation, can be sensitive to the choice of kernel, and the use of a Gaussian kernel may not be appropriate in any situation. 
+Categorical variables, for instance, are not straightforwardly incorporated in the current set-up. 
+To deal with this issue, we created dummy variables for each category in our illustration, which seemed to provide reasonable results. 
+However, other solutions (i.e., different kernels or distance measures) may lead to improvements.
+Additionally, our results showed that the current approach to regularization is not ideal, as the density ratio estimates are shrunk towards an estimated intercept. 
+However, when the intercept deviates from one, the numerator or denominator density are up or downweighted. 
+Rather, one would like to shrink the density ratio estimates towards one, as this is more in line with the interpretation that there is no information to estimate the density ratio. 
+Finally, releasing information from the density ratio may pose an additional privacy risk, as it leaks additional information about the original data. 
+Presumably, solely releasing a global utility measure, such as the Pearson divergence, will yield only additional risk. 
+Releasing the density ratio values themselves may, however, pose unacceptable privacy risks, especially when the density ratio model is sensitive to extreme cases.
+Privatizing density ratio estimation presents an interesting avenue for future research.
+
+Despite these open questions, we have shown that already in its current form, density ratio estimation provides a flexible and informative framework for evaluating synthetic data utility.
+Due to its in-built model selection, it requires little specification on behalf of the user and works well in a wide variety of settings. 
+The framework can be readily incorporated into synthetic data workflows using the `densityratio` package.
+
+# Acknowledgements {.unnumbered}
+
+We are grateful to Joerg Drechsler for sharing the U.S. Current Population Survey data, and to Stef van Buuren, Gerko Vink, Hanne Oberman and Carlos Gonzalez Poses for many valuable discussions.
 
@@
 
-TO DO
+To perform density ratio estimation in `R`, we can use the `densityratio` package, which includes model fitting functions for multiple popular loss functions. 
+Additionally, we can complement the package with standard optimization routines in `R`. 
+Here, we show an example of both approaches.
 
-Small example of ulsif in `densityratio`
+We generate two small toy datasets, `x` and `y`, that serve as numerator and denominator data, respectively.
+Moreover, we set `x` as the centers in the Gaussian kernel, and we pre-define the kernel bandwidth to be $\sigma = 1$ and fix the regularization parameter to $\lambda = 0.5$.
+Note that we do not perform cross-validation for hyperparameter selection to enhance illustrational clarity.
+Subsequently, we fit a density ratio model using the `ulsif()` function from the `densityratio` package.
 
-Do the same with `optim`
+```{r}
+#| label: densityratio-illustration-ulsif
+#| results: false
+#| warning: false
+#| message: false
+#| echo: true
 
-Do the same for a different loss function
+library(densityratio)
+
+set.seed(123)
+
+x <- rnorm(250) |> as.matrix()
+y <- rnorm(250, 1, sqrt(2)) |> as.matrix()
+
+fit_dr <- ulsif(x, y, centers = x, scale = NULL, sigma = 1, lambda = 0.5)
+rhat_dr <- predict(fit_dr)
+```
+
+We can obtain the same solution using standard optimization techniques.
+In order to do this, we first need to define the kernel matrices for the numerator and denominator data, and define a loss function that can be used in optimization.
+We then use the `optim()` function to minimize the loss function.
+Recall that we use a linear model for the density ratio, as per @eq-dr-model, we use a Gaussian kernel as basis function, and we use the least-squares loss of @eq-squared-error-loss.
+
+```{r}
+#| label: optim-illustration-ulsif
+#| results: false
+#| warning: false
+#| message: false
+#| echo: true
+
+Dx <- distance(x, x, TRUE)
+Dy <- distance(y, x, TRUE)
+
+Kx <- kernel_gaussian(Dx, sigma = 1)
+Ky <- kernel_gaussian(Dy, sigma = 1)
+
+loss <- function(alpha, Kx, Ky) {
+  mean((Ky %*% alpha - 1) * Ky %*% alpha - (Ky %*% alpha - 1)^2 / 2) -
+    mean(Kx %*% alpha - 1) +
+    0.5/2 * sum(alpha^2) # ridge penalty
+}
+
+alpha <- runif(ncol(Kx))
+fit_optim <- optim(alpha, loss, Kx = Kx, Ky = Ky, 
+                   method = "BFGS", 
+                   control = list(maxit = 10000))
+
+rhat_optim <- Kx %*% fit_optim$par
+```
+Subsequently, we can show that both procedures yield equivalent results (see @fig-AppA).
+Moreover, the sum of squared difference between the estimated density ratio parameters equals `r format(sum((fit_dr$alpha_opt - fit_optim$par)^2), digits = 3)`, and the sum of squared difference between the predicted density ratios equals `r format(sum((c(rhat_dr) - c(rhat_optim))^2), digits = 3)`.
+
+```{r}
+#| label: fig-AppA
+#| echo: false
+#| fig-dpi: 500
+#| fig-pos: t
+#| fig-height: 3
+#| fig.cap: Density ratio estimation using `ulsif()` from the `densityratio`-package and using standard optimization in `R`.
+
+ggplot(data = NULL, aes(x = x, y = rhat_dr)) +
+  geom_point(col = "steelblue3") +
+  geom_function(aes(x = NULL, y = NULL), 
+                fun = function(x) dnorm(x, 0, 1) / dnorm(x, 1, sqrt(2))) +
+  theme_minimal() +
+  ylab(expression(hat(r)[uLSIF](x))) +
+ggplot(data = NULL, aes(x = x, y = rhat_optim)) +
+  geom_point(col = "steelblue3") +
+  geom_function(aes(x = NULL, y = NULL), 
+                fun = function(x) dnorm(x, 0, 1) / dnorm(x, 1, sqrt(2))) +
+  theme_minimal() +
+  ylab(expression(hat(r)[optim](x)))
+```
+
+If we were to perform cross-validation for hyperparameter selection, we could run the `ulsif()`-function without specifying the `sigma` and `lambda` arguments.
+The function then performs a grid search over a predefined range of hyperparameters.
+Moreover, it is possible to supply a vector of `sigma` and `lamdba` values to `ulsif()`, after which cross-validation is automatically performed.
+
+```{r}
+#| label: illustrate-ulsif-cv
+#| echo: true
+
+fit <- ulsif(x, y, centers = x, scale = NULL)
+fit
+```
+The model output shows the optimal `sigma` parameter out of a vector containing ten values, and the optimal `lambda` parameter out of a vector containing twenty values. 
+The package also includes density ratio estimation techniques, such as `kliep()`, `kmm()`, `spectral()`, `lhss()` and `naive()`, that all work in a similar fashion.
+Additionally, the package contains generic functions to `predict()` with, `plot()`, or summarize (e.g., `summary()`) the fitted models.
+Moreover, computationally intensive code is implemented in `C++` through `Rcpp` [@Rcpp] and most functions allow for parallel computation through the `openmp`-backend in `C++`. 
+The interested reader is referred to the package documentation [@densityratio] for additional information.
+
+# $k$-Nearest neighbor density ratio estimation {#sec-app-B}
+
+We note that density ratio estimation through $k$-nearest neighbor density estimation can also be considered a direct approach to density ratio estimation. 
+Using $k$-nearest neighbors, an estimate of the density can be obtained by
+$$
+\hat{\pobs}(\bx_i) = \frac{k}{n-1} \cdot \frac{\Gamma(d/2 + 1)}{\pi^{d/2} B^d_k(\bx^{(\text{obs})}_i)},
+$$
+where $d$ denotes the dimensionality of the input data, $\Gamma(d/2 + 1) / \pi^{p/2}$ is a normalizing constant, and $B^d_k(\mathbf{x}^{(\text{obs})}_i)$ denotes the Euclidean distance of each observation in $\mathbf{x}^{(\text{obs})}$ to its $k$-th nearest neighbor in $\mathbf{x}^{(\text{obs})}$ raised to the power $d$ [see, e.g., @wang_divergence_2009\;  ].
+Applying the same procedure for the denominator data (i.e., the synthetic data), we can estimate the density ratio as
+$$
+\begin{aligned}
+\hat{r}(\bx_i) &= \frac{\hat{\pobs}(\bx_i)}{\hat{\psyn}(\bx_i)} \\
+&= \frac{\frac{k}{n-1} \cdot \frac{\Gamma(d/2 + 1)}{\pi^{d/2} B^d_k(\mathbf{x}^{(\text{obs})}_i)}}{\frac{k}{m} \cdot \frac{\Gamma(d/2 + 1)}{\pi^{d/2} B^d_k(\mathbf{x}^{(\text{syn})}_i)}} \\
+&= \frac{m}{n-1} \cdot \frac{B^d_k(\mathbf{x}^{(\text{syn})}_i)}{B^d_k(\mathbf{x}^{(\text{obs})}_i)},
+\end{aligned}
+$$
+where $m$ denotes the number of observations in the synthetic data, and $B^d_k(\mathbf{x}^{(\text{syn})}_i)$ denotes the Euclidean distance of each observation in $\mathbf{x}^{(\text{obs})}$ to its $k$-th nearest neighbor in $\mathbf{x}^{(\text{syn})}$ raised to the power $d$.
+Accordingly, a direct estimate of the density ratio can be obtained as
+$$
+\hat{r}(\bx) = \text{diag}\{\tilde{B}^d_k(\mathbf{x}^{(\text{obs})})\}^{-1} \tilde{B}^d_k(\mathbf{x}^{(\text{syn})}),
+$$
+where $\tilde{B}^d_k(\mathbf{x}^{(\text{obs})})$ and $\tilde{B}^d_k(\mathbf{x}^{(\text{syn})})$ denote the vectors with distances of each observation in $\mathbf{x}^{(\text{obs})}$ to their $k$-th nearest neighbor in $\mathbf{x}^{(\text{obs})}$ and $\mathbf{x}^{(\text{syn})}$ raised to the power $d$, multiplied by the respective sample sizes.
+
+Applying this estimate of the density ratio to the simulated data from the univariate simulations, and setting $k = 15 \approx \sqrt n$, as suggested by the authors who introduced $k$-nearest neighbor density estimation [@loftsgaarden_nonparametric_1965], we can compare the performance of the $k$-nearest neighbor density ratio estimation to the density ratio estimation through uLSIF.
+
+```{r}
+#| label: fig-AppB
+#| eval: true
+#| echo: false
+#| warning: false
+#| message: false
+#| fig-dpi: 500
+#| fig-pos: t
+#| out-width: 100%
+#| fig-height: 4
+#| dev: ragg_png
+#| fig.cap: Density ratio estimation through $k$-nearest neighbor density ratio estimation ($k = 15$) and uLSIF.
+
+sims |>
+  mutate(obs_order = map(obs, order),
+         obs = map2(obs, obs_order, ~.x[.y]),
+         rknn = map2(rknn, obs_order, ~.x[.y]),
+         rhat = map2(obs, r, ~predict(.y, .x))) |>
+  unnest(c(obs, rknn, rhat)) |>
+  group_by(model, obs) |>
+  mutate(ymean = mean(rknn)) |>
+  ggplot(aes(x = obs, y = rknn, group = interaction(model, sim))) +
+  geom_line(aes(col = "knn"), alpha = 0.05) +
+  geom_line(aes(x = obs, y = rhat, group = interaction(model, sim),
+                col = "dr"), alpha = 0.05) +
+  geom_line(data = true_ratio |> unnest(c(xpreds, ypreds)),
+            aes(x = xpreds, y = ypreds, group = NULL, col = "r")) +
+  facet_wrap(~model, labeller = label_parsed) +
+  ylim(0, 6) +
+  xlim(-3, 5) +
+  theme_minimal() +
+  xlab(expression(italic(x))) +
+  ylab(expression(hat(italic(r))(x))) +
+  # scale_linetype_manual(values = c("A" = "solid", "B" = "dashed"),
+  #                       labels = c("A" = "True density ratio",
+  #                                  "B" = "Estimated density ratio"),
+  #                       name = "") +
+  scale_color_manual(values = c(knn = "palegreen1", 
+                                dr = "steelblue3", 
+                                r = "black"),
+                     labels = c("dr" = "uLSIF", 
+                                "knn" = "k-NN", 
+                                "r" = "True ratio"),
+                     name = "") +
+  theme(legend.position = "bottom") +
+  guides(colour = guide_legend(override.aes = list(alpha = 1)))
+```
+
+It can be seen that the $k$-nearest neighbor density ratio estimation performs well on average (the green lines in @fig-AppB), especially in the center of the distribution.
+However, it suffers from a similar bias in regions where the number of observations is low (as the density ratio is typically underestimated in the tails).
+Moreover, $k$-nearest neighbor density ratio estimation has much higher variance than density ratio estimation through uLSIF, as can be seen by the fact that the green lines cover a much wider band than the blue lines.

### 2024-08-23 paper/paper.tex
Update paper

@@
 % Make \paragraph and \subparagraph free-standing
+\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
+\makeatother
+
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{241,243,245}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
+\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
 
@@
 \makeatother
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
   \fi
   % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
 \usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 
@@
 \usepackage[T1]{fontenc}
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
+\usepackage{float}
+\floatplacement{table}{t}
 \makeatletter
@@
 \makeatletter
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
@@
 \makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
+
 \ifLuaTeX
@@
 \fi
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\usepackage{bookmark}
+
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
 \begin{abstract}
-TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
-TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
-TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
-TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
-TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
-TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
-TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
-TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
-TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
-TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
-TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO TODO
+Synthetic data generation is a promising technique to facilitate the use
+of sensitive data while mitigating the risk of privacy breaches.
+However, for synthetic data to be useful in downstream analysis tasks,
+it needs to be of sufficient quality. Various methods have been proposed
+to measure the utility of synthetic data, but their results are often
+incomplete or even misleading. In this paper, we propose using density
+ratio estimation to improve quality evaluation for synthetic data, and
+thereby the quality of synthesized datasets. We show how this framework
+relates to and builds on existing measures, yielding global and local
+utility measures that are informative and easy to interpret. We develop
+an estimator which requires little to no manual tuning due to automatic
+selection of a nonparametric density ratio model. Through simulations,
+we find that density ratio estimation yields more accurate estimates of
+global utility than established procedures. A real-world data
+application demonstrates how the density ratio can guide refinements of
+synthesis models and can be used to improve downstream analyses. We
+conclude that density ratio estimation is a valuable tool in synthetic
+data generation workflows and provide these methods in the accessible
+open source R-package densityratio.
 \end{abstract}
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, enhanced]}{\end{tcolorbox}}\fi
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\section{Introduction}\label{introduction}
+
+\linespread{1}
+
+\linespread{2}
 
@@
 that would otherwise be devoted to data collection
-(\protect\hyperlink{ref-ramachandran_open_2021}{Ramachandran, Bugbee,
-and Murphy 2021}). Sharing data in combination with code allows others
-to validate research findings and build upon the work
-(\protect\hyperlink{ref-obels_analysis_2020}{Obels et al. 2020};
-\protect\hyperlink{ref-crosas_automating_2015}{Crosas et al. 2015}).
-Students can benefit from open data, as it fosters education with
-realistic data (\protect\hyperlink{ref-atenas_open_2015}{Atenas,
-Havemann, and Priego 2015}), as well as the general public, through
-stimulating citizen science projects
-(\protect\hyperlink{ref-newman_future_2012}{Newman et al. 2012}).
-However, making data openly available is often (rightfully) hampered by
-official legislation, like the General Data Protection Regulation (GDPR;
-\protect\hyperlink{ref-gdpr}{European Parliament and Council of the
-European Union 2016}), and general privacy concerns. In the worst case,
-sharing data may cause harm to individuals or organizations, which may
-withhold these entities from participating in future research. These
-privacy constraints have been named among the biggest hurdles in the
-advancement of computational social science
-(\protect\hyperlink{ref-lazer_css_2009}{Lazer et al. 2009}), and among
-top reasons for companies to not share their data with researchers
-(\protect\hyperlink{ref-fpf_2017}{Future of Privacy Forum 2017}).
+(\citeproc{ref-ramachandran_open_2021}{Ramachandran, Bugbee, and Murphy
+2021}). Sharing data in combination with code allows others to validate
+research findings and build upon the work
+(\citeproc{ref-obels_analysis_2020}{Obels et al. 2020};
+\citeproc{ref-crosas_automating_2015}{Crosas et al. 2015}). Students can
+benefit from open data, as it fosters education with realistic data
+(\citeproc{ref-atenas_open_2015}{Atenas, Havemann, and Priego 2015}), as
+well as the general public, through stimulating citizen science projects
+(\citeproc{ref-newman_future_2012}{Newman et al. 2012}). However, making
+data openly available is often (rightfully) hampered by official
+legislation, like the General Data Protection Regulation (GDPR;
+\citeproc{ref-gdpr}{European Parliament and Council of the European
+Union 2016}), and general privacy concerns. In the worst case, sharing
+data may cause harm to individuals or organizations, which may withhold
+these entities from participating in future research. These privacy
+constraints have been named among the biggest hurdles in the advancement
+of computational social science (\citeproc{ref-lazer_css_2009}{Lazer et
+al. 2009}), and among top reasons for companies to not share their data
+with researchers (\citeproc{ref-fpf_2017}{Future of Privacy Forum
+2017}).
 
@@
 such as top-coding, record-swapping or adding noise (e.g.,
-\protect\hyperlink{ref-hundepool_disclosure_2012}{Hundepool et al.
-2012}; \protect\hyperlink{ref-willenborg_elements_2001}{Willenborg and
-De Waal 2001}). More recently, synthetic data has gained traction as a
-means to disseminate private data
-(\protect\hyperlink{ref-SIPP_Beta_2006}{Abowd, Stinson, and Benedetto
-2006}; \protect\hyperlink{ref-hawala_synthetic_2008}{Hawala 2008};
-\protect\hyperlink{ref-drechsler2012}{Drechsler 2012};
-\protect\hyperlink{ref-vandewiel2023}{van de Wiel et al. 2023};
-\protect\hyperlink{ref-obermeyer2019}{Obermeyer et al. 2019};
-\protect\hyperlink{ref-zettler2021}{Zettler et al. 2021}), although the
+\citeproc{ref-hundepool_disclosure_2012}{Hundepool et al. 2012};
+\citeproc{ref-willenborg_elements_2001}{Willenborg and De Waal 2001}).
+More recently, synthetic data has gained traction as a means to
+disseminate private data (\citeproc{ref-SIPP_Beta_2006}{Abowd, Stinson,
+and Benedetto 2006}; \citeproc{ref-hawala_synthetic_2008}{Hawala 2008};
+\citeproc{ref-drechsler2012}{Drechsler 2012};
+\citeproc{ref-vandewiel2023}{van de Wiel et al. 2023};
+\citeproc{ref-obermeyer2019}{Obermeyer et al. 2019};
+\citeproc{ref-zettler2021}{Zettler et al. 2021}), although the
 conceptual framework traces back to the previous century
-(\protect\hyperlink{ref-little_statistical_1993}{Little 1993};
-\protect\hyperlink{ref-rubin_statistical_1993}{Rubin 1993}). Simply put,
-the idea of synthetic data is to replace some, or all, of the observed
-values in a data set by synthetic values that are generated from some
-model (e.g., \protect\hyperlink{ref-drechsler2011synthetic}{Drechsler
-2011}). If only some values are replaced, disclosure risks can be
-reduced because the sensitive or identifying values do not correspond to
-their true values anymore. If all values are replaced, there is also no
-one-to-one mapping between the original and the synthetic data, further
-reducing the disclosure risk. However, an increase in privacy typically
-comes at the cost of a decrease in utility. As more of the data is
-altered, the quality of the released data becomes more sensitive to the
-suitability of the generative model. Regardless of the approach to
-disclosure limitation, if the technique used to generate or alter the
-data does not align with the intricacies of the problem at hand, the
-utility of the released data will be further reduced than necessary.
+(\citeproc{ref-little_statistical_1993}{Little 1993};
+\citeproc{ref-rubin_statistical_1993}{Rubin 1993}). Simply put, the idea
+of synthetic data is to replace some, or all, of the observed values in
+a data set by synthetic values that are generated from some model (e.g.,
+\citeproc{ref-drechsler2011synthetic}{Drechsler 2011}). If only some
+values are replaced, disclosure risks can be reduced because the
+sensitive or identifying values do not correspond to their true values
+anymore. If all values are replaced, there is also no one-to-one mapping
+between the original and the synthetic data, further reducing the
+disclosure risk. However, an increase in privacy typically comes at the
+cost of a decrease in utility. As more of the data is altered, the
+quality of the released data becomes more sensitive to the suitability
+of the generative model. Regardless of the approach to disclosure
+limitation, if the technique used to generate or alter the data does not
+align with the intricacies of the problem at hand, the utility of the
+released data will be further reduced than necessary.
 
@@
 In the synthetic data field, three classes of utility measures have been
-distinguished (see \protect\hyperlink{ref-drechsler2023}{Drechsler and
-Haensch 2023} for a thorough review): fit-for-purpose measures,
+distinguished (see \citeproc{ref-drechsler2023}{Drechsler and Haensch
+2023} for a thorough review): fit-for-purpose measures,
 analysis-specific utility measures and global utility measures.
@@
 coefficients of a regression model are (e.g., using the confidence
-interval overlap; \protect\hyperlink{ref-karr_utility_2006}{Karr et al.
-2006}), or whether prediction models trained on the synthetic and
-observed data perform comparably in terms of evaluation metrics.
-However, analysis-specific utility generally does not carry over: high
-specific utility for one analysis does not at all imply high utility for
-another analysis. Since data providers typically do not know which
-analyses will be performed with the synthetic data, it is impossible to
-provide analysis-specific utility measures for all potentially relevant
-analyses (see also
-\protect\hyperlink{ref-drechsler_utility_2022}{Drechsler 2022}).
+interval overlap; \citeproc{ref-karr_utility_2006}{Karr et al. 2006}),
+or whether prediction models trained on the synthetic and observed data
+perform comparably in terms of evaluation metrics. However,
+analysis-specific utility generally does not carry over: high specific
+utility for one analysis does not at all imply high utility for another
+analysis. Since data providers typically do not know which analyses will
+be performed with the synthetic data, it is impossible to provide
+analysis-specific utility measures for all potentially relevant analyses
+(see also \citeproc{ref-drechsler_utility_2022}{Drechsler 2022}).
 
@@
 measure, such as the Kullback-Leibler divergence
-(\protect\hyperlink{ref-karr_utility_2006}{Karr et al. 2006}), or by
-evaluating whether the observed and synthetic data are distinguishable
-using a classification model (a technique called \(pMSE\);
-\protect\hyperlink{ref-Woo_global_2009}{Woo et al. 2009};
-\protect\hyperlink{ref-snoke_utility_2018}{Snoke et al. 2018}). However,
-a common critique of global utility measures is that they tend to be too
-general (\protect\hyperlink{ref-drechsler_utility_2022}{Drechsler
-2022}). That is, analyses on a synthetic data set that is overall quite
-similar to the observed data (i.e., has high global utility), may still
-yield results that are far from the results obtained from the real data.
-Also, commonly used methods for estimating the \(pMSE\), as logistic
+(\citeproc{ref-karr_utility_2006}{Karr et al. 2006}), or by evaluating
+whether the observed and synthetic data are distinguishable using a
+classification model (a technique called \(pMSE\);
+\citeproc{ref-Woo_global_2009}{Woo et al. 2009};
+\citeproc{ref-snoke_utility_2018}{Snoke et al. 2018}). However, a common
+critique of global utility measures is that they tend to be too general
+(\citeproc{ref-drechsler_utility_2022}{Drechsler 2022}). That is,
+analyses on a synthetic data set that is overall quite similar to the
+observed data (i.e., has high global utility), may still yield results
+that are far from the results obtained from the real data. Also,
+commonly used methods for estimating the \(pMSE\), as logistic
 regression and classification and regression trees, tend to become less
@@
 to model misspecification
-(\protect\hyperlink{ref-drechsler_utility_2022}{Drechsler 2022}).
-Lastly, the output of global utility measures can be hard to interpret,
-and say little about the regions in which the synthetic data do not
-resemble the true data accurately enough.
+(\citeproc{ref-drechsler_utility_2022}{Drechsler 2022}). Lastly, the
+output of global utility measures can be hard to interpret, and say
+little about the regions in which the synthetic data do not resemble the
+true data accurately enough.
 
@@
 we propose to use the density ratio estimation framework
-(\protect\hyperlink{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki,
-and Kanamori 2012a}) as a way of evaluating utility. Intuitively, if two
+(\citeproc{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki, and
+Kanamori 2012a}) as a way of evaluating utility. Intuitively, if two
 data sets have similar multivariate distributions, the density ratio
@@
 non-parametric way
-(\protect\hyperlink{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki,
-and Kanamori 2012a}). These non-parametric estimation techniques come
-with automatic model specification, which mitigates the issue of model
+(\citeproc{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki, and
+Kanamori 2012a}). These non-parametric estimation techniques come with
+automatic model specification, which mitigates the issue of model
 specification. This functionality is implemented in the
 \texttt{R}-package \texttt{densityratio}
-(\protect\hyperlink{ref-densityratio}{Volker 2023}). Importantly, the
-density ratio is estimated over the entire range of the data, which
-provides measures of utility at every (possible) point in the data
-space. This point-specific quantification of utility turns out to be a
-useful side-product, as it allows to reweigh analyses on synthetic data
-when further improving the utility directly is not possible.
+(\citeproc{ref-densityratio}{Volker 2023}). Importantly, the density
+ratio is estimated over the entire range of the data, which provides
+measures of utility at every (possible) point in the data space. This
+point-specific quantification of utility turns out to be a useful
+side-product, as it allows to reweigh analyses on synthetic data when
+further improving the utility directly is not possible.
 
@@
 
-\hypertarget{background}{%
-\section{Background}\label{background}}
+\section{Background}\label{background}
 
@@
 of generalized linear models (e.g.,
-\protect\hyperlink{ref-reiter_releasing_2004}{Reiter 2004}), or as
-complex as deep learning models with thousands of parameters (e.g.,
-\protect\hyperlink{ref-xu_ctgan_2019}{Xu et al. 2019}), with many
-options in between. However, the complexity of the generation process is
-not necessarily a good indicator of the quality of the synthetic data.
-That is, relatively simple methods could still capture the most
-important aspects of the data that complex methods fail to capture (and
-vice versa). Hence, data providers typically do not know which synthesis
+\citeproc{ref-reiter_releasing_2004}{Reiter 2004}), or as complex as
+deep learning models with thousands of parameters (e.g.,
+\citeproc{ref-xu_ctgan_2019}{Xu et al. 2019}), with many options in
+between. However, the complexity of the generation process is not
+necessarily a good indicator of the quality of the synthetic data. That
+is, relatively simple methods could still capture the most important
+aspects of the data that complex methods fail to capture (and vice
+versa). Hence, data providers typically do not know which synthesis
 method will provide the highest utility a priori, and might compare
@@
 
-\hypertarget{existing-global-utility-measures}{%
 \subsection{Existing global utility
-measures}\label{existing-global-utility-measures}}
+measures}\label{existing-global-utility-measures}
 
@@
 similarity between the observed and synthetic data samples. The
-intuition is that if two data sets have similar distributions, the data
-sets can be used for the same purposes, and analyses on the two data
-sets should yield similar results. A common way to formalize
-distributional similarity is through the Kullback-Leibler (KL)
-divergence, as proposed in Karr et al.
-(\protect\hyperlink{ref-karr_utility_2006}{2006}). The KL-divergence
-measures the relative entropy from the probability distribution of the
-observed data \(\pobs(\bx)\) to the probability distribution of the
-synthetic data \(\psyn(\bx)\) (with
-\(\bx \in \mathbb{R}^{n \times d}\)), and is defined as
-\begin{equation}\protect\hypertarget{eq-kl-div}{}{
-D_{\text{KL}}(\psyn || \pobs) = 
-\int \psyn(\bx) \log \frac{\psyn(\bx)}{\pobs(\bx)} \text{d} \bx.
-}\label{eq-kl-div}\end{equation} Karr et al.
-(\protect\hyperlink{ref-karr_utility_2006}{2006}) argue that the
-KL-divergence can be constructed from density estimators
-\(\hat{p}_\text{obs}\) and \(\hat{p}_\text{syn}\), after which the
-integral can be approximated using numerical quadrature. Alternatively,
-if the observed and synthetic data are both (multivariate) normally
-distributed, the KL-divergence can be computed analytically. Both
-implementations might be unfeasible: the combination of density
-estimation with numerical quadrature is cumbersome in high dimensional
-settings, and assuming multivariate normality might be too restrictive.
-Yet, we show in later sections that the density ratio estimation
-framework gives rives to an alternative way of estimating the
-KL-divergence that overcomes these challenges.
-
-An alternative way to evaluate whether two distributions are
-statistically indistinguishable is by evaluating whether a
-classification model can tell samples from the two distributions apart
-(see \protect\hyperlink{ref-kim_classification_2021}{Kim et al. 2021},
-who formalize the connection between classification accuracy and
-two-sample testing). In the context of synthetic data, this implies that
-if a classification model can distinguish observed from synthetic
-samples, the distributional similarity is low, and so is the global
-utility. If a classifier cannot distinguish between the observed and
-synthetic data, one would conclude that the global utility is high. The
-propensity score mean-squared error (\(pMSE\)), introduced by Woo et al.
-(\protect\hyperlink{ref-Woo_global_2009}{2009}) and further developed in
-Snoke et al. (\protect\hyperlink{ref-snoke_utility_2018}{2018}),
-formalizes this intuition. Let \(I_i\) denote an indicator variable that
-equals \(1\) if observation \(i\) (\(i \in 1, \dots, N\),
-\(N = \nobs + \nsyn\)) belongs to the synthetic data, and \(0\)
-otherwise. We then train a classifier that outputs the predicted
-probability of observation \(i\) being a synthetic record \(\hat{s}_i\)
-based on the observation's scores on the variables (this can be the set
-of all variables, but also a subset). From these, we can calculate the
-utility statistic \begin{equation}\protect\hypertarget{eq-pMSE}{}{
-pMSE = \frac{1}{N} \sum_{i=1}^N \Big(\hat{s}_i - \frac{\nsyn}{N}\Big)^2,
-}\label{eq-pMSE}\end{equation} which ought to be smaller when the
-synthetic data is more like the observed data. Crucially, the \(pMSE\)
-depends on the classification model used and increases in the
-flexibility of the classification model, making it prone to overfitting
-and hard to interpret. To combat these issues, Snoke et al.
-(\protect\hyperlink{ref-snoke_utility_2018}{2018}) suggests to compare
-the \(pMSE\)-value with its expectation under the null distribution.
-Provided that the classification model is a logistic regression model
-with \(k\) parameters (including the intercept), Snoke et al.
-(\protect\hyperlink{ref-snoke_utility_2018}{2018}) show that the
-expected \(pMSE\) is given by \[
+intuition is that two data sets are realizations from the same
+underlying distribution, the data sets can be used for the same
+purposes, and analyses on the two data sets should yield similar
+results. One way to evaluate distributional similarity is by assessing
+whether a classification model can tell samples from the two
+distributions apart (see \citeproc{ref-kim_classification_2021}{Kim et
+al. 2021}, who formalize the connection between classification accuracy
+and two-sample testing). Hence, if a classification model can
+distinguish between the observed and synthetic data with high accuracy,
+the distributional similarity is low, and so is the global utility. If a
+classifier cannot distinguish between the observed and synthetic data,
+one would conclude that the global utility is high. The propensity score
+mean-squared error (\(pMSE\)), introduced by Woo et al.
+(\citeproc{ref-Woo_global_2009}{2009}) and further developed in Snoke et
+al. (\citeproc{ref-snoke_utility_2018}{2018}), formalizes this
+intuition. Let \(I_i\) denote an indicator variable that equals \(1\) if
+observation \(i\) (\(i \in 1, \dots, N\), \(N = \nobs + \nsyn\)) belongs
+to the synthetic data, and \(0\) otherwise. We then train a classifier
+that outputs the predicted probability of observation \(i\) being a
+synthetic record \(\hat{\pi}_i\) based on the observation's scores on
+the variables (this can be the set of all variables, but also a subset).
+From these, we can calculate the utility statistic
+\begin{equation}\phantomsection\label{eq-pMSE}{
+pMSE = \frac{1}{N} \sum_{i=1}^N \Big(\hat{\pi}_i - \frac{\nsyn}{N}\Big)^2,
+}\end{equation} which ought to be smaller when the synthetic data is
+more like the observed data. Crucially, the \(pMSE\) depends on the
+classification model used and increases in the flexibility of the
+classification model, making it prone to overfitting and hard to
+interpret. To combat these issues, Snoke et al.
+(\citeproc{ref-snoke_utility_2018}{2018}) suggests to compare the
+\(pMSE\)-value with its expected value under the null hypothesis that
+the observed and synthetic data are indistinguishable. Provided that the
+classification model is a logistic regression model with \(k\)
+parameters (including the intercept), Snoke et al.
+(\citeproc{ref-snoke_utility_2018}{2018}) show that the expected
+\(pMSE\) is given by \[
 \mathbb{E}[pMSE] = \Big(\frac{k-1}{N}\Big) \Big(\frac{\nobs}{N}\Big)^2
@@
 by \[
-pMSE\text{-ratio} = \frac{pMSE}{\mathbb{E}[pMSE]},
-\] with values smaller than \(10\) deemed acceptable
-(\protect\hyperlink{ref-raab2021assessing}{Gillian M. Raab, Nowok, and
-Dibben 2021}; although the authors remark that the smaller the better).
-Apart from the \(pMSE\), several other measures exist that can be
-constructed from the estimated propensity scores, such as the percentage
-of records correctly predicted
-(\protect\hyperlink{ref-raab2021assessing}{Gillian M. Raab, Nowok, and
-Dibben 2021}) or the Kolmogorov-Smirnov statistic
-(\protect\hyperlink{ref-Bowen_differentially_2021}{Bowen, Liu, and Su
-2021}), both which are strongly correlated with the \(pMSE\)
-(\protect\hyperlink{ref-raab2021assessing}{Gillian M. Raab, Nowok, and
-Dibben 2021}).
+pMSE\text{-ratio} = \frac{pMSE}{\mathbb{E}[pMSE]}.
+\] Apart from the \(pMSE\), several other measures can be constructed
+from the estimated propensity scores, such as the percentage of records
+correctly predicted (\citeproc{ref-raab2021assessing}{Gillian M. Raab,
+Nowok, and Dibben 2021}) or the Kolmogorov-Smirnov statistic
+(\citeproc{ref-Bowen_differentially_2021}{Bowen, Liu, and Su 2021}),
+both which are strongly correlated with the \(pMSE\)
+(\citeproc{ref-raab2021assessing}{Gillian M. Raab, Nowok, and Dibben
+2021}).
 
@@
 \(pMSE\) as a promising technique to evaluate the quality of synthetic
-data (e.g., \protect\hyperlink{ref-raab2017guidelines}{Gillian M. Raab,
-Nowok, and Dibben 2017};
-\protect\hyperlink{ref-raab2021assessing}{Gillian M. Raab, Nowok, and
-Dibben 2021}; \protect\hyperlink{ref-hu_advancing_2024}{Hu and Bowen
+data (e.g., \citeproc{ref-raab2017guidelines}{Gillian M. Raab, Nowok,
+and Dibben 2017}; \citeproc{ref-raab2021assessing}{Gillian M. Raab,
+Nowok, and Dibben 2021}; \citeproc{ref-hu_advancing_2024}{Hu and Bowen
 2024}). Yet, it is not free of criticism. The usefulness of the \(pMSE\)
@@
 the observed data. Drechsler
-(\protect\hyperlink{ref-drechsler_utility_2022}{2022}) illustrated that
-the utility score is highly dependent on the model used to estimate the
+(\citeproc{ref-drechsler_utility_2022}{2022}) illustrated that the
+utility score is highly dependent on the model used to estimate the
 propensity scores, and that clear improvements in synthesis models are
-not necessarily picked up in the \(pMSE\). Moreover, the \(pMSE\) is
-prone to overfitting and users may find it difficult to select an
-appropriate model for estimating the propensity scores. Hence, current
-methods to evaluate global utility are either difficult to estimate,
-especially in high-dimensional situations, or depend strongly on model
-specification. Density ratio estimation might alleviate these issues. In
-what follows, we explain how the density ratio can be estimated, and how
-it can be used to evaluate the utility of synthetic data.
-
-\hypertarget{density-ratio-estimation}{%
-\subsection{Density ratio estimation}\label{density-ratio-estimation}}
+not necessarily picked up in the \(pMSE\). Moreover, selecting an
+appropriate propensity score model may be difficult, as the challenges
+associated with model selection, such as the bias-variance trade-off,
+typically apply.
+
+Another way to formalize distributional similarity is through the
+Kullback-Leibler (KL) divergence, as proposed in Karr et al.
+(\citeproc{ref-karr_utility_2006}{2006}). The KL-divergence measures the
+relative entropy from the probability distribution of the observed data
+\(\pobs(\bx)\) to the probability distribution of the synthetic data
+\(\psyn(\bx)\) (with \(\bx \in \mathbb{R}^{d}\)), and is defined
+as\footnote{In contrast to Karr et al.
+  (\citeproc{ref-karr_utility_2006}{2006}), we define the KL-divergence
+  as the relative entropy of \(\pobs\) with respect to \(\psyn\), such
+  that it is consistent with our later formulations.}
+\begin{equation}\phantomsection\label{eq-kl-div}{
+D_{\text{KL}}(\pobs || \psyn) = 
+\int \pobs(\bx) \log \frac{\pobs(\bx)}{\psyn(\bx)} \text{d} \bx.
+}\end{equation} Karr et al. (\citeproc{ref-karr_utility_2006}{2006})
+argue that the KL-divergence can be computed by approximating the
+integral using density estimators \(\hat{p}_\text{obs}\) and
+\(\hat{p}_\text{syn}\). An implementation of this approach for
+divergence estimation has been described in Wang, Kulkarni, and Verdu
+(\citeproc{ref-wang_divergence_2009}{2009}), where the densities are
+estimated using nearest-neighbor density estimation. Another approach
+posed by Karr et al. (\citeproc{ref-karr_utility_2006}{2006}) is to
+assume multivariate normality for both \(\psyn\) and \(\pobs\), and
+calculate the KL-divergence in closed form from the sample means and
+covariance matrices. Although all these approaches are easy to
+implement, they have their limitations. Density estimation is
+challenging in high-dimensional settings, and assuming multivariate
+normality might be too restrictive. We show in later sections that the
+KL-divergence naturally fits into the density ratio estimation
+framework.
+
+\subsection{Density ratio estimation}\label{density-ratio-estimation}
 
@@
 distributions (for an overview, see
-\protect\hyperlink{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki,
-and Kanamori 2012a}). The framework has been shown to be applicable to
-prediction (\protect\hyperlink{ref-sugiyama_conditional_2010}{Sugiyama
-et al. 2010};
-\protect\hyperlink{ref-sugiyama_classification_2010}{Sugiyama 2010}),
-outlier detection (\protect\hyperlink{ref-shohei_dre_outlier_2008}{Hido
-et al. 2008}), change-point detection in time-series
-(\protect\hyperlink{ref-liu_change_2013}{Liu et al. 2013}), importance
-weighting under domain adaptation (i.e., sample selection bias;
-\protect\hyperlink{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama
-2009}), and two-sample homogeneity tests
-(\protect\hyperlink{ref-sugiyama_lstst_2011}{Sugiyama et al. 2011}). The
+\citeproc{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki, and
+Kanamori 2012a}). The framework has been shown to be applicable to
+prediction (\citeproc{ref-sugiyama_conditional_2010}{Sugiyama et al.
+2010}; \citeproc{ref-sugiyama_classification_2010}{Sugiyama 2010}),
+outlier detection (\citeproc{ref-shohei_dre_outlier_2008}{Hido et al.
+2008}), change-point detection in time-series
+(\citeproc{ref-liu_change_2013}{Liu et al. 2013}), importance weighting
+under domain adaptation (i.e., sample selection bias;
+\citeproc{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama 2009}),
+and two-sample homogeneity tests
+(\citeproc{ref-sugiyama_lstst_2011}{Sugiyama, Suzuki, et al. 2011}). The
 general idea of density ratio estimation is depicted in
 Figure~\ref{fig-dr-plot}, and boils down to comparing two distributions
-by modelling the density ratio \(r(\bx)\) between the probability
-distributions of the numerator samples, taken from the synthetic data
-distribution, \(\psyn(\bx)\), and the denominator samples, taken from
-the observed data distribution, \(\pobs(\bx)\), such that
-\begin{equation}\protect\hypertarget{eq-dr}{}{
-r(\bx) = \frac{\psyn(\bx)}{\pobs(\bx)}.
-}\label{eq-dr}\end{equation}
+by modelling the density ratio \(r(\bx)\). In our case, we define
+\(r(\bx)\) as the ratio between the probability distributions of the
+numerator samples, taken from the observed data distribution,
+\(\pobs(\bx)\), and the denominator samples, taken from the synthetic
+data distribution, \(\psyn(\bx)\), such that
+\begin{equation}\phantomsection\label{eq-dr}{
+r(\bx) = \frac{\pobs(\bx)}{\psyn(\bx)}.
+}\end{equation}
 
 This specification has the intuitive interpretation that in locations
-where the density ratio takes large values, too many synthetic
-observations will be generated in that region and at the locations where
-the density ratio is small, there will be too few synthetic
-observations, both relative to the observed data. An intuitive approach
-to estimating \(r(\bx)\) from samples of \(\pobs(\bx)\) and
-\(\psyn(\bx)\) would be to estimate the observed and synthetic data
-density separately, for example using kernel density estimation (e.g.,
-see \protect\hyperlink{ref-Scott1992}{Scott 1992} for an overview), and
-subsequently compute the ratio of these estimated densities. However,
-density estimation is one of the hardest tasks in statistical learning,
+where the density ratio takes values larger than 1, too few synthetic
+observations are generated in that region. Conversely, in locations
+where the density ratio takes values smaller than one, too many
+synthetic observations are generated in that region (see
+Figure~\ref{fig-dr-plot}). We put \(\pobs(\bx)\) in the numerator such
+that it plays the role of the actual distribution, while \(\psyn(\bx)\)
+serves as the approximating distribution (we briefly come back to this
+in the next section). A straightforward approach to estimating
+\(r(\bx)\) from samples of \(\pobs(\bx)\) and \(\psyn(\bx)\) would be to
+estimate the observed and synthetic data density separately, for example
+using kernel density estimation (e.g., see
+\citeproc{ref-Scott1992}{Scott 1992} for an overview), and subsequently
+compute the ratio of these estimated densities. However, density
+estimation is one of the hardest tasks in statistical learning,
 unavoidably leading to estimation errors for both densities, especially
-in high dimensions. When subsequently taking the ratio of the estimated
+in high dimensions
+(\citeproc{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki, and
+Kanamori 2012a}). When subsequently taking the ratio of the estimated
 densities, estimation errors tend to be magnified. Direct density ratio
@@
 Extensive simulations on a wide variety of tasks showed that this
-approach typically outperforms density ratio estimation through naive
-kernel density estimation, especially when the dimensionality of the
-data increases (e.g., \protect\hyperlink{ref-Kanamori2012}{Kanamori,
-Suzuki, and Sugiyama 2012b};
-\protect\hyperlink{ref-shohei_dre_outlier_2008}{Hido et al. 2008};
-\protect\hyperlink{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama
-2009}).
+approach typically outperforms density ratio estimation through kernel
+density estimation for each distribution separately, especially when the
+dimensionality of the data increases (e.g.,
+\citeproc{ref-Kanamori2012}{Kanamori, Suzuki, and Sugiyama 2012b};
+\citeproc{ref-shohei_dre_outlier_2008}{Hido et al. 2008};
+\citeproc{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama 2009}).
+
+\linespread{1}
 
@@
 
-{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-dr-plot-1.pdf}
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-dr-plot-1.pdf}
 
@@
 (i.e., \(\psyn(\bx) = N(0,1)\) and \(\pobs(\bx) = N(1,2)\)). The
-function \(r(\bx) = \psyn(\bx)/\pobs(\bx)\) denotes the true density
+function \(r(\bx) = \pobs(\bx)/\psyn(\bx)\) denotes the true density
 ratio, the function \(\hat{r}(\bx)\) denotes an estimate of the density
@@
 
-\end{figure}
-
-\hypertarget{estimating-the-density-ratio}{%
+\end{figure}%
+
+\linespread{2}
+
 \subsubsection{Estimating the density
-ratio}\label{estimating-the-density-ratio}}
+ratio}\label{estimating-the-density-ratio}
 
@@
 have been developed. A large class of these methods attempt to directly
-minimize the error between the true density ratio \(r(\bx)\) and its
-estimate \(\hat{r}(\bx)\). Following this approach, we define a loss
-function \(\mathcal{L}(r(\bx), \hat{r}(\bx))\) that measures the
+minimize the error between the true density ratio \(r(\bx)\) and a
+density ratio model \(\hat{r}(\bx)\). Following this approach, we define
+a loss function \(\mathcal{L}(r(\bx), \hat{r}(\bx))\) that measures the
 discrepancy between the true and estimated density ratio. To give an
 example, consider the following loss based on the squared error
-\begin{equation}\protect\hypertarget{eq-squared-error}{}{
+\begin{equation}\phantomsection\label{eq-squared-error}{
 \begin{aligned}
-\mathcal{L}_S(r(\bx), \hat{r}(\bx)) &= 
-\frac{1}{2} \int (\hat{r}(\bx) - r(\bx))^2 \pobs(\bx) \text{d}\bx \\
-&= \frac{1}{2} \int \hat{r}(\bx)^2 \pobs(\bx) \text{d}\bx - 
-\int \hat{r}(\bx) r(\bx) \pobs(\bx) \text{d}\bx + 
-\frac{1}{2} \int r(\bx)^2 \pobs(\bx) \text{d}\bx.
+\mathcal{L}^*_S(r(\bx), \hat{r}(\bx)) &= 
+\frac{1}{2} \int (\hat{r}(\bx) - r(\bx))^2 \psyn(\bx) \text{d}\bx \\
+&= \frac{1}{2} \int \hat{r}(\bx)^2 \psyn(\bx) \text{d}\bx - 
+\int \hat{r}(\bx) r(\bx) \psyn(\bx) \text{d}\bx + 
+\frac{1}{2} \int r(\bx)^2 \psyn(\bx) \text{d}\bx.
 \end{aligned}
-}\label{eq-squared-error}\end{equation} The second term can be
-rewritten, because the denominator in \(r(\bx)\) cancels with
-\(\pobs(\bx)\), while the third term is a constant with respect to the
-parameters in the density ratio model and can thus be ignored. Hence, we
-are left with the following loss function to minimize
-\begin{equation}\protect\hypertarget{eq-squared-error-loss}{}{
-\mathcal{L}'_S (r(\bx), \hat{r}(\bx)) = \frac{1}{2} \int \hat{r}(\bx)^2 \pobs(\bx) \text{d}\bx - \int \hat{r}(\bx) \psyn(\bx) \text{d}\bx,
-}\label{eq-squared-error-loss}\end{equation} which is termed least
-squares importance fitting (LSIF) by Kanamori, Hido, and Sugiyama
-(\protect\hyperlink{ref-kanamori_ulsif_2009}{2009}).
+}\end{equation} The second term can be rewritten, because the
+denominator in \(r(\bx)\) cancels with \(\psyn(\bx)\), while the third
+term is a constant with respect to the parameters in the density ratio
+model and can thus be ignored. Hence, we are left with the following
+loss function to minimize
+\begin{equation}\phantomsection\label{eq-squared-error-loss}{
+\mathcal{L}_S (r(\bx), \hat{r}(\bx)) = \frac{1}{2} \int \hat{r}(\bx)^2 \psyn(\bx) \text{d}\bx - \int \hat{r}(\bx) \pobs(\bx) \text{d}\bx,
+}\end{equation} which is termed least squares importance fitting (LSIF)
+by Kanamori, Hido, and Sugiyama
+(\citeproc{ref-kanamori_ulsif_2009}{2009}).
 
@@
 loss functions that fall under the family of Bregman divergences (see
-\protect\hyperlink{ref-sugiyama_bregman_2012}{Sugiyama, Suzuki, and
-Kanamori 2012b}; \protect\hyperlink{ref-mohamed2017learning}{Mohamed and
-Lakshminarayanan 2017}). Then, a general class of losses is encompassed
-by the expression
-\begin{equation}\protect\hypertarget{eq-bregman-loss}{}{
-\mathcal{L}_f (r(\bx), \hat{r}(\bx)) =
+\citeproc{ref-sugiyama_bregman_2012}{Sugiyama, Suzuki, and Kanamori
+2012b}; \citeproc{ref-mohamed2017learning}{Mohamed and Lakshminarayanan
+2017}). Then, a general class of losses is encompassed by the expression
+\begin{equation}\phantomsection\label{eq-bregman-loss}{
+\mathcal{L}^*_f (r(\bx), \hat{r}(\bx)) =
 \int \Big(
 f(r(\bx)) - f(\hat{r}(\bx)) - f'(\hat{r}(\bx))(r(\bx) - \hat{r}(\bx))
-\Big) \pobs(\bx) \text{d}\bx,
-}\label{eq-bregman-loss}\end{equation} where \(f\) is a differentiable
-and strictly convex function with derivative \(f'\). Then, ignoring the
-terms independent of \(\hat{r}(\bx)\) and noting that
-\(r(\bx)\pobs(\bx) = \psyn(\bx)\), we obtain the following objective
-\begin{equation}\protect\hypertarget{eq-bregman-objective}{}{
-\mathcal{L}'_f (r(\bx), \hat{r}(\bx)) = \int 
+\Big) \psyn(\bx) \text{d}\bx,
+}\end{equation} where \(f\) is a differentiable and strictly convex
+function with derivative \(f'\). Then, ignoring the terms independent of
+\(\hat{r}(\bx)\) and noting that \(r(\bx)\psyn(\bx) = \pobs(\bx)\), we
+obtain the following objective
+\begin{equation}\phantomsection\label{eq-bregman-objective}{
+\mathcal{L}_f (r(\bx), \hat{r}(\bx)) = \int 
 \Big( 
 f'(\hat{r}(\bx)) \hat{r}(\bx) - f(\hat{r}(\bx))
-\Big)  \pobs(\bx) \text{d}\bx
-- \int f'(\hat{r}(\bx)) \psyn(\bx)
+\Big)  \psyn(\bx) \text{d}\bx
+- \int f'(\hat{r}(\bx)) \pobs(\bx)
 \text{d}\bx.
-}\label{eq-bregman-objective}\end{equation} Minimizing this loss for
-different functions \(f\) yields different estimators for the density
-ratio, that focus on different regions of the density ratio.\footnote{It
-  is easy to see that using \(f(x) = \frac{1}{2}(x-1)^2\) turns the
-  Bregman divergence (Equation~\ref{eq-bregman-objective}) into the
-  squared error (Equation~\ref{eq-squared-error-loss}).} Specifically,
-some estimators place more emphasis on accurately modelling the regions
-in which the true density ratio is large, but less emphasis on
-accurately estimating the smaller density ratios, and vice versa (see
-\protect\hyperlink{ref-sugiyama_bregman_2012}{Sugiyama, Suzuki, and
-Kanamori 2012b}; \protect\hyperlink{ref-menon2016dreloss}{Menon and Ong
-2016}). Importantly, Sugiyama, Suzuki, and Kanamori
-(\protect\hyperlink{ref-sugiyama_bregman_2012}{2012b}) show that the
-Bregman divergence minimization approach to density ratio estimation is
-equivalent to estimating the \(f\)-divergence
-(\protect\hyperlink{ref-ali_silvey_divergence_1966}{Ali and Silvey
-1966}) between the synthetic and observed data distributions. Hence,
+}\end{equation} Minimizing this loss over \(\hat{r}(\bx)\) for different
+functions \(f\) yields different estimators for the density ratio, that
+focus on different regions of the density ratio.\footnote{It is easy to
+  see that using \(f(x) = \frac{1}{2}(x-1)^2\) turns the Bregman
+  divergence (Equation~\ref{eq-bregman-loss}) into the squared error
+  (Equation~\ref{eq-squared-error}).} Specifically, some estimators
+place more emphasis on accurately modelling the regions in which the
+true density ratio is large, but less emphasis on accurately estimating
+the density ratio in regions where it takes small values, and vice versa
+(see \citeproc{ref-sugiyama_bregman_2012}{Sugiyama, Suzuki, and Kanamori
+2012b}; \citeproc{ref-menon2016dreloss}{Menon and Ong 2016}).
+Importantly, Sugiyama, Suzuki, and Kanamori
+(\citeproc{ref-sugiyama_bregman_2012}{2012b}) show that the Bregman
+divergence minimization approach to density ratio estimation is
+equivalent to estimating an \(f\)-divergence
+(\citeproc{ref-ali_silvey_divergence_1966}{Ali and Silvey 1966}) of the
+observed data distribution from the synthetic data distribution. Hence,
 when estimating the density ratio, one implicitly estimates the
-divergence between the synthetic and observed data distributions.
+divergence between the observed and synthetic data distributions.
 
@@
 model, but a common choice is to use a linear model for the density
-ratio (e.g., \protect\hyperlink{ref-huang_kmm_2006}{Huang et al. 2006};
-\protect\hyperlink{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama
-2009}; \protect\hyperlink{ref-izbicki_dre_2014}{Izbicki, Lee, and
-Schafer 2014}; \protect\hyperlink{ref-gruber2024overcoming}{Gruber et
-al. 2024}). That is, we define the density ratio model as
-\begin{equation}\protect\hypertarget{eq-dr-model}{}{
+ratio (e.g., \citeproc{ref-huang_kmm_2006}{Huang et al. 2006};
+\citeproc{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama 2009};
+\citeproc{ref-izbicki_dre_2014}{Izbicki, Lee, and Schafer 2014};
+\citeproc{ref-gruber2024overcoming}{Gruber et al. 2024}). That is, we
+define the density ratio model as
+\begin{equation}\phantomsection\label{eq-dr-model}{
 \hat{r}(\bx) = \boldsymbol{\varphi}(\bx) \boldsymbol{\theta},
-}\label{eq-dr-model}\end{equation} where \(\boldsymbol{\varphi}(\bx)\)
-is a basis function vector, that transforms the data from a
-\(p\)-dimensional to a \(b\)-dimensional space, and
+}\end{equation} where \(\boldsymbol{\varphi}(\bx)\) is a basis function
+vector (optionally including an intercept term), that transforms the
+data from a \(p\)-dimensional to a \(b\)-dimensional space, and
 \(\boldsymbol{\theta}\) is a \(b\)-dimensional parameter vector.
@@
 function \(\mathcal{L}(r(\bx), \hat{r}(\bx))\) in
-Equation~\ref{eq-bregman-objective}.
+Equation~\ref{eq-bregman-objective}. Potentially, the loss function can
+be extended with a regularization term (e.g., a \(L_1\) or \(L_2\) norm
+on the parameters of the model), to decrease the flexibility of the
+density ratio. Parameter estimation can be carried out using standard
+optimization techniques (see Appendix \ref{sec-app-A} for a brief
+demonstration in \texttt{R}).
 
@@
 ranging from an identity function
-(\protect\hyperlink{ref-qin_inferences_1998}{Qin 1998}) to normalizing
-flows (\protect\hyperlink{ref-choi_featurized_2021}{Choi, Liao, and
-Ermon 2021}) or neural networks
-(\protect\hyperlink{ref-tiao2018dre}{Tiao 2018};
-\protect\hyperlink{ref-uehara2016generative}{Uehara et al. 2016}).
-However, a commonly used basis function in the density ratio literature
-is the Gaussian kernel function (e.g.,
-\protect\hyperlink{ref-huang_kmm_2006}{Huang et al. 2006};
-\protect\hyperlink{ref-sugiyama_kliep_2007}{Sugiyama et al. 2007};
-\protect\hyperlink{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama
-2009}; \protect\hyperlink{ref-liu_change_2013}{Liu et al. 2013};
-\protect\hyperlink{ref-gruber2024overcoming}{Gruber et al. 2024}), which
-conveniently allows to model the non-linear density ratio using models
-that are linear in their parameters. The Gaussian kernel function is
-defined as \[
-\boldsymbol{\varphi}(\bx) = \mathcal{K}(\bx_i, \boldsymbol{c}_j) = 
+(\citeproc{ref-qin_inferences_1998}{Qin 1998}) to normalizing flows
+(\citeproc{ref-choi_featurized_2021}{Choi, Liao, and Ermon 2021}) or
+neural networks (\citeproc{ref-tiao2018dre}{Tiao 2018};
+\citeproc{ref-uehara2016generative}{Uehara et al. 2016}). A commonly
+used basis function in the density ratio literature is the Gaussian
+kernel function (e.g., \citeproc{ref-huang_kmm_2006}{Huang et al. 2006};
+\citeproc{ref-sugiyama_kliep_2007}{Sugiyama et al. 2007};
+\citeproc{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama 2009};
+\citeproc{ref-liu_change_2013}{Liu et al. 2013};
+\citeproc{ref-gruber2024overcoming}{Gruber et al. 2024}). The Gaussian
+kernel function is defined as \[
+\boldsymbol{\varphi}(\bx) = \mathcal{K}(\bx, \boldsymbol{c}_j) = 
 \exp\left(
@@
 \right),
-\] where \(\boldsymbol{c}_j\) (\(j \in 1, \dots, J\)) denotes the
+\] where \(\boldsymbol{c}_j\) (\(j \in 1, \dots, b\)) denotes the
 centers of the Gaussian kernel functions, and \(\sigma\) controls the
@@
 information on kernel functions, see, e.g.,
-\protect\hyperlink{ref-murphy_pmlintro_2022}{Murphy 2022}). The
-appropriate width of the kernel can be determined using
-cross-validation. The centers are typically sampled from the data (in
-our case, the observed and synthetic data can both be used, but it is
-also possible to solely use the synthetic samples).
+\citeproc{ref-murphy_pmlintro_2022}{Murphy 2022}). The appropriate width
+of the kernel and the optimal value of the regularization parameter can
+be determined using cross-validation. The centers are typically sampled
+from the data (in our case, both the observed and synthetic data).
 
 After defining the model and the loss function, the density ratio can be
-easily estimated. The \texttt{densityratio} package in R
-(\protect\hyperlink{ref-densityratio}{Volker 2023}) provides an
-implementation of commonly used loss functions with a Gaussian kernel
-basis function. The package comes with an easy-to-use interface,
-automatic cross-validation of hyperparameters and builds on \texttt{C++}
-for fast computation. However, for increased flexibility with respect to
-loss functions and basis functions, one could use classic optimization
-techniques as gradient descent or (quasi-)Newton methods, for example as
-implemented in the \texttt{optim} function in the statistical software
-\texttt{R} (\protect\hyperlink{ref-R}{R Core Team 2023}). Both
-approaches are briefly illustrated in Appendix \ref{sec-app-A} (TODO).
-
-\hypertarget{evaluating-data-utility-with-the-density-ratio}{%
+estimated. The \texttt{densityratio} package in R
+(\citeproc{ref-densityratio}{Volker 2023}) provides an implementation of
+commonly used loss functions with a Gaussian kernel basis function. The
+package comes with an easy-to-use interface, automatic cross-validation
+of hyperparameters and builds on \texttt{C++} for fast computation (see
+Appendix \ref{sec-app-A} for a brief demonstration). The density ratio
+estimation framework thus equips researchers with a measure of utility
+which is accurate, fast to compute and does not require user
+specification of the model.
+
 \subsubsection{Evaluating data utility with the density
-ratio}\label{evaluating-data-utility-with-the-density-ratio}}
-
-After estimating the density ratio, the information provided by the
-density ratio can be summarized in a divergence measure, using the fact
-that estimating the density ratio is equivalent to estimating the
-\(f\)-divergence between the synthetic and observed data distributions
-(\protect\hyperlink{ref-sugiyama_bregman_2012}{Sugiyama, Suzuki, and
-Kanamori 2012b}). Accordingly, we can use the estimated density ratio
-function to estimate an \(f\)-divergence between the observed and
-synthetic data, and use this divergence measure as a utility measure for
-the synthetic data. Although this divergence statistic is difficult to
-interpret in an absolute sense, it can be used as a relative measure of
-quality of the synthetic data. That is, for different synthetic data
-sets, we can calculate the divergence to the observed data, and compare
-these values to determine which synthetic data set is most similar to
-the observed data. In a more formal way, the \(f\)-divergence gives rise
-to a test statistic that can be used to test the null hypothesis that
+ratio}\label{evaluating-data-utility-with-the-density-ratio}
+
+Having estimated the density ratio, we can evaluate both global and
+local utility of the synthetic data. Like other global utility measures,
+the density ratio can be used to construct a single discrepancy measure
+that quantifies the degree of misfit of the synthetic data. The density
+ratio can be used to directly calculate some \(f\)-divergence between
+the observed and synthetic data distributions, which serves as a measure
+of utility for the synthetic data. Although this divergence statistic is
+difficult to interpret in an absolute sense, it can be used as a
+relative measure of the quality of the synthetic data. That is, for
+different synthetic data sets, we can calculate the divergence to the
+observed data, and compare these values to determine which synthetic
+data set is most similar to the observed data. More formally, the
+\(f\)-divergence can be used in a hypothesis test to determine whether
 the synthetic data is generated from the same distribution as the
 observed data. The corresponding \(p\)-value can be calculated by
-comparing the observed test statistic to the null distribution of test
-statistics obtained using random permutations of the observed and
-synthetic data (see, for example,
-\protect\hyperlink{ref-sugiyama_lstst_2011}{Sugiyama et al. 2011};
-\protect\hyperlink{ref-Wornowizki2016}{Wornowizki and Fried 2016};
-\protect\hyperlink{ref-kanamori_divergence_2012}{Kanamori, Suzuki, and
-Sugiyama 2012a}). Lastly, the density ratio function itself can be
-plotted against individual variables or pairs of variables to discover
-whether particular variables are particularly poorly captured in the
-synthetic data. Each of these strategies can be readily executed using
-the \texttt{densityratio} package in \texttt{R}
-(\protect\hyperlink{ref-densityratio}{Volker 2023}), and will be further
-illustrated in the upcoming sections.
-
-\textbf{Ik twijfel nog een beetje over de volgende zaken, en voordat ik
-alles ga zitten uitwerken hoor ik graag eerst jullie mening:}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Ik heb nu het deel over regularisatie eruit gelaten, omdat ik vond dat
-  het wat afleid van de main ideas, maar kan deze ook gewoon toevoegen
-  aan Equation~\ref{eq-bregman-objective}. Het kan ook casual opgemerkt
-  worden wanneer het daadwerkelijk gebruikt wordt. Wat vinden jullie?
-\item
-  Dimension reduction: ik was niet zo zeker hoe ik dit er direct moest
-  bouwen, want effectief is het niet veel anders dan de huidige aanpak,
-  alleen met een andere basis function (namelijk
-  \(\phi(x) = \mathcal{K}(x_i^T U, c_j^T U)\)), waarbij U ook parameters
-  heeft die geschat worden. Anyway, ik kan aan het eind een sectie maken
-  met dat DRE nog steeds lastig kan zijn in hoge dimensies, en dat daar
-  oplossingen voor zijn (bijvoorbeeld een LFDA-stap voor ulsif, of LHSS,
-  of the spectral density ratio estimation methode (laatste twee zitten
-  allebei in het package)). Het punt is vooral dat ik niet wil dat het
-  teveel afleidt van het main punt, namelijk density ratio estimation,
-  maar qua content zou het misschien een eigen sectie verdienen, ook
-  omdat het niet direct enorm aansluit bij de twee secties hierboven.
-\end{enumerate}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{2}
-\item
-  Denken jullie dat dit een goede plek is om wat dieper in te gaan op
-  mogelijke voordelen van density ratio estimation? Ik dacht dat na
-  sectie 2.2.2 een voor de hand liggende plek zou zijn om hier wat
-  dieper op in te gaan, en dan specifiek de volgende punten te benoemen:
-  utility quantifications at every point in multivariate space, easy and
-  flexible model specification through automatic model selection with
-  cross-validation, potential for importance weighting, wellicht dan ook
-  direct het punt van dimension reduction hierbij pakken.
-\item
-  Ik wil graag nog ergens iets zeggen over de relaties tussen density
-  ratio estimation en de andere methodes (KL-divergence, pMSE), maar ik
-  vraag me een beetje af of dat niet beter in de discussie kan.
-\end{enumerate}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{4}
-\tightlist
-\item
-  Is dit echt de goede plaats om iets te zeggen over categorische
-  variabelen? Deze kunnen namelijk gewoon geïncludeerd worden, maar met
-  een Gaussian kernel wordt dat misschien een beetje awkward.
-  Desalniettemin kan het prima, en kunnen we er ook voor kiezen om in de
-  discussie te benoemen dat we een oplossing hebben gekozen, maar dat er
-  ook andere opties zijn (bijv. andere kernels).
-\end{enumerate}
-
-\hypertarget{numerical-illustrations-simulation-studies}{%
+comparing the observed divergence measure to the null distribution of
+divergence measures obtained using random permutations of the observed
+and synthetic data (see, for example,
+\citeproc{ref-sugiyama_lstst_2011}{Sugiyama, Suzuki, et al. 2011};
+\citeproc{ref-Wornowizki2016}{Wornowizki and Fried 2016};
+\citeproc{ref-kanamori_divergence_2012}{Kanamori, Suzuki, and Sugiyama
+2012a}).
+
+An appealing characteristic of the density ratio framework is that it
+also readily allows for evaluating local utility. As said, the estimated
+density ratio shows in which regions the synthetic data diverges from
+the observed data. By inspecting which observations have a high, or
+small, estimated density ratio, we can identify the regions in which the
+synthetic data is most different from the observed data. Visual
+inspection can aid this process, for example by plotting the estimated
+density ratio against variables or pairs of variables in the data. On an
+even finer scale, the density ratio can be used to identify individual
+observations that are far from what can be expected on the basis of the
+observed data. This information could identify flaws in the synthetic
+data model, or could be used directly to remove such observations from
+the synthetic data. An additional advantage of the density ratio lies in
+the fact that for every synthetic observation, the predicted density
+ratio value can be regarded as an importance weight. Releasing the
+predicted density ratio values for each synthetic observation allows
+data users to reweigh analyses on the synthetic data. We illustrate
+these features in Section~\ref{sec-app}.
+
+A final advantage of the density ratio estimation framework lies in its
+flexibility. The combination of a non-parametric kernel-based model with
+automatic cross-validation for hyperparameter selection (e.g., kernel
+and regularization parameters) allows for flexible estimation of the
+density ratio. This flexibility reduces the burden on the user in terms
+of model specification when evaluating the utility of the synthetic
+data. All functionality described above, including visualization,
+divergence estimation (and testing), extracting importance weights and
+automatic cross-validation for hyperparameter selection, is readily
+available in the \texttt{densityratio} \texttt{R}-package
+(\citeproc{ref-densityratio}{Volker 2023}). The package provides fast
+and efficient algorithms for density ratio estimation, and uses default
+implementations that work across a wide range of data distributions,
+which will be shown in the upcoming simulations.
+
 \section{Numerical illustrations: Simulation
-studies}\label{numerical-illustrations-simulation-studies}}
+studies}\label{numerical-illustrations-simulation-studies}
 
 To evaluate the performance of density ratio estimation and compare it
-to existing methods, we conduct a series of simulation studies. We
-consider a variety of scenarios, including univariate and multivariate
-models, varying correlations, number of variables, and sample sizes. To
-perform density ratio estimation, we use unconstrained Least-Squares
-Importance Fitting (uLSIF;
-\protect\hyperlink{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama
-2009}) as implemented in the \texttt{densityratio} package in \texttt{R}
-(\protect\hyperlink{ref-densityratio}{Volker 2023}), because it fast and
-has been shown to perform well in a variety of settings {[}ADD CITATIONS
-{]}. We compare the performance of uLSIF to the performance of the
-\(p\)-MSE as implemented in the \texttt{R}-package \texttt{synthpop}.
-
-\hypertarget{univariate-simulations}{%
-\subsection{Univariate simulations}\label{univariate-simulations}}
-
-For the sake of illustrational clarity, we first evaluate the
-performance of density ratio estimation in a univariate setting where
-its performance can be evaluated using visualizations. In the
-simulations, we generate data according to four data-generating
-mechanisms, (1) a Laplace distribution, (2) a log-normal distribution,
-(3) a location-scale \(t\)-distribution and (4) a normal distribution.
-Subsequently, we approximate the true data generating mechanism using a
-Gaussian model with the same mean and variance as the original data.
-This setting is similar to situations that are commonly encountered in
-the synthetic data field, in the sense that the true data generating
-mechanism is unknown and needs to be approximated using a simpler
-approximating model. The exact specifications of the data generating
-mechanisms are as follows:
+to existing methods, we conduct a series of simulation studies (all code
+is available on GitHub\footnote{\url{https://github.com/thomvolker/dr-utility}}).
+First, we illustrate the approach in a univariate simulation, after
+which we evaluate the performance of the approach in a multivariate
+setting, varying the distributions, number of variables and sample
+sizes. To perform density ratio estimation, we use unconstrained
+Least-Squares Importance Fitting (uLSIF;
+\citeproc{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama 2009})
+as implemented in the \texttt{densityratio} package in \texttt{R}
+(\citeproc{ref-densityratio}{Volker 2023}), because it fast and has been
+shown to perform well in a variety of settings (e.g.,
+\citeproc{ref-kanamori_ulsif_2009}{Kanamori, Hido, and Sugiyama 2009};
+\citeproc{ref-li_application_2010}{Li et al. 2010};
+\citeproc{ref-Kanamori2012}{Kanamori, Suzuki, and Sugiyama 2012b}). We
+compare the performance of uLSIF to the performance of state-of-the-art
+software implementations of the \(pMSE\) (\citeproc{ref-synthpop}{Nowok,
+Raab, and Dibben 2016}) and Kullback-Leibler divergence estimation
+through \(k\)-nearest neighbor density estimation
+(\citeproc{ref-kldest}{Hartung 2024}). Note that Kullback-Leibler
+divergence estimation through \(k\)-nearest neighbor density estimation
+implicitly performs direct density ratio estimation, and the procedure
+can be regarded as a special case of density ratio estimation (see
+Appendix \ref{sec-app-B}).
+
+\subsection{Univariate illustrations}\label{sec-univ-sims}
+
+We first illustrate the behavior of density ratio estimation in a
+univariate setting, which simplifies visualization of the estimated
+density ratios. Specifically, we consider four settings in which we
+generate the real data from a distribution, and generate synthetic data
+from an approximating Gaussian distribution with the same mean and
+variance as the true data generating mechanism. The real data is
+generated from (1) a Laplace distribution, (2) a log-normal
+distribution, (3) a location-scale \(t\)-distribution and (4) a normal
+distribution (see Figure~\ref{fig-densities-sim1}). Subsequently, we
+approximate the true data generating mechanism using a Gaussian model
+with the same mean and variance as the original data. This setting is
+similar to situations commonly encountered in the synthetic data field,
+in the sense that the true data generating mechanism is unknown and
+needs to be approximated using a simpler approximating model. The exact
+specifications of the data generating mechanisms are as follows:
 
@@
   Log-normal distribution with log-mean parameter
-  \(\mu_{\text{log}} = \log \{1/\sqrt{3} \}\) and log-standard deviation
-  parameter \(\sigma_\text{log} = \sqrt{\log 3}\).
+  \(\mu_{\text{log}} = \log \{1/\sqrt{3} \}\) and log-variance parameter
+  \(\sigma^2_\text{log} = \log 3\).
 \item
@@
 same population mean \(\mu = 1\) and variance \(\sigma^2 = 2\). The
-approximating model is a Gaussian distribution with mean \(\mu = 1\) and
-variance \(\sigma^2 = 2\). This distribution has the same mean and
-variance, but differs in higher-order moments, except in the fourth
-scenario. In the last simulation, we model the data generating mechanism
-correctly to obtain some understanding into how density ratio estimation
-behaves under a well-specified synthesis model. In all scenarios, we
-generate \(500\) data sets with \(\nobs = 250\) observations from the
-true data generating mechanism, and \(\nsyn = 250\) synthetic
-observations from the approximating Gaussian model. The density ratio
-model is estimated using uLSIF with a Gaussian kernel and an
-\(L_2\)-regularization parameter. Both the kernel bandwidth and the
+approximating Gaussian distribution has mean \(\mu = 1\) and variance
+\(\sigma^2 = 2\). Hence, the synthetic data has the same population mean
+and variance, but differs in higher-order moments, except in the fourth
+scenario, where it is equal to true data generating mechanism. In all
+scenarios, we generate \(1000\) data sets with \(\nobs = 250\)
+observations from the true data generating mechanism, and
+\(\nsyn = 250\) synthetic observations from the approximating Gaussian
+model. The density ratio model is estimated using uLSIF with the default
+Gaussian kernel and \(L_2\)-penalty on the parameter vector
+\(\boldsymbol{\theta}\). Both the kernel bandwidth and the
 regularization parameter are selected using cross-validation, using the
-default settings as implemented in the \texttt{densityratio} package.
+default settings in the \texttt{densityratio} package.
+
+\linespread{1}
 
@@
 
-{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-densities-sim1-1.pdf}
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-densities-sim1-1.pdf}
 
@@
 
-\end{figure}
+\end{figure}%
+
+\linespread{2}
+
+\linespread{1}
 
@@
 
-{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-sim1-results-1.pdf}
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-sim1-results-1.png}
 
@@
 approximated by a normal distribution with the same mean and variance as
-the sample from the true distribution.}
-
-\end{figure}
-
-TODO: Add interpretation of output; add significance tests and compare
-with \(pMSE\) and kolmogorov-smirnov.
-
-\hypertarget{multivariate-simulations}{%
-\subsection{Multivariate simulations}\label{multivariate-simulations}}
-
-TODO: Adjust section title
-
-Currently, my plan is as follows, specify a (two/four) true data
-generating mechanism: A multivariate normal distribution with a given
-correlation structure, 7 and 17 variables each for two different sample
-size (say n = 500 and n = 2000). Append the multivariate normal
-distribution with three variables with a different distribution that
-depend on the other variables through some non-linear function. Have
-three synthetic data generating mechanisms, a simple multivariate normal
-distribution with the same means and variances but zero covariances, a
-model that uses a multivariate normal distribution with correlation
-structure taken from the real data but misses the non-linear effects
-(and marginal distributions of these variables), and a model that is
-equivalent to the true model. Then, evaluate the quality of the
-synthetic data sets using density ratio estimation and the \(pMSE\), and
-see which of the two ranks the models correctly most of the time.
-
-\hypertarget{application-synthetic-data-generation-for-the-u.s.-current-population-survey}{%
+the sample from the true distribution. Note that the mass of the
+synthetic data distribution in the tails (smaller than \(-2\) or greater
+than \(4\)) is smaller than \(0.034\).}
+
+\end{figure}%
+
+\linespread{2}
+
+Figure~\ref{fig-sim1-results} shows that the estimated density ratios
+(the blue lines) follow the general trend of the true density ratio (the
+solid black line) for each data-generating mechanism. Especially in the
+center of each panel, approximately between \(-2\) and \(4\), the
+average density ratio (the dashed black line) closely matches the true
+density ratio. Moreover, all estimated density ratios model the peak in
+the center of each panel, where the true data generating distribution
+has more mass than the synthetic data distribution (except the bottom
+right panel, where the true and synthetic data distributions are the
+same). When moving from the middle to the sides, where the synthetic
+data distribution has more mass than the true data generating
+distributions, the estimated density ratios still follow the true
+density ratios, in the sense that they are typically smaller than \(1\).
+In the tails of the distribution, where the synthetic data distribution
+has almost no mass (\(P(x \leq -2) = P(x \geq 4) \approx 0.017\)),
+estimating the density ratio becomes difficult, leading to estimated
+density ratios that diverge from the true density ratios In these
+regions, the kernel function yields low values for most observations,
+which, combined with the \(L_2\)-penalty on the density ratio
+parameters, yields a downward bias for the estimated density ratio.
+
+When assessing the panels individually, the following observations can
+be made. In the top-left panel, showing the estimated density ratio of
+the Laplace-distributed samples over the synthetic Gaussian-distributed
+samples, the peak in the center is well approximated. Yet, the
+discontinuity in the peak of the Laplace distribution cannot be captured
+by a Gaussian kernel, which by definition models the density ratio as a
+smooth function. The same can be said about the log-normal distribution,
+where the cut-off at \(0\) is not captured exactly. Moreover, this panel
+clearly shows the shrinkage due to the regularization parameter. When
+there are no samples in one of the two distributions, the regularization
+has relatively more influence and shrinks the density ratio parameters
+towards zero. Accordingly, the estimated density ratio is shrunk towards
+the estimated intercept parameter. The same behavior can be seen in the
+bottom-left panel, showing the estimated density ratios of
+location-scale \(t\)-distribution over the Gaussian samples. In the
+centre of the distributions, the density ratio is well approximated, but
+the density ratio is consistently underestimated in the tails of the
+distribution. Lastly, it can be seen that, although the estimated
+density ratios follow the general trends of the true density ratios,
+there is also some variability around the true values, yielding rather
+wiggly lines in some simulations.
+
+Finally, we briefly compare the performance of uLSIF to \(k\)-nearest
+neighbor density ratio estimation, which is implicitly used in
+Kullback-Leibler divergence estimation through \(k\)-nearest neighbor
+density estimation (see Appendix \ref{sec-app-B} for details). We set
+the number of neighbors to \(k = 15 \approx \sqrt n\) (as suggested by
+\citeproc{ref-loftsgaarden_nonparametric_1965}{Loftsgaarden and
+Quesenberry 1965}). Figure~\ref{fig-AppB} shows that the estimated
+density ratios by uLSIF are less variable than the estimated density
+ratios by \(k\)-nearest neighbor density ratio estimation. Moreover, the
+estimated density ratios by \(k\)-nearest neighbor density ratio
+estimation have similar bias in the tails of the distribution, where
+data is scarce. We conclude that uLSIF is able to provide more accurate
+estimates of the density ratio if there are enough samples in the
+regions of interest.
+
+\subsection{Multivariate simulations}\label{multivariate-simulations}
+
+To further investigate the performance of density ratio estimation, we
+conducted a simulation study in a multivariate setting to compare
+against existing global utility measures. The goal of these simulations
+is to evaluate whether density ratio estimation is able to capture
+improvements in the synthetic data model, and is able to reflect these
+improvements in the utility statistics. In what follows, we describe the
+data-generating mechanism, the synthetic data models, the utility
+measures used to evaluate the quality of the synthetic data models and
+the results of the simulations.
+
+\subsubsection{True data-generating
+mechanism}\label{true-data-generating-mechanism}
+
+We generate the real data according to three data-generating mechanisms,
+consisting of \(D \in \{5, 25, 50\}\) variables. When \(D = 5\), the
+first four variables are multivariate normally distributed. These
+normally distributed variables have mean \(\mu_d = 0\), variance
+\(\sigma^2_{d,d} = 1\) and covariance \(\sigma^2_{d,d'} = 0.5\), for
+\(d = 1, \dots, D-1\). The fifth variable is non-linearly related to the
+first variable, such that
+\(X_5 \sim \mathcal{N}(X_1^2, \hat{\text{Var}}(X_1^2))\). When
+\(D = 25\) or \(D = 50\), the first \(20\) and \(45\) variables are
+distributed normally, again with zero mean, unit variance and covariance
+\(\sigma^2_{d,d'} = 0.5\) for \(d = 1, \dots, D-5\). The last five
+variables have a non-linear relationship with the first five variables,
+defined as
+\(X_{D-5+j} \sim \mathcal{N}(X_j^{j+1}, \hat{\text{Var}}(X_j^j))\), for
+\(j = 1, \dots, 5\). Hence, the first variable with a non-linear
+relationship is obtained by raising the first variable to the power
+\(2\), the second is obtained by raising the second variable to the
+power \(3\), and so on, with an additional variance term that scales
+with the variability due to raising terms to a higher power. For each
+number of variables, we generate \(1000\) data sets, with three
+different numbers of observations (\(n \in \{100, 1000, 2000\}\)).
+
+\subsubsection{Synthetic data
+generation}\label{synthetic-data-generation}
+
+We generate synthetic data via three mechanisms, each incrementally
+closer to the true data generating mechanism. The first synthetic data
+model is a multivariate normal distribution with mean and variance
+estimated from the real data, but neither covariances nor non-linear
+relationships are taken into account. The second synthetic data model is
+a multivariate normal distribution with means, variances and covariances
+estimated from the real data, but without the non-linear relationships.
+The final, and correctly specified, synthetic data model estimates the
+means, variances, covariances and non-linear parameters from the data.
+Each synthetic data set has the same number of observations as the
+observed data.
+
+\subsubsection{Utility measures}\label{utility-measures}
+
+The goal of the simulation study is to evaluate whether the global
+utility measures are able to detect the improvements in the synthesis
+models. To this end, we use three different utility measures, and two
+different specifications of each measure. First, we use uLSIF with
+automatic hyperparameter selection through cross-validation (the default
+in the \texttt{densityratio}-package). We summarize the estimated
+utility of the synthetic data into a single number using the Pearson
+divergence (\(PE\); \citeproc{ref-sugiyama_lstst_2011}{Sugiyama, Suzuki,
+et al. 2011}). Additionally, we use the Kullback-Leibler divergence
+based on \(k\)-nearest neighbor density ratio estimation as implemented
+in the \texttt{R}-package \texttt{kldest} (\citeproc{ref-kldest}{Hartung
+2024}), varying the number of neighbors between \(k = 1\) (the default
+in the software) and \(k = \sqrt{n}\) (as advocated by
+\citeproc{ref-loftsgaarden_nonparametric_1965}{Loftsgaarden and
+Quesenberry 1965}). Finally, we use the \(pMSE\)-ratio, based on a CART
+model using the default settings in \texttt{synthpop}
+(\citeproc{ref-synthpop}{Nowok, Raab, and Dibben 2016}) and a logistic
+regression model without interactions, as not all combinations of number
+of variables and samples allow to include interactions in the model.
+
+\subsubsection{Results}\label{results}
+
+\linespread{1}
+
+\begin{table}
+
+\caption{\label{tbl-sim2-results}Proportion of simulations in which the
+true synthetic data model order is captured by the utility measure. The
+column \(PE\) refers to the Pearson divergence estimated using uLSIF,
+\(KL_1\) and \(KL_{\sqrt{n}}\) refer to the Kullback-Leibler divergence
+estimated using \(k\)-nearest neighbor density ratio estimation with
+respectively \(k = 1\) and \(k = \sqrt{n}\), and \(pMSE_{\text{CART}}\)
+and \(pMSE_{\text{logit}}\) refer to the \(pMSE\)-ratio estimated using
+a CART model and a logistic regression model, respectively.
+\vspace{0.4cm}}
+
+\centering{
+
+\centering
+\begin{tabular}{ll>{\centering\arraybackslash}p{15mm}>{\centering\arraybackslash}p{15mm}>{\centering\arraybackslash}p{15mm}>{\centering\arraybackslash}p{15mm}>{\centering\arraybackslash}p{15mm}>{}p{15mm}}
+\toprule
+$n$ & $D$ & $PE$ & $KL_1$ & $KL_{\sqrt{n}}$ & $pMSE_{\text{CART}}$ & $pMSE_{\text{logit}}$\\
+\midrule
+100 & 5 & 0.583 & 0.844 & 0.904 & 0.369 & 0.188\\
+100 & 25 & 0.450 & 0.838 & 0.933 & 0.364 & 0.250\\
+100 & 50 & 0.363 & 0.673 & 0.786 & 0.348 & 0.352\\
+\addlinespace
+1000 & 5 & 1.000 & 1.000 & 1.000 & 0.887 & 0.173\\
+1000 & 25 & 1.000 & 1.000 & 1.000 & 0.893 & 0.242\\
+1000 & 50 & 0.995 & 1.000 & 1.000 & 0.846 & 0.316\\
+\addlinespace
+2000 & 5 & 1.000 & 1.000 & 1.000 & 0.885 & 0.173\\
+2000 & 25 & 1.000 & 1.000 & 1.000 & 0.822 & 0.247\\
+2000 & 50 & 1.000 & 1.000 & 1.000 & 0.788 & 0.324\\
+\bottomrule
+\end{tabular}
+
+}
+
+\end{table}%
+
+\linespread{2}
+
+Table~\ref{tbl-sim2-results} shows the proportion of simulations in
+which the utility measures correctly order the synthetic data sets in
+terms of quality of the synthesis model. That is, the first synthesis
+model, using only the means and variances of each variable, should have
+the lowest utility (highest divergence or \(pMSE\)-ratio), followed by
+the second synthesis model, using the correct means, variances and
+covariances but not the non-linear relationships, and the third,
+correctly specified synthesis model should obtain the highest utility
+score. Table~\ref{tbl-sim2-results} shows that the \(KL\)-divergence
+based on \(k = \sqrt{n}\) neighbors correctly ranks the synthetic data
+sets most often, followed by the \(KL\)-divergence based on \(k = 1\)
+neighbors and the uLSIF model. Unless the sample size is small, these
+approaches correctly rank the synthetic data sets in almost all
+instances. Both the \(KL\)-divergence and the density ratio based
+Pearson divergence clearly outperform the common \(pMSE\). Additionally,
+Table~\ref{tbl-sim2-results} shows that the performance of all utility
+measures improves with the sample size, and decreases when the
+dimensionality of the data increases. The only exception to the latter
+trend is the logistic regression based \(pMSE\)-ratio model, which
+consistently improves with the dimensionality of the data).
+
+To zoom in on the behavior of the three methods, we visualize the
+results of uLSIF, the \(KL\)-divergence based on \(k = \sqrt{n}\)
+neighbors, and the \(pMSE\)-ratio based on a CART model for \(n = 1000\)
+and \(D = 50\) in Figure~\ref{fig-sim2-results}. For \(PE\) and
+\(KL_{\sqrt{n}}\), the improvements in the synthesis model are
+consistently picked up by the utility measures, whereas for
+\(pMSE_\text{cart}\), the estimated changes in utility are much more
+variable across the simulations. Additionally, the figures show that the
+improvement in utility due to modelling the covariances rather than only
+the means and variances is much larger than the improvement due to also
+modelling the non-linear relationships. That is, the increase in utility
+is much more pronounced when moving from the first to the second
+synthetic data set than moving from the second to the third synthetic
+data set.
+
+\linespread{1}
+
+\begin{figure}[t]
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-sim2-results-1.png}
+
+}
+
+\caption{\label{fig-sim2-results}Utility measures for \(n = 1000\) and
+\(D = 25\) for \(PE\), \(KL_{\sqrt{n}}\) and \(pMSE_{\text{cart}}\) for
+\(1000\) simulations of the three synthetic data models.}
+
+\end{figure}%
+
+\linespread{2}
+
+These results have a remarkable implication for the evaluation of
+synthetic data generation methods. When the aim is to rank candidate
+synthetic data sets using a global utility statistic, using a rather
+flexible model seems advisable. In Section~\ref{sec-univ-sims} and
+Appendix \ref{sec-app-B}, we showed that the \(k\)-nearest neighbor
+density ratio estimate is substantially more variable than the
+uLSIF-based estimate, but when the aim is to quantify the utility of the
+synthetic data, this variability does not seem to be problematic.
+Rather, these simulations seem to suggest that a more flexible model is
+better able to detect discrepancies between the observed and synthetic
+data. Obtaining a stable and accurate estimate of the density ratio (or,
+by analogy, of the propensity scores in the \(pMSE\)) does not seem to
+be required when one is solely interested in globally ranking different
+synthetic data sets. However, local information on the misfit of the
+synthetic data is valuable in itself, which we illustrate in the next
+section.
+
 \section{Application: Synthetic data generation for the U.S. Current
-Population
-Survey}\label{application-synthetic-data-generation-for-the-u.s.-current-population-survey}}
-
-ADD TEXT
+Population Survey}\label{sec-app}
+
+\linespread{1}
+
+\linespread{2}
+
+To further illustrate the use of density ratio estimation for synthetic
+data utility using the March 2000 U.S. Current Population Survey (CPS).
+The data contains the continuous variables age, household income,
+household property taxes and social security payments, and the
+categorical variables sex, race, marital status and education level
+measured on \(n_\text{obs} = 5000\) individuals. We compare two
+different synthesis strategies that increasingly tailor toward the data
+at hand (using the \texttt{R}-package \texttt{synthpop};
+\citeproc{ref-synthpop}{Nowok, Raab, and Dibben 2016}). Note that all
+continuous variables except age are substantially skewed. Moreover, the
+variables household property taxes and social security payments have a
+point-mass at zero (see the first column in
+Figure~\ref{fig-application-distributions}). To account for the
+skewness, we transform the continuous variables to cubic root scale
+(\(f(x) = |x|^{1/3}\cdot\text{sign}(x)\)). Accordingly, the first
+modelling strategies applies a linear regression model on the
+transformed continuous variables. The categorical variables are modelled
+using (multinomial) logistic regression models. The second modelling
+strategy extends the first strategy by applying a semi-continuous model
+to the transformed variables household property taxes and social
+security payments. Here, the point mass is modelled first using a
+logistic regression model, after which the non-zero values are
+synthesized using a linear regression model. All other variables are
+modelled according to the first strategy. We generate \(m = 5\)
+synthetic data sets for both strategies, each with
+\(n_\text{syn} = 5000\) observations, and evaluate their utility using
+unconstrained least-squares importance fitting. We emphasize that
+throughout the illustration, we use the same default settings as in the
+previous sections.
+
+\subsection{Evaluating global utility using density ratio
+estimation}\label{evaluating-global-utility-using-density-ratio-estimation}
+
+\linespread{1}
 
@@
 
-{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-application-distributions-1.pdf}
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-application-distributions-1.pdf}
 
@@
 \caption{\label{fig-application-distributions}Real and synthetic data
-distributions for the variables age, household income (income),
-household property taxes (tax) and social security benefits (social
-security) on a cubic root scale.}
-
-\end{figure}
+distributions for the variables household property taxes and social
+security benefits (social security) for the transformed and
+semi-continuous synthesis strategy. The panel titles display the
+estimated Pearson divergence for each variable between the observed and
+synthetic data, estimated using a density ratio model for each variable
+separately. Note that the y-axis is displayed on a square-root scale to
+enhance visibility.}
+
+\end{figure}%
+
+\linespread{2}
+
+Figure~\ref{fig-application-distributions} displays the distributions of
+the variables social security payments (Social security) and household
+property taxes (Tax) for the real data and the synthetic data according
+to the transformed and semi-continuous modelling strategies. The figure
+shows that the semi-continuous modelling strategy fits the observed data
+better than the transformed modelling strategy, as it better captures
+the point mass and does not yield an excessive number of negative
+values. Note that there are only minor differences for the other
+variables, as there are no differences in synthesis methods. When
+evaluating the global utility of the two synthetic data sets, the
+improvement in synthesis models is reflected in the Pearson divergence.
+That is, the transformed modelling strategy has a higher average
+estimated Pearson divergence (\(\bar{PE}_\text{transformed} = 0.090\),
+\(\text{range} = [0.084, 0.100]\)) than the semi-continuous modelling
+strategy (\(\bar{PE}_\text{semi-continuous} = 0.050\),
+\(\text{range} = [0.047, 0.053]\)). Although the difference may seem
+small in an absolute sense, it is substantial in a relative sense.
+Additionally, the variability in estimated Pearson divergences for each
+synthesis strategy is much smaller than the difference between the two
+strategies. Finally, we remark that when estimating the density ratio on
+the level of the individual variables, the semi-continuous modelling
+strategy obtains substantially lower Pearson divergences than the
+transformed modelling strategy see the panel titles in
+Figure~\ref{fig-application-distributions}. Hence, we conclude that the
+Pearson divergence serves as an adequate measure of synthetic data
+utility.
+
+\subsection{Density ratios as local utility
+estimates}\label{density-ratios-as-local-utility-estimates}
+
+\linespread{1}
 
@@
 
-{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-application-pearson-div-1.pdf}
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-app-utility-results-1.png}
 
@@
 
-\caption{\label{fig-application-pearson-div}Pearson divergence estimates
-after different synthesis strategies for the separate variables and the
-synthetic data as a whole.}
-
-\end{figure}
-
-TODO: Expand example
-
-High-dimensional density ratio estimation
-
-Use density ratio values for weighted analysis
-
-Incorporate point on individual data utility
-
-\begin{itemize}
-\tightlist
-\item
-  High dimensional example
-\item
-  Expansion of discussion points: empirical example with weighted
-  analyses individual data utility
-\end{itemize}
-
-\hypertarget{discussion-and-conclusion}{%
-\section{Discussion and conclusion}\label{discussion-and-conclusion}}
-
-Adapt from previous version
-
-\begin{itemize}
-\tightlist
-\item
-  Privacy remark?
-\item
-  When does it not work?
-\end{itemize}
-
-Density ratios can be used directly to train generative models, see:
-\href{https://arxiv.org/pdf/1610.03483}{Mohamed \& Lakshminarayanan
-(2016)} and \href{https://arxiv.org/pdf/1610.02920.pdf}{Uehara, Sata,
-Suzuki, Nakayama \& Matsuo (2016)}
-
-Density ratio estimation is equivalent to class probability estimation
-in classification problems, only differing in the loss function that is
-employed (assuming the same model class is used for density ratio
-estimation and classification). See
-\href{https://proceedings.mlr.press/v48/menon16.pdf}{Menon \& Ong, 2016}
-
-Logistic regression achieves the minimum asymptotic variance for
-correctly specified models (Qin, Biometrika 1998), but is not reliable
-for misspecified models (Kanamori, Suzuki, Sugiyama, IECE, 2010)
-
-\hypertarget{references}{%
-\section*{References}\label{references}}
+\caption{\label{fig-app-utility-results}Estimated density ratios (on
+log-scale) for the first synthetic data set under the two synthesis
+strategies plotted against the variables social security payments
+(Social security) and household property taxes (Tax) for the observed
+and synthetic data. Note that the density ratios are estimated using a
+single model for the entire data set.}
+
+\end{figure}%
+
+\linespread{2}
+
+Besides quantifying the global utility of synthetic data, the density
+ratio also provides local information about the utility of synthetic
+data. First, the estimated density ratio can provide an intuition about
+which variables are not modelled adequately. That is, by plotting the
+estimated density ratio against each variable or pairs of variables, one
+can identify variables for which there is a clear pattern in the density
+ratio values. Figure~\ref{fig-app-utility-results} shows the estimated
+density ratio values (on a log-scale) for the variables social security
+payments and household property taxes for the first synthetic data set
+under both synthesis strategies. First, the figure shows that the spread
+in density ratio values is substantially larger for the transformed
+modelling strategy than for the semi-continuous modelling strategy.
+Second, the estimated density ratios indicate that the transformed
+modelling strategy misses the point mass at zero for social security,
+but generated too many values slightly larger than zero. Moreover,
+figure shows that the transformed modelling strategy yielded too little
+mass at the values around \(10000\), as indicated by the spike in
+density ratio values around \(10000\). Plotting the density ratio
+against household property taxes is less insightful, as the difference
+in marginal distributions is relatively small. For the semi-continuous
+modelling strategy, there is some spread in density ratio values, but
+there is no clear pattern, and the synthetic and observed data points
+are mostly overlapping. Hence, marginally, the density ratio values do
+not reveal a substantial misfit of the semi-continuous strategy.
+
+Additionally, we note that in the tails of the distribution, the
+estimated density ratios are regularized, such that the high values for
+tax in the observed data do not yield extremely high density ratio
+values. From a utility perspective, this might be undesirable, as misfit
+in the tails of the distribution might go by unnoticed. However, from a
+privacy perspective, this regularization is beneficial, as it prevents
+the density ratio from becoming very large around the location of
+outlying observed data points, potentially revealing the location of
+these points. The regularization thus ensures that the density ratio is
+not overly sensitive to individual observations.
+
+Finally, we illustrate another use of the density ratio as a local
+utility measure, namely, using the density ratio as importance weights
+in downstream analysis tasks. In this case, we consider the synthetic
+data as a sample that suffers from selection bias, and we use the
+estimated density ratio as importance weights to correct for this bias
+when making inferences. Using importance weighting after density ratio
+estimation, it is possible to obtain consistent estimates of the
+population parameters (\citeproc{ref-kanamori_ulsif_2009}{Kanamori,
+Hido, and Sugiyama 2009}; \citeproc{ref-Kanamori2012}{Kanamori, Suzuki,
+and Sugiyama 2012b}), even when the synthesis method is not consistent.
+Consider the case that we deem our semi-continuous synthetic data model
+sufficient from both a privacy and utility perspective. Perhaps using
+more complex synthesis models would yield higher utility, but at the
+cost of a higher disclosure risk. To further improve the utility of the
+semi-continuous synthetic data, we can release the estimated density
+ratio values, such that a third-party can use these values as importance
+weights in their analysis. Releasing these weights obviously adds to the
+disclosure risk as more information about the observed data is provided.
+However, the density ratio values might be less disclosive then
+synthetic data generated with a more complex synthesis model. Although
+this is no formal guarantee, the regularization in the density ratio
+model limits the influence each observation has on the density ratio
+values. We return to this issue in the discussion.
+
+\linespread{1}
+
+\begin{table}
+
+\caption{\label{tbl-application-regression}TODO}
+
+\centering{
+
+\begin{tabular}{lrll}
+\toprule
+Variable & Observed & Synthetic & Reweighted\\
+\midrule
+Intercept & 10.297 & 10.192 & \textbf{10.195}\\
+Age & -0.003 & -0.005 & \textbf{-0.004}\\
+Education: High school & 0.283 & 0.415 & \textbf{ 0.385}\\
+Education: Associate or Bachelor's degree & 0.593 & 0.782 & \textbf{ 0.731}\\
+Education: Master's degree or higher & 1.016 & 1.135 & \textbf{ 1.080}\\
+Marital: Separated & -0.500 & -0.434 & \textbf{-0.437}\\
+Marital: Widowed & -0.467 & \textbf{-0.471} & -0.450\\
+Marital: Single & -0.491 & -0.592 & \textbf{-0.565}\\
+Marital: Widowed or divorced & -0.549 & \textbf{-0.518} & -0.508\\
+Race: Non-white & -0.090 & \textbf{-0.109} & -0.111\\
+Sex: Female & -0.117 & -0.134 & \textbf{-0.126}\\
+log(Tax + 1) & 0.067 & 0.071 & \textbf{ 0.066}\\
+log(Social security + 1) & -0.026 & -0.009 & \textbf{-0.017}\\
+\bottomrule
+\end{tabular}
+
+}
+
+\end{table}%
+
+\linespread{2}
+
+To demonstrate the idea, we fit a linear regression model predicting the
+logarithm of household income using all predictors in the data set
+(i.e., age, household property taxes, social security payments, sex,
+race, marital status and education level) on the observed data and the
+semi-continuous synthetic data. Subsequently, we estimate the model on
+the semi-continuous synthetic data using the estimated density ratio
+values as importance weights. Table~\ref{tbl-application-regression}
+shows the coefficients of these regression models averaged over the
+\(m=5\) synthetic datasets, where boldface text denotes the coefficient
+with smallest absolute bias compared to the estimate obtained from the
+observed data. The table shows that although the synthetic regression
+coefficients are quite close to the observed data coefficients, the
+reweighed coefficients further improve the accuracy. Especially for the
+continuous predictors, age, household property taxes and social security
+payments, using the density ratio weights improves the estimates.
+Averaging the absolute normalized bias over all coefficients and all
+synthetic data sets yields an improvement from approximately
+\(\text{avg}(|\beta_\text{obs}-\beta_\text{syn}|/SE_\text{obs}) = 1.98\)
+to
+\(\text{avg}(|\beta_\text{obs}-\beta_\text{syn-adj}|/SE_\text{obs}) = 1.41\)
+after reweighing. Hence, besides providing a measure of global utility,
+the estimated density ratio can be used to further improve the utility
+of the synthetic data.
+
+\section{Discussion and conclusion}\label{discussion-and-conclusion}
+
+High quality utility measures are essential when creating synthetic data
+for use in downstream tasks. These measures provide an insight into the
+quality of the synthetic data, and can be used to guide the synthesis
+process, fostering higher quality synthetic data. In this paper, we
+provided a framework for evaluating synthetic data utility through
+density ratio estimation. Our simulations and illustration show that the
+density estimation framework can be used in a wide variety of settings,
+and yields more accurate results than established procedures for
+evaluating synthetic data utility under a variety of synthetic data
+models and sample sizes, all using the same default hyperparameter
+specifications. Moreover, we showed that the estimated density ratio,
+besides providing a global utility measure, serves as a measure of local
+utility, and can even be used directly to further improve specific
+utility of the synthetic data. We have made these methods openly
+available through the \texttt{densityratio} package in \texttt{R}
+(\citeproc{ref-densityratio}{Volker 2023}).
+
+Throughout the paper, we have contrasted the density ratio framework
+with existing utility measures as the \(pMSE\) and the Kullback-Leibler
+divergence. Both measures can be considered special cases of the density
+ratio framework. The Kullback-Leibler divergence explicitly contains the
+density ratio in its definition, and Appendix \ref{sec-app-B} describes
+how the \(k\)-nearest neighbor approach to \(KL\)-divergence estimation
+gives rise to a direct estimate of the density ratio. The \(pMSE\) can
+also be seen as a special case of the density ratio framework, by noting
+that the predicted probabilities are easily transformed into density
+ratios (see \citeproc{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama,
+Suzuki, and Kanamori 2012a}). In fact, Menon and Ong
+(\citeproc{ref-menon2016dreloss}{2016}) show that density ratio
+estimation is equivalent to class probability estimation (as performed
+in the \(pMSE\)), in the sense that minimizing a class probability
+estimation loss is equivalent to minimizing a Bregman divergence to the
+true density ratio function. These connections raise the question which
+model class, and which loss function, is most appropriate when the goal
+is to evaluate the utility of synthetic data.
+
+While we leave a formal evaluation of this question to future work, we
+can provide some intuitions. First, utility measures ought to be robust,
+in the sense that they should not be too sensitive to extreme values in
+the observed data, as this may pose a privacy risk. That is, extreme
+cases typically bear higher disclosure risk as they do not blend into
+the crowd (\citeproc{ref-gehrke_crowd_2012}{Gehrke et al. 2012}). If the
+utility measure is too sensitive to these extreme cases, small
+discrepancies that serve to protect privacy may seem to reflect low
+quality synthetic data. Sugiyama, Suzuki, and Kanamori
+(\citeproc{ref-sugiyama_bregman_2012}{2012b}) showed that uLSIF is
+relatively robust to outliers compared to density ratio estimation using
+the \(KL\)-loss, but the robustness can further be improved by using
+different loss functions (such as Basu's power divergence;
+\citeproc{ref-basu_power_1998}{Basu et al. 1998}). Second, utility
+measures should work well in a wide-variety of settings, so users do not
+need to fine-tune utility measures and focus on the problem at hand. To
+this end, non-parametric methods may be preferred over parametric
+methods, as they do not hinge upon an appropriate model specification.
+Although correctly specified parametric models are more efficient in a
+statistical sense, they are less reliable when the model is misspecified
+(\citeproc{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki, and
+Kanamori 2012a}). Throughout the paper, we have shown that uLSIF serves
+as a suitable compromise between flexibility and robustness, although
+the KL-divergence using \(k\)-nearest neighbor density ratio estimation
+may be more sensitive to discrepancies between the observed and
+synthetic data.
+
+By viewing synthetic data utility through the lens of density ratio
+estimation, several appealing properties are obtained. First, the
+density ratio perspective encompasses global and local utility in a
+single framework. The estimated density ratio gives rise to global
+utility measures, but also allows for the identification of variables
+that have not been modelled adequately. Additionally, the estimated
+density ratio can be used directly to improve the utility of the
+synthetic data, by reweighing the synthetic data to better match the
+observed data. These findings are supported by other research showing
+that synthetic data utility can be improved by importance weighting
+(\citeproc{ref-ghalebikesabi_dpiw_2022}{Ghalebikesabi et al. 2022}). The
+proposed framework provide a flexible class of models that are easily
+implemented through the \texttt{densityratio} package
+(\citeproc{ref-densityratio}{Volker 2023}), that tailor to a wide
+variety of settings. Moreover, several extensions are available to deal
+with high-dimensional data, enabling the evaluation of the quality of
+synthetic data with many variables, such as spectral density ratio
+estimation (\citeproc{ref-izbicki_dre_2014}{Izbicki, Lee, and Schafer
+2014}) and least-squares heterodistributional subspace search
+(\citeproc{ref-sugiyama_lhss_2011}{Sugiyama, Yamada, et al. 2011}), both
+implemented in the \texttt{densityratio} package. Finally, the framework
+can be used directly to create synthetic data, as the density ratio can
+serve the role of discriminator in generative models
+(\citeproc{ref-uehara2016generative}{Uehara et al. 2016}).
+
+Despite the appealing characteristics of density ratio estimation,
+several open questions remain. Kernel methods, as typically employed in
+density ratio estimation, can be sensitive to the choice of kernel, and
+the use of a Gaussian kernel may not be appropriate in any situation.
+Categorical variables, for instance, are not straightforwardly
+incorporated in the current set-up. To deal with this issue, we created
+dummy variables for each category in our illustration, which seemed to
+provide reasonable results. However, other solutions (i.e., different
+kernels or distance measures) may lead to improvements. Additionally,
+our results showed that the current approach to regularization is not
+ideal, as the density ratio estimates are shrunk towards an estimated
+intercept. However, when the intercept deviates from one, the numerator
+or denominator density are up or downweighted. Rather, one would like to
+shrink the density ratio estimates towards one, as this is more in line
+with the interpretation that there is no information to estimate the
+density ratio. Finally, releasing information from the density ratio may
+pose an additional privacy risk, as it leaks additional information
+about the original data. Presumably, solely releasing a global utility
+measure, such as the Pearson divergence, will yield only additional
+risk. Releasing the density ratio values themselves may, however, pose
+unacceptable privacy risks, especially when the density ratio model is
+sensitive to extreme cases. Privatizing density ratio estimation
+presents an interesting avenue for future research.
+
+Despite these open questions, we have shown that already in its current
+form, density ratio estimation provides a flexible and informative
+framework for evaluating synthetic data utility. Due to its in-built
+model selection, it requires little specification on behalf of the user
+and works well in a wide variety of settings. The framework can be
+readily incorporated into synthetic data workflows using the
+\texttt{densityratio} package.
+
+\section*{Acknowledgements}\label{acknowledgements}
+\addcontentsline{toc}{section}{Acknowledgements}
+
+We are grateful to Joerg Drechsler for sharing the U.S. Current
+Population Survey data, and to Stef van Buuren, Gerko Vink, Hanne
+Oberman and Carlos Gonzalez Poses for many valuable discussions.
+
+\section*{References}\label{references}
 \addcontentsline{toc}{section}{References}
 
-\hypertarget{refs}{}
+\phantomsection\label{refs}
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-SIPP_Beta_2006}{}}%
+\bibitem[\citeproctext]{ref-SIPP_Beta_2006}
 Abowd, John M., Martha Stinson, and Gary Benedetto. 2006. {``Final
@@
 Program, U.S. Bureau of the Census, Washington, DC.
-\url{https://ecommons.cornell.edu/bitstream/handle/1813/43929/SSAfinal.pdf?sequence=3\&isAllowed=y}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ali_silvey_divergence_1966}{}}%
+\url{https://ecommons.cornell.edu/bitstream/handle/1813/43929/SSAfinal.pdf?sequence=3&isAllowed=y}.
+
+\bibitem[\citeproctext]{ref-ali_silvey_divergence_1966}
 Ali, S. M., and S. D. Silvey. 1966. {``A General Class of Coefficients
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-atenas_open_2015}{}}%
+\bibitem[\citeproctext]{ref-atenas_open_2015}
 Atenas, Javiera, Leo Havemann, and Ernesto Priego. 2015. {``Open Data as
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Bowen_differentially_2021}{}}%
+\bibitem[\citeproctext]{ref-basu_power_1998}
+Basu, Ayanendranath, Ian R. Harris, Nils L. Hjort, and M. C. Jones.
+1998. {``{Robust and efficient estimation by minimising a density power
+divergence}.''} \emph{Biometrika} 85 (3): 549--59.
+\url{https://doi.org/10.1093/biomet/85.3.549}.
+
+\bibitem[\citeproctext]{ref-Bowen_differentially_2021}
 Bowen, Claire McKay, Fang Liu, and Bingyue Su. 2021. {``Differentially
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-choi_featurized_2021}{}}%
+\bibitem[\citeproctext]{ref-choi_featurized_2021}
 Choi, Kristy, Madeline Liao, and Stefano Ermon. 2021. {``Featurized
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-crosas_automating_2015}{}}%
+\bibitem[\citeproctext]{ref-crosas_automating_2015}
 Crosas, Mercè, Gary King, James Honaker, and Latanya Sweeney. 2015.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-drechsler2011synthetic}{}}%
+\bibitem[\citeproctext]{ref-drechsler2011synthetic}
 Drechsler, Jörg. 2011. \emph{Synthetic Datasets for Statistical
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-drechsler2012}{}}%
+\bibitem[\citeproctext]{ref-drechsler2012}
 ---------. 2012. {``New Data Dissemination Approaches in Old Europe
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-drechsler_utility_2022}{}}%
+\bibitem[\citeproctext]{ref-drechsler_utility_2022}
 ---------. 2022. {``Challenges in Measuring Utility for Fully Synthetic
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-drechsler2023}{}}%
+\bibitem[\citeproctext]{ref-drechsler2023}
 Drechsler, Jörg, and Anna-Carolina Haensch. 2023. {``30 Years of
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-gdpr}{}}%
+\bibitem[\citeproctext]{ref-Rcpp}
+Eddelbuettel, Dirk. 2013. \emph{Seamless {R} and {C++} Integration with
+{Rcpp}}. New York: Springer.
+\url{https://doi.org/10.1007/978-1-4614-6868-4}.
+
+\bibitem[\citeproctext]{ref-gdpr}
 European Parliament, and Council of the European Union. 2016.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-fpf_2017}{}}%
+\bibitem[\citeproctext]{ref-fpf_2017}
 Future of Privacy Forum. 2017. {``Understanding Corporate Data Sharing
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-gruber2024overcoming}{}}%
+\bibitem[\citeproctext]{ref-gehrke_crowd_2012}
+Gehrke, Johannes, Michael Hay, Edward Lui, and Rafael Pass. 2012.
+{``Crowd-Blending Privacy.''} In \emph{Advances in Cryptology -- CRYPTO
+2012}, edited by Reihaneh Safavi-Naini and Ran Canetti, 479--96. Berlin,
+Heidelberg: Springer Berlin Heidelberg.
+\url{https://doi.org/10.1007/978-3-642-32009-5_28}.
+
+\bibitem[\citeproctext]{ref-ghalebikesabi_dpiw_2022}
+Ghalebikesabi, Sahra, Harry Wilde, Jack Jewson, Arnaud Doucet, Sebastian
+Vollmer, and Chris Holmes. 2022. {``Mitigating Statistical Bias Within
+Differentially Private Synthetic Data.''} In \emph{Proceedings of the
+Thirty-Eighth Conference on Uncertainty in Artificial Intelligence},
+edited by James Cussens and Kun Zhang, 180:696--705. Proceedings of
+Machine Learning Research. PMLR.
+\url{https://proceedings.mlr.press/v180/ghalebikesabi22a.html}.
+
+\bibitem[\citeproctext]{ref-gruber2024overcoming}
 Gruber, Lukas, Markus Holzleitner, Johannes Lehner, Sepp Hochreiter, and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-hawala_synthetic_2008}{}}%
+\bibitem[\citeproctext]{ref-kldest}
+Hartung, Niklas. 2024. \emph{Kldest: Sample-Based Estimation of
+Kullback-Leibler Divergence}.
+\url{https://CRAN.R-project.org/package=kldest}.
+
+\bibitem[\citeproctext]{ref-hawala_synthetic_2008}
 Hawala, Sam. 2008. \emph{Producing Partially Synthetic Data to Avoid
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-shohei_dre_outlier_2008}{}}%
+\bibitem[\citeproctext]{ref-shohei_dre_outlier_2008}
 Hido, Shohei, Yuta Tsuboi, Hisashi Kashima, Masashi Sugiyama, and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-hu_advancing_2024}{}}%
+\bibitem[\citeproctext]{ref-hu_advancing_2024}
 Hu, Jingchen, and Claire McKay Bowen. 2024. {``Advancing Microdata
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-huang_kmm_2006}{}}%
+\bibitem[\citeproctext]{ref-huang_kmm_2006}
 Huang, Jiayuan, Alexander J. Smola, Arthur Gretton, Karsten M.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-hundepool_disclosure_2012}{}}%
+\bibitem[\citeproctext]{ref-hundepool_disclosure_2012}
 Hundepool, Anco, Josep Domingo-Ferrer, Luisa Franconi, Sarah Giessing,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-izbicki_dre_2014}{}}%
+\bibitem[\citeproctext]{ref-izbicki_dre_2014}
 Izbicki, Rafael, Ann Lee, and Chad Schafer. 2014. {``{High-Dimensional
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-kanamori_ulsif_2009}{}}%
+\bibitem[\citeproctext]{ref-kanamori_ulsif_2009}
 Kanamori, Takafumi, Shohei Hido, and Masashi Sugiyama. 2009. {``A
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-kanamori_divergence_2012}{}}%
+\bibitem[\citeproctext]{ref-kanamori_divergence_2012}
 Kanamori, Takafumi, Taiji Suzuki, and Masashi Sugiyama. 2012a. {``\(f\)
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Kanamori2012}{}}%
+\bibitem[\citeproctext]{ref-Kanamori2012}
 ---------. 2012b. {``Statistical Analysis of Kernel-Based Least-Squares
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-karr_utility_2006}{}}%
+\bibitem[\citeproctext]{ref-karr_utility_2006}
 Karr, Alan F., Christine N. Kohnen, Anna Oganian, Jerome P. Reiter, and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-kim_classification_2021}{}}%
+\bibitem[\citeproctext]{ref-kim_classification_2021}
 Kim, Ilmun, Aaditya Ramdas, Aarti Singh, and Larry Wasserman. 2021.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lazer_css_2009}{}}%
+\bibitem[\citeproctext]{ref-lazer_css_2009}
 Lazer, David, Alex Pentland, Lada Adamic, Sinan Aral, Albert-László
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-little_statistical_1993}{}}%
+\bibitem[\citeproctext]{ref-li_application_2010}
+Li, Yan, Hiroyuki Kambara, Yasuharu Koike, and Masashi Sugiyama. 2010.
+{``Application of Covariate Shift Adaptation Techniques in
+Brain--Computer Interfaces.''} \emph{IEEE Transactions on Biomedical
+Engineering} 57 (6): 1318--24.
+\url{https://doi.org/10.1109/TBME.2009.2039997}.
+
+\bibitem[\citeproctext]{ref-little_statistical_1993}
 Little, Roderick J. A. 1993. {``Statistical Analysis of Masked Data.''}
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-liu_change_2013}{}}%
+\bibitem[\citeproctext]{ref-liu_change_2013}
 Liu, Song, Makoto Yamada, Nigel Collier, and Masashi Sugiyama. 2013.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-menon2016dreloss}{}}%
+\bibitem[\citeproctext]{ref-loftsgaarden_nonparametric_1965}
+Loftsgaarden, D. O., and C. P. Quesenberry. 1965. {``{A Nonparametric
+Estimate of a Multivariate Density Function}.''} \emph{The Annals of
+Mathematical Statistics} 36 (3): 1049--51.
+\url{https://doi.org/10.1214/aoms/1177700079}.
+
+\bibitem[\citeproctext]{ref-menon2016dreloss}
 Menon, Aditya, and Cheng Soon Ong. 2016. {``Linking Losses for Density
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-mohamed2017learning}{}}%
+\bibitem[\citeproctext]{ref-mohamed2017learning}
 Mohamed, Shakir, and Balaji Lakshminarayanan. 2017. {``Learning in
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-murphy_pmlintro_2022}{}}%
+\bibitem[\citeproctext]{ref-murphy_pmlintro_2022}
 Murphy, Kevin P. 2022. \emph{Probabilistic Machine Learning: An
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-newman_future_2012}{}}%
+\bibitem[\citeproctext]{ref-newman_future_2012}
 Newman, Greg, Andrea Wiggins, Alycia Crall, Eric Graham, Sarah Newman,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-obels_analysis_2020}{}}%
+\bibitem[\citeproctext]{ref-synthpop}
+Nowok, Beata, Gillian M. Raab, and Chris Dibben. 2016.
+{``{\emph{Synthpop}:} Bespoke Creation of Synthetic Data in
+{\emph{R}}.''} \emph{Journal of Statistical Software} 74 (11).
+\url{https://doi.org/10.18637/jss.v074.i11}.
+
+\bibitem[\citeproctext]{ref-obels_analysis_2020}
 Obels, Pepijn, Daniël Lakens, Nicholas A. Coles, Jaroslav Gottfried, and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-obermeyer2019}{}}%
+\bibitem[\citeproctext]{ref-obermeyer2019}
 Obermeyer, Ziad, Brian Powers, Christine Vogeli, and Sendhil
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-qin_inferences_1998}{}}%
+\bibitem[\citeproctext]{ref-qin_inferences_1998}
 Qin, Jing. 1998. {``{Inferences for case-control and semiparametric
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-R}{}}%
-R Core Team. 2023. \emph{R: A Language and Environment for Statistical
-Computing}. Vienna, Austria: R Foundation for Statistical Computing.
-\url{https://www.R-project.org/}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-raab2017guidelines}{}}%
+\bibitem[\citeproctext]{ref-raab2017guidelines}
 Raab, Gillian M., Beata Nowok, and Chris Dibben. 2017. {``Guidelines for
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-raab2021assessing}{}}%
+\bibitem[\citeproctext]{ref-raab2021assessing}
 Raab, Gillian M, Beata Nowok, and Chris Dibben. 2021. {``Assessing,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ramachandran_open_2021}{}}%
+\bibitem[\citeproctext]{ref-ramachandran_open_2021}
 Ramachandran, Rahul, Kaylin Bugbee, and Kevin Murphy. 2021. {``From Open
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-reiter_releasing_2004}{}}%
+\bibitem[\citeproctext]{ref-reiter_releasing_2004}
 Reiter, Jerome P. 2004. {``{Releasing Multiply Imputed, Synthetic Public
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rubin_statistical_1993}{}}%
+\bibitem[\citeproctext]{ref-rubin_statistical_1993}
 Rubin, Donald B. 1993. {``Statistical Disclosure Limitation.''}
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Scott1992}{}}%
+\bibitem[\citeproctext]{ref-Scott1992}
 Scott, David W. 1992. \emph{Multivariate Density Estimation: Theory,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-snoke_utility_2018}{}}%
+\bibitem[\citeproctext]{ref-snoke_utility_2018}
 Snoke, Joshua, Gillian M. Raab, Beata Nowok, Chris Dibben, and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-sugiyama_classification_2010}{}}%
+\bibitem[\citeproctext]{ref-sugiyama_classification_2010}
 Sugiyama, Masashi. 2010. {``Superfast-Trainable Multi-Class
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-sugiyama_kliep_2007}{}}%
+\bibitem[\citeproctext]{ref-sugiyama_kliep_2007}
 Sugiyama, Masashi, Shinichi Nakajima, Hisashi Kashima, Paul Buenau, and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-sugiyama_lstst_2011}{}}%
+\bibitem[\citeproctext]{ref-sugiyama_lstst_2011}
 Sugiyama, Masashi, Taiji Suzuki, Yuta Itoh, Takafumi Kanamori, and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-sugiyama_suzuki_kanamori_2012}{}}%
+\bibitem[\citeproctext]{ref-sugiyama_suzuki_kanamori_2012}
 Sugiyama, Masashi, Taiji Suzuki, and Takafumi Kanamori. 2012a.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-sugiyama_bregman_2012}{}}%
+\bibitem[\citeproctext]{ref-sugiyama_bregman_2012}
 ---------. 2012b. {``Density-Ratio Matching Under the Bregman
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-sugiyama_conditional_2010}{}}%
+\bibitem[\citeproctext]{ref-sugiyama_conditional_2010}
 Sugiyama, Masashi, Ichiro Takeuchi, Taiji Suzuki, Takafumi Kanamori,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-tiao2018dre}{}}%
+\bibitem[\citeproctext]{ref-sugiyama_lhss_2011}
+Sugiyama, Masashi, Makoto Yamada, Paul von Bünau, Taiji Suzuki, Takafumi
+Kanamori, and Motoaki Kawanabe. 2011. {``Direct Density-Ratio Estimation
+with Dimensionality Reduction via Least-Squares Hetero-Distributional
+Subspace Search.''} \emph{Neural Networks} 24 (2): 183--98.
+\url{https://doi.org/10.1016/j.neunet.2010.10.005}.
+
+\bibitem[\citeproctext]{ref-tiao2018dre}
 Tiao, Louis C. 2018. {``{D}ensity {R}atio {E}stimation for {KL}
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-uehara2016generative}{}}%
+\bibitem[\citeproctext]{ref-uehara2016generative}
 Uehara, Masatoshi, Issei Sato, Masahiro Suzuki, Kotaro Nakayama, and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-vandewiel2023}{}}%
+\bibitem[\citeproctext]{ref-vandewiel2023}
 van de Wiel, Mark A., Gwenaël G. R. Leday, Jeroen Hoogland, Martijn W.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-densityratio}{}}%
+\bibitem[\citeproctext]{ref-densityratio}
 Volker, Thom Benjamin. 2023. {``Densityratio: Direct Estimation of the
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-willenborg_elements_2001}{}}%
+\bibitem[\citeproctext]{ref-wang_divergence_2009}
+Wang, Qing, Sanjeev R. Kulkarni, and Sergio Verdu. 2009. {``Divergence
+Estimation for Multidimensional Densities via \(k\)-Nearest-Neighbor
+Distances.''} \emph{IEEE Transactions on Information Theory} 55 (5):
+2392--2405. \url{https://doi.org/10.1109/TIT.2009.2016060}.
+
+\bibitem[\citeproctext]{ref-willenborg_elements_2001}
 Willenborg, Leon, and Ton De Waal. 2001. \emph{Elements of Statistical
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Woo_global_2009}{}}%
+\bibitem[\citeproctext]{ref-Woo_global_2009}
 Woo, Mi-Ja, Jerome P. Reiter, Anna Oganian, and Alan F. Karr. 2009.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Wornowizki2016}{}}%
+\bibitem[\citeproctext]{ref-Wornowizki2016}
 Wornowizki, Max, and Roland Fried. 2016. {``Two-Sample Homogeneity Tests
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-xu_ctgan_2019}{}}%
+\bibitem[\citeproctext]{ref-xu_ctgan_2019}
 Xu, Lei, Maria Skoularidou, Alfredo Cuesta-Infante, and Kalyan
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-zettler2021}{}}%
+\bibitem[\citeproctext]{ref-zettler2021}
 Zettler, Ingo, Christoph Schild, Lau Lilleholt, Lara Kroencke, Till
@@
 
-\hypertarget{sec-app-A}{%
 \section{\texorpdfstring{Density ratio estimation in
-\texttt{R}}{Density ratio estimation in R}}\label{sec-app-A}}
-
-TO DO
-
-Small example of ulsif in \texttt{densityratio}
-
-Do the same with \texttt{optim}
-
-Do the same for a different loss function
+\texttt{R}}{Density ratio estimation in R}}\label{sec-app-A}
+
+To perform density ratio estimation in \texttt{R}, we can use the
+\texttt{densityratio} package, which includes model fitting functions
+for multiple popular loss functions. Additionally, we can complement the
+package with standard optimization routines in \texttt{R}. Here, we show
+an example of both approaches.
+
+We generate two small toy datasets, \texttt{x} and \texttt{y}, that
+serve as numerator and denominator data, respectively. Moreover, we set
+\texttt{x} as the centers in the Gaussian kernel, and we pre-define the
+kernel bandwidth to be \(\sigma = 1\) and fix the regularization
+parameter to \(\lambda = 0.5\). Note that we do not perform
+cross-validation for hyperparameter selection to enhance illustrational
+clarity. Subsequently, we fit a density ratio model using the
+\texttt{ulsif()} function from the \texttt{densityratio} package.
+
+\linespread{1}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(densityratio)}
+
+\FunctionTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
+
+\NormalTok{x }\OtherTok{\textless{}{-}} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{250}\NormalTok{) }\SpecialCharTok{|\textgreater{}} \FunctionTok{as.matrix}\NormalTok{()}
+\NormalTok{y }\OtherTok{\textless{}{-}} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{250}\NormalTok{, }\DecValTok{1}\NormalTok{, }\FunctionTok{sqrt}\NormalTok{(}\DecValTok{2}\NormalTok{)) }\SpecialCharTok{|\textgreater{}} \FunctionTok{as.matrix}\NormalTok{()}
+
+\NormalTok{fit\_dr }\OtherTok{\textless{}{-}} \FunctionTok{ulsif}\NormalTok{(x, y, }\AttributeTok{centers =}\NormalTok{ x, }\AttributeTok{scale =} \ConstantTok{NULL}\NormalTok{, }\AttributeTok{sigma =} \DecValTok{1}\NormalTok{, }\AttributeTok{lambda =} \FloatTok{0.5}\NormalTok{)}
+\NormalTok{rhat\_dr }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(fit\_dr)}
+\end{Highlighting}
+\end{Shaded}
+
+\linespread{2}
+
+We can obtain the same solution using standard optimization techniques.
+In order to do this, we first need to define the kernel matrices for the
+numerator and denominator data, and define a loss function that can be
+used in optimization. We then use the \texttt{optim()} function to
+minimize the loss function. Recall that we use a linear model for the
+density ratio, as per Equation~\ref{eq-dr-model}, we use a Gaussian
+kernel as basis function, and we use the least-squares loss of
+Equation~\ref{eq-squared-error-loss}.
+
+\linespread{1}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{Dx }\OtherTok{\textless{}{-}} \FunctionTok{distance}\NormalTok{(x, x, }\ConstantTok{TRUE}\NormalTok{)}
+\NormalTok{Dy }\OtherTok{\textless{}{-}} \FunctionTok{distance}\NormalTok{(y, x, }\ConstantTok{TRUE}\NormalTok{)}
+
+\NormalTok{Kx }\OtherTok{\textless{}{-}} \FunctionTok{kernel\_gaussian}\NormalTok{(Dx, }\AttributeTok{sigma =} \DecValTok{1}\NormalTok{)}
+\NormalTok{Ky }\OtherTok{\textless{}{-}} \FunctionTok{kernel\_gaussian}\NormalTok{(Dy, }\AttributeTok{sigma =} \DecValTok{1}\NormalTok{)}
+
+\NormalTok{loss }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(alpha, Kx, Ky) \{}
+  \FunctionTok{mean}\NormalTok{((Ky }\SpecialCharTok{\%*\%}\NormalTok{ alpha }\SpecialCharTok{{-}} \DecValTok{1}\NormalTok{) }\SpecialCharTok{*}\NormalTok{ Ky }\SpecialCharTok{\%*\%}\NormalTok{ alpha }\SpecialCharTok{{-}}\NormalTok{ (Ky }\SpecialCharTok{\%*\%}\NormalTok{ alpha }\SpecialCharTok{{-}} \DecValTok{1}\NormalTok{)}\SpecialCharTok{\^{}}\DecValTok{2} \SpecialCharTok{/} \DecValTok{2}\NormalTok{) }\SpecialCharTok{{-}}
+    \FunctionTok{mean}\NormalTok{(Kx }\SpecialCharTok{\%*\%}\NormalTok{ alpha }\SpecialCharTok{{-}} \DecValTok{1}\NormalTok{) }\SpecialCharTok{+}
+    \FloatTok{0.5}\SpecialCharTok{/}\DecValTok{2} \SpecialCharTok{*} \FunctionTok{sum}\NormalTok{(alpha}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\CommentTok{\# ridge penalty}
+\NormalTok{\}}
+
+\NormalTok{alpha }\OtherTok{\textless{}{-}} \FunctionTok{runif}\NormalTok{(}\FunctionTok{ncol}\NormalTok{(Kx))}
+\NormalTok{fit\_optim }\OtherTok{\textless{}{-}} \FunctionTok{optim}\NormalTok{(alpha, loss, }\AttributeTok{Kx =}\NormalTok{ Kx, }\AttributeTok{Ky =}\NormalTok{ Ky, }
+                   \AttributeTok{method =} \StringTok{"BFGS"}\NormalTok{, }
+                   \AttributeTok{control =} \FunctionTok{list}\NormalTok{(}\AttributeTok{maxit =} \DecValTok{10000}\NormalTok{))}
+
+\NormalTok{rhat\_optim }\OtherTok{\textless{}{-}}\NormalTok{ Kx }\SpecialCharTok{\%*\%}\NormalTok{ fit\_optim}\SpecialCharTok{$}\NormalTok{par}
+\end{Highlighting}
+\end{Shaded}
+
+\linespread{2}
+
+Subsequently, we can show that both procedures yield equivalent results
+(see Figure~\ref{fig-AppA}). Moreover, the sum of squared difference
+between the estimated density ratio parameters equals 6.5e-11, and the
+sum of squared difference between the predicted density ratios equals
+1.75e-08.
+
+\linespread{1}
+
+\begin{figure}[t]
+
+\centering{
+
+\includegraphics{paper_files/figure-pdf/fig-AppA-1.pdf}
+
+}
+
+\caption{\label{fig-AppA}Density ratio estimation using \texttt{ulsif()}
+from the \texttt{densityratio}-package and using standard optimization
+in \texttt{R}.}
+
+\end{figure}%
+
+\linespread{2}
+
+If we were to perform cross-validation for hyperparameter selection, we
+could run the \texttt{ulsif()}-function without specifying the
+\texttt{sigma} and \texttt{lambda} arguments. The function then performs
+a grid search over a predefined range of hyperparameters. Moreover, it
+is possible to supply a vector of \texttt{sigma} and \texttt{lamdba}
+values to \texttt{ulsif()}, after which cross-validation is
+automatically performed.
+
+\linespread{1}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{ulsif}\NormalTok{(x, y, }\AttributeTok{centers =}\NormalTok{ x, }\AttributeTok{scale =} \ConstantTok{NULL}\NormalTok{)}
+\NormalTok{fit}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+
+Call:
+ulsif(df_numerator = x, df_denominator = y, scale = NULL, centers = x)
+
+Kernel Information:
+  Kernel type: Gaussian with L2 norm distances
+  Number of kernels: 250
+  sigma: num [1:10] 0.0601 0.177 0.2971 0.4228 0.5583 ...
+
+Regularization parameter (lambda): num [1:20] 1000 483.3 233.6 112.9 54.6 ...
+
+Optimal sigma: 0.7045195
+Optimal lambda: 1.43845
+Optimal kernel weights (loocv): num [1:251] 0.03123 0.03174 0.02443 0.00876 0.01296 ...
+
+\end{verbatim}
+
+\linespread{2}
+
+The model output shows the optimal \texttt{sigma} parameter out of a
+vector containing ten values, and the optimal \texttt{lambda} parameter
+out of a vector containing twenty values. The package also includes
+density ratio estimation techniques, such as \texttt{kliep()},
+\texttt{kmm()}, \texttt{spectral()}, \texttt{lhss()} and
+\texttt{naive()}, that all work in a similar fashion. Additionally, the
+package contains generic functions to \texttt{predict()} with,
+\texttt{plot()}, or summarize (e.g., \texttt{summary()}) the fitted
+models. Moreover, computationally intensive code is implemented in
+\texttt{C++} through \texttt{Rcpp} (\citeproc{ref-Rcpp}{Eddelbuettel
+2013}) and most functions allow for parallel computation through the
+\texttt{openmp}-backend in \texttt{C++}. The interested reader is
+referred to the package documentation
+(\citeproc{ref-densityratio}{Volker 2023}) for additional information.
+
+\section{\texorpdfstring{\(k\)-Nearest neighbor density ratio
+estimation}{k-Nearest neighbor density ratio estimation}}\label{sec-app-B}
+
+We note that density ratio estimation through \(k\)-nearest neighbor
+density estimation can also be considered a direct approach to density
+ratio estimation. Using \(k\)-nearest neighbors, an estimate of the
+density can be obtained by \[
+\hat{\pobs}(\bx_i) = \frac{k}{n-1} \cdot \frac{\Gamma(d/2 + 1)}{\pi^{d/2} B^d_k(\bx^{(\text{obs})}_i)},
+\] where \(d\) denotes the dimensionality of the input data,
+\(\Gamma(d/2 + 1) / \pi^{p/2}\) is a normalizing constant, and
+\(B^d_k(\mathbf{x}^{(\text{obs})}_i)\) denotes the Euclidean distance of
+each observation in \(\mathbf{x}^{(\text{obs})}\) to its \(k\)-th
+nearest neighbor in \(\mathbf{x}^{(\text{obs})}\) raised to the power
+\(d\) (see, e.g., \citeproc{ref-wang_divergence_2009}{Wang, Kulkarni,
+and Verdu 2009};). Applying the same procedure for the denominator data
+(i.e., the synthetic data), we can estimate the density ratio as \[
+\begin{aligned}
+\hat{r}(\bx_i) &= \frac{\hat{\pobs}(\bx_i)}{\hat{\psyn}(\bx_i)} \\
+&= \frac{\frac{k}{n-1} \cdot \frac{\Gamma(d/2 + 1)}{\pi^{d/2} B^d_k(\mathbf{x}^{(\text{obs})}_i)}}{\frac{k}{m} \cdot \frac{\Gamma(d/2 + 1)}{\pi^{d/2} B^d_k(\mathbf{x}^{(\text{syn})}_i)}} \\
+&= \frac{m}{n-1} \cdot \frac{B^d_k(\mathbf{x}^{(\text{syn})}_i)}{B^d_k(\mathbf{x}^{(\text{obs})}_i)},
+\end{aligned}
+\] where \(m\) denotes the number of observations in the synthetic data,
+and \(B^d_k(\mathbf{x}^{(\text{syn})}_i)\) denotes the Euclidean
+distance of each observation in \(\mathbf{x}^{(\text{obs})}\) to its
+\(k\)-th nearest neighbor in \(\mathbf{x}^{(\text{syn})}\) raised to the
+power \(d\). Accordingly, a direct estimate of the density ratio can be
+obtained as \[
+\hat{r}(\bx) = \text{diag}\{\tilde{B}^d_k(\mathbf{x}^{(\text{obs})})\}^{-1} \tilde{B}^d_k(\mathbf{x}^{(\text{syn})}),
+\] where \(\tilde{B}^d_k(\mathbf{x}^{(\text{obs})})\) and
+\(\tilde{B}^d_k(\mathbf{x}^{(\text{syn})})\) denote the vectors with
+distances of each observation in \(\mathbf{x}^{(\text{obs})}\) to their
+\(k\)-th nearest neighbor in \(\mathbf{x}^{(\text{obs})}\) and
+\(\mathbf{x}^{(\text{syn})}\) raised to the power \(d\), multiplied by
+the respective sample sizes.
+
+Applying this estimate of the density ratio to the simulated data from
+the univariate simulations, and setting \(k = 15 \approx \sqrt n\), as
+suggested by the authors who introduced \(k\)-nearest neighbor density
+estimation (\citeproc{ref-loftsgaarden_nonparametric_1965}{Loftsgaarden
+and Quesenberry 1965}), we can compare the performance of the
+\(k\)-nearest neighbor density ratio estimation to the density ratio
+estimation through uLSIF.
+
+\linespread{1}
+
+\begin{figure}[t]
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-AppB-1.png}
+
+}
+
+\caption{\label{fig-AppB}Density ratio estimation through \(k\)-nearest
+neighbor density ratio estimation (\(k = 15\)) and uLSIF.}
+
+\end{figure}%
+
+\linespread{2}
+
+It can be seen that the \(k\)-nearest neighbor density ratio estimation
+performs well on average (the green lines in Figure~\ref{fig-AppB}),
+especially in the center of the distribution. However, it suffers from a
+similar bias in regions where the number of observations is low (as the
+density ratio is typically underestimated in the tails). Moreover,
+\(k\)-nearest neighbor density ratio estimation has much higher variance
+than density ratio estimation through uLSIF, as can be seen by the fact
+that the green lines cover a much wider band than the blue lines.
 

### 2024-08-23 code/application.R

@@ -1,0 +1,151 @@
+
+################################################################################
+## Load data and packages
+################################################################################
+
+load("data/cps5000.RData")
+
+library(synthpop)
+library(densityratio)
+library(dplyr)
+library(tidyr)
+library(purrr)
+library(furrr)
+
+plan(multisession, workers = 16)
+
+################################################################################
+## Data wrangling and creating synthetic data
+################################################################################
+
+df <- cps |>
+  select(-csp) |>
+  mutate(educ = as.numeric(as.character(educ)),
+         educ = case_when(educ < 39 ~ 1,
+                          educ < 40 ~ 2,
+                          educ < 44 ~ 3,
+                          educ < 47 ~ 4) |>
+           factor(labels = c("NoHS", "HS", "AoBD", "MoH")),
+         race = factor(race, labels = c("White", "Non-white", "Non-white", "Non-white")),
+         marital = factor(marital, labels = c("Married", "Married", "Separated",
+                                              "Separated", "Widowed", "Single",
+                                              "WidowedORDivorced")),
+         sex = factor(sex, labels = c("Male", "Female")))
+
+method.ini <- c("norm", "norm", "norm", "polyreg", "polyreg", "logreg", "logreg", "norm")
+visit <- c(7, 1, 2, 3, 4, 5, 6, 8)
+m <- 5
+synlist <- list(
+  trans = syn(
+    df |> mutate(across(c(income, tax, age, ss), ~ .x^{1/3})),
+    m = m,
+    method = method.ini,
+    visit.sequence = visit,
+    seed = 1234,
+    print.flag = FALSE
+  ),
+  semi = syn(
+    df |> mutate(across(c(income, tax, age, ss), ~ .x^{1/3})),
+    m = m,
+    method = method.ini,
+    visit.sequence = visit,
+    semicont = list(tax = "0", ss = "0"),
+    seed = 1234,
+    print.flag = FALSE
+  )
+)
+
+synlist$trans$syn <- synlist$trans$syn |>
+  map(~ .x |> mutate(across(c(income, tax, age, ss), ~.x^3)))
+synlist$semi$syn <- synlist$semi$syn |>
+  map(~ .x |> mutate(across(c(income, tax, age, ss), ~.x^3)))
+
+
[…]

### 2024-08-23 code/functions.R

@@ -1,0 +1,71 @@
+
+###############################################################################
+## functions.R containing code to run simulation 1
+###############################################################################
+
+
+# R-script accompanying the paper "A density ratio framework for evaluating the
+# utility of synthetic data"
+# 24-04-2024
+# Thom Benjamin Volker
+
+###############################################################################
+## Functions for generating and evaluating synthetic data
+###############################################################################
+
+
+## Laplace distribution (density function)
+dlaplace <- function(x, mu, sigma) {
+  exp(-abs(x-mu)/(sigma / sqrt(2))) / (2 * (sigma / sqrt(2)))
+}
+
+## Laplace distribution (random number generator)
+rlaplace <- function(n, mu, sigma) {
+  p <- runif(n, min = -0.5, max = 0.5)
+  mu - sigma / sqrt(2) * sign(p) * log(1 - 2 * abs(p))
+}
+
+## Ratio between laplace and normal distribution with same mean and variance
+dratio_lap_norm <- function(x, mu, sigma) {
+  dlaplace(x, mu, sigma) / dnorm(x, mu, sigma)
+}
+
+## Ratio between lognormal and normal distribution with same mean and variance
+dratio_lnorm_norm <- function(x, mu, sigma) {
+  mu_rlnorm    <- log(mu^2 / sqrt(mu^2 + sigma^2))
+  sigma_rlnorm <- sqrt(log(1 + sigma^2 / mu^2))
+  dlnorm(x, mu_rlnorm, sigma_rlnorm) / dnorm(x, mu, sigma)
+}
+
+## Ratio between t-distribution and normal distribution with same mean and variance
+dratio_t_norm <- function(x, mu, sigma) {
+  df <- 2 / (1 - 1/sigma^2)
+  dt(x - mu, df) / dnorm(x - mu, 0, sigma)
+}
+
+## Ratio between normal and normal distribution with same mean and variance
+## (hence, the ratio equals 1 of course, but I'll use the function for
+## consistency nevertheless).
+dratio_norm_norm <- function(x, mu, sigma) {
+  dnorm(x, mu, sigma) / dnorm(x, mu, sigma)
+}
+
[…]

### 2024-08-23 code/sim1.R

@@ -1,0 +1,100 @@
+
+###############################################################################
+## sim1.R containing code to run simulation 1
+###############################################################################
+
+# R-script accompanying the paper "A density ratio framework for evaluating the
+# utility of synthetic data"
+# 24-04-2024
+# Thom Benjamin Volker
+
+###############################################################################
+## Preliminaries: packages, seed and parallel processing
+###############################################################################
+
+# load required packages
+library(purrr)
+library(furrr)
+library(densityratio)
+library(dplyr)
+library(synthpop)
+library(tidyr)
+
+source("code/functions.R")
+
+# set seed for reproducibility
+set.seed(123)
+
+# set up parallel processing
+plan(multisession, workers = 16)
+
+###############################################################################
+## Simulations
+###############################################################################
+
+# parameters
+nsim  <- 1000
+nobs  <- nsyn <- 250
+mu    <- 1
+sigma <- sqrt(2)
+
+# Simulate data
+sims <- expand_grid(model = c("laplace", "lognormal", "t", "normal"),
+                    sim = 1:nsim)
+
+sim_dat <- function(model, n, mu, sigma) {
+  switch(model,
+         laplace = rlaplace(n, mu, sigma),
+         lognormal = rlnorm(n,
+                            log(mu^2 / sqrt(mu^2 + sigma^2)),
+                            sqrt(log(1 + sigma^2 / mu^2))),
+         t = rt(n, 2 / (1 - 1/sigma^2)) + mu,
+         normal = rnorm(n, mu, sigma))
+}
+
+# Create evaluation data for plotting
+x_eval <- seq(-3, 5, length.out = 1600)
+
+# Run simulations
+sims <- sims |>
+  mutate(obs = map(model, ~sim_dat(.x, nobs, mu, sigma)),
+         syn = map(obs, ~rnorm(nsyn, mean = mean(.x), sd = sd(.x))),
+         r = future_map2(obs, syn, ~densityratio::ulsif(.x, .y, progressbar = FALSE),
+                         .options = furrr_options(seed = TRUE),
+                         .progress = TRUE),
+         summary = future_map(r, ~summary(.x, test = FALSE, parallel = FALSE),
+                              .options = furrr_options(seed = TRUE),
+                              .progress = TRUE),
+         model = factor(model,
[…]

### 2024-08-23 code/sim2.R

@@ -1,0 +1,188 @@
+
+## Packages
+library(ggplot2)
+library(densityratio)
+library(synthpop)
+library(purrr)
+library(tidyr)
+library(dplyr)
+library(pbapply)
+
+
+################################################################################
+## Data generation functions
+################################################################################
+
+
+gen_real <- function(n, P, rho = 0.5) {
+
+  if (P < 3) stop("P must be at least 4")
+  Pnonlin <- ifelse(P > 5, 5, 1)
+  Plin    <- P - Pnonlin
+
+  Xlin  <- matrix(rnorm(n*Plin), n) %*% chol(rho + (1-rho) * diag(Plin))
+  Xpoly <- sapply(1:Pnonlin, function(i) {
+    Xi <- Xlin[,i]^{i+1}
+    Xi + rnorm(n, 0, sd(Xi))
+  })
+  cbind(Xlin, Xpoly) |> unname()
+}
+
+gen_syn1 <- function(nsyn, obs) {
+
+  P <- ncol(obs)
+  means <- colMeans(obs)
+  vars  <- diag(diag(var(obs)))
+
+  matrix(rnorm(nsyn*P), nsyn) %*% chol(vars) + matrix(means, nsyn, P, byrow = TRUE)
+}
+
+gen_syn2 <- function(nsyn, obs) {
+
+  P <- ncol(obs)
+  means <- colMeans(obs)
+  vars <- var(obs)
+
+  matrix(rnorm(nsyn*P), nsyn) %*% chol(vars) + matrix(means, nsyn, P, byrow = TRUE)
+}
+
+gen_syn3 <- function(nsyn, obs) {
+
+  P         <- ncol(obs)
+  Pnonlin   <- ifelse(P > 5, 5, 1)
+  Plin      <- P - Pnonlin
+  means_lin <- colMeans(obs[,1:Plin])
+  vars_lin  <- var(obs[,1:Plin])
+
+  Xlin <- matrix(rnorm(nsyn*Plin), nsyn) %*% chol(vars_lin) + matrix(means_lin, nsyn, Plin, byrow = TRUE)
+  Xpoly <- sapply(1:Pnonlin, function(i) {
+    fit <- lm(obs[,Plin + i] ~ I(obs[,i]^{i+1}))
+    cbind(1, Xlin[,i]^{i+1}) %*% coef(fit) + rnorm(nsyn, 0, sd(fit$residuals))
+  })
+
+  cbind(Xlin, Xpoly) |> unname()
+}
+
[…]

### 2024-10-15 paper/paper.qmd
update paper
small change abstract + cover letter

@@
 Various methods have been proposed to measure the utility of synthetic data, but their results are often incomplete or even misleading.
-In this paper, we propose using density ratio estimation to improve quality evaluation for synthetic data, and thereby the quality of synthesized datasets.
+In this paper, we propose to use the density ratio estimation framework to improve quality evaluation for synthetic data, and thereby improve the quality of synthesized datasets.
 We show how this framework relates to and builds on existing measures, yielding global and local utility measures that are informative and easy to interpret.
-We develop an estimator which requires little to no manual tuning due to automatic selection of a nonparametric density ratio model.
+The proposed methodology employs automatic, non-parametric model selection for the density ratio, and thus requires little to no manual model specification.
 Through simulations, we find that density ratio estimation yields more accurate estimates of global utility than established procedures.
 A real-world data application demonstrates how the density ratio can guide refinements of synthesis models and can be used to improve downstream analyses.
-We conclude that density ratio estimation is a valuable tool in synthetic data generation workflows and provide these methods in the accessible open source R-package densityratio.
-keywords: |
-  synthetic data, utility, density ratio, privacy, disclosure limitation
+We conclude that density ratio estimation is a valuable tool in synthetic data generation workflows and make the methodology available through the open source R-package densityratio.
+keywords: synthetic data, utility, density ratio, privacy, disclosure limitation
 bibliography: bibliography.bib

### 2024-10-15 paper/paper.tex
update paper
small change abstract + cover letter

@@
 to measure the utility of synthetic data, but their results are often
-incomplete or even misleading. In this paper, we propose using density
-ratio estimation to improve quality evaluation for synthetic data, and
-thereby the quality of synthesized datasets. We show how this framework
-relates to and builds on existing measures, yielding global and local
-utility measures that are informative and easy to interpret. We develop
-an estimator which requires little to no manual tuning due to automatic
-selection of a nonparametric density ratio model. Through simulations,
-we find that density ratio estimation yields more accurate estimates of
-global utility than established procedures. A real-world data
-application demonstrates how the density ratio can guide refinements of
-synthesis models and can be used to improve downstream analyses. We
-conclude that density ratio estimation is a valuable tool in synthetic
-data generation workflows and provide these methods in the accessible
-open source R-package densityratio.
+incomplete or even misleading. In this paper, we propose to use the
+density ratio estimation framework to improve quality evaluation for
+synthetic data, and thereby improve the quality of synthesized datasets.
+We show how this framework relates to and builds on existing measures,
+yielding global and local utility measures that are informative and easy
+to interpret. The proposed methodology employs automatic, non-parametric
+model selection for the density ratio, and thus requires little to no
+manual model specification. Through simulations, we find that density
+ratio estimation yields more accurate estimates of global utility than
+established procedures. A real-world data application demonstrates how
+the density ratio can guide refinements of synthesis models and can be
+used to improve downstream analyses. We conclude that density ratio
+estimation is a valuable tool in synthetic data generation workflows and
+make the methodology available through the open source R-package
+densityratio.
 \end{abstract}
@@
 
-\caption{\label{tbl-application-regression}TODO}
+\caption{\label{tbl-application-regression}Linear regression
+coefficients for the observed data, synthetic data and reweighted
+synthetic data using the density ratio as importance weights. Boldface
+text indicates which coefficient has smaller absolute bias relative to
+the observed data.}
 

### 2025-09-28 paper/paper.qmd
Update for new submission

@@
 params:
-  runsims: false
+  runsims: true
 ---
@@
 Open data allows third-party researchers to answer research questions with already collected data, freeing up resources that would otherwise be devoted to data collection [@ramachandran_open_2021]. 
-Sharing data in combination with code allows others to validate research findings and build upon the work [@obels_analysis_2020; @crosas_automating_2015].
-Students can benefit from open data, as it fosters education with realistic data [@atenas_open_2015], as well as the general public, through stimulating citizen science projects [@newman_future_2012].
-However, making data openly available is often (rightfully) hampered by official legislation, like the General Data Protection Regulation [GDPR\; @gdpr], and general privacy concerns. 
+Sharing data with the accompanying with code allows others to validate research findings and build upon the work [@obels_analysis_2020; @crosas_automating_2015].
+Students can benefit from open data, as it fosters education with realistic data [@atenas_open_2015], and so can the general public, through stimulating citizen science projects [@newman_future_2012].
+However, making data openly available is (rightfully) hampered by official legislation, like the General Data Protection Regulation [GDPR\; @gdpr], and general privacy concerns. 
 In the worst case, sharing data may cause harm to individuals or organizations, which may withhold these entities from participating in future research.
@@
 If the distributions of the observed and synthetic data are very different, the density ratio should be far from one at those regions where the distributions differ.
-As density estimation is known to be a difficult problem, the density ratio estimation framework provides techniques to directly estimate the density ratio, rather than the two separate densities, in a non-parametric way [@sugiyama_suzuki_kanamori_2012].
-These non-parametric estimation techniques come with automatic model specification, which mitigates the issue of model specification. 
-This functionality is implemented in the `R`-package `densityratio` [@densityratio].
+The density ratio estimation framework approaches the problem by estimating the density ratio directly in a non-parametric way [@sugiyama_suzuki_kanamori_2012].
+This approach alleviates the challenging problem of estimating two densities, which may magnify estimation errors when subsequently taking their ratio.
+The density ratio framework is made readily available in the `R`-package `densityratio` [@densityratio], which attempts to further simplify the procedure by implementing automatic hyperparameter optimization, which mitigates the issue of model specification.
 Importantly, the density ratio is estimated over the entire range of the data, which provides measures of utility at every (possible) point in the data space.
@@
 
-In the remainder of the article, we introduce the density ratio framework and the associated estimation techniques, and connect the framework to traditional utility measures as the $pMSE$ and the Kullback-Leibler divergence.
+In the remainder of the article, we introduce the density ratio framework and connect it to traditional utility measures as the $pMSE$ and the Kullback-Leibler divergence.
 We then present simulations to demonstrate the performance of density ratio estimation in stylized settings and compare it to traditional utility measures.
@@
 The propensity score mean-squared error ($pMSE$), introduced by @Woo_global_2009 and further developed in @snoke_utility_2018, formalizes this intuition.
-Let $I_i$ denote an indicator variable that equals $1$ if observation $i$ ($i \in 1, \dots, N$, $N = \nobs + \nsyn$) belongs to the synthetic data, and $0$ otherwise.
-We then train a classifier that outputs the predicted probability of observation $i$ being a synthetic record $\hat{\pi}_i$ based on the observation's scores on the variables (this can be the set of all variables, but also a subset).
-From these, we can calculate the utility statistic 
+Let $I_i$ denote an indicator variable that equals $1$ if observation $i$ ($i \in 1, \dots, N$; $N = \nobs + \nsyn$) belongs to the synthetic data, and $0$ otherwise.
+These indicators are used in a classifier that outputs the predicted probability of observation $i$ being a synthetic record $\hat{\pi}_i$ based on the observation's scores on the variables (this can be the set of all variables, but also a subset).
+From these, the utility statistic is calculated as
 $$
@@
 The general idea of density ratio estimation is depicted in @fig-dr-plot, and boils down to comparing two distributions by modelling the density ratio $r(\bx)$.
-In our case, we define $r(\bx)$ as the ratio between the probability distributions of the numerator samples, taken from the observed data distribution, $\pobs(\bx)$, and the denominator samples, taken from the synthetic data distribution, $\psyn(\bx)$, such that
+In our case, we define $r(\bx)$ as the ratio between the probability distributions of the numerator samples, taken from the observed data distribution, $\pobs(\bx)$, and the denominator samples, taken from the synthetic data distribution, $\psyn(\bx)$, assumed to have common support, such that
 $$
@@
 $$ {#eq-dr}
-
 This specification has the intuitive interpretation that in locations where the density ratio takes values larger than 1, too few synthetic observations are generated in that region.
@@
 Potentially, the loss function can be extended with a regularization term (e.g., a $L_1$ or $L_2$ norm on the parameters of the model), to decrease the flexibility of the density ratio.
-Parameter estimation can be carried out using standard optimization techniques (see Appendix [-@sec-app-A] for a brief demonstration in `R`).
 
@@
 $$
-where $\boldsymbol{c}_j$ ($j \in 1, \dots, b$) denotes the centers of the Gaussian kernel functions, and $\sigma$ controls the kernel width [i.e., it defines over which distance differences between the observations and the centers are considered relevant\; for more information on kernel functions, see, e.g., @murphy_pmlintro_2022].
+where $\boldsymbol{c}_j$ ($j \in 1, \dots, b$) denotes the centers of the Gaussian kernel functions, and $\sigma$ controls the kernel width [i.e., it defines over which distance similarity between the observations and the centers are considered relevant\; for more information on kernel functions, see, e.g., @murphy_pmlintro_2022].
 The appropriate width of the kernel and the optimal value of the regularization parameter can be determined using cross-validation. 
@@
 The package comes with an easy-to-use interface, automatic cross-validation of hyperparameters and builds on `C++` for fast computation (see Appendix [-@sec-app-A] for a brief demonstration).
+Alternatively, parameter estimation can be carried out using standard optimization techniques (see Appendix [-@sec-app-A] for a brief demonstration of both approaches).
 The density ratio estimation framework thus equips researchers with a measure of utility which is accurate, fast to compute and does not require user specification of the model.
@@
 That is, for different synthetic data sets, we can calculate the divergence to the observed data, and compare these values to determine which synthetic data set is most similar to the observed data.
-More formally, the $f$-divergence can be used in a hypothesis test to determine whether the synthetic data is generated from the same distribution as the observed data.
+More formally, the $f$-divergence can be used in a permutation test to determine whether the synthetic data is generated from the same distribution as the observed data.
 The corresponding $p$-value can be calculated by comparing the observed divergence measure to the null distribution of divergence measures obtained using random permutations of the observed and synthetic data [see, for example, @sugiyama_lstst_2011; @Wornowizki2016; @kanamori_divergence_2012].
@@
 
-1. Laplace distribution with location parameter $\mu = 1$ and scale parameter $b = 1$.
-2. Log-normal distribution with log-mean parameter $\mu_{\text{log}} = \log \{1/\sqrt{3} \}$ and log-variance parameter $\sigma^2_\text{log} = \log 3$.
-3. Location-scale $t$-distribution with location parameter $\mu = 1$, scale parameter $\tau^2 = 1$ and degrees of freedom $\nu = 4$. 
+1. Laplace distribution with location parameter $\mu = 1$ and scale $b = 1$.
+2. Log-normal distribution with log-mean $\mu_{\text{log}} = \log \{1/\sqrt{3} \}$ and log-variance $\sigma^2_\text{log} = \log 3$.
+3. Location-scale $t$-distribution with location parameter $\mu = 1$, scale $\tau^2 = 1$ and degrees of freedom $\nu = 4$. 
 4. Normal distribution with mean $\mu = 1$ and variance $\sigma^2 = 2$.
@@
   ) +
-  scale_y_continuous(breaks = 0:6, limits = c(0,6)) +
+  scale_y_continuous(breaks = 0:7, limits = c(0,7)) +
   theme(legend.position = "bottom",
@@
 #| dev: ragg_png
-#| fig-cap: Utility measures for $n = 1000$ and $D = 25$ for $PE$, $KL_{\sqrt{n}}$ and $pMSE_{\text{cart}}$ for $1000$ simulations of the three synthetic data models.
+#| fig-cap: The estimated utility measures ($PE$, $KL_{\sqrt{n}}$ and $pMSE_{\text{cart}}$) for $n = 1000$ observations and $D = 25$ variables in $1000$ simulations for the three increasingly accurate synthetic data models.
+All metrics show how the estimated utility improves with the refinements of the synthesis models, but $PE$ and $KL_{\sqrt{n}}$ show less variability and catch the synthesis model improvements more consistently.
 
@@
 Despite the appealing characteristics of density ratio estimation, several open questions remain. 
-Kernel methods, as typically employed in density ratio estimation, can be sensitive to the choice of kernel, and the use of a Gaussian kernel may not be appropriate in any situation. 
-Categorical variables, for instance, are not straightforwardly incorporated in the current set-up. 
-To deal with this issue, we created dummy variables for each category in our illustration, which seemed to provide reasonable results. 
+Kernel methods, as typically employed in density ratio estimation, can be sensitive to the choice of kernel, and the use of a Gaussian kernel, while a reasonable and versatile default, may not be appropriate in every situation.
+Categorical variables, for instance, are not naturally incorporated in the current set-up. 
+To deal with this issue, we created dummy variables for the categories of the categorical variables in our illustration, which provided reasonable results. 
 However, other solutions (i.e., different kernels or distance measures) may lead to improvements.
@@
 
-Dx <- distance(x, x, TRUE)
-Dy <- distance(y, x, TRUE)
-
-Kx <- kernel_gaussian(Dx, sigma = 1)
-Ky <- kernel_gaussian(Dy, sigma = 1)
+Dx <- densityratio:::distance(x, x, TRUE)
+Dy <- densityratio:::distance(y, x, TRUE)
+
+Kx <- densityratio:::kernel_gaussian(Dx, sigma = 1)
+Ky <- densityratio:::kernel_gaussian(Dy, sigma = 1)
 

### 2025-09-28 paper/paper.tex
Update for new submission

@@
+% Options for packages loaded elsewhere
 % Options for packages loaded elsewhere
@@
 ]{article}
-
+\usepackage{xcolor}
 \usepackage{amsmath,amssymb}
+\setcounter{secnumdepth}{5}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
@@
 \makeatother
-\usepackage{xcolor}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
@@
 
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\usepackage{longtable,booktabs,array}
 \usepackage{calc} % for calculating minipage widths
@@
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
 \makeatother
+
 % definitions for citeproc citations
@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
 \makeatother
-
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
 \usepackage{bookmark}
-
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+\urlstyle{same}
 \hypersetup{
@@
 (\citeproc{ref-ramachandran_open_2021}{Ramachandran, Bugbee, and Murphy
-2021}). Sharing data in combination with code allows others to validate
-research findings and build upon the work
+2021}). Sharing data with the accompanying with code allows others to
+validate research findings and build upon the work
 (\citeproc{ref-obels_analysis_2020}{Obels et al. 2020};
@@
 benefit from open data, as it fosters education with realistic data
-(\citeproc{ref-atenas_open_2015}{Atenas, Havemann, and Priego 2015}), as
-well as the general public, through stimulating citizen science projects
-(\citeproc{ref-newman_future_2012}{Newman et al. 2012}). However, making
-data openly available is often (rightfully) hampered by official
-legislation, like the General Data Protection Regulation (GDPR;
+(\citeproc{ref-atenas_open_2015}{Atenas, Havemann, and Priego 2015}),
+and so can the general public, through stimulating citizen science
+projects (\citeproc{ref-newman_future_2012}{Newman et al. 2012}).
+However, making data openly available is (rightfully) hampered by
+official legislation, like the General Data Protection Regulation (GDPR;
 \citeproc{ref-gdpr}{European Parliament and Council of the European
@@
 should be far from one at those regions where the distributions differ.
-As density estimation is known to be a difficult problem, the density
-ratio estimation framework provides techniques to directly estimate the
-density ratio, rather than the two separate densities, in a
-non-parametric way
+The density ratio estimation framework approaches the problem by
+estimating the density ratio directly in a non-parametric way
 (\citeproc{ref-sugiyama_suzuki_kanamori_2012}{Sugiyama, Suzuki, and
-Kanamori 2012a}). These non-parametric estimation techniques come with
-automatic model specification, which mitigates the issue of model
-specification. This functionality is implemented in the
-\texttt{R}-package \texttt{densityratio}
-(\citeproc{ref-densityratio}{Volker 2023}). Importantly, the density
-ratio is estimated over the entire range of the data, which provides
-measures of utility at every (possible) point in the data space. This
-point-specific quantification of utility turns out to be a useful
-side-product, as it allows to reweigh analyses on synthetic data when
-further improving the utility directly is not possible.
+Kanamori 2012a}). This approach alleviates the challenging problem of
+estimating two densities, which may magnify estimation errors when
+subsequently taking their ratio. The density ratio framework is made
+readily available in the \texttt{R}-package \texttt{densityratio}
+(\citeproc{ref-densityratio}{Volker 2023}), which attempts to further
+simplify the procedure by implementing automatic hyperparameter
+optimization, which mitigates the issue of model specification.
+Importantly, the density ratio is estimated over the entire range of the
+data, which provides measures of utility at every (possible) point in
+the data space. This point-specific quantification of utility turns out
+to be a useful side-product, as it allows to reweigh analyses on
+synthetic data when further improving the utility directly is not
+possible.
 
 In the remainder of the article, we introduce the density ratio
-framework and the associated estimation techniques, and connect the
-framework to traditional utility measures as the \(pMSE\) and the
-Kullback-Leibler divergence. We then present simulations to demonstrate
-the performance of density ratio estimation in stylized settings and
-compare it to traditional utility measures. Subsequently, we apply
-density ratio estimation in a case study where we evaluate the utility
-of multiple synthetic versions of the U.S. Current Population Survey. We
-conclude with a discussion of the results, highlight the strengths and
-weaknesses of the density ratio framework, and provide recommendations
-for future research.
+framework and connect it to traditional utility measures as the \(pMSE\)
+and the Kullback-Leibler divergence. We then present simulations to
+demonstrate the performance of density ratio estimation in stylized
+settings and compare it to traditional utility measures. Subsequently,
+we apply density ratio estimation in a case study where we evaluate the
+utility of multiple synthetic versions of the U.S. Current Population
+Survey. We conclude with a discussion of the results, highlight the
+strengths and weaknesses of the density ratio framework, and provide
+recommendations for future research.
 
@@
 intuition. Let \(I_i\) denote an indicator variable that equals \(1\) if
-observation \(i\) (\(i \in 1, \dots, N\), \(N = \nobs + \nsyn\)) belongs
-to the synthetic data, and \(0\) otherwise. We then train a classifier
-that outputs the predicted probability of observation \(i\) being a
-synthetic record \(\hat{\pi}_i\) based on the observation's scores on
-the variables (this can be the set of all variables, but also a subset).
-From these, we can calculate the utility statistic
+observation \(i\) (\(i \in 1, \dots, N\); \(N = \nobs + \nsyn\)) belongs
+to the synthetic data, and \(0\) otherwise. These indicators are used in
+a classifier that outputs the predicted probability of observation \(i\)
+being a synthetic record \(\hat{\pi}_i\) based on the observation's
+scores on the variables (this can be the set of all variables, but also
+a subset). From these, the utility statistic is calculated as
 \begin{equation}\phantomsection\label{eq-pMSE}{
@@
 \(\pobs(\bx)\), and the denominator samples, taken from the synthetic
-data distribution, \(\psyn(\bx)\), such that
-\begin{equation}\phantomsection\label{eq-dr}{
+data distribution, \(\psyn(\bx)\), assumed to have common support, such
+that \begin{equation}\phantomsection\label{eq-dr}{
 r(\bx) = \frac{\pobs(\bx)}{\psyn(\bx)}.
-}\end{equation}
-
-This specification has the intuitive interpretation that in locations
-where the density ratio takes values larger than 1, too few synthetic
-observations are generated in that region. Conversely, in locations
-where the density ratio takes values smaller than one, too many
-synthetic observations are generated in that region (see
+}\end{equation} This specification has the intuitive interpretation that
+in locations where the density ratio takes values larger than 1, too few
+synthetic observations are generated in that region. Conversely, in
+locations where the density ratio takes values smaller than one, too
+many synthetic observations are generated in that region (see
 Figure~\ref{fig-dr-plot}). We put \(\pobs(\bx)\) in the numerator such
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-dr-plot-1.pdf}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-dr-plot-1.pdf}
 
@@
 on the parameters of the model), to decrease the flexibility of the
-density ratio. Parameter estimation can be carried out using standard
-optimization techniques (see Appendix \ref{sec-app-A} for a brief
-demonstration in \texttt{R}).
+density ratio.
 
@@
 centers of the Gaussian kernel functions, and \(\sigma\) controls the
-kernel width (i.e., it defines over which distance differences between
+kernel width (i.e., it defines over which distance similarity between
 the observations and the centers are considered relevant; for more
@@
 of hyperparameters and builds on \texttt{C++} for fast computation (see
-Appendix \ref{sec-app-A} for a brief demonstration). The density ratio
-estimation framework thus equips researchers with a measure of utility
-which is accurate, fast to compute and does not require user
-specification of the model.
+Appendix \ref{sec-app-A} for a brief demonstration). Alternatively,
+parameter estimation can be carried out using standard optimization
+techniques (see Appendix \ref{sec-app-A} for a brief demonstration of
+both approaches). The density ratio estimation framework thus equips
+researchers with a measure of utility which is accurate, fast to compute
+and does not require user specification of the model.
 
@@
 data set is most similar to the observed data. More formally, the
-\(f\)-divergence can be used in a hypothesis test to determine whether
+\(f\)-divergence can be used in a permutation test to determine whether
 the synthetic data is generated from the same distribution as the
@@
   Laplace distribution with location parameter \(\mu = 1\) and scale
-  parameter \(b = 1\).
+  \(b = 1\).
 \item
-  Log-normal distribution with log-mean parameter
-  \(\mu_{\text{log}} = \log \{1/\sqrt{3} \}\) and log-variance parameter
+  Log-normal distribution with log-mean
+  \(\mu_{\text{log}} = \log \{1/\sqrt{3} \}\) and log-variance
   \(\sigma^2_\text{log} = \log 3\).
@@
   Location-scale \(t\)-distribution with location parameter \(\mu = 1\),
-  scale parameter \(\tau^2 = 1\) and degrees of freedom \(\nu = 4\).
+  scale \(\tau^2 = 1\) and degrees of freedom \(\nu = 4\).
 \item
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-densities-sim1-1.pdf}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-densities-sim1-1.pdf}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-sim1-results-1.png}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-sim1-results-1.png}
 
@@
 \midrule
-100 & 5 & 0.583 & 0.844 & 0.904 & 0.369 & 0.188\\
-100 & 25 & 0.450 & 0.838 & 0.933 & 0.364 & 0.250\\
-100 & 50 & 0.363 & 0.673 & 0.786 & 0.348 & 0.352\\
+100 & 5 & 0.583 & 0.844 & 0.904 & 0.362 & 0.188\\
+100 & 25 & 0.450 & 0.838 & 0.933 & 0.368 & 0.250\\
+100 & 50 & 0.363 & 0.673 & 0.786 & 0.352 & 0.352\\
 \addlinespace
-1000 & 5 & 1.000 & 1.000 & 1.000 & 0.887 & 0.173\\
-1000 & 25 & 1.000 & 1.000 & 1.000 & 0.893 & 0.242\\
-1000 & 50 & 0.995 & 1.000 & 1.000 & 0.846 & 0.316\\
+1000 & 5 & 1.000 & 1.000 & 1.000 & 0.885 & 0.173\\
+1000 & 25 & 1.000 & 1.000 & 1.000 & 0.898 & 0.242\\
+1000 & 50 & 0.997 & 1.000 & 1.000 & 0.839 & 0.316\\
 \addlinespace
-2000 & 5 & 1.000 & 1.000 & 1.000 & 0.885 & 0.173\\
-2000 & 25 & 1.000 & 1.000 & 1.000 & 0.822 & 0.247\\
-2000 & 50 & 1.000 & 1.000 & 1.000 & 0.788 & 0.324\\
+2000 & 5 & 1.000 & 1.000 & 1.000 & 0.882 & 0.173\\
+2000 & 25 & 1.000 & 1.000 & 1.000 & 0.816 & 0.247\\
+2000 & 50 & 1.000 & 1.000 & 1.000 & 0.780 & 0.324\\
 \bottomrule
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-sim2-results-1.png}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-sim2-results-1.png}
 
@@
 
-\caption{\label{fig-sim2-results}Utility measures for \(n = 1000\) and
-\(D = 25\) for \(PE\), \(KL_{\sqrt{n}}\) and \(pMSE_{\text{cart}}\) for
-\(1000\) simulations of the three synthetic data models.}
+\caption{\label{fig-sim2-results}The estimated utility measures (\(PE\),
+\(KL_{\sqrt{n}}\) and \(pMSE_{\text{cart}}\)) for \(n = 1000\)
+observations and \(D = 25\) variables in \(1000\) simulations for the
+three increasingly accurate synthetic data models. All metrics show how
+the estimated utility improves with the refinements of the synthesis
+models, but \(PE\) and \(KL_{\sqrt{n}}\) show less variability and catch
+the synthesis model improvements more consistently.}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-application-distributions-1.pdf}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-application-distributions-1.pdf}
 
@@
 That is, the transformed modelling strategy has a higher average
-estimated Pearson divergence (\(\bar{PE}_\text{transformed} = 0.090\),
-\(\text{range} = [0.084, 0.100]\)) than the semi-continuous modelling
-strategy (\(\bar{PE}_\text{semi-continuous} = 0.050\),
-\(\text{range} = [0.047, 0.053]\)). Although the difference may seem
+estimated Pearson divergence (\(\bar{PE}_\text{transformed} = 0.085\),
+\(\text{range} = [0.080, 0.091]\)) than the semi-continuous modelling
+strategy (\(\bar{PE}_\text{semi-continuous} = 0.047\),
+\(\text{range} = [0.044, 0.050]\)). Although the difference may seem
 small in an absolute sense, it is substantial in a relative sense.
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-app-utility-results-1.png}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-app-utility-results-1.png}
 
@@
 \midrule
-Intercept & 10.297 & 10.192 & \textbf{10.195}\\
+Intercept & 10.297 & \textbf{10.192} & 10.182\\
 Age & -0.003 & -0.005 & \textbf{-0.004}\\
-Education: High school & 0.283 & 0.415 & \textbf{ 0.385}\\
-Education: Associate or Bachelor's degree & 0.593 & 0.782 & \textbf{ 0.731}\\
-Education: Master's degree or higher & 1.016 & 1.135 & \textbf{ 1.080}\\
-Marital: Separated & -0.500 & -0.434 & \textbf{-0.437}\\
-Marital: Widowed & -0.467 & \textbf{-0.471} & -0.450\\
-Marital: Single & -0.491 & -0.592 & \textbf{-0.565}\\
-Marital: Widowed or divorced & -0.549 & \textbf{-0.518} & -0.508\\
-Race: Non-white & -0.090 & \textbf{-0.109} & -0.111\\
+Education: High school & 0.283 & 0.415 & \textbf{ 0.397}\\
+Education: Associate or Bachelor's degree & 0.593 & 0.782 & \textbf{ 0.744}\\
+Education: Master's degree or higher & 1.016 & 1.135 & \textbf{ 1.092}\\
+Marital: Separated & -0.500 & \textbf{-0.434} & -0.433\\
+Marital: Widowed & -0.467 & \textbf{-0.471} & -0.448\\
+Marital: Single & -0.491 & -0.592 & \textbf{-0.561}\\
+Marital: Widowed or divorced & -0.549 & \textbf{-0.518} & -0.505\\
+Race: Non-white & -0.090 & \textbf{-0.109} & -0.110\\
 Sex: Female & -0.117 & -0.134 & \textbf{-0.126}\\
-log(Tax + 1) & 0.067 & 0.071 & \textbf{ 0.066}\\
-log(Social security + 1) & -0.026 & -0.009 & \textbf{-0.017}\\
+log(Tax + 1) & 0.067 & 0.071 & \textbf{ 0.067}\\
+log(Social security + 1) & -0.026 & -0.009 & \textbf{-0.016}\\
 \bottomrule
@@
 to
-\(\text{avg}(|\beta_\text{obs}-\beta_\text{syn-adj}|/SE_\text{obs}) = 1.41\)
+\(\text{avg}(|\beta_\text{obs}-\beta_\text{syn-adj}|/SE_\text{obs}) = 1.54\)
 after reweighing. Hence, besides providing a measure of global utility,
@@
 density ratio estimation, can be sensitive to the choice of kernel, and
-the use of a Gaussian kernel may not be appropriate in any situation.
-Categorical variables, for instance, are not straightforwardly
-incorporated in the current set-up. To deal with this issue, we created
-dummy variables for each category in our illustration, which seemed to
-provide reasonable results. However, other solutions (i.e., different
-kernels or distance measures) may lead to improvements. Additionally,
-our results showed that the current approach to regularization is not
-ideal, as the density ratio estimates are shrunk towards an estimated
-intercept. However, when the intercept deviates from one, the numerator
-or denominator density are up or downweighted. Rather, one would like to
-shrink the density ratio estimates towards one, as this is more in line
-with the interpretation that there is no information to estimate the
-density ratio. Finally, releasing information from the density ratio may
-pose an additional privacy risk, as it leaks additional information
-about the original data. Presumably, solely releasing a global utility
-measure, such as the Pearson divergence, will yield only additional
-risk. Releasing the density ratio values themselves may, however, pose
+the use of a Gaussian kernel, while a reasonable and versatile default,
+may not be appropriate in every situation. Categorical variables, for
+instance, are not naturally incorporated in the current set-up. To deal
+with this issue, we created dummy variables for the categories of the
+categorical variables in our illustration, which provided reasonable
+results. However, other solutions (i.e., different kernels or distance
+measures) may lead to improvements. Additionally, our results showed
+that the current approach to regularization is not ideal, as the density
+ratio estimates are shrunk towards an estimated intercept. However, when
+the intercept deviates from one, the numerator or denominator density
+are up or downweighted. Rather, one would like to shrink the density
+ratio estimates towards one, as this is more in line with the
+interpretation that there is no information to estimate the density
+ratio. Finally, releasing information from the density ratio may pose an
+additional privacy risk, as it leaks additional information about the
+original data. Presumably, solely releasing a global utility measure,
+such as the Pearson divergence, will yield only additional risk.
+Releasing the density ratio values themselves may, however, pose
 unacceptable privacy risks, especially when the density ratio model is
@@
 \begin{Highlighting}[]
-\NormalTok{Dx }\OtherTok{\textless{}{-}} \FunctionTok{distance}\NormalTok{(x, x, }\ConstantTok{TRUE}\NormalTok{)}
-\NormalTok{Dy }\OtherTok{\textless{}{-}} \FunctionTok{distance}\NormalTok{(y, x, }\ConstantTok{TRUE}\NormalTok{)}
-
-\NormalTok{Kx }\OtherTok{\textless{}{-}} \FunctionTok{kernel\_gaussian}\NormalTok{(Dx, }\AttributeTok{sigma =} \DecValTok{1}\NormalTok{)}
-\NormalTok{Ky }\OtherTok{\textless{}{-}} \FunctionTok{kernel\_gaussian}\NormalTok{(Dy, }\AttributeTok{sigma =} \DecValTok{1}\NormalTok{)}
+\NormalTok{Dx }\OtherTok{\textless{}{-}}\NormalTok{ densityratio}\SpecialCharTok{:::}\FunctionTok{distance}\NormalTok{(x, x, }\ConstantTok{TRUE}\NormalTok{)}
+\NormalTok{Dy }\OtherTok{\textless{}{-}}\NormalTok{ densityratio}\SpecialCharTok{:::}\FunctionTok{distance}\NormalTok{(y, x, }\ConstantTok{TRUE}\NormalTok{)}
+
+\NormalTok{Kx }\OtherTok{\textless{}{-}}\NormalTok{ densityratio}\SpecialCharTok{:::}\FunctionTok{kernel\_gaussian}\NormalTok{(Dx, }\AttributeTok{sigma =} \DecValTok{1}\NormalTok{)}
+\NormalTok{Ky }\OtherTok{\textless{}{-}}\NormalTok{ densityratio}\SpecialCharTok{:::}\FunctionTok{kernel\_gaussian}\NormalTok{(Dy, }\AttributeTok{sigma =} \DecValTok{1}\NormalTok{)}
 
@@
 
-\includegraphics{paper_files/figure-pdf/fig-AppA-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-AppA-1.pdf}}
 
@@
 
-Optimal sigma: 0.7045195
-Optimal lambda: 1.43845
+Optimal sigma (loocv): 0.7045195
+Optimal lambda (loocv): 1.43845
 Optimal kernel weights (loocv): num [1:251] 0.03123 0.03174 0.02443 0.00876 0.01296 ...
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-AppB-1.png}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-AppB-1.png}
 

### 2025-09-28 code/application.R

@@ -126,24 +126,17 @@ syn_coefs_adj <- map2(PE_allvars$semi, synlist$semi$syn, \(w, syn) {
      ) |> coef()
 })
 
-syn_coefs_cart <- lapply(
-  synlist$smoothed$syn,
-  \(syn) lm(log(income) ~ . - tax - ss + log(tax+1) + log(ss+1), syn) |> coef()
-)
-
 coefs <- data.frame(
   true = true_coefs,
   syn_semi = do.call(cbind, syn_coefs_semi) |> rowMeans(),
   syn_adj = do.call(cbind, syn_coefs_adj) |> rowMeans()
-) |>
-  round(5)
+)
 
 
 bias <- data.frame(
   syn_semi = (do.call(cbind, syn_coefs_semi) |> rowMeans() - true_coefs) / true_ses,
   syn_adj = (do.call(cbind, syn_coefs_adj) |> rowMeans() - true_coefs) / true_ses
 )
-
 
 save(list = c("df", "synlist", "PE", "PE_allvars",
                "coefs", "bias"),

### 2025-09-28 code/functions.R

@@ -58,8 +58,8 @@ knn_ratio <- function(nu, de, k) {
   nnu <- nrow(nu)
   nde <- nrow(de)
   p <- ncol(nu)
-  Dnu <- densityratio::distance(nu, nu) |> sqrt()
-  Dde <- densityratio::distance(nu, de) |> sqrt()
+  Dnu <- densityratio:::distance(nu, nu) |> sqrt()
+  Dde <- densityratio:::distance(nu, de) |> sqrt()
   nn_nu <- apply(Dnu, 1, \(x) order(x)[k+1])
   nn_de <- apply(Dde, 1, \(x) order(x)[k])
   Dnn_nu <- t(Dnu)[0:(nnu-1)*ncol(Dnu) + nn_nu]
